### 2015-08-27 notes/notes-linear-system-learning.tex
Added notes.

\documentclass[a4paper,10pt]{article}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{IEEEtrantools}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{subfig}
\usepackage{cite}
\usepackage[margin=1in]{geometry}

\input{linear-system-learning-macros.tex}
\newcommand{\meta}[1]{{\color{red}\em #1}}

%opening
\title{Notes on Learning Linear State Space Models}
\author{Pete Bunch}

\begin{document}

\section{Basic Linear State Space}

\subsection{Definition}

\begin{IEEEeqnarray}{rCl}
 \ls{\ti} & = & \lgtm \ls{\ti-1} + \tn{\ti} \\
 \ob{\ti} & = & \lgom \ls{\ti}   + \on{\ti}
\end{IEEEeqnarray}
%
$\ls{\ti} \in \reals^{\lsd}$ and $\ob{\ti} \in \reals^{\obd}$.
%
\begin{IEEEeqnarray}{rCl}
 \tn{\ti} & \sim & \normaldist{0}{\lgtv} \\
 \on{\ti} & \sim & \normaldist{0}{\lgov}
\end{IEEEeqnarray}
%
$\lgtv$ and $\lgov$ are positive definite covariance matrices.

Equivalently in terms of transition and observation densities:
%
\begin{IEEEeqnarray}{rCl}
 \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) & = & \normalden{\ls{\ti}}{\lgtm\ls{\ti-1}}{\lgtv} \\
 \den(\ob{\ti}|\ls{\ti},\lgom,\lgov)   & = & \normalden{\ob{\ti}}{\lgom\ls{\ti}}{\lgov}
\end{IEEEeqnarray}
%
The initial state $\ls{1}$ may be known or may be assigned a Gaussian prior.

\subsection{Identifiability}

\subsection{Bayesian Learning with Gibbs Sampling}

Target the joint distribution $\den(\lgtm, \lgtv, \ls{1:\timax} | \ob{1:\timax}$. Gibbs sampling can be used, targeting the state and parameter posterior conditionals alternately. State conditional sampled by Kalman filtering and backward simulation.

\subsubsection{Parameter Conditional}

The conjugate prior is:
\begin{equation}
 \den(\lgtm, \lgtv) =  \den(\lgtm | \lgtv) \den(\lgtv)
\end{equation}
\begin{align}
 \lgtv &\sim \iwishartdist{\priordof}{\priorscalematrix} \\
 \lgtm | \lgtv &\sim \matrixnormaldist{\priormeanmatrix}{\lgtv}{\priorcolumnvariance}
\end{align}
%
So the prior densities are:
\begin{align}
\den(\lgtv) &= \frac{ \determ{\priorscalematrix}^{\frac{\priordof}{2}} }{ 2^{\frac{\priordof}{2}} \gammafun\left(\frac{\priordof}{2}\right) } \determ{\lgtv}^{-\frac{\priordof+\lsd+1}{2}} \exp\left( -\half \trace\left[\lgtv\inv\priorscalematrix\right] \right) \\
\den(\lgtm | \lgtv) &= \determ{2 \pi \lgtv}^{-\half} \determ{2 \pi \priorcolumnvariance}^{-\half} \exp\left(-\half \trace\left[ \lgtv\inv (\lgtm-\priormeanmatrix) \priorcolumnvariance\inv (\lgtm-\priormeanmatrix)\tr \right] \right) 
\end{align}
%
$\priordof\in\reals, \priordof>\lsd$. $\priorscalematrix$ and $\priorcolumnvariance$ are $\lsd\times\lsd$ positive definite matrices. $\priormeanmatrix \in \reals^{\lsd\times\lsd}$.

The likelihood function is:
\begin{align}
 \den(\ls{1:\timax}|\lgtm, \lgtv) &= \den(\ls{0}) \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) \nonumber \\
 &\propto \determ{\lgtv}^{-\half(\timax-1)} \exp\left( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \lgtv\inv (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber \\
 &= \determ{\lgtv}^{-\half(\timax-1)} \exp\left(-\half\trace\left[ \lgtv\inv \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \right]\right) \nonumber \\
 &= \determ{\lgtv}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \lgtv\inv \left( \lgtm \suffstats{1} \lgtm\tr - \lgtm \suffstats{2}\tr - \suffstats{2} \lgtm\tr + \suffstats{3} \right) \right]\right)
\end{align}
%
where the sufficient statistics are:
%
\begin{align}
 \suffstats{0} &= \timax - 1 \\
 \suffstats{1} &= \sum_{t=2}^{\timax} \ls{\ti-1}\ls{\ti-1}\tr \\
 \suffstats{2} &= \sum_{t=2}^{\timax} \ls{\ti}\ls{\ti-1}\tr \\
 \suffstats{3} &= \sum_{t=2}^{\timax} \ls{\ti}\ls{\ti}\tr \\
\end{align}

The parameter posterior conditional is:
\begin{equation}
 \den(\lgtm, \lgtv | \ls{1:\timax}, \ob{1:\timax}) =  \den(\lgtm | \lgtv, \ls{1:\timax}) \den(\lgtv | \ls{1:\timax})
\end{equation}
\begin{align}
 \lgtv | \ls{1:\timax} &\sim \iwishartdist{\postdof}{\postscalematrix} \\
 \lgtm | \lgtv, \ls{1:\timax} &\sim \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}
\end{align}

Hyperparameter updates:
\begin{align}
 \postcolumnvariance\inv                 &= \priorcolumnvariance\inv + \suffstats{1} \\
 \postmeanmatrix \postcolumnvariance\inv &= \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2}\\
 \postdof                                &= \priordof + \suffstats{0} \\
 \postscalematrix                        &= \priorscalematrix + \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr - \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr
\end{align}

Proof:
\begin{IEEEeqnarray}{rCl}
 \den(\lgtm, \lgtv | \ls{1:\timax}, \ob{1:\timax}) &\propto& \den(\ls{1:\timax}|\lgtm, \lgtv) \den(\lgtm|\lgtv) \den(\lgtv) \nonumber \\
 &=& \determ{\lgtv}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \lgtv\inv \left( \lgtm \suffstats{1} \lgtm\tr - \lgtm \suffstats{2}\tr - \suffstats{2} \lgtm\tr + \suffstats{3} \right) \right]\right) \nonumber \\
 & & \times \determ{\lgtv}^{-\frac{\priordof+\lsd+1}{2}} \exp\left( -\half \trace\left[\lgtv\inv\priorscalematrix\right] \right) \nonumber \\
 & & \times \determ{\lgtv}^{-\half} \exp\left(-\half \trace\left[ \lgtv\inv (\lgtm-\priormeanmatrix) \priorcolumnvariance\inv (\lgtm-\priormeanmatrix)\tr \right] \right) \nonumber \\ 
 &=& \determ{\lgtv}^{-\frac{(\priordof+\suffstats{0})+\lsd+1}{2}} \exp\left( -\half \trace\left[\lgtv\inv(\priorscalematrix+\suffstats{3})\right] \right) \nonumber \\
 & & \times \determ{\lgtv}^{-\half} \exp\Bigg(-\half \trace\bigg[ \lgtv\inv \bigg( \lgtm (\priorcolumnvariance\inv + \suffstats{1}) \lgtm\tr - \lgtm (\priorcolumnvariance\inv\priormeanmatrix\tr + \suffstats{2}\tr) \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad  -\: (\priormeanmatrix\priorcolumnvariance\inv + \suffstats{2}) \lgtm\tr + \priormeanmatrix\priorcolumnvariance\inv\priormeanmatrix\tr \bigg) \bigg] \Bigg)
\end{IEEEeqnarray}
%
Finish by completing the square and comparing terms.

\section{Degenerate Linear State Space Model}

\subsection{Definition and Complications}

\begin{IEEEeqnarray}{rCl}
 \ls{\ti} & = & \lgtm \ls{\ti-1} + \lgdm \tn{\ti}
\end{IEEEeqnarray}
\begin{IEEEeqnarray}{rCl}
 \tn{\ti} & \sim & \normaldist{0}{\idmat}
\end{IEEEeqnarray}

This is equivalent to the basic case, except that $\lgtv = \lgdm\lgdm\tr$ is only positive semi-definite. This is not in itself a problem. The probability distribution associated with each transition is still well-defined, although it does not have a density on $\reals^{\lsd}$. The state perturbations $(\ls{\ti}-\lgtm\ls{\ti-1})$ must all lie in a linear subspace.

There are two problems with extending the Gibbs sampling methodology to such a model. First, the state sequence and the transition covariance matrix are too strongly related. The sampled state sequence defines the subspace, so we cannot freely change $\lgtv$. Second, the conjugate prior is tricky. The matrix-normal-inverse-wishart can be generalised to the case where $\lgtv$ is not full rank (using singular wishart and singular matrix normal distributions). However, this is not helpful, because it constrains the transition matrix to lie in a linear subspace.

Because Wishart distributions are somewhat more studied than their inverse counterparts, it will be easier in the following section to work with the precision, defined as $\lgtp = (\lgdm\lgdm\tr)\pinv$. We will need the eigen-decomposition of this frequently.

\begin{IEEEeqnarray}{rCl}
 \lgtp &=& \tpvec \tpval \tpvec\tr
\end{IEEEeqnarray}
This is the ``non-singular'' factorisation, where $\tpval$ is a diagonal $\rk\times\rk$ matrix with $\rk = \rank(\lgtp)$, and $\tpvec$ from the appropriate Stiefel manifold. The full decomposition will also be useful at times.

\begin{IEEEeqnarray}{rCl}
 \lgtp &=& \begin{bmatrix}\tpvec & \tpvecorth \end{bmatrix} \begin{bmatrix}\tpval & \zmat \\ \zmat & \zmat \end{bmatrix} \begin{bmatrix}\tpvec\tr \\ \tpvecorth\tr \end{bmatrix}
\end{IEEEeqnarray}

\subsection{Conjugate Prior}

\subsubsection{Wishart Part}

There are two ways to extend the Wishart distribution to positive semi-definite matrices. Either let $\priordof \in \{z \in \naturals | z < \lsd\}$, in which case $\rank(\lgtp) = \priordof$, or let $\priorscalematrix$ be positive semi-definite. The latter makes it tricky to work out what we should set $\priorscalematrix$ to, so we'll go with the former. If we condition on a known value of the rank $r$, then the singular Wishart density is,

\begin{IEEEeqnarray}{rCl}
 \den(\lgtp) &=& \frac{ \pi^{\half\rk(\lsd-\rk)} \determ{\priorscalematrix}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \gammafun[r]\left(\frac{\rk}{2}\right) } \determ{\tpval}^{\half(\rk-\lsd-1)} \exp\left( -\half \lgtp \priorscalematrix \right)    .
\end{IEEEeqnarray}

\cite{Uhlig1994,Diaz-Garcia1997}

Note that $\priorscalematrix$ is defined as the inverse of what it usually is, for consistency with the basic model case.

\subsubsection{Matrix Normal Part}

It's fine for the row-variance matrix in a matrix normal distribution to be less than full rank. That just means that the matrix is constrained to lie in a particular subspace. There's even a density associated with this. See \cite{Diaz-Garcia1997}. But this is no good as a prior. We want to be able to change the transition matrix freely.

Instead we can use,

\begin{IEEEeqnarray}{rCl}
 \lgtm|\lgtp &\sim& \matrixnormaldist{\priormeanmatrix}{\lgtp\pinv+\tpvecorth\priorextratpval\inv\tpvecorth\tr}{\priorcolumnvariance}
\end{IEEEeqnarray}
%
$\priorextratpval$ is a diagonal matrix of positive eigenvalues which relax the constraint to lie in the subspace. They control the rate at which the density decays as we move away from that subspace. Setting these extra eigenvalues might be tricky. I haven't really worked out how to do this well yet. But they can depend on the other ones, so we could take the minimum, maximum or average of the others. Making them really big is uninformative.

\subsection{MCMC for Bayesian Learning}

A sampled state trajectory $\ls{1:\timax}$ defines the subspace in which the state perturbations must lie, which means that neither $\lgtm$ nor $\lgtp$ can be freely changed, and Gibbs sampling alone does not work. However, we can still sample within the space specified by this constraint. Additional MCMC moves will then be needed which allow the subspace to be changed.

\subsubsection{Factorising the Precision}

Using Givens rotations, we can efficiently factorise the precision eigenvector matrix as follows.
%
\begin{IEEEeqnarray}{rCl}
 \begin{bmatrix} \tpvec & \tpvecorth \end{bmatrix} &=& \begin{bmatrix}\tprot & \tprotorth\end{bmatrix}\begin{bmatrix} \tpsign & \zmat \\ \zmat & \tpsignorth \end{bmatrix} \begin{bmatrix} \tprow & \zmat \\ \zmat & \idmat \end{bmatrix}\begin{bmatrix} \idmat & \zmat \\ \zmat & \tpnull \end{bmatrix}
\end{IEEEeqnarray}

So
%
\begin{IEEEeqnarray}{rCl}
 \tpvec     &=& \tprot     \tpsign     \tprow  \\
 \tpvecorth &=& \tprotorth \tpsignorth \tpnull
\end{IEEEeqnarray}

Then defined the following, which is positive definite
%
\begin{equation}
 \tpfull = \tpsign \tprow \tpval \tprow\tr \tpsign\tr
\end{equation}

So
%
\begin{IEEEeqnarray}{rCl}
 \lgtp &=& \tprot \tpfull \tprot\tr
\end{IEEEeqnarray}

\subsubsection{State Trajectory Likelihood}

There is no proper transition density over $\reals^{\lsd}$ for the degenerate model. However, there is a density associated with the underlying state disturbances which allows us to write a likelihood function.

\begin{IEEEeqnarray}{rCl}
 \ls{\ti} &=& \lgtm \ls{\ti-1} + \tprot \tpfull^{-\half} \tn{\ti} \\
 \end{IEEEeqnarray}
 \begin{IEEEeqnarray}{rCl}
 \Rightarrow \tprot\tr(\ls{\ti}-\lgtm\ls{\ti-1}) &\sim& \normaldist{\zmat}{\tpfull\inv}
\end{IEEEeqnarray}
\begin{IEEEeqnarray}{rCl}
 \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) &\propto& \determ{\tpfull}^{\half} \exp\left(-\half (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tprot \tpfull \tprot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right)
\end{IEEEeqnarray}

The likelihood function associated with the state trajectory is:
\begin{align}
 \den(\ls{1:\timax}|\lgtm, \lgtv) &= \den(\ls{0}) \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) \nonumber \\
 &\propto \determ{\tpfull}^{\half(\timax-1)} \exp\left( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tprot \tpfull \tprot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber \\
 &= \determ{\tpfull}^{\half(\timax-1)} \exp\left(-\half\trace\left[ \tpfull \tprot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tprot \right]\right) \nonumber \\
 &= \determ{\tpfull}^{\half\suffstats{0}} \exp\left(-\half\trace\left[ \tpfull \left( \tprot\tr \lgtm \suffstats{1} \lgtm\tr \tprot - \tprot\tr \lgtm \suffstats{2}\tr \tprot - \tprot\tr \suffstats{2} \lgtm\tr \tprot + \tprot\tr \suffstats{3} \tprot \right) \right]\right) \\
 &= \determ{\tpfull}^{\half\suffstats{0}} \exp\left(-\half\trace\left[ \tpfull \left( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tprot - \tprot\tr \suffstats{2} \lgtmrot\tr + \tprot\tr \suffstats{3} \tprot \right) \right]\right)
\end{align}
%
where
%
\begin{equation}
 \lgtmrot = \tprot\tr \lgtm
\end{equation}

This is all subject to the constraint that each state must be reachable from the previous one, which we can write concisely as,
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \lgtm \LS{1:\timax-1} + \tprot Z
\end{IEEEeqnarray}
%
where $Z \in \reals^{\rk\times\lsd}$ and 
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \begin{bmatrix} \ls{2} & \ls{3} & \hdots & \ls{\timax} \end{bmatrix}
\end{IEEEeqnarray}

\subsubsection{Transforming the Prior}

If we transform from $(\lgtm,\lgtp)$ to $(\lgtmrot,\tpfull)$, then the prior on these variables is as follows.

Using the transformation property of the matrix normal distribution:
%
\begin{IEEEeqnarray}{rCl}
 \lgtmrot|\lgtp = \lgtmrot|\tprot,\tpfull &\sim& \matrixnormaldist{\tprot\tr\priormeanmatrix}{\tpfull\inv}{\priorcolumnvariance}
\end{IEEEeqnarray}

Using the transformation property of the Wishart distribution:
%
\begin{IEEEeqnarray}{rCl}
 \tpfull|\tprot &\sim& \wishartdist{\rk}{\tprot\tr\priorscalematrix\tprot}
\end{IEEEeqnarray}

\subsubsection{Within-Subspace Parameter Conditional}

We can now get to the conditional distribution for $\den(\lgtmrot, \tpfull|\tprot,\ls{1:\timax})$ which is well-behaved and can be sampled.

The parameter posterior conditional is:
\begin{equation}
 \den(\lgtmrot, \tpfull | \tprot, \ls{1:\timax}, \ob{1:\timax}) =  \den(\lgtmrot | \tpfull, \tprot, \ls{1:\timax}) \den(\tpfull | \tprot, \ls{1:\timax})
\end{equation}
\begin{align}
 \tpfull | \tprot, \ls{1:\timax} &\sim \wishartdist{\postdof}{\postscalematrix} \\
 \lgtmrot | \tpfull, \tprot, \ls{1:\timax} &\sim \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}
\end{align}

Hyperparameter updates:
\begin{align}
 \postcolumnvariance\inv                 &= \priorcolumnvariance\inv + \suffstats{1} \\
 \postmeanmatrix \postcolumnvariance\inv &= \tprot\tr\left( \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \right) \\
 \postdof                                &= \rk + \suffstats{0} \\
 \postscalematrix                        &= \tprot\tr\left( \priorscalematrix + \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr \right)\tprot - \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr
\end{align}

Proof:
\begin{IEEEeqnarray}{rCl}
 \den(\lgtmrot, \tpfull|\tprot,\ls{1:\timax}) &\propto& \den(\ls{1:\timax}|\lgtmrot, \tpfull, \tprot) \den(\lgtmrot | \tpfull, \tprot) \den(\tpfull|\tprot) \nonumber \\
%
 &\propto& \determ{\tpfull}^{\half\suffstats{0}} \exp\left(-\half\trace\left[ \tpfull \left( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tprot - \tprot\tr \suffstats{2} \lgtmrot\tr + \tprot\tr \suffstats{3} \tprot \right) \right]\right) \nonumber \\
 & & \times \determ{\tpfull}^{\frac{r-r-1}{2}} \exp\left( -\half \trace\left[\tpfull \tprot\tr\priorscalematrix\tprot\right] \right) \nonumber \\
 & & \times \determ{\tpfull}^{\half} \exp\left(-\half \trace\left[ \tpfull (\lgtmrot-\tprot\tr\priormeanmatrix) \priorcolumnvariance\inv (\lgtmrot-\tprot\tr\priormeanmatrix)\tr \right] \right) \nonumber \\ 
%
 &=& \determ{\tpfull}^{\frac{(r+\suffstats{0})-r-1}{2}} \exp\left( -\half \trace\left[\tpfull \tprot\tr(\priorscalematrix+\suffstats{3})\tprot\right] \right) \nonumber \\
 & & \times \determ{\tpfull}^{\half} \exp\Bigg(-\half \trace\bigg[ \tpfull \bigg( \lgtmrot (\priorcolumnvariance\inv + \suffstats{1}) \lgtmrot\tr - \lgtmrot(\priorcolumnvariance\inv\priormeanmatrix\tr + \suffstats{2}\tr) \tprot \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad  -\: \tprot\tr (\priormeanmatrix\priorcolumnvariance\inv + \suffstats{2}) \lgtmrot\tr + \tprot\tr \priormeanmatrix\priorcolumnvariance\inv\priormeanmatrix\tr \tprot \bigg) \bigg] \Bigg)
\end{IEEEeqnarray}
%
Finish by completing the square and comparing terms.

Of course, we want $\lgtm$, not $\lgtmrot$. We can recover the matrix uniquely using the constraint from the likelihood. The previous value of $\lgtm$ from before we sampled must have satisfied this constraint. Let's call that matrix $\lgtm\mcold$. 
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \lgtm\mcold \LS{1:\timax-1} + \tprot Z\mcold \nonumber \\
 \LS{2:\timax} &=& \lgtm \LS{1:\timax-1} + \tprot Z \nonumber \\
 \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) \LS{1:\timax-1} + \tprot(Z-Z\mcold) \\
 \Rightarrow \zmat &=& (\lgtmrot-\lgtmrot\mcold) \LS{1:\timax-1} + (Z-Z\mcold) \\
 \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) \LS{1:\timax-1} - \tprot (\lgtmrot-\lgtmrot\mcold) \LS{1:\timax-1} \\
 \zmat &=& \left[(\lgtm-\lgtm\mcold) - \tprot (\lgtmrot-\lgtmrot\mcold)\right] \LS{1:\timax-1} \\
 \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) - \tprot (\lgtmrot-\lgtmrot\mcold)
\end{IEEEeqnarray}
\begin{IEEEeqnarray}{rCl}
 \lgtm &=& \lgtm\mcold + \tprot (\lgtmrot-\lgtmrot\mcold) \nonumber \\
       &=& \lgtm\mcold + \tprot (\lgtmrot-\tprot\tr\lgtm\mcold) \nonumber \\
       &=& (\idmat - \tprot\tprot\tr) \lgtm\mcold + \tprot \lgtmrot 
\end{IEEEeqnarray}

\appendix

\bibliographystyle{plain}
\bibliography{/home/pete/Dropbox/PhD/bibliographies/OTbib}
\end{document}

### 2015-08-27 learners_mcmc.py

@@ -210,6 +210,7 @@ class MCMCLearnerForBasicModelWithIndependentPriors(AbstractMCMCLearner):
                                                 self.model.parameters['Q'])
 
 
+
 class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
     """
     Container for MCMC system learning algorithm.
@@ -217,7 +218,19 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
     Prior Type: Singular Matrix Normal-Inverse Wishart
     """
     
+    def transition_covariance_prior(self, model):
+        """
+        Prior density of transition model
+        """
+        return smp.singular_wishart_density(model.parameters['val'],
+                                            model.parameters['vec'],
+                                            self.hyperparams['Psi0'])
+    
     def iterate_transition(self, moveType):
+        """
+        MCMC iteration (Metropolis-Hastings) for transition matrix and
+        covariance.
+        """
         
         # Kalman filter
         flt,_,lhood = self.model.kalman_filter(self.observ)
@@ -228,9 +241,21 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
         # Propose change
         if   moveType=='F':
             pass
+        
         elif moveType=='Q':
+            
+            # Make the change
             rotation = smp.sample_cayley(self.model.ds, self.algoparams['Qs'])
             ppsl_model.rotate_transition_covariance(rotation)
+            
+            # Random walk, so forward and backward probabilities are same
+            fwd_prob = 0
+            bwd_prob = 0
+            
+            # Prior terms
+            prior = self.transition_covariance_prior(self.model)
+            ppsl_prior = self.transition_covariance_prior(ppsl_model)
+            
         elif moveType=='rank':
             pass
         else:
@@ -243,7 +268,9 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
         ppsl_flt,_,ppsl_lhood = ppsl_model.kalman_filter(self.observ)
         
         # Decide
-        acceptRatio = (ppsl_lhood-lhood)
+        acceptRatio =   (ppsl_lhood-lhood) \
+                      + (ppsl_prior-prior) \
+                      + (bwd_prob-fwd_prob)
         if self.verbose:
                 print("   Acceptance ratio: {}".format(acceptRatio))
         if np.log(np.random.random())<acceptRatio:
[…]

### 2015-08-27 run_basic_learning.py

@@ -29,6 +29,7 @@ np.random.seed(0)
 state, observ = model.simulate_data(K)
 
 est_params = dict(params)
+est_params['F'] = 0.5*np.identity(ds)
 est_params['rank'] = 2
 est_params['vec'] = np.array([[1,0],[0,1],[0,0]])
 est_params['val'] = np.array([1,1])

### 2015-08-27 sampling.py

@@ -1,10 +1,10 @@
 import numpy as np
 from scipy import linalg as la
 from scipy import stats
+from scipy import special
 #from scipy.stats import multivariate_normal as mvn
 #from scipy import stats as sps
 #from scipy import misc as spm
-#from scipy import special as spec
 #from numpy import random as rnd
 
 def sample_wishart(nu, P):
@@ -51,6 +51,18 @@ def sample_cayley(d, s):
     M = la.solve(I-S,I+S)
     
     return M
+
+def singular_wishart_density(val,vec,P):
+    """
+    Density of a singular wishart distribution at X = vec*diag(val)*vec.T
+    """
+    d,r = vec.shape
+    norm = 0.5*r*(r-d)*np.log(np.pi) - 0.5*r*d*np.log(2.0) \
+           - special.multigammaln(r/2,r) - 0.5*r*la.det(P)
+    prptn = (r-d-1)*np.log(np.prod(val)) \
+           - 0.5*np.trace( np.dot(np.dot(vec.T,la.solve(P,vec)),np.diag(val)) )
+    pdf = norm + prptn
+    return pdf
 
 def evaluate_sufficient_statistics(x):
     """

### 2015-08-27 notes/notes-linear-system-learning.tex
Added correct priors for Q MH moves

@@
 
+\subsection{Metropolis-Hastings for the Precision Subspace}
+
+In order to change $\tprot$, we can target $\den(\lgtp|\lgtm, \ob{1:\timax})$ with Metropolis-Hastings, which constitutes a collapsed Gibbs move.
+
+We sample a rotation matrix $\mhrot$ from some proposal distribution $\mhrotppsl$ and then apply the following transformation,
+%
+\begin{IEEEeqnarray}{rCl}
+ \begin{bmatrix}
+  \lgtp\mcnew \\ \mhrot\mcnew
+ \end{bmatrix}
+ & = &
+ \begin{bmatrix}
+  \mhrot\lgtp\mhrot\tr \\ \mhrot\tr
+ \end{bmatrix} \nonumber     ,
+\end{IEEEeqnarray}
+%
+which is clearly its own inverse.
+Since $\determ{\mhrot}=1$, it is straightforward to show that the Jacobian of this transformation is $1$, and hence using the reversible jump interpretation of Metropolis-Hastings \cite{Green1995,Green2009}, the acceptance probability is,
+%
+\begin{IEEEeqnarray}{rCl}
+ \mhap(\lgtp\to\lgtp\mcnew) & = & \min\left\{1, \frac{ \den(\lgtp\mcnew|\lgtm,\ob{1:\timax})\mhrotppsl(\mhrot\mcnew) }{ \den(\lgtp|\lgtm, \ob{1:\timax})\mhrotppsl(\mhrot) } \right\}  \\
+ & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtp\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtp)} \times \frac{\den(\lgtp\mcnew, \lgtm) }{\den(\lgtp, \lgtm)} \times \frac{\mhrotppsl(\mhrot\mcnew)}{\mhrotppsl(\mhrot)} \right\} \nonumber     .
+\end{IEEEeqnarray}
+%
+The first term is simply a ratio of Kalman filter likelihoods.
+
+There are numerous ways to sample the rotation matrix $\mhrot$ from a suitable proposal distribution $\mhrotppsl$.
+For example, we could use the Cayley transform \cite{Leon2006}, a bijective mapping from the skew-symmetric matrices to the rotation matrices, defined by,
+%
+\begin{IEEEeqnarray}{rCl}
+ P(S) & = & (\idmat - S)\inv(I+S)     .
+\end{IEEEeqnarray}
+%
+To sample from $\mhrotppsl$, we draw $\half\lsd(\lsd-1)$ independent scalar random variables $\{s_{i,j}\}_{0<i<j<\lsd}$ from any zero-mean distribution; a nice choice is,
+%
+\begin{IEEEeqnarray}{rCl}
+ s_{k,l} & \sim & \normaldist{0}{\sigma_s^2} \label{eq:skewsymmetric_proposal}     .
+\end{IEEEeqnarray}
+%
+Use these to construct a skew-symmetric matrix $S$,
+%
+\begin{IEEEeqnarray}{rCl}
+ S_{k,l} & = & \begin{cases}
+                s_{k,l}  & k<l \\
+                -s_{l,k} & k>l \\
+                0        & k=l     ,
+               \end{cases}
+\end{IEEEeqnarray}
+%
+and then set $\mhrot=P(S)$.
+The Cayley transform has the property that $P(-S)=P(S\tr)=P(S)\inv=P(S)\tr$, which implies that $\mhrotppsl(\mhrot)=\mhrotppsl(\mhrot\tr)=\mhrotppsl(\mhrot\mcnew)$, leading to a cancellation in the acceptance probability.
+
+There is an alternative using Givens rotations.
+First sample $i \in [1,\lsd]$, $j \in [1,\lsd]\setminus i$, and $\givrot{} \in [-\pi/2,\pi/2]$ from some zero-mean distribution.
+Form the Givens matrix $\givmat{i}{j}{\givrot{}}$ such that,
+%
+\begin{IEEEeqnarray}{rCl}
+ \left[\givmat{i}{j}{\givrot{}} - \idmat\right]_{k,l} & = & \begin{cases}
+                                                    \cos(\givrot{})-1 & k=l=i \text{ or } k=l=j \\
+                                                    \sin(\givrot{}) & k=i,l=j \\
+                                                    -\sin(\givrot{}) & k=j,l=i \\
+                                                    0 & \text{ otherwise,}
+                                                 \end{cases}
+\end{IEEEeqnarray}
+%
+and use $\mhrot=\givmat{i}{j}{\givrot{}}$.
+This also has the property that $\givmat{i}{j}{-\givrot{}} = \givmat{i}{j}{\givrot{}}\tr$, meaning that we achieve the same cancellation of the proposals as before.
+
 \appendix

### 2015-08-27 learners_mcmc.py

@@ -1,6 +1,7 @@
 # Operational modules
 from abc import ABCMeta, abstractmethod
 import numpy as np
+import scipy.linalg as la
 import matplotlib.pyplot as plt
 
 # Import from other module files
@@ -218,13 +219,22 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
     Prior Type: Singular Matrix Normal-Inverse Wishart
     """
     
-    def transition_covariance_prior(self, model):
+    def transition_prior(self, model):
         """
-        Prior density of transition model
+        Prior density for transition model parameters
         """
-        return smp.singular_wishart_density(model.parameters['val'],
-                                            model.parameters['vec'],
-                                            self.hyperparams['Psi0'])
+        variancePrior = smp.singular_wishart_density(model.parameters['val'],
+                      model.parameters['vec'],la.inv(self.hyperparams['Psi0']))
+        
+        orthVec = model.complete_basis()
+        rowVariance = model.transition_covariance() \
+                      + self.hyperparams['alpha']*np.dot(orthVec,orthVec.T)
+        matrixPrior = smp.matrix_normal_density(model.parameters['F'],
+                                                self.hyperparams['M0'],
+                                                rowVariance,
+                                                self.hyperparams['V0'])
+        
+        return variancePrior + matrixPrior
     
     def iterate_transition(self, moveType):
         """
@@ -253,8 +263,8 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
             bwd_prob = 0
             
             # Prior terms
-            prior = self.transition_covariance_prior(self.model)
-            ppsl_prior = self.transition_covariance_prior(ppsl_model)
+            prior = self.transition_prior(self.model)
+            ppsl_prior = self.transition_prior(ppsl_model)
             
         elif moveType=='rank':
             pass

### 2015-08-27 linear_models.py

@@ -277,6 +277,14 @@ class DegenerateLinearModel(AbstractLinearModel):
         self.parameters['vec'] = np.dot(U, eVec)
         self.parameters['val'] = eVal
     
+    def complete_basis(self):
+        """
+        Returns a matrix of vectors orthogonal to the eigenvectors
+        """
+        Q,R = la.qr(self.parameters['vec'])
+        r = self.parameters['rank']
+        return Q[:,r:]
+    
     def rotate_transition_covariance(self, rotation):
         """
         Rotate noise covariance matrix by multiplying eigenvectors by a

### 2015-08-27 run_basic_learning.py

@@ -40,6 +40,7 @@ hyperparams['nu0'] = params['rank']
 hyperparams['Psi0'] = params['rank']*np.identity(ds)
 hyperparams['M0'] = np.zeros((ds,ds))
 hyperparams['V0'] = np.identity(ds)
+hyperparams['alpha'] = 100
 
 algoparams = dict()
 algoparams['Qs'] = 0.1

### 2015-08-27 sampling.py

@@ -52,16 +52,24 @@ def sample_cayley(d, s):
     
     return M
 
-def singular_wishart_density(val,vec,P):
+def singular_wishart_density(val, vec, P):
     """
     Density of a singular wishart distribution at X = vec*diag(val)*vec.T
     """
     d,r = vec.shape
     norm = 0.5*r*(r-d)*np.log(np.pi) - 0.5*r*d*np.log(2.0) \
            - special.multigammaln(r/2,r) - 0.5*r*la.det(P)
-    prptn = (r-d-1)*np.log(np.prod(val)) \
+    pptn = 0.5*(r-d-1)*np.log(np.prod(val)) \
            - 0.5*np.trace( np.dot(np.dot(vec.T,la.solve(P,vec)),np.diag(val)) )
-    pdf = norm + prptn
+    pdf = norm + pptn
+    return pdf
+
+def matrix_normal_density(X, M, U, V):
+    """Sample from a matrix normal distribution"""
+    norm = - 0.5*la.det(2*np.pi*U) - 0.5*la.det(2*np.pi*V)
+    XM = X-M
+    pptn = np.exp(-0.5*np.trace( np.dot(la.solve(U,XM),la.solve(V,XM.T)) ))
+    pdf = norm + pptn
     return pdf
 
 def evaluate_sufficient_statistics(x):

### 2015-08-27 notes/notes-linear-system-learning.tex
Added MH moves for transition matrix

@@
 
-\subsection{Metropolis-Hastings for the Precision Subspace}
+\subsubsection{Metropolis-Hastings for the Precision Subspace}
 
@@
 
+\subsubsection{Metropolis-Hastings for the Transition Matrix}
+
+Define $\postden$ as the posterior density, conditioning on $\ob{1:\timax}$.
+
+If we modify the covariance matrix slightly so that it is no longer singular, then we can use the existing sampling procedure from the basic linear model and treat the resulting value of $\lgtm$ as a proposal in a Metropolis-Hastings kernel targeting $\postden(\lgtm, \ls{1:\timax}|\lgtp)$.
+
+Suppose we start with sampled values of $\lgtm$,$\lgtp$ and $\ls{1:\timax}$. A suitably modified precision matrix is,
+%
+\begin{IEEEeqnarray}{rCl}
+ \paddedlgtp & = & \left(\lgtp\pinv + \Qpadding \idmat\right)\inv \label{eq:padded_transition_precision}      ,
+\end{IEEEeqnarray}
+%
+where $\Qpadding$ is a small positive constant. We propose a new value of $\lgtm$ by sampling,
+%
+\begin{IEEEeqnarray}{rCl}
+ \lgtm' & \sim & \postden(\lgtm|\paddedlgtp,\ls{1:\timax})     .
+\end{IEEEeqnarray}
+%
+This distribution is simply a Gaussian and we can evaluate the density using \eqref{eq:basic_F_conditional}. Finally, we also sample a new state trajectory using forward filtering backward sampling,
+%
+\begin{IEEEeqnarray}{rCl}
+ \ls{1:\timax}' & \sim & \postden(\ls{1:\timax}|\lgtm', \lgtp)      .
+\end{IEEEeqnarray}
+%
+The acceptance probability is simply,
+%
+\begin{IEEEeqnarray}{rCl}
+ \IEEEeqnarraymulticol{3}{l}{ \mhap\left(\lgtm,\ls{1:\timax}\to\lgtm',\ls{1:\timax}'\right) } \nonumber \\
+ \quad\quad & = & \min\left\{1,\frac{ \postden(\lgtm',\ls{1:\timax}'|\lgtp) \postden(\ls{1:\timax}|\lgtm,\lgtp) \postden(\lgtm|\paddedlgtp,\ls{1:\timax}')   }{ \postden(\lgtm,\ls{1:\timax}|\lgtp) \postden(\ls{1:\timax}'|\lgtm',\lgtp) \postden(\lgtm'|\paddedlgtp,\ls{1:\timax}) }\right\} \nonumber \\
+ & = & \min\left\{1,\frac{ \postden(\lgtm'|\lgtp) \postden(\lgtm|\paddedlgtp,\ls{1:\timax}')   }{ \postden(\lgtm|\lgtp)\postden(\lgtm'|\paddedlgtp,\ls{1:\timax}) }\right\} \nonumber \\
+ & = & \min\left\{1, \frac{ \den(\ob{1:\timax}|\lgtm',\lgtp) }{ \den(\ob{1:\timax}|\lgtm,\lgtp)} \times \frac{ \den(\lgtm'|\lgtp) }{ \den(\lgtm|\lgtp)} \times \frac{ \postden(\lgtm|\paddedlgtp,\ls{1:\timax}')   }{ \postden(\lgtm'|\paddedlgtp,\ls{1:\timax}) }\right\} \nonumber      , \\
+\end{IEEEeqnarray}
+%
+in which the first term is a ratio of Kalman filter likelihoods.
+
 \appendix

### 2015-08-27 learners_mcmc.py

@@ -6,6 +6,7 @@ import matplotlib.pyplot as plt
 
 # Import from other module files
 import sampling as smp
+from linear_models import BasicLinearModel
 
 class AbstractMCMCLearner:
     __metaclass__ = ABCMeta
@@ -250,7 +251,41 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
         
         # Propose change
         if   moveType=='F':
-            pass
+            
+            # Simulate state trajectory
+            x = ppsl_model.backward_simulation(flt)
+            suffStats = smp.evaluate_sufficient_statistics(x)
+            
+            # Pad the transition covariance
+            padded_Q = ppsl_model.transition_covariance() + \
+                               self.algoparams['Fs']*np.identity(ppsl_model.ds)
+            
+            # Sample a new transition matrix
+            ppsl_F,fwd_prob = smp.sample_basic_transition_matrix_mniw_conditional(
+                                            suffStats,
+                                            padded_Q,
+                                            self.hyperparams['M0'],
+                                            self.hyperparams['V0'],
+                                            with_pdf=True)
+            ppsl_model.parameters['F'] = ppsl_F
+            
+            # Sample a new trajectory
+            ppsl_x = ppsl_model.sample_posterior(self.observ)
+            ppsl_suffStats = smp.evaluate_sufficient_statistics(ppsl_x)
+            
+            # Reverse move probaility
+            _,bwd_prob = smp.sample_basic_transition_matrix_mniw_conditional(
+                                            ppsl_suffStats,
+                                            padded_Q,
+                                            self.hyperparams['M0'],
+                                            self.hyperparams['V0'],
+                                            F = self.model.parameters['F'],
+                                            with_pdf=True)
+            ppsl_model.parameters['F'] = ppsl_F
+            
+            # Prior terms
+            prior = self.transition_prior(self.model)
+            ppsl_prior = self.transition_prior(ppsl_model)
         
         elif moveType=='Q':
             

### 2015-08-27 linear_models.py

@@ -173,7 +173,7 @@ class AbstractLinearModel:
         
         # Get system matrices
         F = self.transition_matrix()
-        Q = self.transition_covariance()
+        Q = self.transition_covariance() + 1E-10*np.identity(self.ds)
         
         # Initialise sampled sequence
         num_time_instants = flt.num_time_instants

### 2015-08-27 run_basic_learning.py

@@ -31,7 +31,7 @@ state, observ = model.simulate_data(K)
 est_params = dict(params)
 est_params['F'] = 0.5*np.identity(ds)
 est_params['rank'] = 2
-est_params['vec'] = np.array([[1,0],[0,1],[0,0]])
+est_params['vec'] = params['vec']#np.array([[1,0],[0,1],[0,0]])
 est_params['val'] = np.array([1,1])
 est_model = DegenerateLinearModel(ds, do, prior, est_params)
 
@@ -44,6 +44,7 @@ hyperparams['alpha'] = 100
 
 algoparams = dict()
 algoparams['Qs'] = 0.1
+algoparams['Fs'] = 0.1
 
 learner = MCMCLearnerForDegenerateModelWithMNIWPrior(est_model, observ, hyperparams, algoparams=algoparams, verbose=True)
 
@@ -53,7 +54,10 @@ num_burn = int(num_iter/5)
 for ii in range(num_iter):
     print("Running iteration {} of {}.".format(ii+1,num_iter))
     
-    learner.iterate_transition('Q')
+    if (ii%1)==0:
+        learner.iterate_transition('F')
+    else:
+        learner.iterate_transition('Q')
     learner.save_link()
 
 learner.plot_chain_trace('F', numBurnIn=num_burn)

### 2015-08-27 sampling.py

@@ -115,6 +115,27 @@ def sample_basic_transition_mniw_conditional(suffStats, nu0, Psi0, M0, V0):
 
     return F, Q
 
+def sample_basic_transition_matrix_mniw_conditional(suffStats, Q, M0, V0, F=None, with_pdf=False):
+    """
+    Sample transition matrix from matrix-normal posterior conditional
+    distribution.
+    """
+    invV0 = la.inv(V0)
+    
+    # Posterior hyperparameters    
+    V   = la.inv( invV0 + suffStats[1] )
+    M   = np.dot( np.dot(M0,invV0)+suffStats[2] , V)
+    
+    # Sample
+    if F is None:
+        F = sample_matrix_normal(M, Q, V)
+
+    if not with_pdf:
+        return F
+    else:
+        pdf = matrix_normal_density(F, M, Q, V)
+        return F, pdf
+
 
 
 def sample_degenerate_transition_mniw_conditional(suffStats, U, Fold, nu0, Psi0, M0, V0):

### 2015-08-27 learners_mcmc.py
Removed independent prior learning

@@ -177,41 +177,6 @@ class MCMCLearnerForBasicModelWithMNIWPrior(AbstractMCMCLearner):
                                                     self.hyperparams['M0'],
                                                     self.hyperparams['V0'])
                                                     
-class MCMCLearnerForBasicModelWithIndependentPriors(AbstractMCMCLearner):
-    """
-    Container for MCMC system learning algorithm.
-    Model Type: Basic
-    Prior Type: Independent Matrix Normal and Inverse Wishart
-    """
-        
-    def iterate_transition(self, flt):
-        """
-        MCMC iteration (Gibbs sampling) for transition matrix and covariance
-        """
-        
-        # First sample the state sequence
-        x = self.model.sample_posterior(self.observ)
-        
-        # Calculate sufficient statistics
-        suffStats = smp.evaluate_sufficient_statistics(x)
-        
-        # Sample a new transition covariance
-        self.model.parameters['Q'] = \
-             smp.sample_basic_transition_covariance_independent_conditional(
-                                                suffStats,
-                                                self.hyperparams['nu0'],
-                                                self.hyperparams['Psi0'],
-                                                self.model.parameters['F'])
-        
-        # Sample a new transition matrix
-        self.model.parameters['F'] = \
-             smp.sample_basic_transition_matrix_independent_conditional(
-                                                suffStats,
-                                                self.hyperparams['M0'],
-                                                self.hyperparams['alpha'],
-                                                self.model.parameters['Q'])
-
-
 
 class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
     """

### 2015-08-27 sampling.py

@@ -168,55 +168,6 @@ def sample_degenerate_transition_mniw_conditional(suffStats, U, Fold, nu0, Psi0,
 
 
 
-def sample_basic_transition_matrix_independent_conditional(suffStats, M0, 
-                                                                     alpha, Q):
-    """
-    Sample transition matrix from posterior conditional distribution arising 
-    from isotropic matrix normal prior
-    """
-    ds = Q.shape[0]
-    
-    # Eigendecompositions
-    QeVal,QeVec = la.eigh(Q)
-    ss1eVal, ss1eVec = la.eigh(suffStats[1])
-    
-    # Hyperparameters of rotated transition matrix
-    ss2rot = np.dot(np.dot(QeVec.T, suffStats[2]), ss1eVec)
-    M0rot = np.dot(np.dot(QeVec.T, M0), ss1eVec)
-    P = 1./( 1./alpha + np.kron(1./QeVal, ss1eVal) )
-    m = P*( (np.dot(np.diag(1./QeVal), ss2rot)+M0rot/alpha).flatten() )
-    
-    # Sample rotated, flattened transition matrix
-    Frotflat = stats.norm.rvs(loc=m,scale=np.sqrt(P))
-    
-    # Reshape into a matrix
-    Frot = Frotflat.reshape((ds,ds))
-    
-    # Rotate to get the final transition matrix
-    F = np.dot(np.dot(QeVec, Frot), ss1eVec.T)
-    
-    return F
-
-
-def sample_basic_transition_covariance_independent_conditional(suffStats, nu0,
-                                                                      Psi0, F):
-    """
-    Sample transition covariance from posterior conditional distribution
-    arising from inverse wishart prior
-    """
-    
-    # Update covariance hyperparameters
-    nu  = nu0 + suffStats[0]
-    Psi = Psi0 + suffStats[3] - np.dot(F,suffStats[2].T) \
-                - np.dot(suffStats[2],F.T) + np.dot(np.dot(F,suffStats[1]),F.T)
-    
-    # Sample
-    Q = la.inv(sample_wishart(nu, la.inv(Psi)))
-    
-    return Q
-    
-
-
 
 
 
[…]

### 2015-08-27 test_basiclearning.py

@@ -67,26 +67,3 @@ class MNIWPriorLearning(BasicModelLearningTestCase):
         learner.plot_chain_histogram('F', numBurnIn=self.num_burn, trueValue=self.params['F'])
         learner.plot_chain_histogram('Q', numBurnIn=self.num_burn, trueValue=self.params['Q'])
         
-        
-class IndependentPriorsLearning(BasicModelLearningTestCase):
-    def runTest(self):
-        
-        # Set up learning
-        hyperparams = dict()
-        hyperparams['nu0'] = 3.01
-        hyperparams['Psi0'] = (hyperparams['nu0']-self.ds-1)*np.identity(self.ds)
-        hyperparams['M0'] = np.zeros(self.ds)
-        hyperparams['alpha'] = 100
-        learner = MCMCLearnerForBasicModelWithIndependentPriors(self.est_model, self.observ, hyperparams)
-        
-        # MCMC
-        for ii in range(self.num_iter):
-            print("Running iteration {} of {}.".format(ii+1,self.num_iter))
-            learner.iterate_transition()
-            learner.save_link()
-        
-        learner.plot_chain_trace('F', numBurnIn=self.num_burn)
-        learner.plot_chain_trace('Q', numBurnIn=self.num_burn)
-        
-        learner.plot_chain_histogram('F', numBurnIn=self.num_burn, trueValue=self.params['F'])
-        learner.plot_chain_histogram('Q', numBurnIn=self.num_burn, trueValue=self.params['Q'])

### 2015-08-27 learners_mcmc.py
Refactored sampling.py

@@ -169,14 +169,21 @@ class MCMCLearnerForBasicModelWithMNIWPrior(AbstractMCMCLearner):
         # Calculate sufficient statistics
         suffStats = smp.evaluate_sufficient_statistics(x)
         
-        # Sample a new transition matrix and transition covariance
-        self.model.parameters['F'], self.model.parameters['Q'] = \
-             smp.sample_basic_transition_mniw_conditional(suffStats,
+        # Update hyperparameters
+        nu,Psi,M,V = smp.hyperparam_update_basic_mniw_transition(suffStats,
                                                     self.hyperparams['nu0'],
                                                     self.hyperparams['Psi0'],
                                                     self.hyperparams['M0'],
                                                     self.hyperparams['V0'])
-                                                    
+        
+        # Sample new parameters
+        Q = la.inv(smp.sample_wishart(nu, la.inv(Psi)))
+        F = smp.sample_matrix_normal(M, Q, V)
+        
+        # Update the model
+        self.model.parameters['F'] = F
+        self.model.parameters['Q'] = Q
+                      
 
 class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
     """
@@ -226,12 +233,12 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
                                self.algoparams['Fs']*np.identity(ppsl_model.ds)
             
             # Sample a new transition matrix
-            ppsl_F,fwd_prob = smp.sample_basic_transition_matrix_mniw_conditional(
-                                            suffStats,
-                                            padded_Q,
-                                            self.hyperparams['M0'],
-                                            self.hyperparams['V0'],
-                                            with_pdf=True)
+            M,V = smp.hyperparam_update_basic_mniw_transition_matrix(
+                                                    suffStats,
+                                                    self.hyperparams['M0'],
+                                                    self.hyperparams['V0'],)
+            ppsl_F = smp.sample_matrix_normal(M,padded_Q,V)
+            fwd_prob = smp.matrix_normal_density(ppsl_F,M,padded_Q,V)
             ppsl_model.parameters['F'] = ppsl_F
             
             # Sample a new trajectory
@@ -239,14 +246,12 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
             ppsl_suffStats = smp.evaluate_sufficient_statistics(ppsl_x)
             
[…]

### 2015-08-27 run_basic_learning.py

@@ -54,7 +54,7 @@ num_burn = int(num_iter/5)
 for ii in range(num_iter):
     print("Running iteration {} of {}.".format(ii+1,num_iter))
     
-    if (ii%1)==0:
+    if (ii%2)==0:
         learner.iterate_transition('F')
     else:
         learner.iterate_transition('Q')

### 2015-08-27 sampling.py

@@ -93,75 +93,61 @@ def evaluate_sufficient_statistics(x):
         suffStats[3] += np.outer(x[kk], x[kk])
     
     return suffStats
-    
 
-
-def sample_basic_transition_mniw_conditional(suffStats, nu0, Psi0, M0, V0):
+def hyperparam_update_basic_mniw_transition(suffStats, nu0, Psi0, M0, V0):
     """
-    Sample transition model matrices from matrix-normal-inverse-wishart
-    posterior conditional distribution.
+    Update matrix-normal-inverse-wishart hyperparameters for transition model
+    matrices conditional on observed state trajectory.
     """
-    invV0 = la.inv(V0)
     
     # Posterior hyperparameters    
     nu  = nu0 + suffStats[0]
-    V   = la.inv( invV0 + suffStats[1] )
-    M   = np.dot( np.dot(M0,invV0)+suffStats[2] , V)
-    Psi = Psi0 + suffStats[3] - np.dot(M, la.solve(V,M.T) ) + np.dot(M0, la.solve(V0,M0.T) )
+    V   = la.inv( la.inv(V0) + suffStats[1] )
+    M   = np.dot( la.solve(V0,M0.T).T + suffStats[2] , V)
+    Psi = Psi0 + suffStats[3] - np.dot(M, la.solve(V,M.T) ) \
+                                               + np.dot(M0, la.solve(V0,M0.T) )
     
-    # Sample
-    Q = la.inv(sample_wishart(nu, la.inv(Psi)))
-    F = sample_matrix_normal(M, Q, V)
+    return nu, Psi, M, V
 
-    return F, Q
-
-def sample_basic_transition_matrix_mniw_conditional(suffStats, Q, M0, V0, F=None, with_pdf=False):
+def hyperparam_update_basic_mniw_transition_matrix(suffStats, M0, V0):
     """
-    Sample transition matrix from matrix-normal posterior conditional
-    distribution.
+    Update matrix-normal hyperparameters for transition matrix conditional
+    on observed state trajectory.
     """
-    invV0 = la.inv(V0)
     
     # Posterior hyperparameters    
-    V   = la.inv( invV0 + suffStats[1] )
-    M   = np.dot( np.dot(M0,invV0)+suffStats[2] , V)
+    V   = la.inv( la.inv(V0) + suffStats[1] )
+    M   = np.dot( la.solve(V0,M0.T).T + suffStats[2] , V)
     
-    # Sample
[…]

### 2015-08-27 test_basiclearning.py

@@ -4,7 +4,7 @@ import numpy as np
 
 from kalman import GaussianDensity
 from linear_models import BasicLinearModel
-from learners_mcmc import MCMCLearnerForBasicModelWithMNIWPrior, MCMCLearnerForBasicModelWithIndependentPriors
+from learners_mcmc import MCMCLearnerForBasicModelWithMNIWPrior
 
 class BasicModelLearningTestCase(unittest.TestCase):
     def setUp(self):

### 2015-08-28 learners_mcmc.py
Removed independent prior

@@ -273,6 +273,35 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
             
         elif moveType=='rank':
             pass
+            
+            
+            
+            
+            #TODO ADD THIS STEP
+            
+            
+#            # Add or remove?
+#            if ppsl_model.ds==1:
+#                u = -1                          # 1D model. Cannot do anything
+#            elif ppsl_model.rank==1:
+#                u = 0
+#            elif ppsl_model.rank==ppsl_model.ds:
+#                u = 1
+#            else:
+#                u = int(np.round(np.random.random()))
+#            
+#            if u==0:
+#                
+#                # Increase rank
+#                dcf = ppsl_model.increase_rank(None, None, self.Q_prior_scale)
+#                
+#            elif u==1:
+#                # Decrease rank
+#                dcf,val,vec = ppsl_model.reduce_rank(self.Q_prior_scale)            
+            
+            
+            
+            
         else:
             raise ValueError("Invalid move type")
         
@@ -336,37 +365,6 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
     
     
     
-
-#class MCMCLearnerForDegenerateModelWithIndependentPriors(AbstractMCMCLearner):
-#    """
-#    Container for MCMC system learning algorithm.
-#    Model Type: Degenerate
-#    Prior Type: Independent Matrix Normal and Singular Inverse Wishart
-#    """
-#        
-#    def iterate_transition(self):
-#        """
-#        MCMC iteration (Gibbs sampling) for transition matrix and covariance
-#        """
-#        
-#        # First sample the state sequence
-#        x = self.model.sample_posterior(self.observ)
-#        
-#        # Convert to Givens factorisation form
-#        U,D = self.model.convert_to_givens_form()
-#        
-#        # Sample a new transition matrix and transition covariance
-#        self.model.parameters['F'], D = \
-#             smp.sample_degenerate_transition_independent_conditional(x, U,
-#                                                self.hyperparams['psi0'],
-#                                                self.hyperparams['M0'],
-#                                                self.hyperparams['alpha'],
-#                                                F=self.model.parameters['F'])
-#        
-#        # Convert back to eigen-decomposition form
-#        self.model.update_from_givens_form(U, D)
-
-
 
 
 

### 2015-08-30 notes/notes-linear-system-learning.tex
Changed code to F,Q formulation

--- 2015-08-27 notes/notes-linear-system-learning.tex
@@
 \usepackage[margin=1in]{geometry}
+\usepackage{color}
 
@@
 Second, the conjugate prior is tricky.
-The matrix-normal-inverse-wishart can be generalised to the case where $\lgtv$ is not full rank (using singular wishart and singular matrix normal distributions).
+The matrix-normal-inverse-wishart can be generalised to the case where $\lgtv$ is not full rank (using singular inverse Wishart and singular matrix normal distributions).
 However, this is not helpful, because it constrains the transition matrix to lie in a linear subspace.
 
-Because Wishart distributions are somewhat more studied than their inverse counterparts, it will be easier in the following section to work with the precision, defined as $\lgtp = (\lgdm\lgdm\tr)\pinv$.
-We will need the eigen-decomposition of this frequently.
+We will need the eigen-decomposition of $\lgtv$ frequently.
 
 \begin{IEEEeqnarray}{rCl}
- \lgtp &=& \tpvec \tpval \tpvec\tr
+ \lgtv &=& \tvvec \tvval \tvvec\tr
 \end{IEEEeqnarray}
-This is the ``non-singular'' factorisation, where $\tpval$ is a diagonal $\rk\times\rk$ matrix with $\rk = \rank(\lgtp)$, and $\tpvec$ from the appropriate Stiefel manifold.
+This is the ``non-singular'' part of the factorisation, where $\tvval$ is a diagonal $\rk\times\rk$ matrix with $\rk = \rank(\lgtv)$, and $\tvvec$ from the appropriate Stiefel manifold.
 The full decomposition will also be useful at times.
@@
 \begin{IEEEeqnarray}{rCl}
- \lgtp &=& \begin{bmatrix}\tpvec & \tpvecorth \end{bmatrix} \begin{bmatrix}\tpval & \zmat \\ \zmat & \zmat \end{bmatrix} \begin{bmatrix}\tpvec\tr \\ \tpvecorth\tr \end{bmatrix}
+ \lgtv &=& \begin{bmatrix}\tvvec & \tvvecorth \end{bmatrix} \begin{bmatrix}\tvval & \zmat \\ \zmat & \zmat \end{bmatrix} \begin{bmatrix}\tvvec\tr \\ \tvvecorth\tr \end{bmatrix}
 \end{IEEEeqnarray}
@@
 
-\subsubsection{Wishart Part}
+\subsubsection{Inverse Wishart Part}
 
-There are two ways to extend the Wishart distribution to positive semi-definite matrices.
-Either let $\priordof \in \{z \in \naturals | z < \lsd\}$, in which case $\rank(\lgtp) = \priordof$, or let $\priorscalematrix$ be positive semi-definite.
+There are two ways to extend the inverse Wishart distribution to positive semi-definite matrices.
+Either let $\priordof \in \{z \in \naturals | z < \lsd\}$, in which case $\rank(\lgtv) = \priordof$, or let $\priorscalematrix$ be positive semi-definite.
 The latter makes it tricky to work out what we should set $\priorscalematrix$ to, so we'll go with the former.
-If we condition on a known value of the rank $r$, then the singular Wishart density is,
+If we set $\priordof$ equal to a known value of the rank $\rk$, then the singular Wishart density is,
 
 \begin{IEEEeqnarray}{rCl}
- \den(\lgtp) &=& \frac{ \pi^{\half\rk(\lsd-\rk)} \determ{\priorscalematrix}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \gammafun[r]\left(\frac{\rk}{2}\right) } \determ{\tpval}^{\half(\rk-\lsd-1)} \exp\left( -\half \lgtp \priorscalematrix \right)    .
+ \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrix}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrix \right)    .
 \end{IEEEeqnarray}
 
-\cite{Uhlig1994,Diaz-Garcia1997}
+\begin{itemize}
+ \item \cite{Muirhead1982} covers the basic multivariate statistics for the full-rank case.
+ \item \cite{Uhlig1994} covers singular Wishart distributions, including the eigenvalue distribution, and base measure for positive semi-definite matrices.
+ \item \cite{Diaz-Garcia1997} covers singular and pseudo Wishart distributions in a thorough and unifying manner.
+ \item \cite{Diaz-Garcia2006} covers singular and pseudo inverse Wishart distributions.
+ \item \cite{Bodnar2008} contains a few useful properties.
+
+\end{itemize}
 
@@
 \begin{IEEEeqnarray}{rCl}
- \lgtm|\lgtp &\sim& \matrixnormaldist{\priormeanmatrix}{\lgtp\pinv+\tpvecorth\priorextratpval\inv\tpvecorth\tr}{\priorcolumnvariance}
+ \lgtm|\lgtv &\sim& \matrixnormaldist{\priormeanmatrix}{\lgtv+\tvvecorth\priorextratpval\tvvecorth\tr}{\priorcolumnvariance}
 \end{IEEEeqnarray}
@@
 
-A sampled state trajectory $\ls{1:\timax}$ defines the subspace in which the state perturbations must lie, which means that neither $\lgtm$ nor $\lgtp$ can be freely changed, and Gibbs sampling alone does not work.
+A sampled state trajectory $\ls{1:\timax}$ defines the subspace in which the state perturbations must lie, which means that neither $\lgtm$ nor $\lgtv$ can be freely changed, and Gibbs sampling alone does not work.
 However, we can still sample within the space specified by this constraint.
@@
 \begin{IEEEeqnarray}{rCl}
- \begin{bmatrix} \tpvec & \tpvecorth \end{bmatrix} &=& \begin{bmatrix}\tprot & \tprotorth\end{bmatrix}\begin{bmatrix} \tpsign & \zmat \\ \zmat & \tpsignorth \end{bmatrix} \begin{bmatrix} \tprow & \zmat \\ \zmat & \idmat \end{bmatrix}\begin{bmatrix} \idmat & \zmat \\ \zmat & \tpnull \end{bmatrix}
+ \begin{bmatrix} \tvvec & \tvvecorth \end{bmatrix} &=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix}\begin{bmatrix} \tvsign & \zmat \\ \zmat & \tvsignorth \end{bmatrix} \begin{bmatrix} \tvrow & \zmat \\ \zmat & \idmat \end{bmatrix}\begin{bmatrix} \idmat & \zmat \\ \zmat & \tvnull \end{bmatrix}
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- \tpvec     &=& \tprot     \tpsign     \tprow  \\
- \tpvecorth &=& \tprotorth \tpsignorth \tpnull
+ \tvvec     &=& \tvrot     \tvsign     \tvrow  \\
+ \tvvecorth &=& \tvrotorth \tvsignorth \tvnull
 \end{IEEEeqnarray}
@@
 \begin{equation}
- \tpfull = \tpsign \tprow \tpval \tprow\tr \tpsign\tr
+ \tvfull = \tvsign \tvrow \tvval \tvrow\tr \tvsign\tr
 \end{equation}
@@
 \begin{IEEEeqnarray}{rCl}
- \lgtp &=& \tprot \tpfull \tprot\tr
+ \lgtv &=& \tvrot \tvfull \tvrot\tr
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- \ls{\ti} &=& \lgtm \ls{\ti-1} + \tprot \tpfull^{-\half} \tn{\ti} \\
+ \ls{\ti} &=& \lgtm \ls{\ti-1} + \tvrot \tvfull\msqrt \tn{\ti} \\
  \end{IEEEeqnarray}
  \begin{IEEEeqnarray}{rCl}
- \Rightarrow \tprot\tr(\ls{\ti}-\lgtm\ls{\ti-1}) &\sim& \normaldist{\zmat}{\tpfull\inv}
+ \Rightarrow \tvrot\tr(\ls{\ti}-\lgtm\ls{\ti-1}) &\sim& \normaldist{\zmat}{\tvfull}
 \end{IEEEeqnarray}
 \begin{IEEEeqnarray}{rCl}
- \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) &\propto& \determ{\tpfull}^{\half} \exp\left(-\half (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tprot \tpfull \tprot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right)
+ \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) &\propto& \determ{\tvfull}^{-\half} \exp\left(-\half (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right)
 \end{IEEEeqnarray}
@@
  \den(\ls{1:\timax}|\lgtm, \lgtv) &= \den(\ls{0}) \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) \nonumber \\
- &\propto \determ{\tpfull}^{\half(\timax-1)} \exp\left( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tprot \tpfull \tprot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber \\
- &= \determ{\tpfull}^{\half(\timax-1)} \exp\left(-\half\trace\left[ \tpfull \tprot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tprot \right]\right) \nonumber \\
- &= \determ{\tpfull}^{\half\suffstats{0}} \exp\left(-\half\trace\left[ \tpfull \left( \tprot\tr \lgtm \suffstats{1} \lgtm\tr \tprot - \tprot\tr \lgtm \suffstats{2}\tr \tprot - \tprot\tr \suffstats{2} \lgtm\tr \tprot + \tprot\tr \suffstats{3} \tprot \right) \right]\right) \\
- &= \determ{\tpfull}^{\half\suffstats{0}} \exp\left(-\half\trace\left[ \tpfull \left( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tprot - \tprot\tr \suffstats{2} \lgtmrot\tr + \tprot\tr \suffstats{3} \tprot \right) \right]\right)
+ &\propto \determ{\tvfull}^{-\half(\timax-1)} \exp\left( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber \\
+ &= \determ{\tvfull}^{-\half(\timax-1)} \exp\left(-\half\trace\left[ \tvfull\inv \tvrot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \right]\right) \nonumber \\
+ &= \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \tvrot\tr \lgtm \suffstats{1} \lgtm\tr \tvrot - \tvrot\tr \lgtm \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtm\tr \tvrot + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right) \\
+ &= \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtmrot\tr + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right)
 \end{align}
@@
 \begin{equation}
- \lgtmrot = \tprot\tr \lgtm
+ \lgtmrot = \tvrot\tr \lgtm
 \end{equation}
@@
 \begin{IEEEeqnarray}{rCl}
- \LS{2:\timax} &=& \lgtm \LS{1:\timax-1} + \tprot Z
+ \LS{2:\timax} &=& \lgtm \LS{1:\timax-1} + \tvrot Z
 \end{IEEEeqnarray}
@@
 
-If we transform from $(\lgtm,\lgtp)$ to $(\lgtmrot,\tpfull)$, then the prior on these variables is as follows.
+If we transform from $(\lgtm,\lgtv)$ to $(\lgtmrot,\tvfull)$, then the prior on these variables is as follows.
 
@@
 \begin{IEEEeqnarray}{rCl}
- \lgtmrot|\lgtp = \lgtmrot|\tprot,\tpfull &\sim& \matrixnormaldist{\tprot\tr\priormeanmatrix}{\tpfull\inv}{\priorcolumnvariance}
+ \lgtmrot|\lgtv = \lgtmrot|\tvrot,\tvfull &\sim& \matrixnormaldist{\tvrot\tr\priormeanmatrix}{\tvfull}{\priorcolumnvariance}
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- \tpfull|\tprot &\sim& \wishartdist{\rk}{\tprot\tr\priorscalematrix\tprot}
+ \tvfull|\tvrot &\sim& \iwishartdist{\rk}{(\tvrot\tr\priorscalematrix\inv\tvrot)\inv}
 \end{IEEEeqnarray}
@@
 
-We can now get to the conditional distribution for $\den(\lgtmrot, \tpfull|\tprot,\ls{1:\timax})$ which is well-behaved and can be sampled.
+We can now get to the conditional distribution for $\den(\lgtmrot, \tvfull|\tvrot,\ls{1:\timax})$ which is well-behaved and can be sampled.
 
@@
 \begin{equation}
- \den(\lgtmrot, \tpfull | \tprot, \ls{1:\timax}, \ob{1:\timax}) =  \den(\lgtmrot | \tpfull, \tprot, \ls{1:\timax}) \den(\tpfull | \tprot, \ls{1:\timax})
+ \den(\lgtmrot, \tvfull | \tvrot, \ls{1:\timax}, \ob{1:\timax}) =  \den(\lgtmrot | \tvfull, \tvrot, \ls{1:\timax}) \den(\tvfull | \tvrot, \ls{1:\timax})
 \end{equation}
 \begin{align}
- \tpfull | \tprot, \ls{1:\timax} &\sim \wishartdist{\postdof}{\postscalematrix} \\
- \lgtmrot | \tpfull, \tprot, \ls{1:\timax} &\sim \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}
+ \tvfull | \tvrot, \ls{1:\timax} &\sim \iwishartdist{\postdof}{\postscalematrix} \\
+ \lgtmrot | \tvfull, \tvrot, \ls{1:\timax} &\sim \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}
 \end{align}
@@
  \postcolumnvariance\inv                 &= \priorcolumnvariance\inv + \suffstats{1} \\
- \postmeanmatrix \postcolumnvariance\inv &= \tprot\tr\left( \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \right) \\
+ \postmeanmatrix \postcolumnvariance\inv &= \tvrot\tr\left( \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \right) \\
  \postdof                                &= \rk + \suffstats{0} \\
- \postscalematrix                        &= \tprot\tr\left( \priorscalematrix + \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr \right)\tprot - \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr
+ \postscalematrix                        &= (\tvrot\tr\priorscalematrix\inv\tvrot)\inv + \tvrot\tr\left( \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr \right)\tvrot - \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr
 \end{align}
@@
 \begin{IEEEeqnarray}{rCl}
- \den(\lgtmrot, \tpfull|\tprot,\ls{1:\timax}) &\propto& \den(\ls{1:\timax}|\lgtmrot, \tpfull, \tprot) \den(\lgtmrot | \tpfull, \tprot) \den(\tpfull|\tprot) \nonumber \\
+ \den(\lgtmrot, \tvfull|\tvrot,\ls{1:\timax}) &\propto& \den(\ls{1:\timax}|\lgtmrot, \tvfull, \tvrot) \den(\lgtmrot | \tvfull, \tvrot) \den(\tvfull|\tvrot) \nonumber \\
 %
- &\propto& \determ{\tpfull}^{\half\suffstats{0}} \exp\left(-\half\trace\left[ \tpfull \left( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tprot - \tprot\tr \suffstats{2} \lgtmrot\tr + \tprot\tr \suffstats{3} \tprot \right) \right]\right) \nonumber \\
- & & \times \determ{\tpfull}^{\frac{r-r-1}{2}} \exp\left( -\half \trace\left[\tpfull \tprot\tr\priorscalematrix\tprot\right] \right) \nonumber \\
- & & \times \determ{\tpfull}^{\half} \exp\left(-\half \trace\left[ \tpfull (\lgtmrot-\tprot\tr\priormeanmatrix) \priorcolumnvariance\inv (\lgtmrot-\tprot\tr\priormeanmatrix)\tr \right] \right) \nonumber \\
+ &\propto& \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtmrot\tr + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right) \nonumber \\
+ & & \times \determ{\tvfull}^{-\half(r+r+1)} \exp\left( -\half \trace\left[\tvfull\inv (\tvrot\tr\priorscalematrix\inv\tvrot)\inv \right] \right) \nonumber \\
+ & & \times \determ{\tvfull}^{-\half} \exp\left(-\half \trace\left[ \tvfull\inv (\lgtmrot-\tvrot\tr\priormeanmatrix) \priorcolumnvariance\inv (\lgtmrot-\tvrot\tr\priormeanmatrix)\tr \right] \right) \nonumber \\
 %
- &=& \determ{\tpfull}^{\frac{(r+\suffstats{0})-r-1}{2}} \exp\left( -\half \trace\left[\tpfull \tprot\tr(\priorscalematrix+\suffstats{3})\tprot\right] \right) \nonumber \\
- & & \times \determ{\tpfull}^{\half} \exp\Bigg(-\half \trace\bigg[ \tpfull \bigg( \lgtmrot (\priorcolumnvariance\inv + \suffstats{1}) \lgtmrot\tr - \lgtmrot(\priorcolumnvariance\inv\priormeanmatrix\tr + \suffstats{2}\tr) \tprot \nonumber \\
- & & \qquad\qquad\qquad\qquad\qquad  -\: \tprot\tr (\priormeanmatrix\priorcolumnvariance\inv + \suffstats{2}) \lgtmrot\tr + \tprot\tr \priormeanmatrix\priorcolumnvariance\inv\priormeanmatrix\tr \tprot \bigg) \bigg] \Bigg)
+ &=& \determ{\tvfull}^{-\half\left((r+\suffstats{0})+r+1\right)} \exp\left( -\half \trace\left[\tvfull\inv \left( (\tvrot\tr\priorscalematrix\inv\tvrot)\inv + \tvrot\tr\suffstats{3}\tvrot \right)\right] \right) \nonumber \\
+ & & \times \determ{\tvfull}^{-\half} \exp\Bigg(-\half \trace\bigg[ \tvfull\inv \bigg( \lgtmrot (\priorcolumnvariance\inv + \suffstats{1}) \lgtmrot\tr - \lgtmrot(\priorcolumnvariance\inv\priormeanmatrix\tr + \suffstats{2}\tr) \tvrot \nonumber \\
+ & & \qquad\qquad\qquad\qquad\qquad  -\: \tvrot\tr (\priormeanmatrix\priorcolumnvariance\inv + \suffstats{2}) \lgtmrot\tr + \tvrot\tr \priormeanmatrix\priorcolumnvariance\inv\priormeanmatrix\tr \tvrot \bigg) \bigg] \Bigg)
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- \LS{2:\timax} &=& \lgtm\mcold \LS{1:\timax-1} + \tprot Z\mcold \nonumber \\
- \LS{2:\timax} &=& \lgtm \LS{1:\timax-1} + \tprot Z \nonumber \\
- \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) \LS{1:\timax-1} + \tprot(Z-Z\mcold) \\
+ \LS{2:\timax} &=& \lgtm\mcold \LS{1:\timax-1} + \tvrot Z\mcold \nonumber \\
+ \LS{2:\timax} &=& \lgtm \LS{1:\timax-1} + \tvrot Z \nonumber \\
+ \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) \LS{1:\timax-1} + \tvrot(Z-Z\mcold) \\
  \Rightarrow \zmat &=& (\lgtmrot-\lgtmrot\mcold) \LS{1:\timax-1} + (Z-Z\mcold) \\
- \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) \LS{1:\timax-1} - \tprot (\lgtmrot-\lgtmrot\mcold) \LS{1:\timax-1} \\
- \zmat &=& \left[(\lgtm-\lgtm\mcold) - \tprot (\lgtmrot-\lgtmrot\mcold)\right] \LS{1:\timax-1} \\
- \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) - \tprot (\lgtmrot-\lgtmrot\mcold)
+ \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) \LS{1:\timax-1} - \tvrot (\lgtmrot-\lgtmrot\mcold) \LS{1:\timax-1} \\
+ \zmat &=& \left[(\lgtm-\lgtm\mcold) - \tvrot (\lgtmrot-\lgtmrot\mcold)\right] \LS{1:\timax-1} \\
+ \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) - \tvrot (\lgtmrot-\lgtmrot\mcold)
 \end{IEEEeqnarray}
 \begin{IEEEeqnarray}{rCl}
- \lgtm &=& \lgtm\mcold + \tprot (\lgtmrot-\lgtmrot\mcold) \nonumber \\
-       &=& \lgtm\mcold + \tprot (\lgtmrot-\tprot\tr\lgtm\mcold) \nonumber \\
-       &=& (\idmat - \tprot\tprot\tr) \lgtm\mcold + \tprot \lgtmrot 
+ \lgtm &=& \lgtm\mcold + \tvrot (\lgtmrot-\lgtmrot\mcold) \nonumber \\
+       &=& \lgtm\mcold + \tvrot (\lgtmrot-\tvrot\tr\lgtm\mcold) \nonumber \\
+       &=& (\idmat - \tvrot\tvrot\tr) \lgtm\mcold + \tvrot \lgtmrot 
 \end{IEEEeqnarray}
 
-\subsection{Metropolis-Hastings for the Precision Subspace}
+\subsubsection{Metropolis-Hastings for the Covariance Subspace}
 
-In order to change $\tprot$, we can target $\den(\lgtp|\lgtm, \ob{1:\timax})$ with Metropolis-Hastings, which constitutes a collapsed Gibbs move.
+In order to change $\tvrot$, we can target $\den(\lgtv|\lgtm, \ob{1:\timax})$ with Metropolis-Hastings, which is justified as a collapsed Gibbs move.
 
@@
  \begin{bmatrix}
-  \lgtp\mcnew \\ \mhrot\mcnew
+  \lgtv\mcnew \\ \mhrot\mcnew
  \end{bmatrix}
@@
  \begin{bmatrix}
-  \mhrot\lgtp\mhrot\tr \\ \mhrot\tr
+  \mhrot\lgtv\mhrot\tr \\ \mhrot\tr
  \end{bmatrix} \nonumber     ,
@@
 \begin{IEEEeqnarray}{rCl}
- \mhap(\lgtp\to\lgtp\mcnew) & = & \min\left\{1, \frac{ \den(\lgtp\mcnew|\lgtm,\ob{1:\timax})\mhrotppsl(\mhrot\mcnew) }{ \den(\lgtp|\lgtm, \ob{1:\timax})\mhrotppsl(\mhrot) } \right\}  \\
- & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtp\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtp)} \times \frac{\den(\lgtp\mcnew, \lgtm) }{\den(\lgtp, \lgtm)} \times \frac{\mhrotppsl(\mhrot\mcnew)}{\mhrotppsl(\mhrot)} \right\} \nonumber     .
+ \mhap(\lgtv\to\lgtv\mcnew) & = & \min\left\{1, \frac{ \den(\lgtv\mcnew|\lgtm,\ob{1:\timax})\mhrotppsl(\mhrot\mcnew) }{ \den(\lgtv|\lgtm, \ob{1:\timax})\mhrotppsl(\mhrot) } \right\}  \\
+ & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtv\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtv)} \times \frac{\den(\lgtv\mcnew, \lgtm) }{\den(\lgtv, \lgtm)} \times \frac{\mhrotppsl(\mhrot\mcnew)}{\mhrotppsl(\mhrot)} \right\} \nonumber     .
 \end{IEEEeqnarray}
@@
 
+\subsubsection{Metropolis-Hastings for the Transition Matrix}
+
+Define $\postden$ as the posterior density, conditioning on $\ob{1:\timax}$.
+
+If we modify the covariance matrix slightly so that it is no longer singular, then we can use the existing sampling procedure from the basic linear model and treat the resulting value of $\lgtm$ as a proposal in a Metropolis-Hastings kernel targeting $\postden(\lgtm, \ls{1:\timax}|\lgtv)$.
+
+Suppose we start with sampled values of $\lgtm$,$\lgtv$ and $\ls{1:\timax}$. A suitably modified non-singular covariance matrix is,
+%
+\begin{IEEEeqnarray}{rCl}
+ \paddedlgtv & = & \lgtv + \padding \idmat \label{eq:padded_transition_covariance}      ,
+\end{IEEEeqnarray}
+%
+where $\padding$ is a small positive constant. We propose a new value of $\lgtm$ by sampling,
+%
+\begin{IEEEeqnarray}{rCl}
+ \lgtm' & \sim & \postden(\lgtm|\paddedlgtv,\ls{1:\timax})     .
+\end{IEEEeqnarray}
+%
+This distribution is simply a Gaussian and we can evaluate the density using \eqref{eq:basic_F_conditional}. Finally, we also sample a new state trajectory using forward filtering backward sampling,
+%
+\begin{IEEEeqnarray}{rCl}
+ \ls{1:\timax}' & \sim & \postden(\ls{1:\timax}|\lgtm', \lgtv)      .
+\end{IEEEeqnarray}
+%
+The acceptance probability is simply,
+%
+\begin{IEEEeqnarray}{rCl}
+ \IEEEeqnarraymulticol{3}{l}{ \mhap\left(\lgtm,\ls{1:\timax}\to\lgtm',\ls{1:\timax}'\right) } \nonumber \\
+ \quad\quad & = & \min\left\{1,\frac{ \postden(\lgtm',\ls{1:\timax}'|\lgtv) \postden(\ls{1:\timax}|\lgtm,\lgtv) \postden(\lgtm|\paddedlgtv,\ls{1:\timax}')   }{ \postden(\lgtm,\ls{1:\timax}|\lgtv) \postden(\ls{1:\timax}'|\lgtm',\lgtv) \postden(\lgtm'|\paddedlgtv,\ls{1:\timax}) }\right\} \nonumber \\
+ & = & \min\left\{1,\frac{ \postden(\lgtm'|\lgtv) \postden(\lgtm|\paddedlgtv,\ls{1:\timax}')   }{ \postden(\lgtm|\lgtv)\postden(\lgtm'|\paddedlgtv,\ls{1:\timax}) }\right\} \nonumber \\
+ & = & \min\left\{1, \frac{ \den(\ob{1:\timax}|\lgtm',\lgtv) }{ \den(\ob{1:\timax}|\lgtm,\lgtv)} \times \frac{ \den(\lgtm'|\lgtv) }{ \den(\lgtm|\lgtv)} \times \frac{ \postden(\lgtm|\paddedlgtv,\ls{1:\timax}')   }{ \postden(\lgtm'|\paddedlgtv,\ls{1:\timax}) }\right\} \nonumber      , \\
+\end{IEEEeqnarray}
+%
+in which the first term is a ratio of Kalman filter likelihoods.
+
 \appendix

### 2015-08-30 learners_mcmc.py

@@ -196,7 +196,7 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
         """
         Prior density for transition model parameters
         """
-        variancePrior = smp.singular_wishart_density(model.parameters['val'],
+        variancePrior = smp.singular_inverse_wishart_density(model.parameters['val'],
                       model.parameters['vec'],la.inv(self.hyperparams['Psi0']))
         
         orthVec = model.complete_basis()
@@ -352,7 +352,7 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
                                                     self.hyperparams['Psi0'],
                                                     self.hyperparams['M0'],
                                                     self.hyperparams['V0'])
-        D = smp.sample_wishart(nu, la.inv(Psi))
+        D = la.inv(smp.sample_wishart(nu, la.inv(Psi)))
         FU = smp.sample_matrix_normal(M, D, V)
         
         # Project out

### 2015-08-30 linear_models.py

@@ -249,7 +249,7 @@ class DegenerateLinearModel(AbstractLinearModel):
 
     def transition_covariance(self):
         eVec = self.parameters['vec']
-        eVal = np.diag(1./self.parameters['val'])
+        eVal = np.diag(self.parameters['val'])
         return np.dot(eVec, np.dot(eVal, eVec.T))
         
     def observation_matrix(self):

### 2015-08-30 run_basic_learning.py

@@ -17,7 +17,7 @@ params = dict()
 params['F'] = np.array([[0.9,0.8,0.7],[0,0.9,0.8],[0,0,0.7]])
 params['rank'] = 2
 params['vec'] = (1./np.sqrt(3))*np.array([[1,1],[1,1],[1,-1]])#np.identity(ds)
-params['val'] = np.array([5,2])
+params['val'] = np.array([1./5,1./2])
 #params['Q'] = np.array([[0.1,0.1],[0.1,0.1]])#np.array([[0.1,0.05],[0.05,0.1]])
 params['H'] = np.identity(do)#np.array([[1,0]])
 params['R'] = 0.1*np.identity(do)#np.array([[1]])

### 2015-08-30 sampling.py

@@ -64,6 +64,24 @@ def singular_wishart_density(val, vec, P):
     pdf = norm + pptn
     return pdf
 
+def singular_inverse_wishart_density(val, vec, P):
+    """
+    Density of a singular inverse wishart distribution at
+    X = vec*diag(val)*vec.T
+    """
+    d,r = vec.shape
+    
+    norm = 0.5*r*la.det(P) \
+            -0.5*r*d*np.log(2.0) \
+            -0.5*r*(d-r)*np.log(np.pi) \
+            -special.multigammaln(r/2,r)
+    
+    pptn = 0.5*(3*d-r+1)*np.log(np.prod(val)) \
+            -0.5*np.trace( np.dot(np.dot(vec.T,np.dot(P,vec)),np.diag(val)) )
+    
+    pdf = norm + pptn
+    return pdf
+
 def matrix_normal_density(X, M, U, V):
     """Sample from a matrix normal distribution"""
     norm = - 0.5*la.det(2*np.pi*U) - 0.5*la.det(2*np.pi*V)
@@ -132,8 +150,9 @@ def hyperparam_update_degenerate_mniw_transition(suffStats, U, nu0, Psi0, M0, V0
     nu  = nu0 + suffStats[0]
     V   = la.inv( la.inv(V0) + suffStats[1] )
     M   = np.dot( np.dot(U.T, la.solve(V0,M0.T).T + suffStats[2]) , V)
-    UPsiU = Psi0+suffStats[3]+np.dot(M0,la.solve(V0,M0.T))
-    Psi = np.dot(U.T, np.dot(UPsiU, U)) - np.dot(M, la.solve(V,M.T))
+    UPsi0U = la.inv(np.dot(U.T,la.solve(Psi0,U)))
+    UPsiU = suffStats[3]+np.dot(M0,la.solve(V0,M0.T))
+    Psi = UPsi0U + np.dot(U.T, np.dot(UPsiU, U)) - np.dot(M, la.solve(V,M.T))
     
     return nu, Psi, M, V
 

### 2015-08-31 learners_mcmc.py
Partially added rank-change MH moves.

@@ -2,6 +2,8 @@
 from abc import ABCMeta, abstractmethod
 import numpy as np
 import scipy.linalg as la
+from scipy import stats
+from scipy import special
 import matplotlib.pyplot as plt
 
 # Import from other module files
@@ -252,10 +254,6 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
                                                     self.hyperparams['V0'],)
             bwd_prob = smp.matrix_normal_density(self.model.parameters['F'],
                                                                   M,padded_Q,V)
-            
-            # Prior terms
-            prior = self.transition_prior(self.model)
-            ppsl_prior = self.transition_prior(ppsl_model)
         
         elif moveType=='Q':
             
@@ -267,41 +265,61 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
             fwd_prob = 0
             bwd_prob = 0
             
-            # Prior terms
-            prior = self.transition_prior(self.model)
-            ppsl_prior = self.transition_prior(ppsl_model)
-            
         elif moveType=='rank':
-            pass
             
+            if ppsl_model.ds == 1:
+                switch = -1                                # Do nothing (1D)
+            elif ppsl_model.parameters['rank'] == 1:
+                switch = 1                                 # Must increase rank
+            elif ppsl_model.parameters['rank'] == ppsl_model.ds:
+                switch = 0                                 # Must decrease rank
+            else:
+                switch = np.random.random_integers(0,1)    # choose at random
             
-            
-            
-            #TODO ADD THIS STEP
-            
-            
-#            # Add or remove?
-#            if ppsl_model.ds==1:
-#                u = -1                          # 1D model. Cannot do anything
-#            elif ppsl_model.rank==1:
-#                u = 0
-#            elif ppsl_model.rank==ppsl_model.ds:
-#                u = 1
-#            else:
-#                u = int(np.round(np.random.random()))
-#            
-#            if u==0:
-#                
-#                # Increase rank
-#                dcf = ppsl_model.increase_rank(None, None, self.Q_prior_scale)
-#                
-#            elif u==1:
-#                # Decrease rank
-#                dcf,val,vec = ppsl_model.reduce_rank(self.Q_prior_scale)            
-            
-            
-            
-            
+            if switch == -1:
+                # No options to change rank
+                
+                prior = 0
+                ppsl_prior = 0
[…]

### 2015-08-31 linear_models.py

@@ -271,7 +271,7 @@ class DegenerateLinearModel(AbstractLinearModel):
     
     def update_from_givens_form(self, U, D):
         """
-        Convert from givens to eigen form and update
+        Convert transition covariance from givens to eigen form and update
         """
         eVal, eVec = la.eigh(D)
         self.parameters['vec'] = np.dot(U, eVec)
@@ -279,7 +279,8 @@ class DegenerateLinearModel(AbstractLinearModel):
     
     def complete_basis(self):
         """
-        Returns a matrix of vectors orthogonal to the eigenvectors
+        Returns a matrix of vectors orthogonal to the eigenvectors for the
+        transition covariance
         """
         Q,R = la.qr(self.parameters['vec'])
         r = self.parameters['rank']
@@ -287,8 +288,33 @@ class DegenerateLinearModel(AbstractLinearModel):
     
     def rotate_transition_covariance(self, rotation):
         """
-        Rotate noise covariance matrix by multiplying eigenvectors by a
+        Rotate transition covariance matrix by multiplying eigenvectors by a
         supplied orthoginal matrix
         """
         self.parameters['vec'] = np.dot(rotation, self.parameters['vec'])
+    
+    def add_eigen_value_vector(self, value, vector):
+        """
+        Add an eigenvalue/eigenvector pair to the transition covariance matrix
+        """
+        if self.parameters['rank'] == self.ds:
+            raise ValueError("Covariance matrix is already full rank.")
+        self.parameters['val'] = np.append(self.parameters['val'], value)
+        self.parameters['vec'] = np.append(self.parameters['vec'],
+                                           vector[:,np.newaxis], axis=1)
+        self.parameters['rank'] -= 1
+    
+    def remove_min_eigen_value_vector(self):
+        """
+        Remove the minimum eigenvalue and the corresponding eigenvector from
+        the transition covariance matrix
+        """
+        minIdx = np.argmin(self.parameters['val'])
+        value = self.parameters['val'][minIdx]
+        vector = self.parameters['vec'][:,minIdx]
+        self.parameters['val'] = np.delete(self.parameters['val'], minIdx)
+        self.parameters['vec'] = np.delete(self.parameters['vec'], minIdx,
+                                                                       axis=1)
+        self.parameters['rank'] += 1
+        return value, vector
         

### 2015-08-31 sampling.py

@@ -2,10 +2,6 @@ import numpy as np
 from scipy import linalg as la
 from scipy import stats
 from scipy import special
-#from scipy.stats import multivariate_normal as mvn
-#from scipy import stats as sps
-#from scipy import misc as spm
-#from numpy import random as rnd
 
 def sample_wishart(nu, P):
     """
@@ -51,6 +47,14 @@ def sample_cayley(d, s):
     M = la.solve(I-S,I+S)
     
     return M
+
+def sample_orthogonal_haar(d):
+    """
+    Sample a unit vector uniformly from the (d-1)-sphere
+    """
+    z = stats.multivariate_normal.rvs(mean=np.zeros(d),cov=np.identity(d))
+    z /= la.norm(z)
+    return z
 
 def singular_wishart_density(val, vec, P):
     """

### 2015-08-31 notes/notes-linear-system-learning.tex
Added wishart transformation proof
Added MH moves for covariance rank.

@@
 
+Proof:
+%
+\begin{IEEEeqnarray}{rCl}
+ X &\sim& \wishartdist{\nu}{\Sigma} \nonumber\\
+ \Rightarrow Z &\sim& \matrixnormaldist{\zmat}{\idmat}{\idmat} \nonumber\\
+ \text{with } X &=& \Sigma\msqrt Z Z\tr {\Sigma\msqrt}\tr \nonumber \\
+ \text{and } Z &\in& \reals^{d\times \nu} \nonumber\\
+ Y = X\pinv &=& {\Sigma\msqrt}\inv (Z Z\tr)\pinv {\Sigma\msqrt}\invtr \nonumber \\
+ Y &\sim& \iwishartdist{\nu}{\Sigma} \nonumber \\
+ A Y A\tr &=& A {\Sigma\msqrt}\inv (Z Z\tr)\pinv {\Sigma\msqrt}\invtr A\tr \nonumber \\
+ &=& (A{\Sigma\msqrt}\inv) (Z Z\tr)\pinv (A{\Sigma\msqrt}\inv)\tr \nonumber \\
+ \Rightarrow A Y A\tr &\sim& \iwishartdist{\nu}{ \left((A{\Sigma\msqrt}\inv)(A{\Sigma\msqrt}\inv)\tr\right)\inv } \nonumber \\
+ A Y A\tr &\sim& \iwishartdist{\nu}{ \left((A\Sigma\inv A\tr\right)\inv } \nonumber \\
+\end{IEEEeqnarray}
+
 \subsubsection{Within-Subspace Parameter Conditional}
@@
 
+\subsubsection{Metropolis-Hastings for the Covariance Matrix Rank}
+
+It is unlikely that the rank of $\lgtv$ will be known a priori.
+We can learn this within the MCMC scheme by allowing the sampler to add and remove eigenvalue-eigenvector pairs using reversible jump moves \cite{Green1995,Green2009}.
+This can be achieved with a matching pair of moves.
+The eigenvalues are assumed to be sorted in descending order of magnitude, so that $\eval{1}$ is the largest.
+
+\begin{algorithm}
+\begin{algorithmic}
+ \REQUIRE{$\rk$, $\lgtv$}
+ \STATE Set $\rk\mcnew = \rk + 1$.
+ \STATE Sample a new eigenvector $\evec{\rk\mcnew}$.
+ \STATE Sample a new eigenvalue $\eval{\rk\mcnew} \in [0,\eval{\rk}]$.
+ \STATE $\lgtv\mcnew = \lgtv + \eval{\rk\mcnew}\evec{\rk\mcnew}\evec{\rk\mcnew}\tr$.
+ \RETURN{$\rk\mcnew$, $\lgtv\mcnew$}
+\end{algorithmic}
+\caption{Reversible Jump Move: Increase rank}
+\label{alg:rjmcmc-up}
+\end{algorithm}
+
+\begin{algorithm}
+\begin{algorithmic}
+ \REQUIRE{$\rk\mcnew$, $\lgtv\mcnew$}
+ \STATE $\lgtv = \lgtv\mcnew - \eval{\rk\mcnew}\evec{\rk\mcnew}\evec{\rk\mcnew}\tr$.
+ \STATE Set $\rk = \rk\mcnew - 1$.
+ \RETURN{$\rk$, $\lgtv$}
+\end{algorithmic}
+\caption{Reversible Jump Move: Decrease rank}
+\label{alg:rjmcmc-down}
+\end{algorithm}
+
+The acceptance probabilities are then,
+%
+\begin{IEEEeqnarray}{rCl}
+ \mhap(\rk,\lgtv \to \rk\mcnew,\lgtv\mcnew) &=& \min\left\{1, \mhar \right\} \\
+ \mhap(\rk\mcnew,\lgtv\mcnew \to \rk,\lgtv) &=& \min\left\{1, \mhar\inv \right\}     ,
+\end{IEEEeqnarray}
+%
+with,
+%
+\begin{IEEEeqnarray}{rCl}
+ \mhar &=& \frac{ \postden(\rk\mcnew,\lgtv\mcnew|\lgtm,\ls{1:\timax}) \jacob(\lgtv,\eval{\rk\mcnew},\evec{\rk\mcnew} \to \lgtv\mcnew) }{ \postden(\rk,\lgtv|\lgtm,\ls{1:\timax}) \ppslden{\eval{}}(\eval{\rk\mcnew}|\lgtv) \ppslden{\evec{}}(\evec{\rk\mcnew}|\lgtv) } \\
+ &=& \frac{ \den(\ls{1:\timax}|\lgtm, \lgtv\mcnew) \den(\lgtv\mcnew|\rk\mcnew,\lgtm) \den(\rk\mcnew) \jacob(\lgtv,\eval{\rk\mcnew},\evec{\rk\mcnew} \to \lgtv\mcnew) }{ \den(\ls{1:\timax}|\lgtm, \lgtv) \den(\lgtv|\rk,\lgtm) \den(\rk) \ppslden{\eval{}}(\eval{\rk\mcnew}|\lgtv) \ppslden{\evec{}}(\evec{\rk\mcnew}|\lgtv) } \\
+ &=& \frac{ \den(\ls{1:\timax}|\lgtm, \lgtv\mcnew) \den(\lgtm,\lgtv\mcnew|\rk\mcnew) \den(\rk\mcnew) }{ \den(\ls{1:\timax}|\lgtm, \lgtv) \den(\lgtm,\lgtv|\rk) \den(\rk) } \times 
+ \frac{ \jacob(\lgtv,\eval{\rk\mcnew},\evec{\rk\mcnew} \to \lgtv\mcnew) }{ \ppslden{\eval{}}(\eval{\rk\mcnew}|\lgtv) \ppslden{\evec{}}(\evec{\rk\mcnew}|\lgtv) }      ,
+\end{IEEEeqnarray}
+%
+where $\jacob(\lgtv,\eval{\rk\mcnew},\evec{\rk\mcnew} \to \lgtv\mcnew)$ is the Jacobian of the transformation from $\lgtv$ to $\lgtv\mcnew$, and $\ppslden{\eval{}}$ and $\ppslden{\evec{}}$ are the probability densities for the eigenvalue and eigenvector proposal distributions respectively, with respect to appropriate measures.
+The second term is a tricky beast, particularly the Jacobian.
+
+The volume element (ignoring sign) on the set of positive semi-definite matrices can be written in terms of the eigen-decomposition as follows \cite{Diaz-Garcia1997,Uhlig1994,Diaz-Garcia2006},
+%
+\begin{IEEEeqnarray}{rCl}
+ \volel{d\lgtv} &=& 2^{-\rk} \times \prod_{i=1}^{\rk}\eval{i}^{\lsd-\rk} \times \prod_{j=1}^{r}\prod_{i=1}^{j-1}(\eval{i}-\eval{j}) \times \volel{\tvvec\tr d\tvvec} \wedge \bigwedge_{i=1}^{r}d\eval{i}   ,
+\end{IEEEeqnarray}
+%
+where $\volel{\tvvec\tr d\tvvec}$ is the Haar measure on the appropriate Stiefel manifold \cite{Muirhead1982},
+%
+\begin{IEEEeqnarray}{rCl}
+ \volel{\tvvec\tr d\tvvec} &=& \bigwedge_{i=1}^{\rk}\bigwedge_{j=i+1}^{\lsd} \evec{j}\tr d\evec{i}     .
+\end{IEEEeqnarray}
+%
+Similarly,
+%
+\begin{IEEEeqnarray}{rCl}
+ \volel{d\lgtv\mcnew} &=& 2^{-\rk-1} \times \prod_{i=1}^{\rk+1}\eval{i}^{\lsd-\rk-1} \times \prod_{j=1}^{r+1}\prod_{i=1}^{j-1}(\eval{i}-\eval{j}) \times \volel{{\tvvec\mcnew}\tr d{\tvvec\mcnew}} \wedge \bigwedge_{i=1}^{r+1}d\eval{i} \nonumber \\
+ &=& \left[ 2^{-1} \eval{\rk+1}^{\lsd-\rk-1} \prod_{i=1}^{\rk}\eval{i}^{-1} \prod_{i=1}^{\rk}(\eval{i}-\eval{\rk+1}) \right] \nonumber \\
+ & & \qquad\qquad \times 2^{-\rk} \times \prod_{i=1}^{\rk}\eval{i}^{\lsd-\rk} \times \prod_{j=1}^{r}\prod_{i=1}^{j-1}(\eval{i}-\eval{j}) \times \volel{{\tvvec\mcnew}\tr d{\tvvec\mcnew}} \wedge \bigwedge_{i=1}^{r+1}d\eval{i} \nonumber    ,
+\end{IEEEeqnarray}
+%
+where
+%
+\begin{IEEEeqnarray}{rCl}
+ \volel{{\tvvec\mcnew}\tr d{\tvvec\mcnew}} &=& \bigwedge_{i=1}^{\rk+1}\bigwedge_{j=i+1}^{\lsd} \evec{j}\tr d\evec{i} \nonumber \\
+ &=& \bigwedge_{i=1}^{\rk}\bigwedge_{j=i+1}^{\lsd} \evec{j}\tr d\evec{i} \wedge \bigwedge_{j=\rk+2}^{\lsd} \evec{j}\tr d\evec{i}    .
+\end{IEEEeqnarray}
+
+A suitable base measure for the new eigenvector can be formulated by considering the following generative procedure.
+This new eigenvector $\evec{\rk\mcnew}$ must have unit magnitude and be orthogonal to the existing eigenvectors of $\lgtv$.
+This can be achieved by sampling $u \in \{u \in \reals^{\lsd-\rk} : \determ{u}=1\}$ according to the Haar probability measure on this set and setting $\evec{\rk\mcnew}=\tvvecorth u$.
+Since this is another Stiefel manifold, the volume element for this set is \cite{Muirhead1982},
+%
+\begin{IEEEeqnarray}{rCl}
+ \volel{u\tr du} &=& \bigwedge_{j=2}^{\lsd-\rk} u_{j}\tr du
+\end{IEEEeqnarray}
+%
+where $u_2, \dots, u_{\lsd-\rk}$ is any set of unit vectors orthogonal to $u$. Using the transformation,
+%
+\begin{IEEEeqnarray}{rCl}
+ \volel{\evec{\rk\mcnew}\tr d\evec{\rk\mcnew}} &=& \volel{u\tr \tvvecorth\tr \tvvecorth du} \nonumber \\
+ &=& \volel{u\tr du} \nonumber \\
+ &=& \bigwedge_{j=2}^{\lsd-\rk} u_{j}\tr du \nonumber \\
+ &=& \bigwedge_{j=2}^{\lsd-\rk} u_{j}\tr \tvvecorth\tr d\evec{\rk\mcnew} \nonumber \\
+ &=& \bigwedge_{j=\rk+2}^{\lsd} {\evec{j}\mcnew}\tr d\evec{\rk\mcnew}      ,
+\end{IEEEeqnarray}
+%
+where $\evec{j}\mcnew = \tvvecorth u_{\rk+j}$ are the columns of the new $\tvvecorth\mcnew$.
+
+The base measure for the new eigenvalue is Lesbegue.
+
+Thus, the Jacobian is,
+%
+\begin{IEEEeqnarray}{rCl}
+ \jacob(\lgtv,\eval{\rk\mcnew},\evec{\rk\mcnew} \to \lgtv\mcnew) &=& \frac{ \volel{d\lgtv\mcnew} }{ \volel{d\lgtv} \volel{\evec{\rk\mcnew}\tr d\evec{\rk\mcnew}} \volel{d\eval{\rk\mcnew}} } \nonumber \\
+ &=& 2^{-1} \eval{\rk+1}^{\lsd-\rk-1} \prod_{i=1}^{\rk}\eval{i}^{-1} \prod_{i=1}^{\rk}(\eval{i}-\eval{\rk+1})
+\end{IEEEeqnarray}
+
+The density of the proposal distribution for the new eigenvector is simply the reciprocal of the volume of the set, which is a hypersphere. Thus,
+%
+\begin{IEEEeqnarray}{rCl}
+ \ppslden{\evec{}}(\evec{\rk\mcnew}|\lgtv) &=& \frac{\gammafun\left(\frac{\lsd-\rk}{2}\right)}{2 \pi^{\half(\lsd-\rk)}}      .
+\end{IEEEeqnarray}
+
+For the eigenvalue, we can use any distribution over $[0,\eval{\rk}]$, e.g. uniform.
+
+So,
+%
+\begin{IEEEeqnarray}{rCl}
+ \mhar  &=& \frac{ \den(\ls{1:\timax}|\lgtm, \lgtv\mcnew) \den(\lgtm,\lgtv\mcnew|\rk\mcnew) \den(\rk\mcnew) }{ \den(\ls{1:\timax}|\lgtm, \lgtv) \den(\lgtm,\lgtv|\rk) \den(\rk) } \times 
+ \frac{ \jacob(\lgtv,\eval{\rk\mcnew},\evec{\rk\mcnew} \to \lgtv\mcnew) }{ \ppslden{\eval{}}(\eval{\rk\mcnew}|\lgtv) \ppslden{\evec{}}(\evec{\rk\mcnew}|\lgtv) } \nonumber \\
+ &=& \frac{ \den(\ls{1:\timax}|\lgtm, \lgtv\mcnew) \den(\lgtm,\lgtv\mcnew|\rk\mcnew) \den(\rk\mcnew) }{ \den(\ls{1:\timax}|\lgtm, \lgtv) \den(\lgtm,\lgtv|\rk) \den(\rk) } \times 
+ \frac{ \pi^{\half(\lsd-\rk)} \prod_{i=1}^{\rk}\eval{i}^{-1} }{ \gammafun\left(\frac{\lsd-\rk}{2}\right) } \times
+ \frac{\eval{\rk+1}^{\lsd-\rk-1}\prod_{i=1}^{\rk}(\eval{i}-\eval{\rk+1})}{\ppslden{\eval{}}(\eval{\rk+1}|\lgtv)} \nonumber      .
+\end{IEEEeqnarray}
+
+% The acceptance probability for the increase move is then,
+% %
+% \begin{IEEEeqnarray}{rCl}
+%  \IEEEeqnarraymulticol{3}{l}{ \mhap\left( \tnd,\tnmd,\tnmo \to \tnd',\tnmd',\tnmo' \right) } \nonumber \\
+%  \qquad & = & \min\left\{1, \frac{ \postden(\lgtm, \tnd',\tnmd',\tnmo') }{ \postden(\lgtm, \tnd,\tnmd,\tnmo) } \times \frac{1}{\ppslden{\tnmevec{}}(\tnmevec{}^*)\ppslden{\tnmev{}}(\tnmev{}^*)} \right\} \nonumber \\
+%  & = & \min\Bigg\{1, \frac{\den(\ob{1:\timax} | \lgtm, \tnd',\tnmd',\tnmo') }{ \den(\ob{1:\timax} | \lgtm, \tnd,\tnmd,\tnmo) } \nonumber \\
+%  & & \quad \times \frac{\den(\tnmevec{1},\dots,\tnmevec{\tnd},\tnmevec{}^*)\den(\tnmev{1},\dots,\tnmev{\tnd},\tnmev{}^*|\tnd')\den(\tnd')}{\den(\tnmevec{1},\dots,\tnmevec{\tnd})\den(\tnmev{1},\dots,\tnmev{\tnd}|\tnd)\den(\tnd) \ppslden{\tnmevec{}}(\tnmevec{}^*)\ppslden{\tnmev{}}(\tnmev{}^*)} \Bigg\} \nonumber      ,
+% \end{IEEEeqnarray}
+% %
+% where $\tnmevec{1},\dots,\tnmevec{\tnd}$ are the eigenvectors comprising $\tnmo$ and $\tnmev{1},\dots,\tnmev{\tnd}$ the eigenvalues comprising $\tnmd$.
+The first term is simply a ratio of Kalman filter likelihoods.
+For the decrease move, the ratio is replaced by its reciprocal.
+% 
+% The conditional prior distribution $\den(\tnmevec{}^*|\tnmevec{1},\dots,\tnmevec{\tnd})$ is the normalised Haar measure on the manifold $\tnmo\tr\tnmevec{}^*=0$, conditional on the constraint we imposed to resolve the sign ambiguity \cite{Muirhead1982}.
+We can simulate according to this prior by sampling a standard Gaussian random vector and then using Gram-Schmidt orthogonalisation.
+This implies setting $\ppslden{\tnmevec{}}(\tnmevec{}^*)=\den(\tnmevec{}^*|\tnmevec{1},\dots,\tnmevec{\tnd})$, which results in a cancellation in the acceptance probability.
+
 \appendix

### 2015-08-31 learners_mcmc.py

@@ -202,12 +202,22 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
                       model.parameters['vec'],la.inv(self.hyperparams['Psi0']))
         
         orthVec = model.complete_basis()
+#        relaxEval = self.hyperparams['alpha']
+        relaxEval = np.min(model.parameters['val'])
         rowVariance = model.transition_covariance() \
-                      + self.hyperparams['alpha']*np.dot(orthVec,orthVec.T)
+                      + relaxEval*np.dot(orthVec,orthVec.T)
         matrixPrior = smp.matrix_normal_density(model.parameters['F'],
                                                 self.hyperparams['M0'],
                                                 rowVariance,
                                                 self.hyperparams['V0'])
+        
+#        print(model.parameters['rank'])
+#        print(orthVec)
+#        print(la.det(rowVariance))
+#        print(self.hyperparams['V0'])
+        
+        print(matrixPrior)
+        print(variancePrior)
         
         return variancePrior + matrixPrior
     
@@ -222,6 +232,9 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
         
         # Copy model
         ppsl_model = self.model.copy()
+        
+        if self.verbose:
+            print("Metropolis-Hastings move type: {}".format(moveType))
         
         # Propose change
         if   moveType=='F':
@@ -269,9 +282,9 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
             
             if ppsl_model.ds == 1:
                 switch = -1                                # Do nothing (1D)
-            elif ppsl_model.parameters['rank'] == 1:
+            elif ppsl_model.parameters['rank'][0] == 1:
                 switch = 1                                 # Must increase rank
-            elif ppsl_model.parameters['rank'] == ppsl_model.ds:
+            elif ppsl_model.parameters['rank'][0] == ppsl_model.ds:
                 switch = 0                                 # Must decrease rank
             else:
                 switch = np.random.random_integers(0,1)    # choose at random
@@ -286,45 +299,68 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
                 
             elif switch == 0:
                 # Decrease rank
-                pass
-                #TODO - Add this
+            
+                if self.verbose:
+                    print("   Decreasing covariance rank")
+        
                 
                 # Remove the smallest eigenvalue and associated eigenvector
[…]

### 2015-08-31 linear_models.py

@@ -1,5 +1,6 @@
 # Operational modules
 from abc import ABCMeta, abstractmethod
+from copy import deepcopy
 
 # Numerical modules
 import numpy as np
@@ -35,7 +36,7 @@ class AbstractLinearModel:
         if (state_dimension<1) or (observation_dimension<1):
             raise ValueError("Invalid state or observation dimensions")
         if (initial_state_prior.mn.shape!=(ds,)) \
-                                    or (initial_state_prior.vr.shape!=(ds,ds)):
+                                   or (initial_state_prior.vr.shape!=(ds,ds)):
             raise ValueError("Invalid initial state prior density")
         
         # Store them
@@ -47,7 +48,7 @@ class AbstractLinearModel:
     
     def copy(self):
         return self.__class__(self.ds, self.do, self.initial_state_prior,
-                                                         dict(self.parameters))
+                                                    deepcopy(self.parameters))
     
     
     @abstractmethod
@@ -283,7 +284,7 @@ class DegenerateLinearModel(AbstractLinearModel):
         transition covariance
         """
         Q,R = la.qr(self.parameters['vec'])
-        r = self.parameters['rank']
+        r = self.parameters['rank'][0]
         return Q[:,r:]
     
     def rotate_transition_covariance(self, rotation):
@@ -297,12 +298,12 @@ class DegenerateLinearModel(AbstractLinearModel):
         """
         Add an eigenvalue/eigenvector pair to the transition covariance matrix
         """
-        if self.parameters['rank'] == self.ds:
+        if self.parameters['rank'][0] == self.ds:
             raise ValueError("Covariance matrix is already full rank.")
         self.parameters['val'] = np.append(self.parameters['val'], value)
         self.parameters['vec'] = np.append(self.parameters['vec'],
                                            vector[:,np.newaxis], axis=1)
-        self.parameters['rank'] -= 1
+        self.parameters['rank'][0] += 1
     
     def remove_min_eigen_value_vector(self):
         """
@@ -315,6 +316,6 @@ class DegenerateLinearModel(AbstractLinearModel):
         self.parameters['val'] = np.delete(self.parameters['val'], minIdx)
         self.parameters['vec'] = np.delete(self.parameters['vec'], minIdx,
                                                                        axis=1)
-        self.parameters['rank'] += 1
+        self.parameters['rank'][0] -= 1
         return value, vector
         

### 2015-08-31 run_basic_learning.py

@@ -15,7 +15,7 @@ do = 3
 
 params = dict()
 params['F'] = np.array([[0.9,0.8,0.7],[0,0.9,0.8],[0,0,0.7]])
-params['rank'] = 2
+params['rank'] = np.array([2])
 params['vec'] = (1./np.sqrt(3))*np.array([[1,1],[1,1],[1,-1]])#np.identity(ds)
 params['val'] = np.array([1./5,1./2])
 #params['Q'] = np.array([[0.1,0.1],[0.1,0.1]])#np.array([[0.1,0.05],[0.05,0.1]])
@@ -30,7 +30,7 @@ state, observ = model.simulate_data(K)
 
 est_params = dict(params)
 est_params['F'] = 0.5*np.identity(ds)
-est_params['rank'] = 2
+est_params['rank'] = np.array([2])
 est_params['vec'] = params['vec']#np.array([[1,0],[0,1],[0,0]])
 est_params['val'] = np.array([1,1])
 est_model = DegenerateLinearModel(ds, do, prior, est_params)
@@ -40,7 +40,7 @@ hyperparams['nu0'] = params['rank']
 hyperparams['Psi0'] = params['rank']*np.identity(ds)
 hyperparams['M0'] = np.zeros((ds,ds))
 hyperparams['V0'] = np.identity(ds)
-hyperparams['alpha'] = 100
+hyperparams['alpha'] = 1
 
 algoparams = dict()
 algoparams['Qs'] = 0.1
@@ -48,25 +48,29 @@ algoparams['Fs'] = 0.1
 
 learner = MCMCLearnerForDegenerateModelWithMNIWPrior(est_model, observ, hyperparams, algoparams=algoparams, verbose=True)
 
-num_iter = 200
+num_iter = 20
 num_burn = int(num_iter/5)
 
 for ii in range(num_iter):
     print("Running iteration {} of {}.".format(ii+1,num_iter))
     
-    if (ii%2)==0:
-        learner.iterate_transition('F')
-    else:
+    if (ii%3)==0:
         learner.iterate_transition('Q')
+    elif (ii%3)==1:
+        learner.iterate_transition('rank')
+    else:
+        learner.iterate_transition('F')
     learner.save_link()
 
 learner.plot_chain_trace('F', numBurnIn=num_burn)
+learner.plot_chain_trace('rank', numBurnIn=num_burn)
 #learner.plot_chain_trace('Q', numBurnIn=num_burn)
[…]

### 2015-08-31 sampling.py

@@ -52,8 +52,10 @@ def sample_orthogonal_haar(d):
     """
     Sample a unit vector uniformly from the (d-1)-sphere
     """
-    z = stats.multivariate_normal.rvs(mean=np.zeros(d),cov=np.identity(d))
+    z = stats.multivariate_normal.rvs(mean=np.zeros(d),cov=np.identity(d))    
     z /= la.norm(z)
+    if d == 1:
+        z = z[np.newaxis] # This undoes the squeeze if only 1D
     return z
 
 def singular_wishart_density(val, vec, P):

### 2015-08-31 learners_mcmc.py
Minor plotting fixes

@@ -118,6 +118,9 @@ class AbstractMCMCLearner:
         for idx in np.ndindex(coords.shape):
             samples = [mod.parameters[paramName][coords[idx]] for mod in self.chain]
             axs[idx].plot(samples, 'k')
+            ylims = axs[idx].get_ylim()
+            axs[idx].plot([numBurnIn]*2, ylims, 'r:')
+            axs[idx].set_ylim(ylims)
     
     
     def plot_chain_histogram(self, paramName, numBurnIn=0, dims=None, trueValue=None):
@@ -142,7 +145,7 @@ class AbstractMCMCLearner:
             raise ValueError("Cannot draw plots for this parameter")
         
         for idx in np.ndindex(coords.shape):
-            samples = [mod.parameters[paramName][coords[idx]] for mod in self.chain]
+            samples = [mod.parameters[paramName][coords[idx]] for mod in self.chain[numBurnIn:]]
             axs[idx].hist(samples, color='0.8')
             if trueValue is not None:
                 ylims = axs[idx].get_ylim()
@@ -210,14 +213,6 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
                                                 self.hyperparams['M0'],
                                                 rowVariance,
                                                 self.hyperparams['V0'])
-        
-#        print(model.parameters['rank'])
-#        print(orthVec)
-#        print(la.det(rowVariance))
-#        print(self.hyperparams['V0'])
-        
-        print(matrixPrior)
-        print(variancePrior)
         
         return variancePrior + matrixPrior
     
@@ -368,14 +363,6 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
         # Prior terms
         prior = self.transition_prior(self.model)
         ppsl_prior = self.transition_prior(ppsl_model)
-        
-#        if moveType=='rank':
-#            print(ppsl_lhood)
-#            print(lhood)
-#            print(ppsl_prior)
-#            print(prior)
-#            print(fwd_prob)
-#            print(bwd_prob)
         
         # Decide
         acceptRatio =   (ppsl_lhood-lhood) \

### 2015-08-31 run_basic_learning.py

@@ -48,7 +48,7 @@ algoparams['Fs'] = 0.1
 
 learner = MCMCLearnerForDegenerateModelWithMNIWPrior(est_model, observ, hyperparams, algoparams=algoparams, verbose=True)
 
-num_iter = 20
+num_iter = 200
 num_burn = int(num_iter/5)
 
 for ii in range(num_iter):

### 2015-08-31 basic.py
Removed old commented code

@@ -1,361 +1,0 @@
-# Operational modules
-from abc import ABCMeta, abstractmethod
-import collections
-
-# Numerical modules
-import numpy as np
-from scipy.stats import multivariate_normal as mvn
-
-
-
-
-# Import Kalman filter/smoother operations
-import kalman as kal
-
-
-
-
-
-
-
-
-
-
-
-#class SparseLinearModel(LinearModel):
-#    """Sparse Transition Matrix Linear-Gaussian Model Class"""
-#    
-#    def __init__(self, P1, A, B, G, H, R):
-#        self.A = A.copy()           # Dense transition matrix
-#        self.B = B.copy()           # Transition matrix mask
-#        self.G = G.copy()           # Transition noise matrix
-#        self.H = H.copy()           # Observation matrix
-#        self.R = R.copy()           # Observation variance
-#        self.P1 = P1.copy()         # First state prior
-#        self.ds = A.shape[0]        # State dimension
-#        self.do = H.shape[0]        # State dimension
-#        
-#    def copy(self):
-#        return SparseLinearModel(
-#                               self.P1, self.A, self.B, self.G, self.H, self.R)
-#    
-#    @property
-#    def F(self):
-#        """Sparse Transition Matrix"""
-#        return np.multiply(self.A,self.B)
-#    @F.setter
-#    def F(self, value):
-#        raise AttributeError("Can't set attribute")        
-#    @F.deleter
-#    def F(self):
-#        raise AttributeError("Can't delete attribute")
-#
-#
-#
-#class DegenerateLinearModel(LinearModel):
-#    """Degenerate Transition Covariance Linear-Gaussian Model Class"""
-#    
-#    def __init__(self, P1, F, D, givens, H, R):
-#        self.F = F.copy()           # Transition matrix
-#        self.H = H.copy()           # Observation matrix
-#        self.R = R.copy()           # Observation variance
-#        self.P1 = P1.copy()         # First state prior
-#        self.ds = F.shape[0]        # State dimension
-#        self.do = H.shape[0]        # State dimension
-#        self.rank = D.shape[0]      # Rank of noise
-#        
-#        self._D = D.copy()              # Postive definite part of transition noise matrix
[…]

### 2015-08-31 learners_mcmc.py

@@ -417,60 +417,6 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
         self.model.update_from_givens_form(U, D)
     
     
-    
-
-
-
-#import matplotlib.pyplot as plt
-#import numpy as np
-#import linear_models_sampling as sysl
-#from scipy import linalg as la
-#from scipy import stats
-#from scipy.stats import multivariate_normal as mvn
-#import statsmodels.tsa.stattools as st
-#from linear_models import LinearModel, SparseLinearModel, DegenerateLinearModel
-#
-#class LinearModelMCMC:
-#    """Container Class for Linear-Gaussian Model Learning with MCMC"""
-#
-#    def __init__(self, model_type, init_model_est, observ, A_prior_vr=None, B_prior_prob=None, Q_prior_scale=None, Q_prior_dof=None, giv_std=None, pad_std=None, rot_std=None):
-#        self.model_type = model_type
-#        
-#        self.model = init_model_est.copy()
-#        self.observ = observ
-#
-#        self.chain = []
-#        self.chain.append(self.model.copy())        
-#        
-#        self.A_prior_vr = A_prior_vr
-#        self.B_prior_prob = B_prior_prob
-#        self.Q_prior_scale = Q_prior_scale # This is the scalar p such that the Q^{-1} is Wishart with mean nu*p*I (i.e. phi in the paper)
-#        self.Q_prior_dof = Q_prior_dof
-#        
-#        if model_type==2:
-#            
-##            worst_rank = int(self.model.ds/2)
-##            most_givens = worst_rank*self.model.ds-worst_rank
-#            
-##            self.giv_std = [giv_std]*(most_givens)
-#            self.giv_std = giv_std
-#            self.pad_std = pad_std
-#            self.rot_std = rot_std
-#            
-#            self.F_acc = []
-#            self.pad_std_chain = []
-#            
-#            self.rot_acc = []
-#            self.rot_std_chain = []
-#            
-#            self.givens_acc = []
-#            self.giv_std_chain = []
-##            self.givens_acc = [[] for gg in range(most_givens)]
-##            self.giv_std_chain = [[] for gg in range(most_givens)]
-##            for gg in range(most_givens):
-##                self.giv_std_chain[gg].append(self.giv_std[gg])
-#                
-#            self.rank_acc = []
 #            
 #        
[…]

### 2015-09-01 notes/notes-linear-system-learning.tex
Minor fiddles and fixes

@@
 
-\subsubsection{Factorising the Precision}
+\subsubsection{Factorising the Covariance}
 
-Using Givens rotations, we can efficiently factorise the precision eigenvector matrix as follows.
+Using Givens rotations, we can efficiently factorise the covariance eigenvector matrix as follows.
 %
@@
 
-We sample a rotation matrix $\mhrot$ from some proposal distribution $\mhrotppsl$ and then apply the following transformation,
+We sample a rotation matrix $\mhrot$ from some proposal distribution with density $\mhrotppsl$ (w.r.t. Haar measure on orthogonal matrices) and then apply the following transformation,
 %
@@
 %
+The matrix exponential would also work. Note that the Jacobian of the Cayley transformation is derived in \cite{Olkin1953}, but this is not needed as it cancels out.
+
 To sample from $\mhrotppsl$, we draw $\half\lsd(\lsd-1)$ independent scalar random variables $\{s_{i,j}\}_{0<i<j<\lsd}$ from any zero-mean distribution; a nice choice is,
@@
 
-The density of the proposal distribution for the new eigenvector is simply the reciprocal of the volume of the set, which is a hypersphere. Thus,
+The density of the proposal distribution for the new eigenvector is simply the reciprocal of the volume of the set, which is a half of a hypersphere (since we have to resolve the eigenvector sign ambiguity).
+Thus,
 %
 \begin{IEEEeqnarray}{rCl}
- \ppslden{\evec{}}(\evec{\rk\mcnew}|\lgtv) &=& \frac{\gammafun\left(\frac{\lsd-\rk}{2}\right)}{2 \pi^{\half(\lsd-\rk)}}      .
+ \ppslden{\evec{}}(\evec{\rk\mcnew}|\lgtv) &=& \frac{\gammafun\left(\frac{\lsd-\rk}{2}\right)}{\pi^{\half(\lsd-\rk)}}      .
 \end{IEEEeqnarray}
@@
  &=& \frac{ \den(\ls{1:\timax}|\lgtm, \lgtv\mcnew) \den(\lgtm,\lgtv\mcnew|\rk\mcnew) \den(\rk\mcnew) }{ \den(\ls{1:\timax}|\lgtm, \lgtv) \den(\lgtm,\lgtv|\rk) \den(\rk) } \times 
- \frac{ \pi^{\half(\lsd-\rk)} \prod_{i=1}^{\rk}\eval{i}^{-1} }{ \gammafun\left(\frac{\lsd-\rk}{2}\right) } \times
+ \frac{ \pi^{\half(\lsd-\rk)} \prod_{i=1}^{\rk}\eval{i}^{-1} }{ 2 \: \gammafun\left(\frac{\lsd-\rk}{2}\right) } \times
  \frac{\eval{\rk+1}^{\lsd-\rk-1}\prod_{i=1}^{\rk}(\eval{i}-\eval{\rk+1})}{\ppslden{\eval{}}(\eval{\rk+1}|\lgtv)} \nonumber      .
@@
 
+\section{Additional Notes}
+
+I originally focussed on methods where the prior dependence between $\lgtm$ and $\lgtv$ was broken, and the two were sampled alternately.
+To do this efficiently (with matrix normal distribution for the transition matrix) requires using eigen-decompositions in a manner inspired by this paper: \cite{Stegle2011}.
+
 \appendix

### 2015-09-01 learners_mcmc.py

@@ -342,6 +342,7 @@ class MCMCLearnerForDegenerateModelWithMNIWPrior(AbstractMCMCLearner):
                 # Calculate the Jacobian
                 constJac = 0.5*nullDims*np.log(np.pi) \
                          - np.sum(np.log(ppsl_model.parameters['val'])) \
+                         - np.log(2.0) \
                          - special.gammaln(nullDims/2)
                 varJac = nullDims*np.log(newValue) \
                        + np.sum(np.log(ppsl_model.parameters['val']-newValue))

### 2015-09-01 run_basic_learning.py

@@ -48,7 +48,7 @@ algoparams['Fs'] = 0.1
 
 learner = MCMCLearnerForDegenerateModelWithMNIWPrior(est_model, observ, hyperparams, algoparams=algoparams, verbose=True)
 
-num_iter = 200
+num_iter = 2000
 num_burn = int(num_iter/5)
 
 for ii in range(num_iter):

### 2015-09-01 givens.py
New givens factorisation for comparison

@@ -16,6 +16,16 @@ def plane_rotation(x):
     g = np.arctan(G[0,1]/G[0,0])
     return y,g,G
     
+def plane_rotation2(x):
+    """Find the plane rotation and Givens angle to zero a component"""
+    r = la.norm(x)
+    if r > 0:
+        G = np.hstack((x[:,np.newaxis],x[::-1,np.newaxis]))/r
+        G[1,0] *= -1
+    else:
+        G = np.identity(2)
+    return G
+    
 def givens_matrix(g,ii,jj,d):
     """create a givens rotation matrix"""
     G = np.identity(d)
@@ -42,6 +52,8 @@ def givensise(U):
     for rr in reversed(range(r)):
         for cc in range(rr):
             
+            print((rr,cc))
+            
             # Get the next pair of elements to be rotated
             v = U[rr,[rr,cc]]
             
@@ -53,12 +65,18 @@ def givensise(U):
 
             # Perform the rotation 
             #TODO This is inefficient rotating the whole array - should do it on just the 4 elements affected.
+            tmp = U.copy()
             U = np.dot(U,G);
             Ur = np.dot(G.T,Ur)
+            
+            #print(G)
+            print(U)
             
     # Cross loop
     for cc in range(r):
         for rr in reversed(range(r,d)):
+            
+            print((rr,cc))
             
             # Get the next pair of elements to be rotated
             v = U[[cc,rr],cc]
@@ -71,11 +89,81 @@ def givensise(U):
             
             # Perform the rotation 
             #TODO This is inefficient rotating the whole array - should do it on just the 4 elements affected.
+            tmp = U.copy()
             U = np.dot(G,U)
             Uc = np.dot(Uc,G.T)
+            
+            #print(G)
+            print(U)
     
     # The remaining elements are the signs
     E = U
     
     return Uc,E,Ur
+
+
+def givensise2(U):
+    """
+    Factorise a matrix of orthonormal columns into row space and cross 
+    rotations and a sign matrix, such that U = Ur x E x Uc
+    """
     
+    # Copy it so we don't break it
+    U = U.copy()
+    
+    # Get the shape
+    d,r = U.shape
+    
+    # Create arrays for the two rotation matrices
+    Ur = np.identity(r)
+    Uc = np.identity(d)
+    
+    
+    print("Row space")
+    
+    # Row space loop
+    for rr in reversed(range(r)):
+        for cc in range(rr):
[…]

### 2015-09-01 givens.py
Improved givens factorisation implementation and testing.

@@ -2,38 +2,23 @@
 
 import numpy as np
 import scipy.linalg as la
-
+    
 def plane_rotation(x):
     """Find the plane rotation and Givens angle to zero a component"""
-    if x[1] != 0:
-        r = la.norm(x)
-        G = np.hstack((np.expand_dims(x,1),np.expand_dims(x[::-1],1)))/r
-        G[1,0] *= -1
-        y = np.array([r,0])
-    else:
-        G = np.identity(2)
-        y = x
-    g = np.arctan(G[0,1]/G[0,0])
-    return y,g,G
-    
-def plane_rotation2(x):
-    """Find the plane rotation and Givens angle to zero a component"""
     r = la.norm(x)
+    x /= r
+    if x[0] < 0:
+        x *= -1
     if r > 0:
-        G = np.hstack((x[:,np.newaxis],x[::-1,np.newaxis]))/r
+        G = np.hstack((x[:,np.newaxis],x[::-1,np.newaxis]))
         G[1,0] *= -1
     else:
         G = np.identity(2)
-    return G
     
-def givens_matrix(g,ii,jj,d):
-    """create a givens rotation matrix"""
-    G = np.identity(d)
-    G[ii,ii] = np.cos(g)
-    G[jj,jj] = np.cos(g)
-    G[ii,jj] = np.sin(g)
-    G[jj,ii] =-np.sin(g)
+    #print(np.arctan2(x[1],x[0]))
+    
     return G
+
 
 def givensise(U):
     """
@@ -41,129 +26,55 @@ def givensise(U):
     rotations and a sign matrix, such that U = Ur x E x Uc
     """
     
-    # Get the shape
-    d,r = U.shape
-    
-    # Create arrays for the two rotation matrices
-    Ur = np.identity(r)
-    Uc = np.identity(d)
-    
-    # Row space loop
-    for rr in reversed(range(r)):
-        for cc in range(rr):
-            
-            print((rr,cc))
-            
-            # Get the next pair of elements to be rotated
-            v = U[rr,[rr,cc]]
-            
-            # Find the plane rotation to zero the first element
-            _,g,_ = plane_rotation(v)
-            
-            # Build the Givens matrix
-            G = givens_matrix(g,cc,rr,r)
-
-            # Perform the rotation 
-            #TODO This is inefficient rotating the whole array - should do it on just the 4 elements affected.
-            tmp = U.copy()
-            U = np.dot(U,G);
-            Ur = np.dot(G.T,Ur)
[…]

### 2015-09-01 test_givens.py

@@ -1,0 +1,34 @@
+import unittest
+
+import numpy as np
+from scipy import linalg
+
+from givens import givensise
+
+class LinearModelOperationsTestCase(unittest.TestCase):
+    def setUp(self):
+        
+        # Test parameters
+        self.dMax = 10
+
+
+class GivensiseTestCase(LinearModelOperationsTestCase):
+    def runTest(self):
+        
+        for d in range(1,self.dMax+1):
+            for r in range(1,d):
+                
+                # Random matrix of orthogonal columns
+                X = np.random.randn(d,d)
+                A = np.dot(X,X.T)
+                eVal,eVec = linalg.eigh(A)
+                U = eVec[:,:2]
+                
+                # Factorise it
+                Uc,E,Ur = givensise(U)
+                
+                # Reconstruct
+                shouldBeU = np.dot(Uc,np.dot(E,Ur))
+                
+                # Test
+                np.testing.assert_almost_equal(U,shouldBeU)                

### 2015-09-02 givens.py
Added diagonal observation covariance learning

@@ -15,8 +15,6 @@ def plane_rotation(x):
     else:
         G = np.identity(2)
     
-    #print(np.arctan2(x[1],x[0]))
-    
     return G
 
 

### 2015-09-02 learners_mcmc.py

@@ -20,14 +20,18 @@ class AbstractMCMCLearner:
     """
     
     def __init__(self, initial_model_estimate, observ, hyperparams,
-                                             algoparams=dict(), verbose=False):
+                                            algoparams=dict(), verbose=False):
         self.model = initial_model_estimate
         self.observ = observ
         self.hyperparams = hyperparams
         self.algoparams = algoparams
+        
         self.chain = []
+        self.chain_model = []
+        self.chain_state = []
+        self.chain_lhood = []
+        
         self.verbose = verbose
-    
     
     def save_link(self):
         """Save the current state of the model as a link in the chain"""
@@ -111,24 +115,26 @@ class AbstractMCMCLearner:
             if dims is None:
                 dims = (None,None)
             fig, axs, coords = self._create_2d_plot_axes(paramName, dims[0],
-                                                                       dims[1])
+                                                                      dims[1])
         else:
             raise ValueError("Cannot draw plots for this parameter")
         
         for idx in np.ndindex(coords.shape):
-            samples = [mod.parameters[paramName][coords[idx]] for mod in self.chain]
+            samples = [mod.parameters[paramName][coords[idx]] \
+                                                        for mod in self.chain]
             axs[idx].plot(samples, 'k')
             ylims = axs[idx].get_ylim()
             axs[idx].plot([numBurnIn]*2, ylims, 'r:')
             axs[idx].set_ylim(ylims)
     
     
-    def plot_chain_histogram(self, paramName, numBurnIn=0, dims=None, trueValue=None):
+    def plot_chain_histogram(self, paramName, numBurnIn=0, dims=None,
+                                                              trueValue=None):
         """
         Make Markov chain histograms for a chosen parameter
         
-        dims is a tuple of two lists specificy which rows and columns should be
-        plotted. If empty then all are plotted.
+        dims is a tuple of two lists specificy which rows and columns should
+        be plotted. If empty then all are plotted.
         """
         
         # Get the parameter shape
@@ -140,16 +146,18 @@ class AbstractMCMCLearner:
             if dims is None:
                 dims = (None,None)
             fig, axs, coords = self._create_2d_plot_axes(paramName, dims[0],
-                                                                       dims[1])
+                                                                      dims[1])
         else:
[…]

### 2015-09-02 linear_models.py

@@ -139,10 +139,12 @@ class AbstractLinearModel:
                 prd_kk = self.initial_state_prior
             prd.set_instant(kk, prd_kk)
             
-            # Correction
-            flt_kk,innov = kal.correct(prd.get_instant(kk), observ[kk], H, R)
-            flt.set_instant(kk, flt_kk)
-            lhood = lhood + mvn.logpdf(observ[kk], innov.mn, innov.vr)
+            # Correction - skip if there are NaNs, indicating missing data
+            y = observ[kk]
+            if not np.any(np.isnan(y)):
+                flt_kk,innov = kal.correct(prd.get_instant(kk), y, H, R)
+                flt.set_instant(kk, flt_kk)
+                lhood = lhood + mvn.logpdf(observ[kk], innov.mn, innov.vr)
             
         return flt, prd, lhood
 

### 2015-09-02 run_basic_learning.py

@@ -33,6 +33,7 @@ est_params['F'] = 0.5*np.identity(ds)
 est_params['rank'] = np.array([2])
 est_params['vec'] = params['vec']#np.array([[1,0],[0,1],[0,0]])
 est_params['val'] = np.array([1,1])
+est_params['R'] = np.identity(do)
 est_model = DegenerateLinearModel(ds, do, prior, est_params)
 
 hyperparams = dict()
@@ -40,7 +41,8 @@ hyperparams['nu0'] = params['rank']
 hyperparams['Psi0'] = params['rank']*np.identity(ds)
 hyperparams['M0'] = np.zeros((ds,ds))
 hyperparams['V0'] = np.identity(ds)
-hyperparams['alpha'] = 1
+hyperparams['a0'] = 1
+hyperparams['b0'] = 0.2
 
 algoparams = dict()
 algoparams['Qs'] = 0.1
@@ -48,7 +50,7 @@ algoparams['Fs'] = 0.1
 
 learner = MCMCLearnerForDegenerateModelWithMNIWPrior(est_model, observ, hyperparams, algoparams=algoparams, verbose=True)
 
-num_iter = 2000
+num_iter = 200
 num_burn = int(num_iter/5)
 
 for ii in range(num_iter):
@@ -60,16 +62,19 @@ for ii in range(num_iter):
         learner.iterate_transition('rank')
     else:
         learner.iterate_transition('F')
+    learner.iterate_diagonal_observation_covariance()
     learner.save_link()
 
 learner.plot_chain_trace('F', numBurnIn=num_burn)
 learner.plot_chain_trace('rank', numBurnIn=num_burn)
+learner.plot_chain_trace('R', numBurnIn=num_burn)
 #learner.plot_chain_trace('Q', numBurnIn=num_burn)
 #learner.plot_chain_trace('val', numBurnIn=num_burn)
 #learner.plot_chain_trace('vec', numBurnIn=num_burn)
 
 learner.plot_chain_histogram('F', numBurnIn=num_burn, trueValue=params['F'])
 learner.plot_chain_histogram('rank', numBurnIn=num_burn, trueValue=params['rank'])
+learner.plot_chain_histogram('R', numBurnIn=num_burn, trueValue=params['R'])
 #learner.plot_chain_histogram('Q', numBurnIn=num_burn, trueValue=params['Q'])
 #learner.plot_chain_histogram('val', numBurnIn=num_burn, trueValue=params['val'])
[…]

### 2015-09-02 sampling.py

@@ -96,9 +96,12 @@ def matrix_normal_density(X, M, U, V):
     pdf = norm + pptn
     return pdf
 
-def evaluate_sufficient_statistics(x):
+
+
+def evaluate_transition_sufficient_statistics(x):
     """
     Calculate the required sufficient statistics from the state trajectory
+    for the transition model sampling operations.
     """
     K,ds = x.shape
     
@@ -118,6 +121,32 @@ def evaluate_sufficient_statistics(x):
     
     return suffStats
 
+def evaluate_observation_sufficient_statistics(x, y):
+    """
+    Calculate the required sufficient statistics from the state trajectory
+    for the observation model sampling operations.
+    """
+    K,ds = x.shape
+    _,do = y.shape
+    
+    # Create arrays
+    suffStats = list()
+    suffStats.append( 0 )
+    suffStats.append( np.zeros((ds,ds)) )
+    suffStats.append( np.zeros((do,ds)) )
+    suffStats.append( np.zeros((do,do)) )
+    
+    # Loop through time incrementing stats
+    for kk in range(K):
+        suffStats[0] += 1
+        suffStats[1] += np.outer(x[kk], x[kk])
+        suffStats[2] += np.outer(y[kk], x[kk])
+        suffStats[3] += np.outer(y[kk], y[kk])
+    
+    return suffStats
+
+
+
 def hyperparam_update_basic_mniw_transition(suffStats, nu0, Psi0, M0, V0):
     """
     Update matrix-normal-inverse-wishart hyperparameters for transition model
@@ -133,6 +162,26 @@ def hyperparam_update_basic_mniw_transition(suffStats, nu0, Psi0, M0, V0):
     
     return nu, Psi, M, V
 
+
+
+def hyperparam_update_basic_ig_observation_variance(suffStats, H, a0, b0):
+    """
+    Update inverse-gamma hyperparameters for the scale factor on the 
+    observation covariance (which is assumed to be diagonal).
+    """
+    dy = suffStats[3].shape[0]
+    Hss2 = np.dot(H,suffStats[2].T)
+    sumSquares = suffStats[3] - Hss2 - Hss2.T \
[…]

### 2015-09-02 test_basiclearning.py

@@ -58,7 +58,12 @@ class MNIWPriorLearning(BasicModelLearningTestCase):
         # MCMC
         for ii in range(self.num_iter):
             print("Running iteration {} of {}.".format(ii+1,self.num_iter))
-            learner.iterate_transition()
+            if (ii%3)==0:
+                learner.iterate_transition()
+            elif (ii%3)==1:
+                learner.iterate_transition_matrix()
+            elif (ii%3)==2:
+                learner.iterate_transition_covariance()
             learner.save_link()
         
         learner.plot_chain_trace('F', numBurnIn=self.num_burn)

### 2015-09-02 learners_mcmc.py
Added autocorrelation and derived parameter plots. Re-structured learning classes.

@@ -1,19 +1,21 @@
 # Operational modules
 from abc import ABCMeta, abstractmethod
+import types
+
 import numpy as np
 import scipy.linalg as la
 from scipy import stats
 from scipy import special
+from statsmodels.tsa import stattools
 import matplotlib.pyplot as plt
 
 # Import from other module files
 import sampling as smp
-from linear_models import BasicLinearModel
 
-class AbstractMCMCLearner:
+class BaseMCMCLearner:
     __metaclass__ = ABCMeta
     """
-    Abstract Container class for MCMC system learning algorithms.
+    Basic methods for MCMC system learning algorithms.
     Sub-class this and implement methods to perform individual MCMC sampling 
     steps. These will depend on the type of linear model and the choice of
     prior distribusion.
@@ -21,12 +23,16 @@ class AbstractMCMCLearner:
     
     def __init__(self, initial_model_estimate, observ, hyperparams,
                                             algoparams=dict(), verbose=False):
-        self.model = initial_model_estimate
         self.observ = observ
         self.hyperparams = hyperparams
+        
         self.algoparams = algoparams
         
-        self.chain = []
+        self.model = initial_model_estimate
+        self.flt,_,self.lhood = self.model.kalman_filter(self.observ)
+        self.state = self.model.backward_simulation(self.flt)
+        self.filter_current = True
+        
         self.chain_model = []
         self.chain_state = []
         self.chain_lhood = []
@@ -35,16 +41,34 @@ class AbstractMCMCLearner:
     
     def save_link(self):
         """Save the current state of the model as a link in the chain"""
-        self.chain.append(self.model.copy())
+        
+        # Need to make sure the likelihood value is current
+        if not self.filter_current:
+            self.flt,_,self.lhood = self.model.kalman_filter(self.observ)
+            self.filter_current = True
+        
+        self.chain_model.append(self.model.copy())
+        self.chain_state.append(self.state.copy())
+        self.chain_lhood.append(self.lhood)
+    
+    def sample_state_trajectory(self):
+        """
+        Sample a state trajectory conditional on the current model parameters
+        """
+        # Need to make sure the likelihood value is current
+        if not self.filter_current:
+            self.flt,_,self.lhood = self.model.kalman_filter(self.observ)
+            self.filter_current = True
+        self.state = self.model.backward_simulation(self.flt)
     
     
-    def _create_2d_plot_axes(self, paramName, index1=None, index2=None):
+    def _create_2d_plot_axes(self, param, index1=None, index2=None):
         """
[…]

### 2015-09-02 run_basic_learning.py

@@ -5,7 +5,14 @@ from matplotlib import pyplot as plt
 
 from kalman import GaussianDensity
 from linear_models import BasicLinearModel, DegenerateLinearModel
-from learners_mcmc import MCMCLearnerForDegenerateModelWithMNIWPrior
+from learners_mcmc import BaseMCMCLearner, MCMCLearnerObservationDiagonalCovarianceWithIGPrior, MCMCLearnerTransitionBasicModelWithMNIWPrior, MCMCLearnerTransitionDegenerateModelWithMNIWPrior
+
+# Create a learner class for this run
+class MCMCLearner(BaseMCMCLearner,
+                    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+                    MCMCLearnerTransitionDegenerateModelWithMNIWPrior):
+    pass
+            
 
 plt.close('all')
 
@@ -48,34 +55,36 @@ algoparams = dict()
 algoparams['Qs'] = 0.1
 algoparams['Fs'] = 0.1
 
-learner = MCMCLearnerForDegenerateModelWithMNIWPrior(est_model, observ, hyperparams, algoparams=algoparams, verbose=True)
+learner = MCMCLearner(est_model, observ, hyperparams, algoparams=algoparams, verbose=True)
 
-num_iter = 200
-num_burn = int(num_iter/5)
+num_iter = 800
+num_burn = int(num_iter/2)
+num_hold = num_burn/2
 
 for ii in range(num_iter):
     print("Running iteration {} of {}.".format(ii+1,num_iter))
     
     if (ii%3)==0:
-        learner.iterate_transition('Q')
+        learner.sample_transition_covariance('rotate')
     elif (ii%3)==1:
-        learner.iterate_transition('rank')
+        learner.sample_transition_covariance('rank')
     else:
-        learner.iterate_transition('F')
-    learner.iterate_diagonal_observation_covariance()
+        learner.sample_transition_matrix('F')
+    learner.sample_transition_within_subspace()
+    if ii > num_hold:
+        learner.sample_observation_diagonal_covariance()
+    learner.sample_state_trajectory()
     learner.save_link()
 
-learner.plot_chain_trace('F', numBurnIn=num_burn)
-learner.plot_chain_trace('rank', numBurnIn=num_burn)
-learner.plot_chain_trace('R', numBurnIn=num_burn)
[…]

### 2015-09-02 sampling.py

@@ -182,7 +182,7 @@ def hyperparam_update_basic_ig_observation_variance(suffStats, H, a0, b0):
     
     
 
-def hyperparam_update_basic_mniw_transition_matrix(suffStats, M0, V0):
+def hyperparam_update_basic_mn_transition_matrix(suffStats, M0, V0):
     """
     Update matrix-normal hyperparameters for transition matrix conditional
     on observed state trajectory.
@@ -196,19 +196,19 @@ def hyperparam_update_basic_mniw_transition_matrix(suffStats, M0, V0):
 
 
 
-#def hyperparam_update_basic_mniw_transition_covariance(suffStats, F, nu0, Psi0):
-#    """
-#    Update inverse-wishart hyperparameters for transition covariance matrix
-#    conditional on observed state trajectory.
-#    """
-#    
-#    # Posterior hyperparameters    
-#    Fss2 = np.dot(F,suffStats[2].T)
-#    nu  = nu0 + suffStats[0]
-#    Psi = Psi0 + suffStats[3] - Fss2 - Fss2.T \
-#                                          + np.dot(np.dot(F,suffStats[1]),F.T)
-#    
-#    return nu, Psi
+def hyperparam_update_basic_iw_transition_covariance(suffStats, F, nu0, Psi0):
+    """
+    Update inverse-wishart hyperparameters for transition covariance matrix
+    conditional on observed state trajectory.
+    """
+    
+    # Posterior hyperparameters    
+    Fss2 = np.dot(F,suffStats[2].T)
+    nu  = nu0 + suffStats[0]
+    Psi = Psi0 + suffStats[3] - Fss2 - Fss2.T \
+                                          + np.dot(np.dot(F,suffStats[1]),F.T)
+    
+    return nu, Psi
 
 
 

### 2015-09-02 test_basiclearning.py

@@ -49,8 +49,8 @@ class MNIWPriorLearning(BasicModelLearningTestCase):
         
         # Set up learning
         hyperparams = dict()
-        hyperparams['nu0'] = 3.01
-        hyperparams['Psi0'] = (hyperparams['nu0']-self.ds-1)*np.identity(self.ds)
+        hyperparams['nu0'] = self.ds-1
+        hyperparams['Psi0'] = hyperparams['nu0']*0.1*np.identity(self.ds)
         hyperparams['M0'] = np.zeros(self.ds)
         hyperparams['V0'] = np.identity(self.ds)
         learner = MCMCLearnerForBasicModelWithMNIWPrior(self.est_model, self.observ, hyperparams)
@@ -58,12 +58,8 @@ class MNIWPriorLearning(BasicModelLearningTestCase):
         # MCMC
         for ii in range(self.num_iter):
             print("Running iteration {} of {}.".format(ii+1,self.num_iter))
-            if (ii%3)==0:
-                learner.iterate_transition()
-            elif (ii%3)==1:
-                learner.iterate_transition_matrix()
-            elif (ii%3)==2:
-                learner.iterate_transition_covariance()
+            learner.sample_state_trajectory()
+            learner.sample_transition()
             learner.save_link()
         
         learner.plot_chain_trace('F', numBurnIn=self.num_burn)

### 2015-09-02 learners_mcmc.py
Added MH proposal parameter adaptation

@@ -37,6 +37,9 @@ class BaseMCMCLearner:
         self.chain_state = []
         self.chain_lhood = []
         
+        self.chain_accept = dict()
+        self.chain_algoparams = dict()
+        
         self.verbose = verbose
     
     def save_link(self):
@@ -50,6 +53,21 @@ class BaseMCMCLearner:
         self.chain_model.append(self.model.copy())
         self.chain_state.append(self.state.copy())
         self.chain_lhood.append(self.lhood)
+
+        
+    def adapt_algorithm_parameters(self, batchLength=20, targetRate=0.44,
+                                                                minReact=0.1):
+        """
+        Adapt algorithm parameters based on recent acceptance history.
+        """
+        for param in list(self.algoparams.keys()):
+            length = np.minimum(batchLength,len(self.chain_accept[param]))
+            actualRate = np.mean(np.array(self.chain_accept[param][-length:]))
+            expnt = np.min(minReact,1.0/np.sqrt(length))
+            if actualRate < targetRate:
+                expnt *= -1
+            self.algoparams[param] *= np.exp(expnt)
+                    
     
     def sample_state_trajectory(self):
         """
@@ -122,6 +140,44 @@ class BaseMCMCLearner:
         return fig, axs, coords
     
     
+    def plot_chain_accept(self, paramName=None, numBurnIn=0):
+        """
+        Make Markov chain accpetance plots for algorithm parameters
+        """
+        
+        if paramName is None:
+            paramName = list(self.chain_accept.keys())
+        numParams = len(paramName)
+        
+        fig, axs = plt.subplots(nrows=1, ncols=numParams, squeeze=False)
+        
+        for ii in range(numParams):
+            samples = self.chain_accept[paramName[ii]]
+            axs[0,ii].plot(np.cumsum(samples), 'k')
+            axs[0,ii].set_title(paramName[ii])
+        
+        plt.show()
+        
+        
+    def plot_chain_adapt(self, paramName=None, numBurnIn=0):
+        """
+        Make Markov chain adaptive algorithm parameters
+        """
+        
+        if paramName is None:
+            paramName = list(self.chain_algoparams.keys())
+        numParams = len(paramName)
+        
+        fig, axs = plt.subplots(nrows=1, ncols=numParams, squeeze=False)
+        
+        for ii in range(numParams):
[…]

### 2015-09-02 run_basic_learning.py

@@ -52,14 +52,14 @@ hyperparams['a0'] = 1
 hyperparams['b0'] = 0.2
 
 algoparams = dict()
-algoparams['Qs'] = 0.1
-algoparams['Fs'] = 0.1
+algoparams['rotate'] = 0.1
+algoparams['perturb'] = 0.1
 
 learner = MCMCLearner(est_model, observ, hyperparams, algoparams=algoparams, verbose=True)
 
-num_iter = 800
+num_iter = 1000
 num_burn = int(num_iter/2)
-num_hold = num_burn/2
+num_hold = int(num_burn/2)
 
 for ii in range(num_iter):
     print("Running iteration {} of {}.".format(ii+1,num_iter))
@@ -69,12 +69,15 @@ for ii in range(num_iter):
     elif (ii%3)==1:
         learner.sample_transition_covariance('rank')
     else:
-        learner.sample_transition_matrix('F')
+        learner.sample_transition_matrix()
     learner.sample_transition_within_subspace()
     if ii > num_hold:
         learner.sample_observation_diagonal_covariance()
     learner.sample_state_trajectory()
     learner.save_link()
+    
+    if ((ii+1)%20)==0:
+        learner.adapt_algorithm_parameters()
 
 learner.plot_chain_trace('F', numBurnIn=num_burn, trueModel=model)
 learner.plot_chain_trace('transition_covariance', numBurnIn=num_burn, trueModel=model, derived=True)
@@ -88,3 +91,6 @@ learner.plot_chain_histogram('R', numBurnIn=num_burn, trueModel=model)
 
 learner.plot_chain_acf('F', numBurnIn=num_burn)
 learner.plot_chain_acf('R', numBurnIn=num_burn)
+
+learner.plot_chain_accept()
+learner.plot_chain_adapt()

### 2015-09-02 learners_mcmc.py
Sorted out simple unit tests

@@ -1,6 +1,5 @@
 # Operational modules
-from abc import ABCMeta, abstractmethod
-import types
+from abc import ABCMeta
 
 import numpy as np
 import scipy.linalg as la
@@ -12,49 +11,49 @@ import matplotlib.pyplot as plt
 # Import from other module files
 import sampling as smp
 
-class BaseMCMCLearner:
+class BaseMCMCLearner():
     __metaclass__ = ABCMeta
     """
     Basic methods for MCMC system learning algorithms.
-    Sub-class this and implement methods to perform individual MCMC sampling 
+    Sub-class this and implement methods to perform individual MCMC sampling
     steps. These will depend on the type of linear model and the choice of
     prior distribusion.
     """
-    
+
     def __init__(self, initial_model_estimate, observ, hyperparams,
                                             algoparams=dict(), verbose=False):
         self.observ = observ
         self.hyperparams = hyperparams
-        
+
         self.algoparams = algoparams
-        
+
         self.model = initial_model_estimate
         self.flt,_,self.lhood = self.model.kalman_filter(self.observ)
         self.state = self.model.backward_simulation(self.flt)
         self.filter_current = True
-        
+
         self.chain_model = []
         self.chain_state = []
         self.chain_lhood = []
-        
+
         self.chain_accept = dict()
         self.chain_algoparams = dict()
-        
+
         self.verbose = verbose
-    
+
     def save_link(self):
         """Save the current state of the model as a link in the chain"""
-        
+
         # Need to make sure the likelihood value is current
         if not self.filter_current:
             self.flt,_,self.lhood = self.model.kalman_filter(self.observ)
             self.filter_current = True
-        
+
         self.chain_model.append(self.model.copy())
         self.chain_state.append(self.state.copy())
         self.chain_lhood.append(self.lhood)
 
-        
+
     def adapt_algorithm_parameters(self, batchLength=20, targetRate=0.44,
                                                                 minReact=0.1):
         """
@@ -67,8 +66,8 @@ class BaseMCMCLearner:
             if actualRate < targetRate:
                 expnt *= -1
             self.algoparams[param] *= np.exp(expnt)
-                    
-    
+
+
     def sample_state_trajectory(self):
         """
         Sample a state trajectory conditional on the current model parameters
@@ -78,16 +77,16 @@ class BaseMCMCLearner:
             self.flt,_,self.lhood = self.model.kalman_filter(self.observ)
             self.filter_current = True
         self.state = self.model.backward_simulation(self.flt)
-    
[…]

### 2015-09-02 linear_models.py

@@ -19,55 +19,55 @@ class AbstractLinearModel:
     Abstract Linear-Gaussian Model Class.
     This implements various standard procedures (such as filtering and
     sampling) which are common to any parameterisation. Derived classes should
-    implement the abstract methods which supply the system matrices. Parameters
-    are held in the the parameters dictionary and should each be a numpy array
-    (even if they are scalars).
+    implement the abstract methods which supply the system matrices.
+    Parameters are held in the the parameters dictionary and should each be a
+    numpy array (even if they are scalars).
     """
-    
+
     def __init__(self,
                  state_dimension,
                  observation_dimension,
                  initial_state_prior,
                  parameters):
         """Initialise with model parameters"""
-        
+
         # Consistency checks
         ds = state_dimension
-        if (state_dimension<1) or (observation_dimension<1):
+        if (state_dimension < 1) or (observation_dimension < 1):
             raise ValueError("Invalid state or observation dimensions")
-        if (initial_state_prior.mn.shape!=(ds,)) \
-                                   or (initial_state_prior.vr.shape!=(ds,ds)):
+        if (initial_state_prior.mn.shape != (ds,)) \
+                                 or (initial_state_prior.vr.shape != (ds,ds)):
             raise ValueError("Invalid initial state prior density")
-        
+
         # Store them
         self.ds = state_dimension
         self.do = observation_dimension
         self.initial_state_prior = initial_state_prior
         self.parameters = parameters
-    
-    
+
+
     def copy(self):
         return self.__class__(self.ds, self.do, self.initial_state_prior,
                                                     deepcopy(self.parameters))
-    
-    
+
+
     @abstractmethod
     def transition_matrix(self):
         pass
-    
+
     @abstractmethod
     def transition_covariance(self):
         pass
-    
+
     @abstractmethod
     def observation_matrix(self):
         pass
-    
+
     @abstractmethod
     def observation_covariance(self):
         pass
-    
-    
+
+
     def simulate_data(self, num_time_instants):
         """Sample data a priori according to the model"""
         state = self.sample_state(num_time_instants)
@@ -77,111 +77,111 @@ class AbstractLinearModel:
 
     def sample_state(self, num_time_instants):
         """Sample a sequence of states from the prior"""
-        
+
         # Get system matices
[…]

### 2015-09-02 run_basic_learning.py

@@ -1,96 +1,76 @@
-from timeit import default_timer as timer
+from copy import deepcopy
 
 import numpy as np
 from matplotlib import pyplot as plt
 
 from kalman import GaussianDensity
-from linear_models import BasicLinearModel, DegenerateLinearModel
-from learners_mcmc import BaseMCMCLearner, MCMCLearnerObservationDiagonalCovarianceWithIGPrior, MCMCLearnerTransitionBasicModelWithMNIWPrior, MCMCLearnerTransitionDegenerateModelWithMNIWPrior
+from linear_models import BasicLinearModel
+from learners_mcmc import (
+    BaseMCMCLearner,
+    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+    MCMCLearnerTransitionBasicModelWithMNIWPrior)
 
 # Create a learner class for this run
 class MCMCLearner(BaseMCMCLearner,
                     MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
-                    MCMCLearnerTransitionDegenerateModelWithMNIWPrior):
+                    MCMCLearnerTransitionBasicModelWithMNIWPrior):
     pass
-            
+
 
 plt.close('all')
 
-K = 100
+K = 200
 ds = 3
 do = 3
 
 params = dict()
-params['F'] = np.array([[0.9,0.8,0.7],[0,0.9,0.8],[0,0,0.7]])
-params['rank'] = np.array([2])
-params['vec'] = (1./np.sqrt(3))*np.array([[1,1],[1,1],[1,-1]])#np.identity(ds)
-params['val'] = np.array([1./5,1./2])
-#params['Q'] = np.array([[0.1,0.1],[0.1,0.1]])#np.array([[0.1,0.05],[0.05,0.1]])
-params['H'] = np.identity(do)#np.array([[1,0]])
+params['F'] = np.array([[0.9,0.8,0.7],[0,0.9,0.8],[0,0,0.7]])#0.9*np.identity(ds)#
+params['Q'] = np.array([[0.1,0.05,0],[0.05,0.1,0.05],[0,0.05,0.1]])
+params['H'] = np.identity(do)#np.array([[1,0,0],[0,1,0]])#
 params['R'] = 0.1*np.identity(do)#np.array([[1]])
 
 prior = GaussianDensity(np.zeros(ds), 100*np.identity(ds))
-model = DegenerateLinearModel(ds, do, prior, params)
[…]

### 2015-09-02 run_degenerate_learning.py

@@ -1,0 +1,98 @@
+from copy import deepcopy
+
+import numpy as np
+from matplotlib import pyplot as plt
+
+from kalman import GaussianDensity
+from linear_models import DegenerateLinearModel
+from learners_mcmc import (
+    BaseMCMCLearner, 
+    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+    MCMCLearnerTransitionDegenerateModelWithMNIWPrior)
+    
+# Create a learner class for this run
+class MCMCLearner(BaseMCMCLearner,
+                    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+                    MCMCLearnerTransitionDegenerateModelWithMNIWPrior):
+    pass
+            
+
+plt.close('all')
+
+K = 100
+ds = 3
+do = 3
+
+params = dict()
+params['F'] = np.array([[0.9,0.8,0.7],[0,0.9,0.8],[0,0,0.7]])
+params['rank'] = np.array([2])
+params['vec'] = (1./np.sqrt(3))*np.array([[1,1],[1,1],[1,-1]])
+params['val'] = np.array([1./5,1./2])
+params['H'] = np.identity(do)
+params['R'] = 0.1*np.identity(do)
+
+prior = GaussianDensity(np.zeros(ds), 100*np.identity(ds))
+model = DegenerateLinearModel(ds, do, prior, params)
+
+np.random.seed(0)
+state, observ = model.simulate_data(K)
+
+est_params = deepcopy(params)
+est_params['F'] = 0.5*np.identity(ds)
+est_params['rank'] = np.array([2])
+est_params['vec'] = np.array([[1,0],[0,1],[0,0]])
+est_params['val'] = np.array([1,1])
+est_params['R'] = np.identity(do)
+est_model = DegenerateLinearModel(ds, do, prior, est_params)
+
+hyperparams = dict()
+hyperparams['nu0'] = params['rank']
+hyperparams['Psi0'] = params['rank']*np.identity(ds)
+hyperparams['M0'] = np.zeros((ds,ds))
+hyperparams['V0'] = np.identity(ds)
+hyperparams['a0'] = 1
+hyperparams['b0'] = 0.2
+
+algoparams = dict()
+algoparams['rotate'] = 0.1
+algoparams['perturb'] = 0.1
+
+learner = MCMCLearner(est_model, observ, hyperparams, algoparams=algoparams, verbose=True)
+
[…]

### 2015-09-02 sampling.py

@@ -6,7 +6,7 @@ from scipy import special
 def sample_wishart(nu, P):
     """
     Sample from a Wishart distribution.
-    
+
     N.B. There is a function in scipy 0.17, the development version,
     but not the current 0.16. Probably worth updating when it gets released.
     """
@@ -34,25 +34,25 @@ def sample_cayley(d, s):
     Sample a random orthogonal matrix using normal variates and the Cayley
     transformation
     """
-    
+
     # Random skew-symmetric matrix
     S = np.zeros((d,d))
     for dd in range(d-1):
         y = stats.norm.rvs(loc=np.zeros(d-dd-1), scale=s*np.ones(d-dd-1))
         S[dd,dd+1:] = y
         S[dd+1:,dd] = -y
-    
+
     # Cayley transformation
     I = np.identity(d)
     M = la.solve(I-S,I+S)
-    
+
     return M
 
 def sample_orthogonal_haar(d):
     """
     Sample a unit vector uniformly from the (d-1)-sphere
     """
-    z = stats.multivariate_normal.rvs(mean=np.zeros(d),cov=np.identity(d))    
+    z = stats.multivariate_normal.rvs(mean=np.zeros(d),cov=np.identity(d))
     z /= la.norm(z)
     if d == 1:
         z = z[np.newaxis] # This undoes the squeeze if only 1D
@@ -64,9 +64,9 @@ def singular_wishart_density(val, vec, P):
     """
     d,r = vec.shape
     norm = 0.5*r*(r-d)*np.log(np.pi) - 0.5*r*d*np.log(2.0) \
-           - special.multigammaln(r/2,r) - 0.5*r*la.det(P)
+          - special.multigammaln(r/2,r) - 0.5*r*la.det(P)
     pptn = 0.5*(r-d-1)*np.log(np.prod(val)) \
-           - 0.5*np.trace( np.dot(np.dot(vec.T,la.solve(P,vec)),np.diag(val)) )
+          - 0.5*np.trace( np.dot(np.dot(vec.T,la.solve(P,vec)),np.diag(val)) )
     pdf = norm + pptn
     return pdf
 
@@ -76,15 +76,15 @@ def singular_inverse_wishart_density(val, vec, P):
     X = vec*diag(val)*vec.T
     """
     d,r = vec.shape
-    
+
     norm = 0.5*r*la.det(P) \
             -0.5*r*d*np.log(2.0) \
             -0.5*r*(d-r)*np.log(np.pi) \
             -special.multigammaln(r/2,r)
-    
+
[…]

### 2015-09-02 test_basiclearning.py

@@ -4,41 +4,50 @@ import numpy as np
 
 from kalman import GaussianDensity
 from linear_models import BasicLinearModel
-from learners_mcmc import MCMCLearnerForBasicModelWithMNIWPrior
+from learners_mcmc import (
+    BaseMCMCLearner, 
+    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+    MCMCLearnerTransitionBasicModelWithMNIWPrior)
+
+# Create a learner class for this test
+class MCMCLearner(BaseMCMCLearner,
+                    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+                    MCMCLearnerTransitionBasicModelWithMNIWPrior):
+    pass
 
 class BasicModelLearningTestCase(unittest.TestCase):
     def setUp(self):
-        
+
         # Basic parameters
         self.K = 100
-        self.ds = 2
-        self.do = 2
-        
+        self.ds = 3
+        self.do = 3
+
         # System matrices
         params = dict()
-        params['F'] = np.array([[0.9,0.81],[0,0.9]])
-        params['Q'] = np.array([[0.1,0],[0,0.1]])
-        params['H'] = np.array([[1,0],[0,1]])
-        params['R'] = np.array([[0.1,0],[0,0.1]])
+        params['F'] = np.array([[0.9,0.8,0.7],[0,0.9,0.8],[0,0,0.7]])
+        params['Q'] = np.array([[0.1,0.05,0],[0.05,0.1,0.05],[0,0.05,0.1]])
+        params['H'] = np.identity(self.do)
+        params['R'] = 0.1*np.identity(self.do)
         self.params = params
-        
+
         # Create model
-        prior = GaussianDensity(np.array([0,0]), np.array([[100,0],[0,100]]))
-        self.model = BasicLinearModel(self.ds, self.do, prior, params)
-        
+        prior = GaussianDensity(np.zeros(self.ds), np.identity(self.ds))
+        self.model = BasicLinearModel(self.ds, self.do, prior, self.params)
+
         # Simulate data
-        np.random.seed(0)
+        np.random.seed(1)
         self.state, self.observ = self.model.simulate_data(self.K)
-        
+
         # Create initial estimated model
         est_params = dict()
-        est_params['F'] = np.array([[0.5,0],[0,0.5]])
-        est_params['Q'] = 0.3*np.identity(self.ds)
[…]

### 2015-09-02 test_degeneratelearning.py

@@ -1,0 +1,138 @@
+import unittest
+
+import numpy as np
+
+from kalman import GaussianDensity
+from linear_models import DegenerateLinearModel
+from learners_mcmc import (
+    BaseMCMCLearner, 
+    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+    MCMCLearnerTransitionDegenerateModelWithMNIWPrior)
+
+# Create a learner class for this test
+class MCMCLearner(BaseMCMCLearner,
+                    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+                    MCMCLearnerTransitionDegenerateModelWithMNIWPrior):
+    pass
+
+class DegenerateModelLearningTestCase(unittest.TestCase):
+    def setUp(self):
+
+        # Basic parameters
+        self.K = 100
+        self.ds = 3
+        self.do = 3
+
+        # System matrices
+        params = dict()
+        params['F'] = np.array([[0.9,0.8,0.7],[0,0.9,0.8],[0,0,0.7]])
+        params['rank'] = np.array([2])
+        params['vec'] = (1./np.sqrt(3))*np.array([[1,1],[1,1],[1,-1]])
+        params['val'] = np.array([1./5,1./2])
+        params['H'] = np.identity(self.do)
+        params['R'] = 0.1*np.identity(self.do)
+        self.params = params
+
+        # Create model
+        prior = GaussianDensity(np.zeros(self.ds), np.identity(self.ds))
+        self.model = DegenerateLinearModel(self.ds, self.do, prior, 
+                                                                  self.params)
+
+        # Simulate data
+        np.random.seed(1)
+        self.state, self.observ = self.model.simulate_data(self.K)
+
+        # Create initial estimated model
+        est_params = dict()
+        est_params['F'] = 0.5*np.identity(self.ds)
+        est_params['rank'] = np.array([2])
+        est_params['vec'] = np.array([[1.0,0.0],[0.0,1.0],[0.0,0.0]])
+        est_params['val'] = np.array([1,1])
+        est_params['H'] = np.identity(self.do)
+        est_params['R'] = np.identity(self.do)
+        est_model = DegenerateLinearModel(self.ds, self.do, prior, est_params)
+        self.est_model = est_model
+
+        # Set MCMC parameters
+        self.num_iter = 2000
+        self.num_burn = int(self.num_iter/5)
+
+
+class MNIWPriorLearning(DegenerateModelLearningTestCase):
[…]

### 2015-09-02 learners_mcmc.py
Added motion capture demo

@@ -342,7 +342,12 @@ class MCMCLearnerObservationDiagonalCovarianceWithIGPrior():
                                     self.hyperparams['b0'])
 
         # Sample new parameter
-        r = stats.invgamma.rvs(a, scale=b)
+        try:
+            r = stats.invgamma.rvs(a, scale=b)
+        except ValueError as e:
+            print(a)
+            print(b)
+            raise e
         self.model.parameters['R'] = r*np.identity(self.model.do)
 
         # self.flt and self.lhood are no longer up-to-date

### 2015-09-02 linear_models.py

@@ -143,8 +143,10 @@ class AbstractLinearModel:
             y = observ[kk]
             if not np.any(np.isnan(y)):
                 flt_kk,innov = kal.correct(prd.get_instant(kk), y, H, R)
-                flt.set_instant(kk, flt_kk)
                 lhood = lhood + mvn.logpdf(observ[kk], innov.mn, innov.vr)
+            else:
+                flt_kk = prd_kk
+            flt.set_instant(kk, flt_kk)
 
         return flt, prd, lhood
 

### 2015-09-17 learners_smc.py
Moved the resample-move to after the resampling.

@@ -186,9 +186,9 @@ class DegenerateSMCLearner():
         d = chain_model[0]['F'].shape[0]
         r = chain_model[0]['val'].shape[0]
         self.approx[d] = DegenerateModelSMCApproximation(N, d, r)
-        self.state[d] = np.zeros((N, 1, K, d))
+        self.state[d-1] = np.zeros((N, 1, K, d)) # Note that the (r-1)th entry of state stores samples corresponding to the (r)th rank
         for nn in range(N):
-            self.state[d][nn,0,:,:] = chain_state[nn].copy()
+            self.state[d-1][nn,0,:,:] = chain_state[nn].copy()
             self.approx[d].F[nn,:,:] = chain_model[nn]['F'].copy()
             self.approx[d].val[nn,:] = chain_model[nn]['val'].copy()
             self.approx[d].vec[nn,:,:] = chain_model[nn]['vec'].copy()
@@ -201,7 +201,7 @@ class DegenerateSMCLearner():
                                                         hyperparams)
 
 
-
+        self.filters = None
 
         self.observ = observ
         self.hyperparams = hyperparams
@@ -230,6 +230,15 @@ class DegenerateSMCLearner():
             raise ValueError("Need to do rank {} first.".format(rank+1))
         self.approx[rank] = DegenerateModelSMCApproximation(self.num_samples,
                                                             self.ds, rank)
+
+        # Create space to store the state trajectories
+        self.state[rank] = np.zeros((self.num_samples,
+                                     self.num_rejuv,
+                                     self.K,
+                                     self.ds))
+
+        # Create a store for the filter results for RM in the next iteration
+        filters = []
 
         # Resampling
         w = self.approx[rank+1].weight.copy()
@@ -264,6 +273,30 @@ class DegenerateSMCLearner():
                                           self.initial_state_prior,
                                           parameters)
 
+            # Resample-move with Gibbs sampling to improve diversity
+            if (self.filters is not None) and (self.num_rejuv > 0):
+                flt = self.filters[ai]
+                for ii in range(self.num_rejuv):
+                    state = model.backward_simulation(flt)
+                    model = sample_transition_within_subspace(model, state,
+                                                              self.hyperparams)
+                    model = sample_observation_diagonal_covariance(
+                                                            model,
+                                                            state,
+                                                            self.observ,
+                                                            self.hyperparams)
+                    flt,_,old_lhood = model.kalman_filter(self.observ)
+                    self.state[rank][nn,ii,:,:] = state
+                old_prior = transition_prior(rank,
[…]

### 2015-09-17 run_degenerate_learning_smc.py

@@ -62,7 +62,7 @@ algoparams = dict()
 
 learner = MCMCLearner(est_model, observ, hyperparams, algoparams=algoparams, verbose=True)
 
-num_iter = 1000
+num_iter = 20
 num_burn = int(num_iter/2)
 
 # MCMC with full rank
@@ -96,12 +96,12 @@ for rr in reversed(range(1,ds)):
 
 fig = plt.figure()
 ax = fig.add_subplot(1,1,1)
-ax.plot(range(ds), mlhoodratios)
+ax.plot(range(1,ds), np.cumsum(mlhoodratios[::-1])[::-1][1:])
 plt.show()
 
 fig = plt.figure()
 ax = fig.add_subplot(1,1,1)
-ax.plot(range(ds), effsampsizes)
+ax.plot(range(1,ds), effsampsizes[1:])
 plt.show()
 
 #learner.plot_chain_trace('F', numBurnIn=num_burn, trueModel=model)

### 2015-09-17 run_mocap_smc.py

@@ -27,10 +27,10 @@ np.random.seed(0)
 data_path = './mocap-data/'
 test_data_file = './results/mocap-test-data.p'
 
-num_iter = 1000
+num_iter = 750
 num_burn = 500
 thin = 5
-num_rejuv = 1
+num_rejuv = 2
 
 # Import marker data
 markers = np.genfromtxt(data_path+'downsampled_head_markers.csv', delimiter=',')

### 2015-09-17 learners_mcmc_augmented.py
Added augmented state MCMC learner for degenerate linear model

@@ -1,0 +1,446 @@
+# Operational modules
+from abc import ABCMeta
+import pickle
+
+import numpy as np
+import scipy.linalg as la
+from scipy import stats
+from scipy import special
+from statsmodels.tsa import stattools
+import matplotlib.pyplot as plt
+
+# Import from other module files
+from learners_mcmc import BaseMCMCLearner
+import sampling as smp
+
+
+def complete_basis(vec):
+    """
+    Returns a matrix of vectors orthogonal to the eigenvectors for the
+    transition covariance
+    """
+    d,r = vec.shape
+    Q,_ = la.qr(vec)
+    return Q[:,r:]
+
+
+def effective_sample_size(weight):
+    """
+    Calculate effective sample size for a set of unnormalised importance
+    log-weights.
+    """
+    w = weight.copy()
+    w -= np.max(w)
+    w = np.exp(w)
+    w /= np.sum(w)
+    return 1.0/(np.sum(w**2))
+
+
+def normalising_constant_estimate(weight):
+    """
+    Estimate normalising constant from a set of unnormalised importance
+    log-weights
+    """
+    w = weight.copy()
+    wmax = np.max(w)
+    w -= wmax
+    w = np.exp(w)
+    return np.log(np.sum(w)) + wmax
+
+
+def transition_prior(rank, val, vec, F, hyperparams):
+    """
+    Prior density for transition model parameters
+    """
+    Psi0 = rank*hyperparams['rPsi0']
+    variancePrior = smp.singular_inverse_wishart_density(val, vec, Psi0)
+
+    orthVec = complete_basis(vec)
+#    relaxEval = self.hyperparams['alpha']
+#    relaxEval = np.min(model.parameters['val'])
+    relaxEval = np.max(val)
+    rowVariance = np.dot(vec, np.dot(np.diag(val), vec.T)) \
+                  + relaxEval*np.dot(orthVec,orthVec.T)
+    matrixPrior = smp.matrix_normal_density(F,
+                                            hyperparams['M0'],
+                                            rowVariance,
+                                            hyperparams['V0'])
+
+    return variancePrior + matrixPrior
+
+
+def extended_density(extraVal, extraVec, val):
+    """
+    Artificial conditional 'extension' density for extra eigenvalue/vector.
+    This is currently uniform over the interval [0,l] for the eigenvalue
[…]

### 2015-09-17 learners_smc.py

@@ -186,9 +186,9 @@ class DegenerateSMCLearner():
         d = chain_model[0]['F'].shape[0]
         r = chain_model[0]['val'].shape[0]
         self.approx[d] = DegenerateModelSMCApproximation(N, d, r)
-        self.state[d-1] = np.zeros((N, 1, K, d)) # Note that the (r-1)th entry of state stores samples corresponding to the (r)th rank
+        self.state[d] = np.zeros((N, 1, K, d)) # Note that the (r-1)th entry of state stores samples corresponding to the (r)th rank
         for nn in range(N):
-            self.state[d-1][nn,0,:,:] = chain_state[nn].copy()
+            self.state[d][nn,0,:,:] = chain_state[nn].copy()
             self.approx[d].F[nn,:,:] = chain_model[nn]['F'].copy()
             self.approx[d].val[nn,:] = chain_model[nn]['val'].copy()
             self.approx[d].vec[nn,:,:] = chain_model[nn]['vec'].copy()
@@ -323,10 +323,10 @@ class DegenerateSMCLearner():
                      - jac \
                      + exten
 
-            print(lhood-old_lhood)
-            print(prior-old_prior)
-            print(exten)
-            print(jac)
+#            print(lhood-old_lhood)
+#            print(prior-old_prior)
+#            print(exten)
+#            print(jac)
 
             if self.verbose:
                 print("Particle log-weight: {}".format(weight))
@@ -356,13 +356,13 @@ class DegenerateSMCLearner():
 
 
 
-    def estimate_state_trajectory(self, numBurnIn=0):
+    def estimate_state_trajectory(self, rank):
         """
         Estimate of the state trajectory (mean and standard deviation) using
-        all the samples in the chain
+        the samples from a particular rank of covariance matrix
         """
-        #TODO Rewrite for SMC
-        samples = np.array(self.chain_state[numBurnIn:])
+        shape = (self.num_rejuv*self.num_samples, self.K, self.ds)
+        samples = np.reshape(self.state[rank-1], shape,order='F')
         mn = np.mean(samples, axis=0)
         sd = np.std(samples, axis=0)
         return mn, sd

### 2015-09-17 run_degenerate_learning_mcmc_augmented.py

@@ -1,0 +1,94 @@
+from copy import deepcopy
+
+import numpy as np
+from matplotlib import pyplot as plt
+
+from kalman import GaussianDensity
+from linear_models import DegenerateLinearModel
+from learners_mcmc_augmented import CleverDegenerateMCMCLearner
+
+plt.close('all')
+filename = './results/toy-mcmc-degenerate.p'
+
+K = 100
+ds = 3
+do = 3
+
+params = dict()
+params['F'] = np.array([[0.9,0.8,0.7],[0,0.9,0.8],[0,0,0.7]])
+params['rank'] = np.array([2])
+params['vec'] = (1./np.sqrt(3))*np.array([[1,1],[1,1],[1,-1]])
+params['val'] = np.array([1./5,1./2])
+params['H'] = np.identity(do)
+params['R'] = 0.1*np.identity(do)
+
+prior = GaussianDensity(np.zeros(ds), 100*np.identity(ds))
+model = DegenerateLinearModel(ds, do, prior, params)
+
+np.random.seed(0)
+state, observ = model.simulate_data(K)
+
+est_params = deepcopy(params)
+est_params['F'] = 0.5*np.identity(ds)
+est_params['rank'] = np.array([ds])
+est_params['vec'] = np.identity(ds)
+est_params['val'] = np.array([1./3, 1./4, 1./5])
+est_params['R'] = np.identity(do)
+est_model = DegenerateLinearModel(ds, do, prior, est_params)
+
+hyperparams = dict()
+hyperparams['nu0'] = params['rank']
+hyperparams['rPsi0'] = np.identity(ds)
+hyperparams['M0'] = np.zeros((ds,ds))
+hyperparams['V0'] = np.identity(ds)
+hyperparams['a0'] = 1
+hyperparams['b0'] = 0.2
+hyperparams['alpha'] = 0.001
+
+algoparams = dict()
+algoparams['rotate'] = 1E-1
+algoparams['perturb'] = 1E-6
+
+learner = CleverDegenerateMCMCLearner(est_model, observ, hyperparams,
+                                      algoparams, verbose=True)
+
+num_iter = 200
+num_burn = int(num_iter/2)
+num_warm = 20
+
+for ii in range(num_warm):
+    learner.sample_transition_warm_up()
+
+for ii in range(num_iter):
+    print("")
+    print("Running iteration {} of {}.".format(ii+1,num_iter))
+
+    rchange = np.random.random_integers(-1,1)
[…]

### 2015-09-17 run_degenerate_learning_smc.py

@@ -87,12 +87,16 @@ smclearner = DegenerateSMCLearner(learner.chain_model[num_burn:],
 
 mlhoodratios = np.zeros(ds)
 effsampsizes = np.zeros(ds)
+rmse = np.zeros(ds)
 for rr in reversed(range(1,ds)):
     print("")
     print("Running SMC sampler for rank {}.".format(rr))
     smclearner.smc_reduce_rank(rr)
     mlhoodratios[rr] = normalising_constant_estimate(smclearner.approx[rr].weight)
     effsampsizes[rr] = effective_sample_size(smclearner.approx[rr].weight)
+    if rr > 2:
+        state_mn, state_sd = smclearner.estimate_state_trajectory(rr)
+        rmse[rr] = np.sqrt(np.sum((state-state_mn)**2))
 
 fig = plt.figure()
 ax = fig.add_subplot(1,1,1)
@@ -102,6 +106,11 @@ plt.show()
 fig = plt.figure()
 ax = fig.add_subplot(1,1,1)
 ax.plot(range(1,ds), effsampsizes[1:])
+plt.show()
+
+fig = plt.figure()
+ax = fig.add_subplot(1,1,1)
+ax.plot(range(1,ds), rmse[1:])
 plt.show()
 
 #learner.plot_chain_trace('F', numBurnIn=num_burn, trueModel=model)

### 2015-09-17 run_mocap_mcmc.py

@@ -10,6 +10,7 @@ from learners_mcmc import (
     MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
     MCMCLearnerTransitionBasicModelWithMNIWPrior,
     MCMCLearnerTransitionDegenerateModelWithMNIWPrior)
+from learners_mcmc_augmented import CleverDegenerateMCMCLearner
 
 # Create a learner class for this run
 class MCMCBasicLearner(
@@ -35,12 +36,12 @@ np.random.seed(0)
 data_path = './mocap-data/'
 test_data_file = './results/mocap-test-data.p'
 
-model_type = 'basic'#'degenerate'#'naive'#'basic'#
-num_iter = 20000
+model_type = 'degenerate_augmented'#'degenerate'#'naive'#'basic'#
+num_iter = 2000
 
-num_burn = 10000#int(num_iter-10000)
+num_burn = 1000#int(num_iter-10000)
 num_hold = 0#min(100,int(num_burn/2))
-num_warm = 50
+num_warm = 10
 
 
 # Import marker data
@@ -114,6 +115,9 @@ hyperparams['b0'] = 0.001
 algoparams = dict()
 algoparams['rotate'] = 1E-4
 algoparams['perturb'] = 1E-8
+algoparams['pseudo_dof'] = 1000
+#algoparams['pseudo_sd'] = 0.0001
+#algoparams['pseudo_shape'] = 1000
 
 if model_type == 'naive':
 
@@ -187,10 +191,31 @@ elif model_type == 'degenerate':
             learner.plot_chain_adapt()
             learner.plot_chain_trace('R', dims=([0],[0]))
 
+elif model_type == 'degenerate_augmented':
+
+    learner = CleverDegenerateMCMCLearner(est_degenerate_model, markers,
+                                          hyperparams, algoparams,
+                                          verbose=True)
+
+    for ii in range(num_warm):
+        print("Warm up iteration {} of {}.".format(ii+1,num_warm))
+        learner.sample_transition_warm_up()
+
+    for ii in range(num_iter):
+        print("")
+        print("Running iteration {} of {}.".format(ii+1,num_iter))
+
+        rchange = np.random.random_integers(-1,1)
+        rank = max(min(learner.model.parameters['rank'][0] + rchange, ds),1)
+        learner.sample_transition(rank)
+
+        learner.save_link()
+        print("Current rank: {}".format(learner.model.parameters['rank'][0]))
+
[…]

### 2015-09-17 run_mocap_smc.py

@@ -117,14 +117,20 @@ smclearner = DegenerateSMCLearner(learner.chain_model[num_burn::thin],
                                   num_rejuv,
                                   verbose=True)
 
-mlhoodratios = np.zeros(ds)
-effsampsizes = np.zeros(ds)
 for rr in reversed(range(1,ds)):
     print("")
     print("Running SMC sampler for rank {}.".format(rr))
     smclearner.smc_reduce_rank(rr)
+
+mlhoodratios = np.zeros(ds)
+effsampsizes = np.zeros(ds)
+rmse = np.zeros(ds)
+for rr in reversed(range(1,ds)):
     mlhoodratios[rr] = normalising_constant_estimate(smclearner.approx[rr].weight)
     effsampsizes[rr] = effective_sample_size(smclearner.approx[rr].weight)
+    if rr > 2:
+        state_mn, state_sd = smclearner.estimate_state_trajectory(rr)
+        rmse[rr] = np.sqrt(np.sum((markers-state_mn)**2))
 
 fig = plt.figure()
 ax = fig.add_subplot(1,1,1)
@@ -134,6 +140,11 @@ plt.show()
 fig = plt.figure()
 ax = fig.add_subplot(1,1,1)
 ax.plot(range(ds), effsampsizes)
+plt.show()
+
+fig = plt.figure()
+ax = fig.add_subplot(1,1,1)
+ax.plot(range(1,ds), rmse[1:])
 plt.show()
 
 filename = "./results/mocap-smc-degenerate.p"

### 2015-09-17 sampling.py

@@ -84,7 +84,7 @@ def singular_inverse_wishart_density(val, vec, P):
 
     pptn = -0.5*(3*d-r+1)*np.sum(np.log(val)) \
            -0.5*np.trace( np.dot(np.dot(vec.T,np.dot(P,vec)),np.diag(1/val)) )
-    
+
     pdf = norm + pptn
     return pdf
 
@@ -222,6 +222,25 @@ def hyperparam_update_degenerate_mniw_transition(suffStats, U, nu0, Psi0,
     Psi = UPsi0U + np.dot(U.T, np.dot(UPsiU, U)) - np.dot(M, la.solve(V,M.T))
 
     return nu, Psi, M, V
+
+def hyperparam_update_degenerate_iw_transition_covariance(suffStats, U, F,
+                                                          nu0, Psi0):
+    """
+    Update inverse-wishart hyperparameters for within-subspace
+    comonents of the transition covariance conditional on observed state
+    trajectory.
+    """
+
+    # Posterior hyperparameters
+    nu = nu0 + suffStats[0]
+    UPsi0U = la.inv(np.dot(U.T,la.solve(Psi0,U)))
+    UPsiU = np.dot(F,np.dot(suffStats[1],F.T)) \
+            - np.dot(F, suffStats[2].T) \
+            - np.dot(suffStats[2], F.T) \
+            + suffStats[3]
+    Psi = UPsi0U + np.dot(U.T, np.dot(UPsiU, U))
+
+    return nu, Psi
 
 def project_degenerate_transition_matrix(Fold, FU, U):
     """

### 2015-09-18 notes/paper-degenerate-linear-systems.tex
Added paper draft

\documentclass[journal,10pt]{IEEEtran}
% \documentclass[11pt,draftcls,onecolumn]{IEEEtran}
% \documentclass[a4paper,10pt]{article}

\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{IEEEtrantools}
\usepackage{algorithm}
\usepackage{algorithmic}
% \usepackage{graphicx}
\usepackage{hyperref}
\usepackage{subfig}
\usepackage{cite}

\usepackage{tikz}
\usepackage{pgfplots}
\usepgfplotslibrary{groupplots}
 \usetikzlibrary{plotmarks}
 \pgfplotsset{compat=newest}
 \pgfplotsset{plot coordinates/math parser=false}
 \usepgfplotslibrary{external}
 \tikzexternalize[prefix=tikz/]

\input{linear-system-learning-macros.tex}
\newcommand{\meta}[1]{{\color{red}\em #1}}

%opening
\title{Bayesian Learning of Degenerate Linear~Gaussian~State~Space~Models using~Markov~Chain~Monte~Carlo}
\author{Pete Bunch$^*$, James Murphy and Simon Godsill
  \thanks{Department of Engineering, University of Cambridge, Cambridge, UK.}
  \thanks{Email: $\{$pb404,jm362,sjg30$\}$@cam.ac.uk}%
}

\begin{document}

\maketitle

\begin{abstract}
Linear Gaussian state space models are ubiquitous in signal processing, and an important procedure is that of estimating system parameters from observed data. Rather than making a single point estimate, it is often desirable to conduct Bayesian learning, in which the entire posterior distribution of the unknown parameters is sought. This can be achieved using Markov chain Monte Carlo. On some occasions it is possible deduce the form of the unknown system matrices in terms of a small number of scalar parameters, by considering the underlying physical processes involved. Here we study the case where this is not possible, and the entire matrices must be treated as unknowns. An efficient Gibbs sampling algorithm exists for the basic formulation of linear model. We extend this to the more challenging situation where the transition model is degenerate. Appropriate Markov kernels are devised and demonstrated with simulations.
\end{abstract}

\section{Introduction}

State space models are frequently used to describe time-varying systems, and Linear Gaussian models in particular are ubiquitous throughout signal processing. The physical phenomena modelled within this framework include the kinematics of moving targets \cite{Bar-Shalom2002}, audio signals \cite{Godsill1998}, genetic networks \cite{Beal2005}, and many others. The popularity of linear Gaussian state space models stems in part from their analytic tractability. Inference tasks can be performed in closed form using algorithms such as the Kalman filter \cite{Kalman1960} and Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965}.

A linear Gaussian state space model is specified by a number of system matrices which govern the latent state and observation processes. An important consideration is how to learn these fixed matrices from observed data. The parametric approach to this problem is to deduce the form of the system matrices in terms of a small number of scalar parameters by considering the physical mechanisms underlying the system, and then attempt to learn these. (See e.g. \cite{Kantas2009,Andrieu2010}.) However, in many systems there may not be an obvious parametric form, and it may be better to treat each system matrix in its entirety as an unknown variable to be learned. This type of non-parametric learning for linear Gaussian state space models is the focus of this paper. There is a large body of research on this topic, which has generally concentrated on point estimate methods. These include subspace identification approaches \cite{VanOverschee1991,Viberg1995}, and maximum likelihood estimates using either gradient-based algorithms (see e.g. \cite{Cappe2005,Sarkka2013}) or Expectation-Maximisation (EM) \cite{Shumway1982,Digalakis1993,Ghahramani1996,Gibson2005,Li2009}.

More recently, Bayesian approaches have been developed for non-parametric learning of linear Gaussian state space models, in which full posterior distributions for the system parameters are calculated. Since this is not possible analytically (see discussion in \cite{Beal2003}), approximations must be employed. One possibility is to take a variational approach, approximating the full joint posterior distribution as a product of independent marginals \cite{Ghahramani2001,Beal2003,Barber2007}. Alternatively, the true joint posterior can be targeted using Markov chain Monte Carlo (MCMC). This has the advantage of providing consistent estimates not only of the system matrices but also of any function of these quantities (e.g. phase margins, system poles). The price we pay is in computation; it is necessary to allow the Markov chains to run until convergence in order to form good estimates. Generic Metropolis-Hastings strategies such as those adopted in \cite{Ninness2010} are liable to mix very slowly and exact too great a computational demand. However, if conjugate priors are assumed for the various system matrices, then a Gibbs sampler may be implemented which attains substantially improved efficiency and requires almost no algorithm tuning \cite{Wills2012}.

The Gibbs sampler approach developed in \cite{Wills2012} is effective on the basic flavour of linear Gaussian state space models. However, it is not able to handle the case in which the transition model is degenerate, i.e. when the transition covariance matrix is not full rank. In this paper we extend the MCMC learning framework to encompass this scenario. This is achieved by appropriate factorisations of the covariance matrix which allow us still to exploit the fast-mixing Gibbs kernel. Furthermore, an extended distribution construction is introduced which enables the sampler to learn the rank of the transition covariance matrix.

%and parameterising the attainable subspace using a set of Givens rotations \cite{Anderson1987,Shalit2014,Yang1994,Cao2011,Cron2014}.

The paper is structured as follows. In section~\ref{sec:linear_gaussian_models} we review the Gibbs sampler for basic linear Gaussian state space systems, focusing on the transition model. In section~\ref{sec:degenerate_transition_models} we describe modifications for degenerate transitions. Simulations on a toy model, and a motion capture application are presented in section~\ref{sec:applications}.

% Citations which might be useful
% Linear systems with stability constraints: \cite{Boots2008}
% Givens rotations: \cite{Yang1994,Cao2011,Cron2014} (for covariance) \cite{Anderson1987,Shalit2014} (for generic orthogonal matrixes)

\section{Basic Linear Gaussian Models} \label{sec:linear_gaussian_models}
State space models are used to represent time-varying systems, and consist of a latent state process $\{\ls{\ti}\}_{\ti=1:\timax}$ and a related observation process $\{\ob{\ti}\}_{\ti=1:\timax}$. The challenge is to infer the sequence of unknown latent states, and learn parameters of the model, from the sequence of known observations. It is usually assumed that the state process is Markovian (i.e. $\ls{\ti}|\ls{\ti-1}$ is independent of $\ls{1:\ti-2}$) and that each observation depends only on the current state (i.e. $\ob{\ti}|\ls{\ti}$ is independent of $\ls{1:\ti-1}$ and $\ls{\ti+1:\timax}$).

A basic linear state space model therefore obeys the following recursive system equations,
%
\begin{IEEEeqnarray}{rCl}
 \ls{\ti} & = & \lgtm \ls{\ti-1} + \tn{\ti} \\
 \ob{\ti} & = & \lgom \ls{\ti}   + \on{\ti}       ,
\end{IEEEeqnarray}
%
where $\ls{\ti} \in \reals^{\lsd}$ and $\ob{\ti} \in \reals^{\obd}$. In addition, we assume the state disturbance and observation noise variables have a Gaussian distribution,
%
\begin{IEEEeqnarray}{rCl}
 \tn{\ti} & \sim & \normalden{\tn{\ti}}{0}{\lgtv} \\
 \on{\ti} & \sim & \normalden{\on{\ti}}{0}{\lgov}     ,
\end{IEEEeqnarray}
%
where $\lgtv$ and $\lgov$ are positive definite covariance matrices. This model can be written equivalently in terms of transition and observation densities,
%
\begin{IEEEeqnarray}{rCl}
 \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) & = & \normalden{\ls{\ti}}{\lgtm\ls{\ti-1}}{\lgtv} \\
 \den(\ob{\ti}|\ls{\ti},\lgom,\lgov)   & = & \normalden{\ob{\ti}}{\lgom\ls{\ti}}{\lgov}      .
\end{IEEEeqnarray}
%
The initial state $\ls{1}$ may be known or may be assigned a Gaussian prior.

The system is parameterised by the four matrices $\left\{\lgtm, \lgom, \lgtv, \lgov\right\}$. In this paper we focus on learning the transition model, i.e. $\lgtm$ and $\lgtv$, and treat $\lgom$ and $\lgov$ as fixed and known throughout. Conditional on these matrices, state inference may be carried out analytically. Posterior filtering and smoothing densities may be calculated using the Kalman filter \cite{Kalman1960} and Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965}. Furthermore, the Kalman filter may also be used to evaluate the marginal likelihood $\den(\ob{1:\timax}|\lgtm, \lgtv)$, and it is possible to draw posterior samples from the state posterior $\den(\ls{1:\timax}|\ob{1:\timax}, \lgtm, \lgtv)$ using the forward-filtering-backward-sampling method \cite{Chib1996}. The purpose of a learning algorithm is estimate unknown values of $\left\{\lgtm, \lgtv\right\}$ from a sequence of observations. Within the Bayesian framework this means calculating or approximating the posterior distribution $\den(\lgtm, \lgtv|\ob{1:\timax})$, which we can achieve using MCMC.

In order to conduct Bayesian learning, we first need to define prior distributions for the unknown system matrices $\lgtm$ and $\lgtv$. Although these should be selected to reflect prior belief about the system in question, there is substantial benefit in using conjugate priors, as these enable us to use Gibbs sampling moves. Typically, we will not have strong prior beliefs about the parameters, so it is reasonable to use an uninformative conjugate prior. However, if we do have prior knowledge to take into account which means that a conjugate prior is not appropriate, then we can treat each Gibbs move as a proposal in a Metropolis-Hastings scheme and use an accept/reject stage to account for the difference between the true prior and the conjugate prior used for the sampling \cite{Wills2012}.

\subsection{System Matrix Priors}

The conjugate prior for $\lgtm, \lgtv$ is a matrix normal--inverse Wishart distribution \cite{Wills2012},
%
\begin{align}
 \lgtv &\sim \iwishartdist{\priordof}{\priorscalematrix} \\
 \lgtm | \lgtv &\sim \matrixnormaldist{\priormeanmatrix}{\lgtv}{\priorcolumnvariance}     ,
\end{align}
%
with the following density function,
%
\begin{equation}
 \den(\lgtm, \lgtv) =  \den(\lgtm | \lgtv) \den(\lgtv)
\end{equation}
%
\begin{align}
\den(\lgtv) &= \frac{ \determ{\priorscalematrix}^{\frac{\priordof}{2}} }{ 2^{\frac{\priordof}{2}} \gammafun\left(\frac{\priordof}{2}\right) } \determ{\lgtv}^{-\frac{\priordof+\lsd+1}{2}} \exp\left( -\half \trace\left[\lgtv\inv\priorscalematrix\right] \right) \\
\den(\lgtm | \lgtv) &= \determ{2 \pi \lgtv}^{-\half} \determ{2 \pi \priorcolumnvariance}^{-\half} \exp\left(-\half \trace\left[ \lgtv\inv (\lgtm-\priormeanmatrix) \priorcolumnvariance\inv (\lgtm-\priormeanmatrix)\tr \right] \right) 
\end{align}
%
$\priordof\in\reals, \priordof>\lsd-1$. $\priorscalematrix$ and $\priorcolumnvariance$ are $\lsd\times\lsd$ positive definite matrices. $\priormeanmatrix \in \reals^{\lsd\times\lsd}$.

\subsection{MCMC for Basic Linear Gaussian Models}

Our principal interest is to learn the system matrices $\lgtm$ and $\lgtv$ from a sequence of observations $\ob{1:\timax}$. The appropriate posterior distribution is not amenable to efficient MCMC sampling, so instead we introduce the latent state sequence $\ls{1:\timax}$ as a nuisance variable and target the joint posterior,
%
\begin{IEEEeqnarray}{rCl}
 \postden(\lgtm, \lgtv, \ls{1:\timax}) & = & \den(\lgtm, \lgtv, \ls{1:\timax} | \ob{1:\timax}) \\
 & \propto & \den(\ob{1:\timax}|\ls{1:\timax}) \den(\ls{1:\timax}|\lgtm,\lgtv) \den(\lgtm|\lgtv) \den(\lgtv) \nonumber      .
\end{IEEEeqnarray}

An appropriate Gibbs sampler can be constructed by sampling alternately from the following conditional posterior distributions,
%
\begin{IEEEeqnarray}{l}
 \postden(\ls{1:\timax}|\lgtm, \lgtv) \nonumber \\
 \postden(\lgtm, \lgtv| \ls{1:\timax}) \nonumber      .
\end{IEEEeqnarray}
%
The distribution of the sampled values will then converge to the target posterior distribution \cite{Roberts1994}. The state conditional $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ can be sampled using the forward-filtering-backward-sampling algorithm \cite{Chib1996,Wills2012}, which works by first running a Kalman filter forwards through the data, then running backwards in time sampling each state conditional on those in the future. The parameter conditional may also be sampled directly. This depends on the state sequence probability,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm,\lgtv) \propto \prod_{\ti=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) } \\
 \quad\quad & \propto & \determ{\lgtv}^{-\frac{\timax-1}{2}} \exp\left( -\half \sum_{\ti=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \lgtv\inv (\ls{\ti}-\lgtm\ls{\ti-1}) \right)      .
\end{IEEEeqnarray}

It is straightforward to show that the required conditional distribution is also a matrix normal--inverse Wishart distribution \cite{Wills2012},
%
\begin{IEEEeqnarray}{rCl}
 \lgtv | \ls{1:\timax} &\sim& \iwishartdist{\postdof}{\postscalematrix} \\
 \lgtm | \lgtv, \ls{1:\timax} &\sim& \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}     ,
\end{IEEEeqnarray}
%
with the following updated hyperparameters,
%
\begin{align}
 \postcolumnvariance\inv                 &= \priorcolumnvariance\inv + \suffstats{1} \\
 \postmeanmatrix \postcolumnvariance\inv &= \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2}\\
 \postdof                                &= \priordof + \suffstats{0} \\
 \postscalematrix                        &= \priorscalematrix + \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr - \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr    ,
\end{align}
%
in which the following sufficient statistics are used,
%
\begin{align}
 \suffstats{0} &= \timax - 1 \\
 \suffstats{1} &= \sum_{t=2}^{\timax} \ls{\ti-1}\ls{\ti-1}\tr \\
 \suffstats{2} &= \sum_{t=2}^{\timax} \ls{\ti}\ls{\ti-1}\tr \\
 \suffstats{3} &= \sum_{t=2}^{\timax} \ls{\ti}\ls{\ti}\tr      .
\end{align}

Both matrix normal and inverse Wishart distributions may be sampled using standard methods. Hence we have all the components necessary to implement a Gibbs sampler for this model.

\section{Degenerate Transitions} \label{sec:degenerate_transition_models}

A commonly encountered variation on the basic linear Gaussian state space model is as follows,
%
\begin{IEEEeqnarray}{rCl}
 \ls{\ti} & = & \lgtm \ls{\ti-1} + \lgdm \tn{\ti} \label{eq:degenerate_transition} \\
 \tn{\ti} & \sim & \normalden{\tn{\ti}}{0}{\idmat}      .
\end{IEEEeqnarray}
%
This is almost equivalent, with $\lgtv=\lgdm\lgdm\tr$. However, whereas before $\lgtv$ was required to be positive definite, this is not necessarily the case for $\lgdm\lgdm\tr$. When this occurs, the conventional transition density $\den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgdm)$ is not defined, and the sampling operations underlying our Gibbs sampler are no longer valid. Degenerate models such as these arise in tracking applications (see \cite{Maskell2004} for a discussion of this phenomenon); when non-Markovian models, such as autoregressive process, are converted to state space form; and when it is natural to parameterise a system with more latent states than there are degrees of freedom (see section~\ref{sec:mocap}). In this section, we introduce a parameterisation for such degenerate transition models which can uniquely represent any such system, and show how MCMC moves may be implemented for learning.

Attempting to learn $\lgdm$ directly is problematic, as this matrix is not identifiable. The distribution of $\lgdm\tn{\ti}$ is identical to that of $\lgdm \Xi \tn{\ti}$ where $\Xi$ is an arbitrary orthogonal matrix. Therefore, there is not a unique value of $\lgdm$ which maximises the probability of the state sequence $\den(\ls{1:\timax}|\lgtm,\lgdm)$. Instead, as in the basic model, we parameterise the system with a $\lsd\times\lsd$ precision matrix $\lgtp$ which is positive semi-definite with rank $\tnd$. For a particular value of $\lgtp$, $\lgdm$ may be any matrix such that,
%
\begin{IEEEeqnarray}{rCl}
 \lgdm\lgdm\tr & = & \lgtp\pinv     ,
\end{IEEEeqnarray}
%
where $\pinv$ denote the pseudo-inverse.

Learning degenerate transition models is challenging because each transition, $(\ls{\ti}-\lgtm\ls{\ti-1})$, is constrained to lie in a particular $\tnd$-dimensional subspace within the state space; specifically, within the column space of $\lgdm$. This subspace is fixed conditional on any sampled state trajectory $\ls{1:\timax}$ (provided that $\timax>\tnd$). This makes standard Gibbs sampling from $\postden(\lgtp|\lgtm,\ls{1:\timax})$ and $\postden(\lgtm|\lgtp,\ls{1:\timax})$ insufficient, since there are values of $\lgtm$ and $\lgtp$ which can never be reached.

In order to allow the Markov chain to fully explore the parameter space, we introduce Markov kernels which target alternative conditional distributions, specifically $\postden(\lgtm, \ls{1:\timax}|\lgtp)$ and $\postden(\lgtp|\lgtm)$. Drawing from these distributions exactly is not possible, but we can use Metropolis-Hastings. Naively, we might simply try to propose changes to the elements of $\lgtm$ and $\lgtp$, but this is very inefficient. Better schemes can be formulated by exploiting the following factorisations.

For a rank-$\tnd$ positive semi-definite precision matrix $\lgtp$,
%
\begin{IEEEeqnarray}{rCl}
 \lgtp & = & \tnmo \tnmd \tnmo\tr \nonumber      ,
\end{IEEEeqnarray}
%
where $\tnmd = \diag([\tnmev{1},\tnmev{2},\dots])$ is the diagonal matrix of $\tnd$ \emph{positive} eigenvalues (such that $\tnmev{1} > \tnmev{2} > \dots > \tnmev{\tnd}$), and $\tnmo$ is a $\lsd\times\tnd$ matrix of orthonormal eigenvectors. This is a standard eigendecomposition. As in \cite{Muirhead1982}, the first non-zero element in each column of $\tnmo$ is required to be positive for each $i$. This resolves the sign ambiguity in the eigenvectors and thus ensures that the transformation from $\lgtp$ to $\{\tnmo,\tnmd\}$ is bijective.

An alternative decomposition will also be useful. For a rank-$\tnd$ positive semi-definite precision matrix $\lgtp$,
%
\begin{IEEEeqnarray}{rCl}
 \lgtp & = & \tnmocrossred \tnmf \tnmocrossred\tr \label{eq:degenerate_covariance_factorisation}      ,
\end{IEEEeqnarray}
%
where,
%
$\tnmf$ is a $\tnd\times\tnd$ positive definite matrix and $\tnmocrossred$ is a $\lsd\times\tnd$ matrix of orthonormal vectors subject to certain conditions. The transformation from $\lgtp$ to $\{\tnmocrossred,\tnmf\}$ is also bijective. See appendix~\ref{app:givens-factorisation} for details.

\subsection{Degenerate Transition Priors}

In the basic linear model, when $\lgtp$ was positive definite, we used a Wishart prior for this matrix. The singular Wishart distribution is an extension over the space of positive semi-definite matrices \cite{Uhlig1994}, and has the following density with respect to the Lebesgue measure on the distinct matrix elements, conditional on a known value of $\tnd$,
%
\begin{IEEEeqnarray}{rCl}
 \den(\lgtp|\tnd) & = & \wishartden{\lgtp}{\tnd}{\Ppscale} \nonumber \\
 & \propto & \determ{\tnmd}^{\frac{\tnd-\lsd-1}{2}} \exp\left( -\half \trace\left[ \Ppscale\inv \lgtp \right] \right)     .
\end{IEEEeqnarray}
%
The prior is completed by specifying a distribution on the integer $\tnd \in [1,\lsd]$.

In order to exploit the two transformations of $\lgtp$ introduced in the previous section, we will need the prior distributions induced on $\{\tnmo,\tnmd\}$ and $\{\tnmocrossred,\tnmf\}$ by the singular Wishart distribution on $\lgtp$.

For the first, if we constrain the scale matrix to be proportional to the identity matrix, $\Ppscale = \Ppscalescalar \idmat$, then it can be shown (see \cite{Uhlig1994,Srivastava2003} and \cite[chapter 3]{Muirhead1982}) that the eigenvectors and eigenvalues of $\lgtp$ are independent. Furthermore, the distribution for the eigenvector matrix is uniform (i.e. a normalised Haar measure, conditional on the first element being positive), and for the eigenvalues,
%
\begin{IEEEeqnarray}{rCl}
 \den(\tnmd|\tnd) & = & \frac{ \pi^{\half\tnd^2} }{ (2\Ppscalescalar)^{\half\tnd\lsd} \gammafun[\tnd](\frac{\tnd}{2}) \gammafun[\tnd](\frac{\lsd}{2}) } \left[ \prod_{i<j}^{\tnd} (\tnmev{i}-\tnmev{j}) \right] \nonumber \\
 & & \qquad \times \left[ \prod_{i=1}^{\tnd} \tnmev{i}^{\half(\lsd-\tnd-1)} \exp\left(-\frac{1}{2\Ppscalescalar} \tnmev{i}\right) \right]      . \nonumber \\ \label{eq:eigenvalue_prior}
\end{IEEEeqnarray}

For the second factorisation of $\lgtp$, since $\tnmf = \tnmocrossred\tr \lgtp \tnmocrossred$, we find that,
%
\begin{IEEEeqnarray}{rCl}
 \den(\tnmf|\tnmocrossred, \tnd) & = & \wishartden{\tnmf}{\tnd}{\Ppscalescalar \idmat}      .
\end{IEEEeqnarray}
%
(See \cite[theorem 3.2.5]{Muirhead1982}.) It is possible to derive a corresponding prior distribution for a parameterisation of $\tnmocrossred$ using a set of Givens rotations, but this will not be needed.

\subsection{MCMC for Degenerate Transitions}

As discussed above, standard Gibbs sampling from $\postden(\lgtp|\lgtm,\ls{1:\timax})$ and $\postden(\lgtm|\lgtp,\ls{1:\timax})$ does not allow the Markov chain to fully explore the parameter space because of the constraint that each transition $(\ls{\ti}-\lgtm\ls{\ti-1})$ must lie in the $\tnd$-dimensional column space of $\lgdm$. Nevertheless, we can sample from these distributions exactly, and doing so allows the chain to efficiently explore values of the transition matrix and transition covariance within the permitted regions.

In addition to the within-subspace Gibbs moves, we introduce three efficient Metropolis-Hastings moves which target $\postden(\lgtp|\lgtm)$ and $\postden(\lgtm, \ls{1:\timax}|\lgtp)$, thus allowing the transition subspace to be changed and the parameter space to be fully explored by the Markov chain.

\subsubsection{Gibbs Sampling within the Subspace}

It is here that we exploit the transformation from $\lgtp$ to $\{\tnmocrossred,\tnmf\}$ introduced in \eqref{eq:degenerate_covariance_factorisation}. In the degenerate case, the transition density $\den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtp)$ with respect to the Lebesgue measure does not exist. However, since the set of possible values for $\ls{\ti}$ constitutes a hyperplane, we can project this onto a lower dimensional space in which a density does exist. From \eqref{eq:degenerate_transition} and \eqref{eq:degenerate_covariance_factorisation}, 
%
\begin{IEEEeqnarray}{rCl}
 \tnmocrossred\tr(\ls{\ti}-\lgtm\ls{\ti-1}) & \sim & \normaldist{0}{\tnmf\inv}        .
\end{IEEEeqnarray}
%
The likelihood of the state trajectory can then be written as,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm,\tnmf,\tnmocross) } \nonumber \\
 & \propto & \begin{cases}
                                  \determ{\tnmf}^{\frac{\timax-1}{2}} \exp\left( -\half \sum\limits_{\ti=1}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr\tnmocrossred \tnmf \tnmocrossred^T(\ls{\ti}-\lgtm\ls{\ti-1}) \right) \\ \hphantom{0} \qquad \text{if }\forall \,\, \ti, (\ls{\ti}-\lgtm\ls{\ti-1}) \text{ lies in the column space of } \tnmocrossred \\
                                  0 \qquad \text{otherwise}     .
                                                        \end{cases} \nonumber 
\end{IEEEeqnarray}

We now re-formulate the posterior conditional for $\lgtp$ in terms of $\tnmocrossred$ and $\tnmf$,
%
\begin{IEEEeqnarray}{rCl}
 \postden(\tnmocrossred, \tnmf | \lgtm,\ls{1:\timax}) & = & \postden(\tnmf | \tnmocrossred, \lgtm,\ls{1:\timax}) \postden(\tnmocrossred | \lgtm,\ls{1:\timax})    .
\end{IEEEeqnarray}
%
Because of the subspace constraint, $\tnmocrossred$ is uniquely defined conditional on $\ls{1:\timax}$, and thus sampling from $\postden(\tnmocrossred | \lgtm,\ls{1:\timax})$ simply implies keeping the existing value of $\tnmocrossred$. This can be proved using a ``degrees of freedom'' argument. The set of points $\{(\ls{\ti}-\lgtm\ls{\ti-1})\}_{t=1}^{\timax}$ lie in a $\tnd$-dimensional subspace, characterised by $\tnd(\lsd-\tnd)$ independent parameters. This is the same number as the parameters used to describe $\tnmocrossred$. (See appendix~\ref{app:givens-factorisation}.)

For $\tnmf$ the appropriate conditional is,
%
\begin{IEEEeqnarray}{rCl}
 \postden(\tnmf|\lgtm, \tnmocrossred, \tnd, \ls{1:\timax}) &\propto& \den(\ls{1:\timax}|\lgtm,\tnmf,\tnmocrossred) \den(\tnmf) \nonumber \\
% \qquad & \propto & \determ{\tnmf}^{\frac{\timax-2}{2}} \exp\left( -\half \trace\left[ \tnmf \left( \frac{1}{\Ppscalescalar}\idmat + \sum_{\ti=2}^{\timax} \tnmocrossred\tr(\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr\tnmocrossred \right) \right] \right)  \nonumber \\
 & = & \wishartden{\tnmf}{\Ppdof''}{\Ppscale''}      ,
\end{IEEEeqnarray}
%
where,
%
\begin{IEEEeqnarray}{rCl}
 \Ppdof''   & = & \tnd + \timax - 1 \nonumber \\
 \Ppscale'' & = & \left[\frac{1}{\Ppscalescalar}\idmat + \sum_{\ti=2}^{\timax} \tnmocrossred\tr(\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr\tnmocrossred \right]\inv \label{eq:D_posterior_params}      ,
\end{IEEEeqnarray}
%
which closely resembles that for the precision matrix in the full rank case.

For $\lgtm$, if the current value is $\lgtm^*$, then the values which meet the constraint after sampling from $\postden(\lgtm|\tnmf,\tnmocrossred, \tnd, \ls{1:\timax})$ can be written as,
%
\begin{IEEEeqnarray}{rCl}
 \lgtm & = & \lgtm^* + \tnmocrossred \lgtmchange
\end{IEEEeqnarray}
%
where $\lgtmchange$ is an arbitrary $\tnd\times\lsd$ matrix. We can then re-frame the sampling operation to draw $\lgtmchange$ from an appropriate conditional. The state sequence probability becomes,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm^*,\lgtmchange,\tnmf,\tnmocross) } \nonumber \\
 & \propto & \exp\left( -\half \left[ \sum_{\ti=1}^{\timax} \ls{\ti-1}\tr \lgtmchange\tr \tnmf \lgtmchange \ls{\ti-1} \right.\right. \nonumber \\
 & & \qquad\qquad \left.\left.\vphantom{\sum_{\ti=1}^{\timax}}  -\:2 (\ls{\ti}-\lgtm^*\ls{\ti-1})\tr \tnmocrossred \tnmf \lgtmchange \ls{\ti-1} \right] \right)     . 
\end{IEEEeqnarray}
%
Since $\lgtmchange = \tnmocrossred\tr(\lgtm-\lgtm^*)$, the prior for $\lgtmchange$ is simply a Gaussian, with mean and variance transformed from those of $\lgtm$. The prior and likelihood may be combined using the vectorisation method in a manner analogous to \eqref{eq:vectorisation_trick}, leading to a Gaussian distribution for $\lgtmchange$ which is closely related to that for $\lgtm$ in the full rank case.

We will write $\mcmckern{\lgtp,\lgtm}(\cdot|\lgtm,\lgtp,\ls{1:\timax})$ to represent the Markov kernel which corresponds to transforming from $\lgtp$ to $\{\tnmocrossred,\tnmf\}$, sampling $\lgtm$ and $\tnmf$ from their posterior conditionals, and then transforming back.

\subsubsection{A Markov Kernel for the Precision Matrix}

In order to allow the Markov chain to fully explore the parameter space, we use Metropolis-Hastings moves to target the marginal distribution $\postden(\lgtp|\lgtm)$. The resulting kernel constitutes a \emph{collapsed Gibbs} move \cite{Dyk2008}. For $\lgtp$, we can construct an efficient Metropolis-Hastings kernel as follows. We sample a rotation matrix $\mhrot$ from some proposal distribution $\mhrotppsl$ and then apply the following transformation,
%
\begin{IEEEeqnarray}{rCl}
 \begin{bmatrix}
  \lgtp\mcnew \\ \mhrot\mcnew
 \end{bmatrix}
 & = &
 \begin{bmatrix}
  \mhrot\lgtp\mhrot\tr \\ \mhrot\tr
 \end{bmatrix} \nonumber     ,
\end{IEEEeqnarray}
%
which is clearly its own inverse. Since $\determ{\mhrot}=1$, it is straightforward to show that the Jacobian of this transformation is $1$, and hence using the reversible jump interpretation of Metropolis-Hastings \cite{Green1995,Green2009}, the acceptance probability is,
%
\begin{IEEEeqnarray}{rCl}
 \mhap(\lgtp\to\lgtp\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtp\mcnew|\lgtm)\mhrotppsl(\mhrot\mcnew) }{ \postden(\lgtp|\lgtm)\mhrotppsl(\mhrot) } \right\}  \\
 & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtp\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtp)} \times \frac{\den(\lgtp\mcnew) \mhrotppsl(\mhrot\mcnew)}{\den(\lgtp) \mhrotppsl(\mhrot)} \right\} \nonumber     .
\end{IEEEeqnarray}
%
The first term is simply a ratio of Kalman filter likelihoods. We will write $\mcmckern{\lgtp,1}(\cdot|\lgtm,\lgtp)$ to represent this Markov kernel.

There are numerous ways to sample the rotation matrix $\mhrot$ from a suitable proposal distribution $\mhrotppsl$. For example, we could use the Cayley transform \cite{Leon2006}, a bijective mapping from the skew-symmetric matrices to the rotation matrices, defined by,
%
\begin{IEEEeqnarray}{rCl}
 P(S) & = & (\idmat - S)\inv(I+S)     .
\end{IEEEeqnarray}
%
To sample from $\mhrotppsl$, we draw $\half\lsd(\lsd-1)$ independent scalar random variables $\{s_{i,j}\}_{0<i<j<\lsd}$ from any zero-mean distribution; a nice choice is,
%
\begin{IEEEeqnarray}{rCl}
 s_{k,l} & \sim & \normaldist{0}{\sigma_s^2} \label{eq:skewsymmetric_proposal}     .
\end{IEEEeqnarray}
%
Use these to construct a skew-symmetric matrix $S$,
%
\begin{IEEEeqnarray}{rCl}
 S_{k,l} & = & \begin{cases}
                s_{k,l}  & k<l \\
                -s_{l,k} & k>l \\
                0        & k=l     ,
               \end{cases}
\end{IEEEeqnarray}
%
and then set $\mhrot=P(S)$. The Cayley transform has the property that $P(-S)=P(S\tr)=P(S)\inv=P(S)\tr$, which implies that $\mhrotppsl(\mhrot)=\mhrotppsl(\mhrot\tr)=\mhrotppsl(\mhrot\mcnew)$, leading to a cancellation in the acceptance probability.

There is an alternative using Givens rotations. First sample $i \in [1,\lsd]$, $j \in [1,\lsd]\setminus i$, and $\givrot{} \in [-\pi/2,\pi/2]$ from some zero-mean distribution. Form the Givens matrix $\givmat{i}{j}{\givrot{}}$ such that,
%
\begin{IEEEeqnarray}{rCl}
 \left[\givmat{i}{j}{\givrot{}} - \idmat\right]_{k,l} & = & \begin{cases}
                                                    \cos(\givrot{})-1 & k=l=i \text{ or } k=l=j \\
                                                    \sin(\givrot{}) & k=i,l=j \\
                                                    -\sin(\givrot{}) & k=j,l=i \\
                                                    0 & \text{ otherwise,}
                                                 \end{cases}
\end{IEEEeqnarray}
%
and use $\mhrot=\givmat{i}{j}{\givrot{}}$. This also has the property that $\givmat{i}{j}{-\givrot{}} = \givmat{i}{j}{\givrot{}}\tr$, meaning that we achieve the same cancellation of the proposals as before.

\subsubsection{A Markov Kernel for the Precision Rank}

It is unlikely that the rank of $\lgtp$ will be known a priori. We can learn this within the MCMC scheme by using the transformation of $\lgtp$ to $\{\tnmo,\tnmd\}$ and allowing the sampler to add and remove eigenvalue-eigenvector pairs using reversible jump moves \cite{Green1995,Green2009}. This can be achieved with a matching pair of moves.

\begin{algorithm}
\begin{algorithmic}
 \REQUIRE{$\tnd'$, $\tnmd'$, $\tnmo'$}
 \STATE Set $\tnd = \tnd' - 1$.
 \STATE Remove largest eigenvalue $\tnmev{1}$. Form $\tnmd$ from the remaining eigenvalues.
 \STATE Remove corresponding eigenvector $\tnmevec{1}$. Form $\tnmo$ from the remaining eigenvectors.
 \RETURN{$\tnd$, $\tnmd$, $\tnmo$}
\end{algorithmic}
\caption{Reversible Jump Move: Decrease rank}
\label{alg:rjmcmc-down}
\end{algorithm}

\begin{algorithm}
\begin{algorithmic}
 \REQUIRE{$\tnd$, $\tnmd$, $\tnmo$}
 \STATE Set $\tnd' = \tnd + 1$.
 \STATE Sample an eigenvalue $\tnmev{}^* \in (\tnmev{1},\infty)$ from $\ppslden{\tnmev{}}(\tnmev{})$. Form $\tnmd'$ from the new set of eigenvalues.
 \STATE Sample a corresponding eigenvector $\tnmevec{}^*$ from $\ppslden{\tnmevec{}}(\tnmevec{})$. Form $\tnmo'$ from the new set of eigenvectors.
 \RETURN{$\tnd'$, $\tnmd'$, $\tnmo'$}
\end{algorithmic}
\caption{Reversible Jump Move: Increase rank}
\label{alg:rjmcmc-up}
\end{algorithm}

The acceptance probability for the increase move is then,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \mhap\left( \tnd,\tnmd,\tnmo \to \tnd',\tnmd',\tnmo' \right) } \nonumber \\
 \qquad & = & \min\left\{1, \frac{ \postden(\lgtm, \tnd',\tnmd',\tnmo') }{ \postden(\lgtm, \tnd,\tnmd,\tnmo) } \times \frac{1}{\ppslden{\tnmevec{}}(\tnmevec{}^*)\ppslden{\tnmev{}}(\tnmev{}^*)} \right\} \nonumber \\
 & = & \min\Bigg\{1, \frac{\den(\ob{1:\timax} | \lgtm, \tnd',\tnmd',\tnmo') }{ \den(\ob{1:\timax} | \lgtm, \tnd,\tnmd,\tnmo) } \nonumber \\
 & & \quad \times \frac{\den(\tnmevec{1},\dots,\tnmevec{\tnd},\tnmevec{}^*)\den(\tnmev{1},\dots,\tnmev{\tnd},\tnmev{}^*|\tnd')\den(\tnd')}{\den(\tnmevec{1},\dots,\tnmevec{\tnd})\den(\tnmev{1},\dots,\tnmev{\tnd}|\tnd)\den(\tnd) \ppslden{\tnmevec{}}(\tnmevec{}^*)\ppslden{\tnmev{}}(\tnmev{}^*)} \Bigg\} \nonumber      ,
\end{IEEEeqnarray}
%
where $\tnmevec{1},\dots,\tnmevec{\tnd}$ are the eigenvectors comprising $\tnmo$ and $\tnmev{1},\dots,\tnmev{\tnd}$ the eigenvalues comprising $\tnmd$. The first term is simply a ratio of Kalman filter likelihoods. For the decrease move, the ratio is replaced by its reciprocal.

The conditional prior distribution $\den(\tnmevec{}^*|\tnmevec{1},\dots,\tnmevec{\tnd})$ is the normalised Haar measure on the manifold $\tnmo\tr\tnmevec{}^*=0$, conditional on the constraint we imposed to resolve the sign ambiguity \cite{Muirhead1982}. We can simulate according to this prior by sampling a standard Gaussian random vector and then using Gram-Schmidt orthogonalisation. This implies setting $\ppslden{\tnmevec{}}(\tnmevec{}^*)=\den(\tnmevec{}^*|\tnmevec{1},\dots,\tnmevec{\tnd})$, which results in a cancellation in the acceptance probability.

For the eigenvalues, inspection of the joint density \eqref{eq:eigenvalue_prior} suggests the following truncated gamma proposal,
%
\begin{IEEEeqnarray}{rCl}
 \ppslden{\tnmev{}}(\tnmev{}^*) & \propto & \begin{cases}
                                               \gammaden{\tnmev{}^*}{\frac{\lsd-\tnd}{2}}{2\Ppscalescalar} & \tnmev{}^*>\tnmev{1} \\
                                               0 & \text{otherwise}
                                              \end{cases} \nonumber \\
 & = & \begin{cases} 
 \frac{ (\tnmev{}^*)^{\frac{\lsd-\tnd}{2}-1} \exp\left( -\frac{\tnmev{}^*}{2\Ppscalescalar} \right) }{(2\Ppscalescalar)^{\frac{\lsd-\tnd}{2}} \compnormincgammafun\left(\frac{\lsd-\tnd}{2},\frac{\tnmev{1}}{2\Ppscalescalar}\right) }  & \tnmev{}^*>\tnmev{1} \\
 0 & \text{otherwise}     ,
 \end{cases}
\end{IEEEeqnarray}
%
where $\compnormincgammafun(\cdot,\cdot)$ is the complement of the normalised incomplete gamma function.
%
This leads to the following cancellation,
%
\begin{IEEEeqnarray}{rCl}
 \frac{\den(\tnmev{1},\dots,\tnmev{\tnd},\tnmev{}^*|\tnd')}{\den(\tnmev{1},\dots,\tnmev{\tnd}|\tnd)\ppslden{\tnmev{}}(\tnmev{}^*)} & = & \frac{ \pi^{\half} \compnormincgammafun\left(\frac{\lsd-\tnd}{2},\frac{\tnmev{1}}{2\Ppscalescalar}\right) }{ (2\Ppscalescalar)^{\frac{\tnd}{2}} \gammafun\left(\frac{\tnd+1}{2}\right) } \prod_{i=1}^{\tnd} \tnmev{i}^{-\half}(\tnmev{}^*-\tnmev{i}) \nonumber \\
\end{IEEEeqnarray}
%
Thus simplifying the acceptance probability,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \mhap\left( \tnd,\tnmd,\tnmo \to \tnd',\tnmd',\tnmo' \right) } \nonumber \\
 \qquad & = & \min\Bigg\{1, \frac{\den(\ob{1:\timax} | \lgtm, \tnd',\tnmd',\tnmo') }{ \den(\ob{1:\timax} | \lgtm, \tnd,\tnmd,\tnmo) } \times \frac{\den(\tnd')}{\den(\tnd)} \nonumber \\
 & & \qquad\qquad \times \frac{ \pi^{\half} \compnormincgammafun\left(\frac{\lsd-\tnd}{2},\frac{\tnmev{1}}{2\Ppscalescalar}\right) }{ (2\Ppscalescalar)^{\frac{\tnd}{2}} \gammafun\left(\frac{\tnd+1}{2}\right) } \prod_{i=1}^{\tnd} \tnmev{i}^{-\half}(\tnmev{}^*-\tnmev{i}) \Bigg\} \nonumber      ,
\end{IEEEeqnarray}
%
We will write $\mcmckern{\lgtp,2}(\cdot|\lgtm,\lgtp)$ for the Markov kernel which corresponds to transforming from $\lgtp$ to $\{\tnmo,\tnmd\}$, conducting a reversible jump move as described above, and then transforming back.

\subsubsection{A Markov Kernel for the Transition Matrix}

Next we turn to $\lgtm$. If we modify the covariance matrix slightly so that it is no longer singular, then we can use the existing sampling procedure from the basic linear model and treat the resulting value of $\lgtm$ as a proposal in a Metropolis-Hastings kernel targeting $\postden(\lgtm, \ls{1:\timax}|\lgtp)$.

Suppose we start with sampled values of $\lgtm$,$\lgtp$ and $\ls{1:\timax}$. A suitably modified precision matrix is,
%
\begin{IEEEeqnarray}{rCl}
 \paddedlgtp & = & \left(\lgtp\pinv + \Qpadding \idmat\right)\inv \label{eq:padded_transition_precision}      ,
\end{IEEEeqnarray}
%
where $\Qpadding$ is a small positive constant. We propose a new value of $\lgtm$ by sampling,
%
\begin{IEEEeqnarray}{rCl}
%  \lgtm' & \sim & \ppslden{\lgtm}(\lgtm|\lgtp,\ls{1:\timax}) & = & \postden(\lgtm|\paddedlgtp,\ls{1:\timax})     .
 \lgtm' & \sim & (\lgtm|\paddedlgtp,\ls{1:\timax})     .
\end{IEEEeqnarray}
%
This distribution is simply a Gaussian and we can evaluate the density using \eqref{eq:basic_F_conditional}. Finally, we also sample a new state trajectory using forward filtering backward sampling,
%
\begin{IEEEeqnarray}{rCl}
 \ls{1:\timax}' & \sim & \postden(\ls{1:\timax}|\lgtm', \lgtp)      .
\end{IEEEeqnarray}
%
The acceptance probability is simply,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \mhap\left(\lgtm,\ls{1:\timax}\to\lgtm',\ls{1:\timax}'\right) } \nonumber \\
 \quad\quad & = & \min\left\{1,\frac{ \postden(\lgtm',\ls{1:\timax}'|\lgtp) \postden(\ls{1:\timax}|\lgtm,\lgtp) \postden(\lgtm|\paddedlgtp,\ls{1:\timax}')   }{ \postden(\lgtm,\ls{1:\timax}|\lgtp) \postden(\ls{1:\timax}'|\lgtm',\lgtp) \postden(\lgtm'|\paddedlgtp,\ls{1:\timax}) }\right\} \nonumber \\
 & = & \min\left\{1,\frac{ \postden(\lgtm'|\lgtp) \postden(\lgtm|\paddedlgtp,\ls{1:\timax}')   }{ \postden(\lgtm|\lgtp)\postden(\lgtm'|\paddedlgtp,\ls{1:\timax}) }\right\} \nonumber \\
 & = & \min\left\{1, \frac{ \den(\ob{1:\timax}|\lgtm',\lgtp) }{ \den(\ob{1:\timax}|\lgtm,\lgtp)} \times \frac{ \den(\lgtm') }{ \den(\lgtm)} \times \frac{ \postden(\lgtm|\paddedlgtp,\ls{1:\timax}')   }{ \postden(\lgtm'|\paddedlgtp,\ls{1:\timax}) }\right\} \nonumber      , \\
\end{IEEEeqnarray}
%
in which the first term is a ratio of Kalman filter likelihoods. We will write $\mcmckern{\lgtm,\ls{}}(\cdot|\lgtm,\lgtp,\ls{1:\timax})$ for this Markov kernel.

\subsubsection{The Complete Sampler}

In this section we have described four Markov kernels which can be composed in order to learn a degenerate linear state space model. Each one by construction targets some conditional of the full posterior distribution and together they allow us to fully explore the parameter space. One iteration of the final algorithm consists of the following cycle. Starting with $\{\lgtm,\lgtp,\ls{1:\timax}\}$,
%
\begin{IEEEeqnarray}{rCl}
 \lgtm',\ls{1:\timax} & \sim & \mcmckern{\lgtm,\ls{}}(\cdot|\lgtm,\lgtp,\ls{1:\timax}) \nonumber \\
 \lgtp' & \sim & \mcmckern{\lgtp,1}(\cdot|\lgtm',\lgtp) \nonumber \\
 \ls{1:\timax}'' & \sim & \postden(\cdot|\lgtm',\lgtp') \nonumber \\
 \lgtp'' & \sim & \mcmckern{\lgtp,2}(\cdot|\lgtm',\lgtp') \nonumber \\
 \ls{1:\timax}''' & \sim & \postden(\ls{1:\timax}|\lgtm',\lgtp'') \nonumber \\
 \lgtm'',\lgtp''' & \sim & \mcmckern{\lgtp,\lgtm}(\cdot|\lgtm',\lgtp'',\ls{1:\timax}''') \nonumber     .
\end{IEEEeqnarray}

The only tuning required is to select the Metropolis-Hastings proposal distributions for $\mcmckern{\lgtp,1}$ and $\mcmckern{\lgtm,\ls{}}$. In both cases these can be reduced to a single scale parameter; $\Qpadding$ in \eqref{eq:padded_transition_precision} for $\mcmckern{\lgtm,\ls{}}$ and $\sigma_s$ in \eqref{eq:skewsymmetric_proposal} for $\mcmckern{\lgtp,1}$. An adaptive MCMC scheme, such as those discussed in \cite{Roberts2009}, can be used to modify these values as the chain runs for maximum efficiency.

\subsection{Example}

The algorithm was tested using the transition matrix specified in section~\ref{sec:example_basic}, but with the transition covariance changed to,
%
\begin{IEEEeqnarray}{rCl}
 \lgdm\lgdm\tr & = & \begin{bmatrix}
              0.26 &  0.33 &  0.26 & -0.42 \\
              0.33 &  1.14 &  0.46 & -0.85 \\ 
              0.26 &  0.46 &  0.29 & -0.48 \\
             -0.42 & -0.85 & -0.48 &  0.81
             \end{bmatrix} \nonumber      ,
\end{IEEEeqnarray}
%
(showing $2$ decimal places) which has a rank of $2$.

The MCMC algorithm was run using an adaptation scheme from \cite{Roberts2009}, adjusting the two proposal parameters based on the acceptance rate of the associated Metropolis-Hastings moves. Posterior histograms for $\lgtm$, $\lgtv$, and for the rank of $\lgtv$, are shown in figure~\ref{fig:degenerate_FQ_histograms} and~\ref{fig:degenerate_rank_histograms} for one MCMC run.

\begin{figure}
 \centering
 \subfloat[]{\includegraphics[width=0.7\columnwidth]{../code/degenerate_F_with_rank.pdf} \label{subf:degenerate_FQ_histograms:F}} \\
 \subfloat[]{\includegraphics[width=0.7\columnwidth]{../code/degenerate_Q_with_rank.pdf} \label{subf:degenerate_FQ_histograms:Q}}
 \caption{MCMC Posterior histograms for \protect\subref{subf:degenerate_FQ_histograms:F} $\lgtm$ and \protect\subref{subf:degenerate_FQ_histograms:Q} $\lgtv$. Degenerate transition model. Unknown rank.}
 \label{fig:degenerate_FQ_histograms}
\end{figure}

\begin{figure}
 \centering
 \includegraphics[width=0.7\columnwidth]{../code/degenerate_rank_with_rank.pdf} 
 \caption{MCMC Posterior histogram for $\tnd$. Degenerate transition model.}
 \label{fig:degenerate_rank_histograms}
\end{figure}

\section{Applications} \label{sec:applications}

\subsection{Example} \label{sec:example_basic}

We demonstrate the efficacy of the Gibbs sampler with a simple simulation example on a 4-dimensional model. The following transition matrix is used,
%
\begin{IEEEeqnarray}{rCl}
 \lgtm & = & \begin{bmatrix}
              0.9 & 0.7 & 0.7  & 0    \\
              0   & 0.9 & -0.5 & 0.1  \\
              0   & 0   & 1.6  & -0.8 \\
              0   & 0   & 1    & 0
             \end{bmatrix} \nonumber      .
\end{IEEEeqnarray}
%
The transfer function associated with $\lgtm$ has two real and two complex poles, all on or close to the unit circle. The transition covariance matrix is,
%
\begin{IEEEeqnarray}{rCl}
 \lgtv & = & \begin{bmatrix}
              1.0  & 0.5  & 0.5  & 0.5 \\
              0.5  & 1.0  & 0.5  & 0.5 \\ 
              0.5  & 0.5  & 1.0  & 0.5 \\
              0.5  & 0.5  & 0.5  & 1.0
             \end{bmatrix} \nonumber      .
\end{IEEEeqnarray}
%
For the observation model, $\lgom = \idmat$ and $\lgov = \idmat{}$.

A data set with $\timax=100$ time instants simulated from the model is shown in figure~\ref{fig:data_fullrank}. Running the Gibbs sampler results in rapid convergence. The resulting posterior histograms are shown in figure~\ref{fig:basic_FQ_histograms}, using $5000$ samples following a burn in of $1000$ iterations.

\begin{figure}
 \centering
% \includegraphics[width=0.9\columnwidth]{../code/data_fullrank.pdf}
 \caption{An example simulated data set. States (solid line) and observations (dots) of the four state components.}
 \label{fig:data_fullrank}
\end{figure}

\begin{figure}
 \centering
% \subfloat[]{\includegraphics[width=0.7\columnwidth]{../code/basic_F.pdf} \label{subf:basic_FQ_histograms:F}} \\
% \subfloat[]{\includegraphics[width=0.7\columnwidth]{../code/basic_Q.pdf} \label{subf:basic_FQ_histograms:Q}}
 \caption{MCMC Posterior histograms for \protect\subref{subf:basic_FQ_histograms:F} $\lgtm$ and \protect\subref{subf:basic_FQ_histograms:Q} $\lgtv$. Basic linear transition model. Vertical lines indicate true values}
 \label{fig:basic_FQ_histograms}
\end{figure}

\subsection{Rigid-Body Modelling with Motion Capture} \label{sec:mocap}

In this section, the MCMC algorithm for degenerate transition models is applied to a simplified motion capture task. The data used is from \cite{Aristidou2013}. A motion capture system was used to measure the $3$-dimensional coordinates of a number of markers attached to a subject's body. Here we consider the four markers attached to the head. Since these are all attached to the same rigid body, we expect there to be substantial correlation in their motion. The linear model used is slightly modified from that introduced before. The latent state $\ls{\ti}$ consists of two parts,
%
\begin{IEEEeqnarray}{rCl}
 \ls{\ti} & = & \begin{bmatrix}
                 r_{\ti} \\ v_{\ti}
                \end{bmatrix} \nonumber 
\end{IEEEeqnarray}
%
where the vector $r_{\ti}$ denotes the positions of the markers and $v_{\ti}$ their corresponding velocities. A constant velocity model \cite{Li2003} is used with the following transition equation,
%
\begin{IEEEeqnarray}{rCl}
 \begin{bmatrix} r_{\ti} \\ v_{\ti} \end{bmatrix} & = & \begin{bmatrix} \idmat & \idmat \\ \zmat & \idmat \end{bmatrix} \begin{bmatrix} r_{\ti-1} \\ v_{\ti-1} \end{bmatrix} + \begin{bmatrix} \zmat \\ \lgdm \end{bmatrix} \tn{\ti} \nonumber \\
 \tn{\ti} & \sim & \normalden{\tn{\ti}}{0}{\idmat}      .
\end{IEEEeqnarray}
%
Our objective is to learn $\lgdm$. Defining $\lgtp$ as before, the MCMC kernels derived in section~\ref{sec:degenerate_transition_models} can all be applied. The measured marker positions are modelled as independent noisy observations of $r_t$, i.e. with,
%
\begin{IEEEeqnarray}{rCl}
 \lgov & = & \frac{1}{\xi} \idmat \nonumber     .
\end{IEEEeqnarray}
%
It is straightforward to Gibbs sample $\xi$; the posterior conditional is a Gamma distribution.

The original data is down-sampled to $5$Hz, and consists of $250$ time instants with occasional measurements missing due to occlusion. The full data set is shown in figure~\ref{fig:mocap_data}.

\begin{figure}
 \centering
% \includegraphics[width=0.9\columnwidth]{../code/mocap_data.pdf}
 \caption{Motion capture data. Showing the three position coordinates over time for each of the four head markers.}
 \label{fig:mocap_data}
\end{figure}

In order to test the learning algorithm, we discard the observations from one marker at a time in $4$ sections each of $20$ time instants (i.e. $4$s each). The remaining observations are used to learn the system parameters with MCMC. Kalman smoothing with the resulting model is then used to estimate the marker positions in the sections of missing data, and a root-mean-square-error may be calculated by comparing the estimated marker positions to the true values.

The MCMC learning is run with $1000$ iterations, of which $500$ are discarded as burn-in. The sampler rapidly settles on a rank of $6$ and remains there for the rest of the chain. This accords with the fact that specifying the position of a solid body requires $6$ degrees of freedom. Figure~\ref{fig:mocap_Qhist} shows posterior histograms for a $6\times6$ block of the full covariance matrix.

\begin{figure}
 \centering
% \includegraphics[width=0.9\columnwidth]{../mocap/mocap_Qhist.pdf}
 \caption{Motion capture MCMC model learning. Posterior histograms for the elements of $\lgtv$ corresponding to the first two markers.}
 \label{fig:mocap_Qhist}
\end{figure}

Two other baseline methods are used for comparison: a Kalman smoother with $\lgdm=\idmat$, which corresponds to independent interpolation of each missing variable; and the missing value singular value decomposition algorithm \cite{Srebro2003,Liu2006,Li2009}, in which we use components sufficient to capture $95\%$ of the energy in each iteration. Results are shown in table~\ref{tab:mocap_rmse}. Figure~\ref{fig:missing_marker} shows one of the missing data sections and the estimates generated by the various methods.

\begin{table}[t]
\centering
\begin{tabular}{l|c}
 Algorithm                              & RMSE \\
 \hline
 Independent interpolation              & $0.505$ \\
 MSVD                                   & $0.101$ \\
 MCMC model mearning + Kalman Smoother  & $0.019$
\end{tabular}
\caption{RMSE results for missing marker estimation.}
\label{tab:mocap_rmse}
\end{table}

\begin{figure}
 \centering
% \includegraphics[width=0.9\columnwidth]{../code/missing_marker.pdf}
 \caption{First position coordinate of marker $1$, with data missing from time instant $90$ to $110$, showing ground truth (solid), Kalman smoother estimate with the learned model (mean as dashed, $\pm$2 standard deviations as dotted), independent interpolation (dash-dotted, upper), and MSVD (dash-dotted, lower).}
 \label{fig:missing_marker}
\end{figure}

\section{Conclusions}

Markov chain Monte Carlo algorithms are an effective method for Bayesian learning of linear state space systems. In this paper, we have introduced a number of new MCMC kernels for the linear system toolkit. New parameterisations have enabled us to formulate algorithms for learning transition models when it is known a priori that the model is (i) sparse or (ii) degenerate. The new algorithms are efficient and require minimal tuning, and have been verified with simulations.

\section*{Acknowledgements}
The authors are supported by the EPSRC BTaRoT grant. Our thanks to Michael Burke for advice and assistance with the motion capture application, and to Rich Wareham for his geometric insights.

\appendices

\section{A~Matrix~Factorisation for Positive~Semidefinite~Matrices} \label{app:givens-factorisation}

For a $\lsd\times\lsd$, rank-$\tnd$ positive semi-definite precision matrix $\lgtp$, we write the full eigendecomposition as,
%
\begin{IEEEeqnarray}{rCl}
 \lgtp & = & \tnmofull \begin{bmatrix}
                        \tnmd & \zmat \\
                        \zmat & \zmat
                       \end{bmatrix} \tnmofull\tr \nonumber      .
\end{IEEEeqnarray}
%
where $\tnmofull$ is a square orthogonal matrix and $\tnmd$ is a diagonal matrix of $\tnd$ eigenvalues.% This is not a unique factorisation, since $\lsd-\tnd$ of the columns of $\tnmofull$ do not contribute to $\lgtp$ at all.

Next we consider factorising $\tnmofull$ using Givens rotations. A Givens rotation matrix has the following structure,
%
\begin{IEEEeqnarray}{rCl}
 \left[\givmat{i}{j}{\givrot{}} - \idmat\right]_{k,l} & = & \begin{cases}
                                                    \cos(\givrot{})-1 & k=l=i \text{ or } k=l=j \\
                                                    \sin(\givrot{}) & k=i,l=j \\
                                                    -\sin(\givrot{}) & k=j,l=i \\
                                                    0 & \text{ otherwise}     ,
                                                 \end{cases}
\end{IEEEeqnarray}
%
where $\givrot{} \in [-\pi/2,\pi/2]$ is a rotation in the plane of the $i$ and $j$ coordinate directions. Any orthogonal matrix may be written as a product of $\half d(d-1)$ such rotations in the following way \cite{Anderson1987},
%
\begin{IEEEeqnarray}{rCl}
\tnmosign &\times& \left[ \givmat{1}{2}{\givrot{1,2}}\right] \times \dotsm \nonumber\\
&\times& \left[\givmat{1}{d-1}{\givrot{1,d-1}} \dotsm \givmat{d-2}{d-1}{\givrot{d-2,d-1}} \right] \nonumber\\
&\times& \left[\givmat{1}{d}{\givrot{1,d}} \dotsm \givmat{d-1}{d}{\givrot{d-1,d}} \right] \label{eq:standard_givens}     ,
\end{IEEEeqnarray}
%
where $\tnmosign$ is a diagonal matrix in which the diagonal elements are $\pm1$. This factorisation may be derived (and also calculated) by iteratively multiplying the original matrix by a Givens rotation matrix such that one element is set to $0$. The matrix remaining once all the elements below the diagonal have been eliminated is $\tnmosign$, and the factorisation above results from a straightforward rearrangement. See \cite{Anderson1987} for details.

The order of the rotation matrices in the Givens factorisation is not unique. By successively eliminating elements using both left and right multiplication with Givens matrices, it is possible in the same way to show that there is a unique factorisation,
%
\begin{IEEEeqnarray}{rCl}
 \tnmofull & = & \tnmocross \tnmosign \tnmorow \tnmonull      ,
\end{IEEEeqnarray}
%
where
%
\begin{IEEEeqnarray}{rCl}
 \tnmonull & = & \left[\givmat{r+1}{r+2}{\givrot{r+1,r+2}}\right] \times \dotsm \nonumber \\
 & & \qquad\qquad \times \left[\givmat{r+1}{d}{\givrot{r+1,d}} \dots \givmat{d-1}{d}{\givrot{d-1,d}}\right] \nonumber \\
 \tnmorow  & = & \left[\givmat{1}{2}{\givrot{1,2}}\right] \times \dotsm \nonumber \\
 & & \qquad\qquad \times \left[\givmat{1}{r}{\givrot{1,r}} \dots \givmat{r-1}{r}{\givrot{r-1,r}}\right] \nonumber \\
 \tnmocross & = & \left[\givmat{1}{r+1}{\givrot{1,r+1}}\dots\givmat{1}{d}{\givrot{1,d}}\right] \times \dotsm \nonumber \\
 & & \qquad\qquad \times \left[\givmat{r}{r+1}{\givrot{r,r+1}} \dots \givmat{r}{d}{\givrot{r,d}}\right] \nonumber      ,
\end{IEEEeqnarray}
%
and $\tnmosign$ is as before. The terms $\tnmonull$ and $\tnmorow$ have the following particular form,
%
\begin{IEEEeqnarray}{rCl}
 \tnmonull & = & \begin{bmatrix}
                  \idmat[r\times r] & \zmat[r\times(d-r)] \\
                  \zmat[(d-r)\times r] & \tnmonullred
                 \end{bmatrix} \nonumber \\
 \tnmorow & = & \begin{bmatrix}
                  \tnmorowred & \zmat[r\times(d-r)] \\
                  \zmat[(d-r)\times r] & \idmat[(d-r)\times(d-r)]
                 \end{bmatrix} \nonumber      .
\end{IEEEeqnarray}
%
Substituting into the eigendecomposition,
%
\begin{IEEEeqnarray}{rCl}
 \lgtp & = & \tnmofull \begin{bmatrix}
                        \tnmd & \zmat \\
                        \zmat & \zmat
                       \end{bmatrix} \tnmofull\tr \nonumber \\
 & = & \tnmocross \tnmosign \tnmorow \tnmonull \begin{bmatrix}
                        \tnmd & \zmat \\
                        \zmat & \zmat
                       \end{bmatrix} \tnmonull\tr \tnmorow\tr \tnmosign\tr \tnmocross\tr \nonumber \\
 & = & \tnmocross \begin{bmatrix}
                        \tnmosignred\tnmorowred\tnmd\tnmorowred\tr\tnmosignred\tr & \zmat \\
                        \zmat & \zmat
                       \end{bmatrix} \tnmocross\tr \nonumber      ,
\end{IEEEeqnarray}
%
where $\tnmosignred$ is the top-left $r\times r$ block of $\tnmosign$. Comparing with \eqref{eq:standard_givens}, $\tnmosignred\tnmorowred$ can represent any $r\times r$ orthogonal matrix and hence $\tnmosignred\tnmorowred\tnmd\tnmorowred\tr\tnmosignred\tr$ is a unique positive definite matrix, which we write $\tnmf$. Finally, define $\tnmocrossred$ as the first $r$ columns of $\tnmocross$, and we reach,
%
\begin{IEEEeqnarray}{rCl}
 \lgtp & = & \tnmocrossred \tnmf \tnmocrossred\tr      .
\end{IEEEeqnarray}
%
Note that the space of possible values for $\tnmocrossred$ is restricted by the Givens factorisation. In particular, note that it is defined by $r(d-r)$ independent parameters. However, since we keep $\tnmocrossred$ fixed when using this factorisation, this restriction does not matter.

\bibliographystyle{IEEEtran}
\bibliography{/home/pete/Dropbox/PhD/bibliographies/Cleanbib}
% \bibliography{/users/pete/Dropbox/PhD/Cleanbib}
% \bibliography{/users/pete/Dropbox/PhD/OTbib}
% \bibliography{/home/pete/Dropbox/PhD/OTbib}
\end{document}

\end{document}

### 2015-09-18 learners_mcmc.py
Added naive model scale learning. Fixed hyperparameters.

@@ -388,6 +388,40 @@ class MCMCLearnerObservationDiagonalCovarianceWithIGPrior():
         self.filter_current = False
 
 
+class MCMCLearnerNCVMTransitionCovarianceWithIGPrior():
+    __metaclass__ = ABCMeta
+    """
+    Container for MCMC system learning algorithm.
+    Model Type: Basic
+    Unknown Parameters: Q (transition covariance, assumed scaled NCVM)
+    Prior Type: Inverse Gamma
+    """
+
+    def sample_transition_ncvm_covariance(self):
+        """
+        MCMC iteration (Gibbs sampling) for diagonal observation covariance
+        matrix with inverse-gamma prior
+        """
+
+        # Calculate sufficient statistics using current state trajectory
+        suffStats = smp.evaluate_transition_sufficient_statistics(self.state)
+
+        # Update hyperparameters
+        Qs = self.model.parameters['Q']
+        Qs /= Qs[0,0]
+        a,b = smp.hyperparam_update_ncvm_ig_transition_variance(
+                                    suffStats,
+                                    self.model.parameters['F'],
+                                    Qs,
+                                    self.hyperparams['a0'],
+                                    self.hyperparams['b0'])  # This uses the same hyperparameters as the observation variance because I'm lazy and want to go to bed.
+
+        # Sample new parameter
+        s = stats.invgamma.rvs(a, scale=b)
+        self.model.parameters['Q'] = s*Qs
+
+        # self.flt and self.lhood are no longer up-to-date
+        self.filter_current = False
 
 
 class MCMCLearnerTransitionBasicModelWithMNIWPrior():

### 2015-09-18 run_mocap_mcmc.py

@@ -8,6 +8,7 @@ from linear_models import BasicLinearModel, DegenerateLinearModel
 from learners_mcmc import (
     BaseMCMCLearner,
     MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+    MCMCLearnerNCVMTransitionCovarianceWithIGPrior,
     MCMCLearnerTransitionBasicModelWithMNIWPrior,
     MCMCLearnerTransitionDegenerateModelWithMNIWPrior)
 from learners_mcmc_augmented import CleverDegenerateMCMCLearner
@@ -27,6 +28,7 @@ class MCMCDegenerateLearner(
 
 class MCMCNaiveLearner(
                 BaseMCMCLearner,
+                MCMCLearnerNCVMTransitionCovarianceWithIGPrior,
                 MCMCLearnerObservationDiagonalCovarianceWithIGPrior):
     pass
 
@@ -36,12 +38,11 @@ np.random.seed(0)
 data_path = './mocap-data/'
 test_data_file = './results/mocap-test-data.p'
 
-model_type = 'degenerate_augmented'#'degenerate'#'naive'#'basic'#
-num_iter = 2000
+model_type = 'naive'#'degenerate_augmented'#'degenerate'#'naive'#'basic'#
+num_iter = 20000
 
-num_burn = 1000#int(num_iter-10000)
-num_hold = 0#min(100,int(num_burn/2))
-num_warm = 10
+num_burn = 10000#int(num_iter-10000)
+num_warm = 100
 
 
 # Import marker data
@@ -85,7 +86,7 @@ Zmat = np.zeros((d,d))
 
 est_params['F'] = np.vstack((np.hstack((Imat,Imat)), np.hstack((Zmat,Imat))))
 
-est_params['Q'] = np.vstack((np.hstack((Imat,Imat)), np.hstack((Imat,Imat))))
+est_params['Q'] = 0.001*np.vstack((np.hstack((Imat,Imat)), np.hstack((Imat,Imat))))
 est_params['val'], est_params['vec'] = la.eigh(est_params['Q'])
 est_params['rank'] = np.array([ds])
 #val,vec = la.eigh(est_params['Q'])
@@ -104,8 +105,8 @@ est_naive_model = BasicLinearModel(ds, do, prior, est_params)
 # Hyperparameters
 hyperparams = dict()
 hyperparams['nu0'] = ds
[…]

### 2015-09-18 run_mocap_smc.py

@@ -27,9 +27,9 @@ np.random.seed(0)
 data_path = './mocap-data/'
 test_data_file = './results/mocap-test-data.p'
 
-num_iter = 750
-num_burn = 500
-thin = 5
+num_iter = 200
+num_burn = 100
+thin = 1
 num_rejuv = 2
 
 # Import marker data
@@ -72,7 +72,7 @@ Zmat = np.zeros((d,d))
 
 
 est_params['F'] = np.vstack((np.hstack((Imat,Imat)), np.hstack((Zmat,Imat))))
-est_params['Q'] = np.vstack((np.hstack((Imat,Imat)), np.hstack((Imat,Imat))))
+est_params['Q'] = 0.001*np.vstack((np.hstack((Imat,Imat)), np.hstack((Imat,Imat))))
 est_params['val'], est_params['vec'] = la.eigh(est_params['Q'])
 est_params['rank'] = np.array([ds])
 est_params['H'] = np.hstack((np.identity(d),np.zeros((d,d))))
@@ -85,7 +85,7 @@ est_naive_model = BasicLinearModel(ds, do, prior, est_params)
 
 # Hyperparameters
 hyperparams = dict()
-hyperparams['rPsi0'] = np.identity(ds)
+hyperparams['rPsi0'] = 0.001*np.identity(ds)
 hyperparams['M0'] = np.zeros((ds,ds))
 hyperparams['V0'] = 100*np.identity(ds)
 hyperparams['a0'] = 1

### 2015-09-18 sampling.py

@@ -179,6 +179,24 @@ def hyperparam_update_basic_ig_observation_variance(suffStats, H, a0, b0):
 
     return a,b
 
+def hyperparam_update_ncvm_ig_transition_variance(suffStats, F, Qs, a0, b0):
+    """
+    Update inverse gamma hyperparameters for the scale factor on the
+    transition covariance (which is assumed to be a fixed matrix, Qs, times
+    the scale factor, like an NCVM)
+    """
+    dx = F.shape[0]
+    Fss2 = np.dot(F,suffStats[2].T)
+    sumSquares = suffStats[3] - Fss2 - Fss2.T \
+                                          + np.dot(np.dot(F,suffStats[1]),F.T)
+
+    # Posterior hyperparameters
+    a = a0 + 0.5*dx*suffStats[0]
+    b = b0 + 0.5*np.trace( sumSquares )
+
+    return a,b
+
+
 def hyperparam_update_basic_mn_transition_matrix(suffStats, M0, V0):
     """
     Update matrix-normal hyperparameters for transition matrix conditional

### 2015-09-18 notes/paper-degenerate-linear-systems.tex
About halfwy through rewrite

@@
 This is achieved by appropriate factorisations of the covariance matrix which allow us still to exploit the fast-mixing Gibbs kernel.
-Furthermore, an extended distribution construction is introduced which enables the sampler to learn the rank of the transition covariance matrix.
+Furthermore, reversible jump construction \cite{Green1995,Green2009} is introduced which enables the sampler to learn the rank of the transition covariance matrix.
 
@@
 
-\subsection{MCMC for Basic Linear Gaussian Models}
+\subsection{Gibbs Sampling for Basic Linear Gaussian Models}
 
@@
 
-\section{Degenerate Transitions} \label{sec:degenerate_transition_models}
+\section{Degenerate Transition Models} \label{sec:degenerate_transition_models}
 
@@
  \ls{\ti} & = & \lgtm \ls{\ti-1} + \lgdm \tn{\ti} \label{eq:degenerate_transition} \\
- \tn{\ti} & \sim & \normalden{\tn{\ti}}{0}{\idmat}      .
+ \tn{\ti} & \sim & \normalden{\tn{\ti}}{0}{\idmat}      ,
 \end{IEEEeqnarray}
 %
-This is almost equivalent, with $\lgtv=\lgdm\lgdm\tr$.
-However, whereas before $\lgtv$ was required to be positive definite, this is not necessarily the case for $\lgdm\lgdm\tr$.
+in which $\tn{\ti}$ as dimension $\rk$, and $\lgdm$ is $\lsd\times\rk$.
+This is almost equivalent to the previous, with $\lgtv=\lgdm\lgdm\tr$.
+However, whereas before $\lgtv$ was required to be positive definite, this is no longer the case.
 When this occurs, the conventional transition density $\den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgdm)$ is not defined, and the sampling operations underlying our Gibbs sampler are no longer valid.
@@
 
-Attempting to learn $\lgdm$ directly is problematic, as this matrix is not identifiable.
-The distribution of $\lgdm\tn{\ti}$ is identical to that of $\lgdm \Xi \tn{\ti}$ where $\Xi$ is an arbitrary orthogonal matrix.
-Therefore, there is not a unique value of $\lgdm$ which maximises the probability of the state sequence $\den(\ls{1:\timax}|\lgtm,\lgdm)$.
-Instead, as in the basic model, we parameterise the system with a $\lsd\times\lsd$ precision matrix $\lgtp$ which is positive semi-definite with rank $\tnd$.
-For a particular value of $\lgtp$, $\lgdm$ may be any matrix such that,
+The matrix $\lgdm$ is not uniquely identifiable from the state sequence, because the distribution of $\lgdm\tn{\ti}$ is identical to that of $\lgdm \Xi \tn{\ti}$ where $\Xi$ is an arbitrary orthogonal matrix.
+To avoid this ambiguity, it is better to focus on learning $\lgtv$ as before, and to define $\lgdm$ as some matrix square root of $\lgtv$.
+In this new setting, $\lgtv$ is a positive semi-definite matrix with rank $\rk$.
+
+There are two difficulties with extending the methodology from the basic model to the degenerate case.
+First, the conjugate matrix normal--inverse Wishart distribution becomes singular.
+This places an undesirable constraint on the transition matrix which needs to be removed.
+Second, each transition, $(\ls{\ti}-\lgtm\ls{\ti-1})$, is now constrained to lie in a particular $\rk$-dimensional subspace within the state space; specifically, within the column space of $\lgdm$.
+This subspace is fixed conditional on any sampled state trajectory $\ls{1:\timax}$ (provided that $\timax>\rk$).
+This makes standard Gibbs sampling from $\postden(\lgtm, \lgtv|\ls{1:\timax})$ insufficient, since there are values of $\lgtm$ and $\lgtv$ which can never be reached.
+
+There is an obvious way to avoid being constrained to a particular subspace by the sampled state trajectory; simply return to targeting the marginal posterior $\postden(\lgtm, \lgtv|\ob{1:\timax})$ and fall back to using a generic Metropolis-Hastings approach.
+However, with a complex and high dimensional parameter space, designing efficient proposal distributions for Metropolis-Hastings is difficult.
+We would prefer to attain the automated, fast-mixing properties displayed by the Gibbs sampler on the full rank model.
+In this section we show how a Gibbs kernel may still be applied to sample a component of the parameters, with the remainder handled using Metropolis-Hastings.
+
+In formulating an MCMC algorithm for degenerate transition models, we will make use of two factorisations of the transition covariance matrix. First, an eigendecomposition,
 %
 \begin{IEEEeqnarray}{rCl}
- \lgdm\lgdm\tr & = & \lgtp\pinv     ,
+ \lgtv &=& \tvvec \tvval \tvvec\tr     ,
 \end{IEEEeqnarray}
 %
-where $\pinv$ denote the pseudo-inverse.
+in which $\tvval$ is a diagonal $\rk\times\rk$ matrix with $\rk = \rank(\lgtv)$, and $\tvvec$ is a $\lsd\times\rk$ matrix of orthonormal columns from the appropriate Stiefel manifold.
+This is the ``non-singular'' part of the factorisation.
+In order to resolve ambiguities in this representation, we follow \cite{Muirhead1982} and require that the first element in each column of $\tvval$ should be positive, and that eigenvalues in $\tvval$ be arranged in decreasing order of magnitude.
+The transformation from $\lgtv$ to $\{\tvval, \tvvec\}$ is then bijective.
 
-Learning degenerate transition models is challenging because each transition, $(\ls{\ti}-\lgtm\ls{\ti-1})$, is constrained to lie in a particular $\tnd$-dimensional subspace within the state space; specifically, within the column space of $\lgdm$.
-This subspace is fixed conditional on any sampled state trajectory $\ls{1:\timax}$ (provided that $\timax>\tnd$).
-This makes standard Gibbs sampling from $\postden(\lgtp|\lgtm,\ls{1:\timax})$ and $\postden(\lgtm|\lgtp,\ls{1:\timax})$ insufficient, since there are values of $\lgtm$ and $\lgtp$ which can never be reached.
-
-In order to allow the Markov chain to fully explore the parameter space, we introduce Markov kernels which target alternative conditional distributions, specifically $\postden(\lgtm, \ls{1:\timax}|\lgtp)$ and $\postden(\lgtp|\lgtm)$.
-Drawing from these distributions exactly is not possible, but we can use Metropolis-Hastings.
-Naively, we might simply try to propose changes to the elements of $\lgtm$ and $\lgtp$, but this is very inefficient.
-Better schemes can be formulated by exploiting the following factorisations.
-
-For a rank-$\tnd$ positive semi-definite precision matrix $\lgtp$,
+We can also write the ``full'' factorisation by finding a set of $\lsd-\rk$ additional orthonormal vectors and appending them to $\tvvec$, along with matching zeros to $\tvval$, as follows,
 %
 \begin{IEEEeqnarray}{rCl}
- \lgtp & = & \tnmo \tnmd \tnmo\tr \nonumber      ,
+ \lgtv &=& \begin{bmatrix}\tvvec & \tvvecorth \end{bmatrix} \begin{bmatrix}\tvval & \zmat \\ \zmat & \zmat \end{bmatrix} \begin{bmatrix}\tvvec\tr \\ \tvvecorth\tr \end{bmatrix}
 \end{IEEEeqnarray}
 %
-where $\tnmd = \diag([\tnmev{1},\tnmev{2},\dots])$ is the diagonal matrix of $\tnd$ \emph{positive} eigenvalues (such that $\tnmev{1} > \tnmev{2} > \dots > \tnmev{\tnd}$), and $\tnmo$ is a $\lsd\times\tnd$ matrix of orthonormal eigenvectors.
-This is a standard eigendecomposition.
-As in \cite{Muirhead1982}, the first non-zero element in each column of $\tnmo$ is required to be positive for each $i$.
-This resolves the sign ambiguity in the eigenvectors and thus ensures that the transformation from $\lgtp$ to $\{\tnmo,\tnmd\}$ is bijective.
+Note that $\tvvecorth$ is not unique.
 
-An alternative decomposition will also be useful. For a rank-$\tnd$ positive semi-definite precision matrix $\lgtp$,
+A further factorisation will also be useful, which we will refer to as the Givens decomposition (on account of its method of computation).
+For the rank-$\rk$ positive semi-definite covariance matrix $\lgtp$,
 %
-\begin{IEEEeqnarray}{rCl}
- \lgtp & = & \tnmocrossred \tnmf \tnmocrossred\tr \label{eq:degenerate_covariance_factorisation}      ,
+\begin{IEEEeqnarray}{rClCl}
+ \lgtv &=& \tvrot \tvfull \tvrot\tr &=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix} \begin{bmatrix}\tvfull & \zmat \\ \zmat & \zmat\end{bmatrix} \begin{bmatrix}\tvrot\tr \\ \tvrotorth\tr\end{bmatrix} \label{eq:givens_decomposition}
 \end{IEEEeqnarray}
@@
 %
-$\tnmf$ is a $\tnd\times\tnd$ positive definite matrix and $\tnmocrossred$ is a $\lsd\times\tnd$ matrix of orthonormal vectors subject to certain conditions.
-The transformation from $\lgtp$ to $\{\tnmocrossred,\tnmf\}$ is also bijective.
+$\tvfull$ is a $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors subject to certain constraints.
+The transformation from $\lgtv$ to $\{\tvfull, \tvrot\}$ is also bijective.
 See appendix~\ref{app:givens-factorisation} for details.
@@
 
-In the basic linear model, when $\lgtp$ was positive definite, we used a Wishart prior for this matrix.
-The singular Wishart distribution is an extension over the space of positive semi-definite matrices \cite{Uhlig1994}, and has the following density with respect to the Lebesgue measure on the distinct matrix elements, conditional on a known value of $\tnd$,
+The matrix normal--inverse Wishart conjugate prior will need some modifications in order to apply it to degenerate models.
+
+An appropriate prior distribution for the degenerate transition covariance matrix may be defined by setting the degrees of freedom of an inverse Wishart distribution to the desired rank, $\priordof=\rk$.
+This results in a singular inverse Wishart distribution over the rank-$\rk$ positive semi-definite matrices \cite{Diaz-Garcia2006}.
+The density (with respect to the Lesbegue measure on the distinct elements),
 %
 \begin{IEEEeqnarray}{rCl}
- \den(\lgtp|\tnd) & = & \wishartden{\lgtp}{\tnd}{\Ppscale} \nonumber \\
- & \propto & \determ{\tnmd}^{\frac{\tnd-\lsd-1}{2}} \exp\left( -\half \trace\left[ \Ppscale\inv \lgtp \right] \right)     .
+ \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrix}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrix \right)    ,
 \end{IEEEeqnarray}
 %
-The prior is completed by specifying a distribution on the integer $\tnd \in [1,\lsd]$.
+in which $\lgtv\pinv$ denotes the Moore-Penrose pseudoinverse of $\lgtv$.
 
-In order to exploit the two transformations of $\lgtp$ introduced in the previous section, we will need the prior distributions induced on $\{\tnmo,\tnmd\}$ and $\{\tnmocrossred,\tnmf\}$ by the singular Wishart distribution on $\lgtp$.
+The matrix normal component of the prior is more problematic.
+Although the distribution it is still well-defined, the fact that $\lgtv$ is singular means that the space of possible values is restricted to a subspace of $\reals^{\lsd\times\lsd}$, placing an unwanted constraint on our model.
+(See \cite{Diaz-Garcia2006} for a discussion of the singular matrix normal distribution.)
 
-For the first, if we constrain the scale matrix to be proportional to the identity matrix, $\Ppscale = \Ppscalescalar \idmat$, then it can be shown (see \cite{Uhlig1994,Srivastava2003} and \cite[chapter 3]{Muirhead1982}) that the eigenvectors and eigenvalues of $\lgtp$ are independent.
-Furthermore, the distribution for the eigenvector matrix is uniform (i.e.
-a normalised Haar measure, conditional on the first element being positive), and for the eigenvalues,
+In order to relax this unwanted constraint, the following non-singular prior distribution may be used,
 %
 \begin{IEEEeqnarray}{rCl}
- \den(\tnmd|\tnd) & = & \frac{ \pi^{\half\tnd^2} }{ (2\Ppscalescalar)^{\half\tnd\lsd} \gammafun[\tnd](\frac{\tnd}{2}) \gammafun[\tnd](\frac{\lsd}{2}) } \left[ \prod_{i<j}^{\tnd} (\tnmev{i}-\tnmev{j}) \right] \nonumber \\
- & & \qquad \times \left[ \prod_{i=1}^{\tnd} \tnmev{i}^{\half(\lsd-\tnd-1)} \exp\left(-\frac{1}{2\Ppscalescalar} \tnmev{i}\right) \right]      . \nonumber \\ \label{eq:eigenvalue_prior}
+ \lgtm|\lgtv &\sim& \matrixnormaldist{\priormeanmatrix}{\lgtv+\tvvecorth\priorextratpval\tvvecorth\tr}{\priorcolumnvariance}
+\end{IEEEeqnarray}
+%
+in which $\priorextratpval$ is a diagonal matrix of positive eigenvalues. The matrix $\lgtv+\tvvecorth\priorextratpval\tvvecorth\tr$ is then guaranteed to be positive definite.
+
+Using this prior, we need a mechanism for choosing $\priorextratpval$.
+These extra eigenvalues control the rate at which probability decays as the matrix moves away from the subspace defined by $\lgtv$.
+\meta{Add something about how to set these.}
+
+Finally, the prior is completed by specifying a distribution over the possible values of the rank $\den(\rk)$.
+
+\subsection{Constrained Gibbs Sampling for Degenerate Transitions}
+
+As discussed, standard Gibbs sampling using $\postden(\lgtm, \lgtv|\ls{1:\timax})$ does not allow the Markov chain to fully explore the parameter space because of the constraint that each transition $(\ls{\ti}-\lgtm\ls{\ti-1})$ must lie in the $\rk$-dimensional column space of $\lgdm$.
+Nevertheless, we can sample from these distributions exactly, and doing so allows the chain to efficiently explore values of the transition matrix and transition covariance within the permitted regions.
+These constrained Gibbs sampling steps rely on the Givens decomposition \eqref{eq:givens_decomposition}.
+
+\subsubsection{Likelihood and Subspace Constraints}
+There is no transition density over $\reals^{\lsd}$ for the degenerate model.
+However, a density does exist with respect to an appropriate measure on the reachable subspace, which provides us with a likelihood function \cite{Diaz-Garcia2006}.
+We can derive this from the underlying state transition equation,
+%
+\begin{IEEEeqnarray}{rCl}
+ \ls{\ti} &=& \lgtm \ls{\ti-1} + \tvrot \tvfull\msqrt \tn{\ti} \nonumber
+ \end{IEEEeqnarray}
+ \begin{IEEEeqnarray}{rCl}
+ \Rightarrow \tvrot\tr(\ls{\ti}-\lgtm\ls{\ti-1}) &\sim& \normaldist{\zmat}{\tvfull} \nonumber
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
+ \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) &\propto& \determ{\tvfull}^{-\half} \exp\left(-\half (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right)     .
+\end{IEEEeqnarray}
+%
+The subspace over which this is defined is,
+%
+\begin{IEEEeqnarray}{rCl}
+ \{\ls{\ti} : \ls{\ti} = \lgtm \ls{\ti-1} + \tvrot z, z \in \reals^{\rk}\}     .
 \end{IEEEeqnarray}
 
-For the second factorisation of $\lgtp$, since $\tnmf = \tnmocrossred\tr \lgtp \tnmocrossred$, we find that,
+The likelihood function for the entire state trajectory is thus,
+%
+\begin{align}
+ \den(\ls{1:\timax}|\lgtm, \lgtv) &= \den(\ls{0}) \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) \nonumber \\
+ &\propto \determ{\tvfull}^{-\half(\timax-1)} \exp\left( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber \\
+ &= \determ{\tvfull}^{-\half(\timax-1)} \exp\left(-\half\trace\left[ \tvfull\inv \tvrot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \right]\right) \nonumber \\
+ &= \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \tvrot\tr \lgtm \suffstats{1} \lgtm\tr \tvrot - \tvrot\tr \lgtm \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtm\tr \tvrot + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right) \\
+ &= \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtmrot\tr + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right)
+\end{align}
+%
+in which the sufficient statistics are as before and,
+%
+\begin{equation}
+ \lgtmrot = \tvrot\tr \lgtm      .
+\end{equation}
+
+Writing all the subspace constraints concisely in a single equation,
 %
 \begin{IEEEeqnarray}{rCl}
- \den(\tnmf|\tnmocrossred, \tnd) & = & \wishartden{\tnmf}{\tnd}{\Ppscalescalar \idmat}      .
+ \LS{2:\timax} &=& \lgtm \LS{1:\timax-1} + \tvrot Z
 \end{IEEEeqnarray}
 %
-(See \cite[theorem 3.2.5]{Muirhead1982}.) It is possible to derive a corresponding prior distribution for a parameterisation of $\tnmocrossred$ using a set of Givens rotations, but this will not be needed.
-
-\subsection{MCMC for Degenerate Transitions}
-
-As discussed above, standard Gibbs sampling from $\postden(\lgtp|\lgtm,\ls{1:\timax})$ and $\postden(\lgtm|\lgtp,\ls{1:\timax})$ does not allow the Markov chain to fully explore the parameter space because of the constraint that each transition $(\ls{\ti}-\lgtm\ls{\ti-1})$ must lie in the $\tnd$-dimensional column space of $\lgdm$.
-Nevertheless, we can sample from these distributions exactly, and doing so allows the chain to efficiently explore values of the transition matrix and transition covariance within the permitted regions.
-
-In addition to the within-subspace Gibbs moves, we introduce three efficient Metropolis-Hastings moves which target $\postden(\lgtp|\lgtm)$ and $\postden(\lgtm, \ls{1:\timax}|\lgtp)$, thus allowing the transition subspace to be changed and the parameter space to be fully explored by the Markov chain.
-
-\subsubsection{Gibbs Sampling within the Subspace}
-
-It is here that we exploit the transformation from $\lgtp$ to $\{\tnmocrossred,\tnmf\}$ introduced in \eqref{eq:degenerate_covariance_factorisation}.
-In the degenerate case, the transition density $\den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtp)$ with respect to the Lebesgue measure does not exist.
-However, since the set of possible values for $\ls{\ti}$ constitutes a hyperplane, we can project this onto a lower dimensional space in which a density does exist.
-From \eqref{eq:degenerate_transition} and \eqref{eq:degenerate_covariance_factorisation},
+where $Z \in \reals^{\rk\times\lsd}$ and 
 %
 \begin{IEEEeqnarray}{rCl}
- \tnmocrossred\tr(\ls{\ti}-\lgtm\ls{\ti-1}) & \sim & \normaldist{0}{\tnmf\inv}        .
+ \LS{2:\timax} &=& \begin{bmatrix} \ls{2} & \ls{3} & \hdots & \ls{\timax} \end{bmatrix}    .
+\end{IEEEeqnarray}
+
+Now introduce,
+%
+\begin{IEEEeqnarray}{rCl}
+ \lgtmorth &=& (\idmat-\tvrot\tvrot\tr) \lgtm     ,
 \end{IEEEeqnarray}
 %
-The likelihood of the state trajectory can then be written as,
+such that the transition matrix may be decomposed into orthogonal components,
 %
 \begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm,\tnmf,\tnmocross) } \nonumber \\
- & \propto & \begin{cases}
-                                  \determ{\tnmf}^{\frac{\timax-1}{2}} \exp\left( -\half \sum\limits_{\ti=1}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr\tnmocrossred \tnmf \tnmocrossred^T(\ls{\ti}-\lgtm\ls{\ti-1}) \right) \\ \hphantom{0} \qquad \text{if }\forall \,\, \ti, (\ls{\ti}-\lgtm\ls{\ti-1}) \text{ lies in the column space of } \tnmocrossred \\
-                                  0 \qquad \text{otherwise}     .
-                                                        \end{cases} \nonumber 
+ \lgtm &=& \tvrot \lgtmrot + \lgtmorth    .
 \end{IEEEeqnarray}
-
-We now re-formulate the posterior conditional for $\lgtp$ in terms of $\tnmocrossred$ and $\tnmf$,
+%
+Using this, the constraint equation becomes,
 %
 \begin{IEEEeqnarray}{rCl}
- \postden(\tnmocrossred, \tnmf | \lgtm,\ls{1:\timax}) & = & \postden(\tnmf | \tnmocrossred, \lgtm,\ls{1:\timax}) \postden(\tnmocrossred | \lgtm,\ls{1:\timax})    .
+ \LS{2:\timax} &=& (\tvrot \lgtmrot + \lgtmorth) \LS{1:\timax-1} + \tvrot Z \nonumber \\
+ \Rightarrow \LS{2:\timax} &=& \lgtmorth \LS{1:\timax-1} + \tvrot Z' \nonumber      .
 \end{IEEEeqnarray}
 %
-Because of the subspace constraint, $\tnmocrossred$ is uniquely defined conditional on $\ls{1:\timax}$, and thus sampling from $\postden(\tnmocrossred | \lgtm,\ls{1:\timax})$ simply implies keeping the existing value of $\tnmocrossred$.
-This can be proved using a ``degrees of freedom'' argument.
-The set of points $\{(\ls{\ti}-\lgtm\ls{\ti-1})\}_{t=1}^{\timax}$ lie in a $\tnd$-dimensional subspace, characterised by $\tnd(\lsd-\tnd)$ independent parameters.
-This is the same number as the parameters used to describe $\tnmocrossred$.
-(See appendix~\ref{app:givens-factorisation}.)
+Notice that $\lgtmrot$ and $\tvfull$ do not appear in this equation, and thus may be freely altered by the sampler.
+In contrast, $\lgtmorth$ and $\tvrot$ are uniquely determined.
+Sampling from $\postden(\lgtm,\lgtv|\ls{1:\timax})$ thus entails drawing new values from $\postden(\lgtmrot,\tvfull|\lgtmorth,\tvrot,\ls{1:\timax})$.
 
-For $\tnmf$ the appropriate conditional is,
+\subsubsection{Transforming the Prior}
+We will need the distribution on $\{\lgtmrot, \tvfull\}$ implied by the prior we chose for $\{\lgtm,\lgtv\}$.
+These follow from simple properties of the wishart and matrix normal distributions (see \cite{Muirhead1982}).
+For the matrix normal part,
 %
 \begin{IEEEeqnarray}{rCl}
- \postden(\tnmf|\lgtm, \tnmocrossred, \tnd, \ls{1:\timax}) &\propto& \den(\ls{1:\timax}|\lgtm,\tnmf,\tnmocrossred) \den(\tnmf) \nonumber \\
-% \qquad & \propto & \determ{\tnmf}^{\frac{\timax-2}{2}} \exp\left( -\half \trace\left[ \tnmf \left( \frac{1}{\Ppscalescalar}\idmat + \sum_{\ti=2}^{\timax} \tnmocrossred\tr(\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr\tnmocrossred \right) \right] \right)  \nonumber \\
- & = & \wishartden{\tnmf}{\Ppdof''}{\Ppscale''}      ,
+ \lgtmrot|\tvrot,\tvfull &\sim& \matrixnormaldist{\tvrot\tr\priormeanmatrix}{\tvfull}{\priorcolumnvariance}     .
 \end{IEEEeqnarray}
 %
-where,
+For the inverse Wishart part,
 %
 \begin{IEEEeqnarray}{rCl}
- \Ppdof''   & = & \tnd + \timax - 1 \nonumber \\
- \Ppscale'' & = & \left[\frac{1}{\Ppscalescalar}\idmat + \sum_{\ti=2}^{\timax} \tnmocrossred\tr(\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr\tnmocrossred \right]\inv \label{eq:D_posterior_params}      ,
+ \tvfull|\tvrot &\sim& \iwishartdist{\rk}{(\tvrot\tr\priorscalematrix\inv\tvrot)\inv}     .
 \end{IEEEeqnarray}
 %
-which closely resembles that for the precision matrix in the full rank case.
+Notice that this is no longer a singular Inverse Wishart distribution, since $\tvfull$ is $\rk\times\rk$.
 
-For $\lgtm$, if the current value is $\lgtm^*$, then the values which meet the constraint after sampling from $\postden(\lgtm|\tnmf,\tnmocrossred, \tnd, \ls{1:\timax})$ can be written as,
+\subsubsection{Sampling the Conditional}
+Using the constrained likelihood function and the transformed prior, we now obtain the required posterior conditional distribution $\postden(\lgtmrot,\tvfull|\lgtmorth,\tvrot,\ls{1:\timax})$.
+This follows the same algebraic steps as in the full rank case and leads to a similar matrix normal--inverse Wishart distribution,
+%
+\begin{align}
+ \tvfull | \tvrot, \ls{1:\timax} &\sim \iwishartdist{\postdof}{\postscalematrix} \\
+ \lgtmrot | \tvfull, \tvrot, \ls{1:\timax} &\sim \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}     ,
+\end{align}
+%
+with the following updated hyperparameters,
+%
+\begin{align}
+ \postcolumnvariance\inv                 &= \priorcolumnvariance\inv + \suffstats{1} \\
+ \postmeanmatrix \postcolumnvariance\inv &= \tvrot\tr\left( \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \right) \\
+ \postdof                                &= \rk + \suffstats{0} \\
+ \postscalematrix                        &= (\tvrot\tr\priorscalematrix\inv\tvrot)\inv + \tvrot\tr\left( \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr \right)\tvrot - \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr
+\end{align}
+
+Once new values of $\lgtmrot$ and $\tvfull$ have been sampled, the corresponding values of $\lgtm$ and $\lgtv$ may be calculated using the existing $\lgtmorth$ and $\tvrot$.
+
+\subsection{Metropolis-Hastings on the Posterior Marginal}
+
+Gibbs sampling is very efficient and requires no algorithm tuning but it only allows the Markov chain to explore a subset of the parameter space.
+In this section we introduce two Metropolis-Hastings kernels which target the marginal posterior $\postden(\lgtm,\lgtv)$ and allow the sampler to fully explore the target distribution.
+These sampling operations are each followed by a draw of the state trajectory from $\postden(\ls{1:\timax}|\lgtm,\lgtv)$, thus constituting a collapsed Gibbs move \cite{Dyk2008}.
+
+\subsubsection{Covariance Random Walk}
+
+The covariance matrix is specified by its eigenvalues and eigenvectors.
+The sampler is already able to explore the possible eigenvalues using the constrained Gibbs moves, but the eigenvector matrix is partially fixed during this process.
+We can allow the sampler to reach any possible eigenvector matrix using a rotational random walk.
+Suppose the current value of the transition covariance is $\lgtv\mcold$, a new value of $\lgtv$ is proposed using the following procedure:
+%
+\begin{itemize}
+ \item Sample an orthogonal matrix $\mhrot$ from some distribution $\mhrotppsl$.
+ \item Set $\lgtv\mcnew = \mhrot \lgtv\mcold \mhrot\tr$.
+\end{itemize}
+%
+The transformation is invertible and has a Jacobian of $1$.
+If in addition we require that $\mhrotppsl(\mhrot) = \mhrotppsl(\mhrot\tr)$, then the proposal distribution is symmetric and the acceptance probability is simply,
 %
 \begin{IEEEeqnarray}{rCl}
- \lgtm & = & \lgtm^* + \tnmocrossred \lgtmchange
+ \mhap(\lgtp\mcold\to\lgtp\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtm, \lgtp\mcnew) }{ \postden(\lgtm,\lgtp\mcold) } \right\} \\
+  & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtp\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtp\mcold)} \times \frac{\den(\lgtm,\lgtp\mcnew)}{\den(\lgtm,\lgtp\mcold)} \right\} \nonumber     .
 \end{IEEEeqnarray}
-%
-where $\lgtmchange$ is an arbitrary $\tnd\times\lsd$ matrix. We can then re-frame the sampling operation to draw $\lgtmchange$ from an appropriate conditional. The state sequence probability becomes,
-%
-\begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm^*,\lgtmchange,\tnmf,\tnmocross) } \nonumber \\
- & \propto & \exp\left( -\half \left[ \sum_{\ti=1}^{\timax} \ls{\ti-1}\tr \lgtmchange\tr \tnmf \lgtmchange \ls{\ti-1} \right.\right. \nonumber \\
- & & \qquad\qquad \left.\left.\vphantom{\sum_{\ti=1}^{\timax}}  -\:2 (\ls{\ti}-\lgtm^*\ls{\ti-1})\tr \tnmocrossred \tnmf \lgtmchange \ls{\ti-1} \right] \right)     . 
-\end{IEEEeqnarray}
-%
-Since $\lgtmchange = \tnmocrossred\tr(\lgtm-\lgtm^*)$, the prior for $\lgtmchange$ is simply a Gaussian, with mean and variance transformed from those of $\lgtm$.
-The prior and likelihood may be combined using the vectorisation method in a manner analogous to \eqref{eq:vectorisation_trick}, leading to a Gaussian distribution for $\lgtmchange$ which is closely related to that for $\lgtm$ in the full rank case.
-
-We will write $\mcmckern{\lgtp,\lgtm}(\cdot|\lgtm,\lgtp,\ls{1:\timax})$ to represent the Markov kernel which corresponds to transforming from $\lgtp$ to $\{\tnmocrossred,\tnmf\}$, sampling $\lgtm$ and $\tnmf$ from their posterior conditionals, and then transforming back.
-
-\subsubsection{A Markov Kernel for the Precision Matrix}
-
-In order to allow the Markov chain to fully explore the parameter space, we use Metropolis-Hastings moves to target the marginal distribution $\postden(\lgtp|\lgtm)$.
-The resulting kernel constitutes a \emph{collapsed Gibbs} move \cite{Dyk2008}.
-For $\lgtp$, we can construct an efficient Metropolis-Hastings kernel as follows.
-We sample a rotation matrix $\mhrot$ from some proposal distribution $\mhrotppsl$ and then apply the following transformation,
-%
-\begin{IEEEeqnarray}{rCl}
- \begin{bmatrix}
-  \lgtp\mcnew \\ \mhrot\mcnew
- \end{bmatrix}
- & = &
- \begin{bmatrix}
-  \mhrot\lgtp\mhrot\tr \\ \mhrot\tr
- \end{bmatrix} \nonumber     ,
-\end{IEEEeqnarray}
-%
-which is clearly its own inverse.
-Since $\determ{\mhrot}=1$, it is straightforward to show that the Jacobian of this transformation is $1$, and hence using the reversible jump interpretation of Metropolis-Hastings \cite{Green1995,Green2009}, the acceptance probability is,
-%
-\begin{IEEEeqnarray}{rCl}
- \mhap(\lgtp\to\lgtp\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtp\mcnew|\lgtm)\mhrotppsl(\mhrot\mcnew) }{ \postden(\lgtp|\lgtm)\mhrotppsl(\mhrot) } \right\}  \\
- & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtp\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtp)} \times \frac{\den(\lgtp\mcnew) \mhrotppsl(\mhrot\mcnew)}{\den(\lgtp) \mhrotppsl(\mhrot)} \right\} \nonumber     .
-\end{IEEEeqnarray}
-%
-The first term is simply a ratio of Kalman filter likelihoods. We will write $\mcmckern{\lgtp,1}(\cdot|\lgtm,\lgtp)$ to represent this Markov kernel.
 
@@
 %
-and then set $\mhrot=P(S)$.
-The Cayley transform has the property that $P(-S)=P(S\tr)=P(S)\inv=P(S)\tr$, which implies that $\mhrotppsl(\mhrot)=\mhrotppsl(\mhrot\tr)=\mhrotppsl(\mhrot\mcnew)$, leading to a cancellation in the acceptance probability.
+and then set $\mhrot=P(S)$. The Cayley transform has the property that $P(-S)=P(S)\inv=P(S)\tr$, which implies that $\mhrotppsl(\mhrot)=\mhrotppsl(\mhrot\tr)$ as required.
 
@@
 %
-and use $\mhrot=\givmat{i}{j}{\givrot{}}$.
-This also has the property that $\givmat{i}{j}{-\givrot{}} = \givmat{i}{j}{\givrot{}}\tr$, meaning that we achieve the same cancellation of the proposals as before.
+and use $\mhrot=\givmat{i}{j}{\givrot{}}$. This also has the property that $\givmat{i}{j}{-\givrot{}} = \givmat{i}{j}{\givrot{}}\tr$, implying a symmetric proposal.
+
+\subsubsection{Transition Matrix Random Walk}
+
+\subsection{Reversible Jump for Rank Learning}
+
+\section{Simulations}
+
+\subsection{A Toy Model}
+
+\subsection{Motion Capture Interpolation}
+
+%%% REWRITING %%%
+
+\section{Old Material}
+
+\subsubsection{Transforming the Prior}
+
+In order to exploit the two transformations of $\lgtp$ introduced in the previous section, we will need the prior distributions induced on $\{\tnmo,\tnmd\}$ and $\{\tnmocrossred,\tnmf\}$ by the singular Wishart distribution on $\lgtp$.
+
+For the first, if we constrain the scale matrix to be proportional to the identity matrix, $\Ppscale = \Ppscalescalar \idmat$, then it can be shown (see \cite{Uhlig1994,Srivastava2003} and \cite[chapter 3]{Muirhead1982}) that the eigenvectors and eigenvalues of $\lgtp$ are independent.
+Furthermore, the distribution for the eigenvector matrix is uniform (i.e.
+a normalised Haar measure, conditional on the first element being positive), and for the eigenvalues,
+%
+\begin{IEEEeqnarray}{rCl}
+ \den(\tnmd|\rk) & = & \frac{ \pi^{\half\rk^2} }{ (2\Ppscalescalar)^{\half\rk\lsd} \gammafun[\rk](\frac{\rk}{2}) \gammafun[\rk](\frac{\lsd}{2}) } \left[ \prod_{i<j}^{\rk} (\tnmev{i}-\tnmev{j}) \right] \nonumber \\
+ & & \qquad \times \left[ \prod_{i=1}^{\rk} \tnmev{i}^{\half(\lsd-\rk-1)} \exp\left(-\frac{1}{2\Ppscalescalar} \tnmev{i}\right) \right]      . \nonumber \\ \label{eq:eigenvalue_prior}
+\end{IEEEeqnarray}
 
@@
 \begin{algorithmic}
- \REQUIRE{$\tnd'$, $\tnmd'$, $\tnmo'$}
- \STATE Set $\tnd = \tnd' - 1$.
+ \REQUIRE{$\rk'$, $\tnmd'$, $\tnmo'$}
+ \STATE Set $\rk = \rk' - 1$.
  \STATE Remove largest eigenvalue $\tnmev{1}$. Form $\tnmd$ from the remaining eigenvalues.
  \STATE Remove corresponding eigenvector $\tnmevec{1}$. Form $\tnmo$ from the remaining eigenvectors.
- \RETURN{$\tnd$, $\tnmd$, $\tnmo$}
+ \RETURN{$\rk$, $\tnmd$, $\tnmo$}
 \end{algorithmic}
@@
 \begin{algorithmic}
- \REQUIRE{$\tnd$, $\tnmd$, $\tnmo$}
- \STATE Set $\tnd' = \tnd + 1$.
+ \REQUIRE{$\rk$, $\tnmd$, $\tnmo$}
+ \STATE Set $\rk' = \rk + 1$.
  \STATE Sample an eigenvalue $\tnmev{}^* \in (\tnmev{1},\infty)$ from $\ppslden{\tnmev{}}(\tnmev{})$. Form $\tnmd'$ from the new set of eigenvalues.
  \STATE Sample a corresponding eigenvector $\tnmevec{}^*$ from $\ppslden{\tnmevec{}}(\tnmevec{})$. Form $\tnmo'$ from the new set of eigenvectors.
- \RETURN{$\tnd'$, $\tnmd'$, $\tnmo'$}
+ \RETURN{$\rk'$, $\tnmd'$, $\tnmo'$}
 \end{algorithmic}
@@
 \begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{ \mhap\left( \tnd,\tnmd,\tnmo \to \tnd',\tnmd',\tnmo' \right) } \nonumber \\
- \qquad & = & \min\left\{1, \frac{ \postden(\lgtm, \tnd',\tnmd',\tnmo') }{ \postden(\lgtm, \tnd,\tnmd,\tnmo) } \times \frac{1}{\ppslden{\tnmevec{}}(\tnmevec{}^*)\ppslden{\tnmev{}}(\tnmev{}^*)} \right\} \nonumber \\
- & = & \min\Bigg\{1, \frac{\den(\ob{1:\timax} | \lgtm, \tnd',\tnmd',\tnmo') }{ \den(\ob{1:\timax} | \lgtm, \tnd,\tnmd,\tnmo) } \nonumber \\
- & & \quad \times \frac{\den(\tnmevec{1},\dots,\tnmevec{\tnd},\tnmevec{}^*)\den(\tnmev{1},\dots,\tnmev{\tnd},\tnmev{}^*|\tnd')\den(\tnd')}{\den(\tnmevec{1},\dots,\tnmevec{\tnd})\den(\tnmev{1},\dots,\tnmev{\tnd}|\tnd)\den(\tnd) \ppslden{\tnmevec{}}(\tnmevec{}^*)\ppslden{\tnmev{}}(\tnmev{}^*)} \Bigg\} \nonumber      ,
+ \IEEEeqnarraymulticol{3}{l}{ \mhap\left( \rk,\tnmd,\tnmo \to \rk',\tnmd',\tnmo' \right) } \nonumber \\
+ \qquad & = & \min\left\{1, \frac{ \postden(\lgtm, \rk',\tnmd',\tnmo') }{ \postden(\lgtm, \rk,\tnmd,\tnmo) } \times \frac{1}{\ppslden{\tnmevec{}}(\tnmevec{}^*)\ppslden{\tnmev{}}(\tnmev{}^*)} \right\} \nonumber \\
+ & = & \min\Bigg\{1, \frac{\den(\ob{1:\timax} | \lgtm, \rk',\tnmd',\tnmo') }{ \den(\ob{1:\timax} | \lgtm, \rk,\tnmd,\tnmo) } \nonumber \\
+ & & \quad \times \frac{\den(\tnmevec{1},\dots,\tnmevec{\rk},\tnmevec{}^*)\den(\tnmev{1},\dots,\tnmev{\rk},\tnmev{}^*|\rk')\den(\rk')}{\den(\tnmevec{1},\dots,\tnmevec{\rk})\den(\tnmev{1},\dots,\tnmev{\rk}|\rk)\den(\rk) \ppslden{\tnmevec{}}(\tnmevec{}^*)\ppslden{\tnmev{}}(\tnmev{}^*)} \Bigg\} \nonumber      ,
 \end{IEEEeqnarray}
 %
-where $\tnmevec{1},\dots,\tnmevec{\tnd}$ are the eigenvectors comprising $\tnmo$ and $\tnmev{1},\dots,\tnmev{\tnd}$ the eigenvalues comprising $\tnmd$.
+where $\tnmevec{1},\dots,\tnmevec{\rk}$ are the eigenvectors comprising $\tnmo$ and $\tnmev{1},\dots,\tnmev{\rk}$ the eigenvalues comprising $\tnmd$.
 The first term is simply a ratio of Kalman filter likelihoods.
@@
 
-The conditional prior distribution $\den(\tnmevec{}^*|\tnmevec{1},\dots,\tnmevec{\tnd})$ is the normalised Haar measure on the manifold $\tnmo\tr\tnmevec{}^*=0$, conditional on the constraint we imposed to resolve the sign ambiguity \cite{Muirhead1982}.
+The conditional prior distribution $\den(\tnmevec{}^*|\tnmevec{1},\dots,\tnmevec{\rk})$ is the normalised Haar measure on the manifold $\tnmo\tr\tnmevec{}^*=0$, conditional on the constraint we imposed to resolve the sign ambiguity \cite{Muirhead1982}.
 We can simulate according to this prior by sampling a standard Gaussian random vector and then using Gram-Schmidt orthogonalisation.
-This implies setting $\ppslden{\tnmevec{}}(\tnmevec{}^*)=\den(\tnmevec{}^*|\tnmevec{1},\dots,\tnmevec{\tnd})$, which results in a cancellation in the acceptance probability.
+This implies setting $\ppslden{\tnmevec{}}(\tnmevec{}^*)=\den(\tnmevec{}^*|\tnmevec{1},\dots,\tnmevec{\rk})$, which results in a cancellation in the acceptance probability.
 
@@
  \ppslden{\tnmev{}}(\tnmev{}^*) & \propto & \begin{cases}
-                                               \gammaden{\tnmev{}^*}{\frac{\lsd-\tnd}{2}}{2\Ppscalescalar} & \tnmev{}^*>\tnmev{1} \\
+                                               \gammaden{\tnmev{}^*}{\frac{\lsd-\rk}{2}}{2\Ppscalescalar} & \tnmev{}^*>\tnmev{1} \\
                                                0 & \text{otherwise}
@@
  & = & \begin{cases} 
- \frac{ (\tnmev{}^*)^{\frac{\lsd-\tnd}{2}-1} \exp\left( -\frac{\tnmev{}^*}{2\Ppscalescalar} \right) }{(2\Ppscalescalar)^{\frac{\lsd-\tnd}{2}} \compnormincgammafun\left(\frac{\lsd-\tnd}{2},\frac{\tnmev{1}}{2\Ppscalescalar}\right) }  & \tnmev{}^*>\tnmev{1} \\
+ \frac{ (\tnmev{}^*)^{\frac{\lsd-\rk}{2}-1} \exp\left( -\frac{\tnmev{}^*}{2\Ppscalescalar} \right) }{(2\Ppscalescalar)^{\frac{\lsd-\rk}{2}} \compnormincgammafun\left(\frac{\lsd-\rk}{2},\frac{\tnmev{1}}{2\Ppscalescalar}\right) }  & \tnmev{}^*>\tnmev{1} \\
  0 & \text{otherwise}     ,
@@
 \begin{IEEEeqnarray}{rCl}
- \frac{\den(\tnmev{1},\dots,\tnmev{\tnd},\tnmev{}^*|\tnd')}{\den(\tnmev{1},\dots,\tnmev{\tnd}|\tnd)\ppslden{\tnmev{}}(\tnmev{}^*)} & = & \frac{ \pi^{\half} \compnormincgammafun\left(\frac{\lsd-\tnd}{2},\frac{\tnmev{1}}{2\Ppscalescalar}\right) }{ (2\Ppscalescalar)^{\frac{\tnd}{2}} \gammafun\left(\frac{\tnd+1}{2}\right) } \prod_{i=1}^{\tnd} \tnmev{i}^{-\half}(\tnmev{}^*-\tnmev{i}) \nonumber \\
+ \frac{\den(\tnmev{1},\dots,\tnmev{\rk},\tnmev{}^*|\rk')}{\den(\tnmev{1},\dots,\tnmev{\rk}|\rk)\ppslden{\tnmev{}}(\tnmev{}^*)} & = & \frac{ \pi^{\half} \compnormincgammafun\left(\frac{\lsd-\rk}{2},\frac{\tnmev{1}}{2\Ppscalescalar}\right) }{ (2\Ppscalescalar)^{\frac{\rk}{2}} \gammafun\left(\frac{\rk+1}{2}\right) } \prod_{i=1}^{\rk} \tnmev{i}^{-\half}(\tnmev{}^*-\tnmev{i}) \nonumber \\
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{ \mhap\left( \tnd,\tnmd,\tnmo \to \tnd',\tnmd',\tnmo' \right) } \nonumber \\
- \qquad & = & \min\Bigg\{1, \frac{\den(\ob{1:\timax} | \lgtm, \tnd',\tnmd',\tnmo') }{ \den(\ob{1:\timax} | \lgtm, \tnd,\tnmd,\tnmo) } \times \frac{\den(\tnd')}{\den(\tnd)} \nonumber \\
- & & \qquad\qquad \times \frac{ \pi^{\half} \compnormincgammafun\left(\frac{\lsd-\tnd}{2},\frac{\tnmev{1}}{2\Ppscalescalar}\right) }{ (2\Ppscalescalar)^{\frac{\tnd}{2}} \gammafun\left(\frac{\tnd+1}{2}\right) } \prod_{i=1}^{\tnd} \tnmev{i}^{-\half}(\tnmev{}^*-\tnmev{i}) \Bigg\} \nonumber      ,
+ \IEEEeqnarraymulticol{3}{l}{ \mhap\left( \rk,\tnmd,\tnmo \to \rk',\tnmd',\tnmo' \right) } \nonumber \\
+ \qquad & = & \min\Bigg\{1, \frac{\den(\ob{1:\timax} | \lgtm, \rk',\tnmd',\tnmo') }{ \den(\ob{1:\timax} | \lgtm, \rk,\tnmd,\tnmo) } \times \frac{\den(\rk')}{\den(\rk)} \nonumber \\
+ & & \qquad\qquad \times \frac{ \pi^{\half} \compnormincgammafun\left(\frac{\lsd-\rk}{2},\frac{\tnmev{1}}{2\Ppscalescalar}\right) }{ (2\Ppscalescalar)^{\frac{\rk}{2}} \gammafun\left(\frac{\rk+1}{2}\right) } \prod_{i=1}^{\rk} \tnmev{i}^{-\half}(\tnmev{}^*-\tnmev{i}) \Bigg\} \nonumber      ,
 \end{IEEEeqnarray}
@@
  \includegraphics[width=0.7\columnwidth]{../code/degenerate_rank_with_rank.pdf} 
- \caption{MCMC Posterior histogram for $\tnd$. Degenerate transition model.}
+ \caption{MCMC Posterior histogram for $\rk$. Degenerate transition model.}
  \label{fig:degenerate_rank_histograms}
@@
 
-For a $\lsd\times\lsd$, rank-$\tnd$ positive semi-definite precision matrix $\lgtp$, we write the full eigendecomposition as,
+For a $\lsd\times\lsd$, rank-$\rk$ positive semi-definite precision matrix $\lgtp$, we write the full eigendecomposition as,
 %
@@
 %
-where $\tnmofull$ is a square orthogonal matrix and $\tnmd$ is a diagonal matrix of $\tnd$ eigenvalues.% This is not a unique factorisation, since $\lsd-\tnd$ of the columns of $\tnmofull$ do not contribute to $\lgtp$ at all.
+where $\tnmofull$ is a square orthogonal matrix and $\tnmd$ is a diagonal matrix of $\rk$ eigenvalues.% This is not a unique factorisation, since $\lsd-\rk$ of the columns of $\tnmofull$ do not contribute to $\lgtp$ at all.
 

### 2015-09-18 learners_mcmc.py
Changed to fixed padding hyperparameter and increased column variance.

@@ -481,9 +481,9 @@ class MCMCLearnerTransitionDegenerateModelWithMNIWPrior():
                                             Psi0)
 
         orthVec = model.complete_basis()
-#        relaxEval = self.hyperparams['alpha']
+        relaxEval = self.hyperparams['alpha']
 #        relaxEval = np.min(model.parameters['val'])
-        relaxEval = np.max(model.parameters['val'])
+#        relaxEval = np.max(model.parameters['val'])
         rowVariance = model.transition_covariance() \
                       + relaxEval*np.dot(orthVec,orthVec.T)
         matrixPrior = smp.matrix_normal_density(model.parameters['F'],

### 2015-09-18 run_mocap_mcmc.py

@@ -38,7 +38,7 @@ np.random.seed(0)
 data_path = './mocap-data/'
 test_data_file = './results/mocap-test-data.p'
 
-model_type = 'naive'#'degenerate_augmented'#'degenerate'#'naive'#'basic'#
+model_type = 'degenerate'#'degenerate_augmented'#'degenerate'#'naive'#'basic'#
 num_iter = 20000
 
 num_burn = 10000#int(num_iter-10000)
@@ -108,7 +108,8 @@ hyperparams['nu0'] = ds
 hyperparams['rPsi0'] = 0.001*np.identity(ds)
 hyperparams['Psi0'] = ds*hyperparams['rPsi0']
 hyperparams['M0'] = np.zeros((ds,ds))
-hyperparams['V0'] = 100*np.identity(ds)
+hyperparams['V0'] = 1E4*np.identity(ds)
+hyperparams['alpha'] = 0.001
 hyperparams['a0'] = 1
 hyperparams['b0'] = 0.001
 
@@ -116,7 +117,7 @@ hyperparams['b0'] = 0.001
 algoparams = dict()
 algoparams['rotate'] = 1E-4
 algoparams['perturb'] = 1E-8
-algoparams['pseudo_dof'] = None#1000
+#algoparams['pseudo_dof'] = None#1000
 #algoparams['pseudo_sd'] = 0.0001
 #algoparams['pseudo_shape'] = 1000
 

### 2015-09-21 notes/paper-degenerate-linear-systems.tex
Mocap parameters optimised. Added rank learning section to paper.

@@
 
+\meta{Handle these consistently.}
+
 The matrix normal--inverse Wishart conjugate prior will need some modifications in order to apply it to degenerate models.
@@
 We can allow the sampler to reach any possible eigenvector matrix using a rotational random walk.
-Suppose the current value of the transition covariance is $\lgtv\mcold$, a new value of $\lgtv$ is proposed using the following procedure:
+Suppose the current value of the transition covariance is $\lgtv\mcold$, a new value is proposed using the following procedure:
 %
@@
 \begin{IEEEeqnarray}{rCl}
- \mhap(\lgtp\mcold\to\lgtp\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtm, \lgtp\mcnew) }{ \postden(\lgtm,\lgtp\mcold) } \right\} \\
-  & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtp\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtp\mcold)} \times \frac{\den(\lgtm,\lgtp\mcnew)}{\den(\lgtm,\lgtp\mcold)} \right\} \nonumber     .
+ \mhap(\lgtv\mcold\to\lgtv\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtm, \lgtv\mcnew) }{ \postden(\lgtm,\lgtv\mcold) } \right\} \\
+  & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtv\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtv\mcold)} \times \frac{\den(\lgtm,\lgtv\mcnew)}{\den(\lgtm,\lgtv\mcold)} \right\} \nonumber     .
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- s_{k,l} & \sim & \normaldist{0}{\sigma_s^2} \label{eq:skewsymmetric_proposal}     .
+ s_{k,l} & \sim & \normaldist{0}{\sigma_Q^2} \label{eq:skewsymmetric_proposal}     .
 \end{IEEEeqnarray}
@@
 
-There is an alternative using Givens rotations.
+An alternative is to use Givens rotations.
 First sample $i \in [1,\lsd]$, $j \in [1,\lsd]\setminus i$, and $\givrot{} \in [-\pi/2,\pi/2]$ from some zero-mean distribution.
@@
 
+A more conventional Metropolis-Hastings kernel can be used for the transition matrix; for example, using a simple Gaussian random walk as the proposal distribution.
+If the current value is $\lgtm\mcold$, a new matrix is sampled from a matrix normal distribution,
+%
+\begin{IEEEeqnarray}{rCl}
+ \lgtm\mcnew &\sim& \matrixnormaldist{\lgtm\mcold}{\sigma_F^2\idmat}{\idmat}     ,
+\end{IEEEeqnarray}
+%
+and accepted with probability,
+%
+\begin{IEEEeqnarray}{rCl}
+ \mhap(\lgtm\mcold\to\lgtm\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtm\mcnew, \lgtv) }{ \postden(\lgtm\mcold,\lgtv) } \right\} \\
+  & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm\mcnew,\lgtv)}{\den(\ob{1:\timax}|\lgtm\mcold,\lgtv)} \times \frac{\den(\lgtm\mcnew,\lgtv)}{\den(\lgtm\mcold,\lgtv)} \right\} \nonumber     .
+\end{IEEEeqnarray}
+
 \subsection{Reversible Jump for Rank Learning}
+
+Since it is unlikely that the rank of $\lgtv$ will be known a priori, a mechanism is needed which will allow the sampler to learn this parameters as well.
+Changing the rank of $\lgtv$ changes the number of independent elements, and therefore the dimension of the target probability distribution.
+Thus, we will need to use reversible jump MCMC for this task \cite{Green1995,Green2009}.
+
+A reversible jump sampler is an MCMC algorithm on a family of probability distributions on spaces of different dimensions.
+Sampling operations are designed in pairs which allow the sampler to move between the different distributions.
+Each move begins by sampling additional variables (if required) and then applying a reversible transformation.
+Finally the newly proposed state is accepted or rejected according to a Metropolis-Hastings acceptance probability.
+
+For this problem, there is a distribution for each possible value of the rank.
+Two moves are required, to increase and decrease the rank.
+The moves suggested here are by no means unique, but are intuitive, simple to implement, and have proven effective in simulations.
+
+The reversible transformation can be accomplished in terms of eigenvectors and eigenvalues.
+Suppose the current value of the transition covariance is $\lgtv = \tvvec \tvval \tvvec\tr$, which has rank $\rk$.
+To increase the rank, first we sample an additional eigenvalue $\eval{\rk+1} \in [0,\eval{\rk}]$ (recall that the elements of $\tvval$ are sorted in decreasing order of magnitude, so $\eval{\rk+1} < \eval{\rk} < \dots < \eval{1}$) and a corresponding eigenvector $\evec{\rk+1} \in \evecset{\rk+1}$, from the set of vectors which could legitimately be appended to $\tvvec$ (i.e.
+orthonormal to the existing eigenvectors with positive sign of the first element),
+%
+\begin{IEEEeqnarray}{rCl}
+ \evecset{\rk+1} &=& \{\evec{} : |\evec{}|=1,\:\tvvec\tr\evec{}=0,\:\sign(\evec{\rk+1,1}=+1 \}    .
+\end{IEEEeqnarray}
+%
+A new covariance matrix is then generated using the transformation,
+%
+\begin{IEEEeqnarray}{rCl}
+ \lgtv\mcnew &=& \lgtv + \eval{\rk+1} \evec{\rk+1}\evec{\rk+1}\tr \\
+ &=& \begin{bmatrix} \tvvec & \evec{\rk+1} \end{bmatrix} \begin{bmatrix} \tvval & \zmat \\ \zmat & \eval{\rk+1} \end{bmatrix} \begin{bmatrix} \tvvec\tr \\ \evec{\rk+1}\tr \end{bmatrix}
+\end{IEEEeqnarray}
+%
+The inverse transformation, from $\lgtv\mcnew$ to $\{\lgtv, \eval{\rk+1}, \evec{\rk+1}\}$ simply requires identifying and removing the smallest eigenvalue and its corresponding eigenvector.
+This is an effective choice because the component with the smallest eigenvalue has the least effect on the likelihood of the observed data, and will thus have the highest probability of being accepted.
+
+For the sake of simplicity we suggest sampling the new eigenvalue and eigenvector independently from uniform distributions. For the eigenvalue,
+%
+\begin{IEEEeqnarray}{rCl}
+ \ppslden{\eval{}}(\eval{\rk+1}|\lgtv) = \begin{cases}
+                                          \frac{1}{\eval{\rk}} & 0 < \eval{\rk+1} < \eval{\rk} \\
+                                          0 & \text{ otherwise}    .
+                                         \end{cases}
+\end{IEEEeqnarray}
+%
+For the eigenvector,
+%
+\begin{IEEEeqnarray}{rCl}
+ \ppslden{\evec{}}(\evec{\rk+1}|\lgtv) = \begin{cases}
+                                          \frac{1}{C} & \evec{\rk+1} \in \evecset{\rk+1} \\
+                                          0 & \text{ otherwise}    .
+                                         \end{cases}
+\end{IEEEeqnarray}
+%
+$C$ is the volume of the manifold, which is half of a $(d-r-1)$-sphere (half because of the requirement that the sign of the first element be positive) \cite{Muirhead1982}. Hence,
+%
+\begin{IEEEeqnarray}{rCl}
+ C &=& \frac{\pi^{\half(\lsd-\rk)}}{\gammafun\left(\frac{\lsd-\rk}{2}\right)}
+\end{IEEEeqnarray}
+
+The acceptance probability for this move and its reverse respectively are \cite{Green1995},
+%
+\begin{IEEEeqnarray}{rCl}
+ \mhap(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) &=& \min\left\{1, \beta \right\} \\
+ \mhap(\lgtv\mcnew\to\lgtv,\eval{\rk+1},\evec{\rk+1}) &=& \min\left\{1, \beta\inv \right\} 
+\end{IEEEeqnarray}
+%
+\begin{IEEEeqnarray}{rCl}
+ \beta &=& \frac{ \postden(\lgtm, \lgtv\mcnew) \jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) }{ \postden(\lgtm,\lgtv) \ppslden{\eval{}}(\eval{\rk+1}|\lgtv) \ppslden{\evec{}}(\evec{\rk+1}|\lgtv) }     ,
+\end{IEEEeqnarray}
+%
+where $\jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew)$ is the Jacobian of the transformation.
+
+The Jacobian of the eigendecomposition for a positive semidefinite matrix is \cite{Diaz-Garcia2006},
+%
+\begin{IEEEeqnarray}{rCl}
+ \jac(\tvval,\tvvec\to\lgtv) &=& 2^{-\rk} \times \prod_{i=1}^{\rk}\eval{i}^{\lsd-\rk} \times \prod_{j=1}^{r}\prod_{i=1}^{j-1}(\eval{i}-\eval{j})      .
+\end{IEEEeqnarray}
+%
+Repeating this for $\lgtv\mcnew$ and taking the ratio of the two, we obtain,
+%
+\begin{IEEEeqnarray}{rCl}
+ \jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) &=& 2^{-1} \eval{\rk+1}^{\lsd-\rk-1} \prod_{i=1}^{\rk}\eval{i}^{-1} \prod_{i=1}^{\rk}(\eval{i}-\eval{\rk+1})     .
+\end{IEEEeqnarray}
+
+\meta{I don't have $\rk$ in the distributions here. Sort that out.}
 
@@
 
-\subsubsection{Transforming the Prior}
-
-In order to exploit the two transformations of $\lgtp$ introduced in the previous section, we will need the prior distributions induced on $\{\tnmo,\tnmd\}$ and $\{\tnmocrossred,\tnmf\}$ by the singular Wishart distribution on $\lgtp$.
-
-For the first, if we constrain the scale matrix to be proportional to the identity matrix, $\Ppscale = \Ppscalescalar \idmat$, then it can be shown (see \cite{Uhlig1994,Srivastava2003} and \cite[chapter 3]{Muirhead1982}) that the eigenvectors and eigenvalues of $\lgtp$ are independent.
-Furthermore, the distribution for the eigenvector matrix is uniform (i.e.
-a normalised Haar measure, conditional on the first element being positive), and for the eigenvalues,
-%
-\begin{IEEEeqnarray}{rCl}
- \den(\tnmd|\rk) & = & \frac{ \pi^{\half\rk^2} }{ (2\Ppscalescalar)^{\half\rk\lsd} \gammafun[\rk](\frac{\rk}{2}) \gammafun[\rk](\frac{\lsd}{2}) } \left[ \prod_{i<j}^{\rk} (\tnmev{i}-\tnmev{j}) \right] \nonumber \\
- & & \qquad \times \left[ \prod_{i=1}^{\rk} \tnmev{i}^{\half(\lsd-\rk-1)} \exp\left(-\frac{1}{2\Ppscalescalar} \tnmev{i}\right) \right]      . \nonumber \\ \label{eq:eigenvalue_prior}
-\end{IEEEeqnarray}
-
-\subsubsection{A Markov Kernel for the Precision Rank}
-
-It is unlikely that the rank of $\lgtp$ will be known a priori.
-We can learn this within the MCMC scheme by using the transformation of $\lgtp$ to $\{\tnmo,\tnmd\}$ and allowing the sampler to add and remove eigenvalue-eigenvector pairs using reversible jump moves \cite{Green1995,Green2009}.
-This can be achieved with a matching pair of moves.
-
-\begin{algorithm}
-\begin{algorithmic}
- \REQUIRE{$\rk'$, $\tnmd'$, $\tnmo'$}
- \STATE Set $\rk = \rk' - 1$.
- \STATE Remove largest eigenvalue $\tnmev{1}$. Form $\tnmd$ from the remaining eigenvalues.
- \STATE Remove corresponding eigenvector $\tnmevec{1}$. Form $\tnmo$ from the remaining eigenvectors.
- \RETURN{$\rk$, $\tnmd$, $\tnmo$}
-\end{algorithmic}
-\caption{Reversible Jump Move: Decrease rank}
-\label{alg:rjmcmc-down}
-\end{algorithm}
-
-\begin{algorithm}
-\begin{algorithmic}
- \REQUIRE{$\rk$, $\tnmd$, $\tnmo$}
- \STATE Set $\rk' = \rk + 1$.
- \STATE Sample an eigenvalue $\tnmev{}^* \in (\tnmev{1},\infty)$ from $\ppslden{\tnmev{}}(\tnmev{})$. Form $\tnmd'$ from the new set of eigenvalues.
- \STATE Sample a corresponding eigenvector $\tnmevec{}^*$ from $\ppslden{\tnmevec{}}(\tnmevec{})$. Form $\tnmo'$ from the new set of eigenvectors.
- \RETURN{$\rk'$, $\tnmd'$, $\tnmo'$}
-\end{algorithmic}
-\caption{Reversible Jump Move: Increase rank}
-\label{alg:rjmcmc-up}
-\end{algorithm}
-
-The acceptance probability for the increase move is then,
-%
-\begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{ \mhap\left( \rk,\tnmd,\tnmo \to \rk',\tnmd',\tnmo' \right) } \nonumber \\
- \qquad & = & \min\left\{1, \frac{ \postden(\lgtm, \rk',\tnmd',\tnmo') }{ \postden(\lgtm, \rk,\tnmd,\tnmo) } \times \frac{1}{\ppslden{\tnmevec{}}(\tnmevec{}^*)\ppslden{\tnmev{}}(\tnmev{}^*)} \right\} \nonumber \\
- & = & \min\Bigg\{1, \frac{\den(\ob{1:\timax} | \lgtm, \rk',\tnmd',\tnmo') }{ \den(\ob{1:\timax} | \lgtm, \rk,\tnmd,\tnmo) } \nonumber \\
- & & \quad \times \frac{\den(\tnmevec{1},\dots,\tnmevec{\rk},\tnmevec{}^*)\den(\tnmev{1},\dots,\tnmev{\rk},\tnmev{}^*|\rk')\den(\rk')}{\den(\tnmevec{1},\dots,\tnmevec{\rk})\den(\tnmev{1},\dots,\tnmev{\rk}|\rk)\den(\rk) \ppslden{\tnmevec{}}(\tnmevec{}^*)\ppslden{\tnmev{}}(\tnmev{}^*)} \Bigg\} \nonumber      ,
-\end{IEEEeqnarray}
-%
-where $\tnmevec{1},\dots,\tnmevec{\rk}$ are the eigenvectors comprising $\tnmo$ and $\tnmev{1},\dots,\tnmev{\rk}$ the eigenvalues comprising $\tnmd$.
-The first term is simply a ratio of Kalman filter likelihoods.
-For the decrease move, the ratio is replaced by its reciprocal.
-
-The conditional prior distribution $\den(\tnmevec{}^*|\tnmevec{1},\dots,\tnmevec{\rk})$ is the normalised Haar measure on the manifold $\tnmo\tr\tnmevec{}^*=0$, conditional on the constraint we imposed to resolve the sign ambiguity \cite{Muirhead1982}.
-We can simulate according to this prior by sampling a standard Gaussian random vector and then using Gram-Schmidt orthogonalisation.
-This implies setting $\ppslden{\tnmevec{}}(\tnmevec{}^*)=\den(\tnmevec{}^*|\tnmevec{1},\dots,\tnmevec{\rk})$, which results in a cancellation in the acceptance probability.
-
-For the eigenvalues, inspection of the joint density \eqref{eq:eigenvalue_prior} suggests the following truncated gamma proposal,
-%
-\begin{IEEEeqnarray}{rCl}
- \ppslden{\tnmev{}}(\tnmev{}^*) & \propto & \begin{cases}
-                                               \gammaden{\tnmev{}^*}{\frac{\lsd-\rk}{2}}{2\Ppscalescalar} & \tnmev{}^*>\tnmev{1} \\
-                                               0 & \text{otherwise}
-                                              \end{cases} \nonumber \\
- & = & \begin{cases} 
- \frac{ (\tnmev{}^*)^{\frac{\lsd-\rk}{2}-1} \exp\left( -\frac{\tnmev{}^*}{2\Ppscalescalar} \right) }{(2\Ppscalescalar)^{\frac{\lsd-\rk}{2}} \compnormincgammafun\left(\frac{\lsd-\rk}{2},\frac{\tnmev{1}}{2\Ppscalescalar}\right) }  & \tnmev{}^*>\tnmev{1} \\
- 0 & \text{otherwise}     ,
- \end{cases}
-\end{IEEEeqnarray}
-%
-where $\compnormincgammafun(\cdot,\cdot)$ is the complement of the normalised incomplete gamma function.
-%
-This leads to the following cancellation,
-%
-\begin{IEEEeqnarray}{rCl}
- \frac{\den(\tnmev{1},\dots,\tnmev{\rk},\tnmev{}^*|\rk')}{\den(\tnmev{1},\dots,\tnmev{\rk}|\rk)\ppslden{\tnmev{}}(\tnmev{}^*)} & = & \frac{ \pi^{\half} \compnormincgammafun\left(\frac{\lsd-\rk}{2},\frac{\tnmev{1}}{2\Ppscalescalar}\right) }{ (2\Ppscalescalar)^{\frac{\rk}{2}} \gammafun\left(\frac{\rk+1}{2}\right) } \prod_{i=1}^{\rk} \tnmev{i}^{-\half}(\tnmev{}^*-\tnmev{i}) \nonumber \\
-\end{IEEEeqnarray}
-%
-Thus simplifying the acceptance probability,
-%
-\begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{ \mhap\left( \rk,\tnmd,\tnmo \to \rk',\tnmd',\tnmo' \right) } \nonumber \\
- \qquad & = & \min\Bigg\{1, \frac{\den(\ob{1:\timax} | \lgtm, \rk',\tnmd',\tnmo') }{ \den(\ob{1:\timax} | \lgtm, \rk,\tnmd,\tnmo) } \times \frac{\den(\rk')}{\den(\rk)} \nonumber \\
- & & \qquad\qquad \times \frac{ \pi^{\half} \compnormincgammafun\left(\frac{\lsd-\rk}{2},\frac{\tnmev{1}}{2\Ppscalescalar}\right) }{ (2\Ppscalescalar)^{\frac{\rk}{2}} \gammafun\left(\frac{\rk+1}{2}\right) } \prod_{i=1}^{\rk} \tnmev{i}^{-\half}(\tnmev{}^*-\tnmev{i}) \Bigg\} \nonumber      ,
-\end{IEEEeqnarray}
-%
-We will write $\mcmckern{\lgtp,2}(\cdot|\lgtm,\lgtp)$ for the Markov kernel which corresponds to transforming from $\lgtp$ to $\{\tnmo,\tnmd\}$, conducting a reversible jump move as described above, and then transforming back.
-
-\subsubsection{A Markov Kernel for the Transition Matrix}
-
-Next we turn to $\lgtm$.
-If we modify the covariance matrix slightly so that it is no longer singular, then we can use the existing sampling procedure from the basic linear model and treat the resulting value of $\lgtm$ as a proposal in a Metropolis-Hastings kernel targeting $\postden(\lgtm, \ls{1:\timax}|\lgtp)$.
-
-Suppose we start with sampled values of $\lgtm$,$\lgtp$ and $\ls{1:\timax}$. A suitably modified precision matrix is,
-%
-\begin{IEEEeqnarray}{rCl}
- \paddedlgtp & = & \left(\lgtp\pinv + \Qpadding \idmat\right)\inv \label{eq:padded_transition_precision}      ,
-\end{IEEEeqnarray}
-%
-where $\Qpadding$ is a small positive constant. We propose a new value of $\lgtm$ by sampling,
-%
-\begin{IEEEeqnarray}{rCl}
-%  \lgtm' & \sim & \ppslden{\lgtm}(\lgtm|\lgtp,\ls{1:\timax}) & = & \postden(\lgtm|\paddedlgtp,\ls{1:\timax})     .
- \lgtm' & \sim & (\lgtm|\paddedlgtp,\ls{1:\timax})     .
-\end{IEEEeqnarray}
-%
-This distribution is simply a Gaussian and we can evaluate the density using \eqref{eq:basic_F_conditional}. Finally, we also sample a new state trajectory using forward filtering backward sampling,
-%
-\begin{IEEEeqnarray}{rCl}
- \ls{1:\timax}' & \sim & \postden(\ls{1:\timax}|\lgtm', \lgtp)      .
-\end{IEEEeqnarray}
-%
-The acceptance probability is simply,
-%
-\begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{ \mhap\left(\lgtm,\ls{1:\timax}\to\lgtm',\ls{1:\timax}'\right) } \nonumber \\
- \quad\quad & = & \min\left\{1,\frac{ \postden(\lgtm',\ls{1:\timax}'|\lgtp) \postden(\ls{1:\timax}|\lgtm,\lgtp) \postden(\lgtm|\paddedlgtp,\ls{1:\timax}')   }{ \postden(\lgtm,\ls{1:\timax}|\lgtp) \postden(\ls{1:\timax}'|\lgtm',\lgtp) \postden(\lgtm'|\paddedlgtp,\ls{1:\timax}) }\right\} \nonumber \\
- & = & \min\left\{1,\frac{ \postden(\lgtm'|\lgtp) \postden(\lgtm|\paddedlgtp,\ls{1:\timax}')   }{ \postden(\lgtm|\lgtp)\postden(\lgtm'|\paddedlgtp,\ls{1:\timax}) }\right\} \nonumber \\
- & = & \min\left\{1, \frac{ \den(\ob{1:\timax}|\lgtm',\lgtp) }{ \den(\ob{1:\timax}|\lgtm,\lgtp)} \times \frac{ \den(\lgtm') }{ \den(\lgtm)} \times \frac{ \postden(\lgtm|\paddedlgtp,\ls{1:\timax}')   }{ \postden(\lgtm'|\paddedlgtp,\ls{1:\timax}) }\right\} \nonumber      , \\
-\end{IEEEeqnarray}
-%
-in which the first term is a ratio of Kalman filter likelihoods. We will write $\mcmckern{\lgtm,\ls{}}(\cdot|\lgtm,\lgtp,\ls{1:\timax})$ for this Markov kernel.
-
-\subsubsection{The Complete Sampler}
-
-In this section we have described four Markov kernels which can be composed in order to learn a degenerate linear state space model.
-Each one by construction targets some conditional of the full posterior distribution and together they allow us to fully explore the parameter space.
-One iteration of the final algorithm consists of the following cycle.
-Starting with $\{\lgtm,\lgtp,\ls{1:\timax}\}$,
-%
-\begin{IEEEeqnarray}{rCl}
- \lgtm',\ls{1:\timax} & \sim & \mcmckern{\lgtm,\ls{}}(\cdot|\lgtm,\lgtp,\ls{1:\timax}) \nonumber \\
- \lgtp' & \sim & \mcmckern{\lgtp,1}(\cdot|\lgtm',\lgtp) \nonumber \\
- \ls{1:\timax}'' & \sim & \postden(\cdot|\lgtm',\lgtp') \nonumber \\
- \lgtp'' & \sim & \mcmckern{\lgtp,2}(\cdot|\lgtm',\lgtp') \nonumber \\
- \ls{1:\timax}''' & \sim & \postden(\ls{1:\timax}|\lgtm',\lgtp'') \nonumber \\
- \lgtm'',\lgtp''' & \sim & \mcmckern{\lgtp,\lgtm}(\cdot|\lgtm',\lgtp'',\ls{1:\timax}''') \nonumber     .
-\end{IEEEeqnarray}
-
 The only tuning required is to select the Metropolis-Hastings proposal distributions for $\mcmckern{\lgtp,1}$ and $\mcmckern{\lgtm,\ls{}}$.
@@
 An adaptive MCMC scheme, such as those discussed in \cite{Roberts2009}, can be used to modify these values as the chain runs for maximum efficiency.
-
-\subsection{Example}
-
-The algorithm was tested using the transition matrix specified in section~\ref{sec:example_basic}, but with the transition covariance changed to,
-%
-\begin{IEEEeqnarray}{rCl}
- \lgdm\lgdm\tr & = & \begin{bmatrix}
-              0.26 &  0.33 &  0.26 & -0.42 \\
-              0.33 &  1.14 &  0.46 & -0.85 \\ 
-              0.26 &  0.46 &  0.29 & -0.48 \\
-             -0.42 & -0.85 & -0.48 &  0.81
-             \end{bmatrix} \nonumber      ,
-\end{IEEEeqnarray}
-%
-(showing $2$ decimal places) which has a rank of $2$.
-
-The MCMC algorithm was run using an adaptation scheme from \cite{Roberts2009}, adjusting the two proposal parameters based on the acceptance rate of the associated Metropolis-Hastings moves.
-Posterior histograms for $\lgtm$, $\lgtv$, and for the rank of $\lgtv$, are shown in figure~\ref{fig:degenerate_FQ_histograms} and~\ref{fig:degenerate_rank_histograms} for one MCMC run.
-
-\begin{figure}
- \centering
- \subfloat[]{\includegraphics[width=0.7\columnwidth]{../code/degenerate_F_with_rank.pdf} \label{subf:degenerate_FQ_histograms:F}} \\
- \subfloat[]{\includegraphics[width=0.7\columnwidth]{../code/degenerate_Q_with_rank.pdf} \label{subf:degenerate_FQ_histograms:Q}}
- \caption{MCMC Posterior histograms for \protect\subref{subf:degenerate_FQ_histograms:F} $\lgtm$ and \protect\subref{subf:degenerate_FQ_histograms:Q} $\lgtv$.
-Degenerate transition model.
-Unknown rank.}
- \label{fig:degenerate_FQ_histograms}
-\end{figure}
-
-\begin{figure}
- \centering
- \includegraphics[width=0.7\columnwidth]{../code/degenerate_rank_with_rank.pdf} 
- \caption{MCMC Posterior histogram for $\rk$. Degenerate transition model.}
- \label{fig:degenerate_rank_histograms}
-\end{figure}
 
@@
  \label{fig:basic_FQ_histograms}
+\end{figure}
+
+\subsection{Example}
+
+The algorithm was tested using the transition matrix specified in section~\ref{sec:example_basic}, but with the transition covariance changed to,
+%
+\begin{IEEEeqnarray}{rCl}
+ \lgdm\lgdm\tr & = & \begin{bmatrix}
+              0.26 &  0.33 &  0.26 & -0.42 \\
+              0.33 &  1.14 &  0.46 & -0.85 \\ 
+              0.26 &  0.46 &  0.29 & -0.48 \\
+             -0.42 & -0.85 & -0.48 &  0.81
+             \end{bmatrix} \nonumber      ,
+\end{IEEEeqnarray}
+%
+(showing $2$ decimal places) which has a rank of $2$.
+
+The MCMC algorithm was run using an adaptation scheme from \cite{Roberts2009}, adjusting the two proposal parameters based on the acceptance rate of the associated Metropolis-Hastings moves.
+Posterior histograms for $\lgtm$, $\lgtv$, and for the rank of $\lgtv$, are shown in figure~\ref{fig:degenerate_FQ_histograms} and~\ref{fig:degenerate_rank_histograms} for one MCMC run.
+
+\begin{figure}
+ \centering
+ \subfloat[]{\includegraphics[width=0.7\columnwidth]{../code/degenerate_F_with_rank.pdf} \label{subf:degenerate_FQ_histograms:F}} \\
+ \subfloat[]{\includegraphics[width=0.7\columnwidth]{../code/degenerate_Q_with_rank.pdf} \label{subf:degenerate_FQ_histograms:Q}}
+ \caption{MCMC Posterior histograms for \protect\subref{subf:degenerate_FQ_histograms:F} $\lgtm$ and \protect\subref{subf:degenerate_FQ_histograms:Q} $\lgtv$.
+Degenerate transition model.
+Unknown rank.}
+ \label{fig:degenerate_FQ_histograms}
+\end{figure}
+
+\begin{figure}
+ \centering
+ \includegraphics[width=0.7\columnwidth]{../code/degenerate_rank_with_rank.pdf} 
+ \caption{MCMC Posterior histogram for $\rk$. Degenerate transition model.}
+ \label{fig:degenerate_rank_histograms}
 \end{figure}

### 2015-09-21 run_mocap_mcmc.py

@@ -108,7 +108,7 @@ hyperparams['nu0'] = ds
 hyperparams['rPsi0'] = 0.001*np.identity(ds)
 hyperparams['Psi0'] = ds*hyperparams['rPsi0']
 hyperparams['M0'] = np.zeros((ds,ds))
-hyperparams['V0'] = 1E4*np.identity(ds)
+hyperparams['V0'] = 1E2*np.identity(ds)
 hyperparams['alpha'] = 0.001
 hyperparams['a0'] = 1
 hyperparams['b0'] = 0.001

### 2015-09-21 notes/paper-degenerate-linear-systems.tex
Added figure plotting. Done most of rewrite.

@@
 In this section, we introduce a parameterisation for such degenerate transition models which can uniquely represent any such system, and show how MCMC moves may be implemented for learning.
+Here we treat $\rk$ as fixed and known; in the following section mechanisms will be introduced to learn this too.
 
 The matrix $\lgdm$ is not uniquely identifiable from the state sequence, because the distribution of $\lgdm\tn{\ti}$ is identical to that of $\lgdm \Xi \tn{\ti}$ where $\Xi$ is an arbitrary orthogonal matrix.
-To avoid this ambiguity, it is better to focus on learning $\lgtv$ as before, and to define $\lgdm$ as some matrix square root of $\lgtv$.
+To avoid this ambiguity, it is better to focus on learning $\lgtv$ as before, and to define $\lgdm$ as some matrix square root of $\lgtv$ (e.g.
+the Cholesky factor or the principal matrix square root).
 In this new setting, $\lgtv$ is a positive semi-definite matrix with rank $\rk$.
@@
 This is the ``non-singular'' part of the factorisation.
-In order to resolve ambiguities in this representation, we follow \cite{Muirhead1982} and require that the first element in each column of $\tvval$ should be positive, and that eigenvalues in $\tvval$ be arranged in decreasing order of magnitude.
+In order to resolve ambiguities in this representation, we follow \cite{Muirhead1982} and require that the first element in each column of $\tvvec$ should be positive, and that eigenvalues in $\tvval$ be arranged in decreasing order of magnitude.
 The transformation from $\lgtv$ to $\{\tvval, \tvvec\}$ is then bijective.
@@
 
-\meta{Handle these consistently.}
-
 The matrix normal--inverse Wishart conjugate prior will need some modifications in order to apply it to degenerate models.
@@
 An appropriate prior distribution for the degenerate transition covariance matrix may be defined by setting the degrees of freedom of an inverse Wishart distribution to the desired rank, $\priordof=\rk$.
-This results in a singular inverse Wishart distribution over the rank-$\rk$ positive semi-definite matrices \cite{Diaz-Garcia2006}.
-The density (with respect to the Lesbegue measure on the distinct elements),
+This results in a singular inverse Wishart distribution over the rank-$\rk$ positive semi-definite matrices \cite{Diaz-Garcia2006},
 %
 \begin{IEEEeqnarray}{rCl}
- \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrix}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrix \right)    ,
+ \lgtv|\rk &\sim& \iwishartdist{\rk}{\priorscalematrixrank{\rk}}     .
+\end{IEEEeqnarray}
+%
+The density (with respect to the Lesbegue measure on the distinct elements) is,
+%
+\begin{IEEEeqnarray}{rCl}
+ \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrixrank{\rk}}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrixrank{\rk} \right)    ,
 \end{IEEEeqnarray}
@@
 in which $\lgtv\pinv$ denotes the Moore-Penrose pseudoinverse of $\lgtv$.
+A prior estimate of $\lgtv$ is $\frac{1}{\rk}\priorscalematrixrank{\rk}$.
+Thus, in order to maintain consistency between the different ranks, we should set $\priorscalematrixrank{\rk} = \rk \priorscalematrixbase$, where $\priorscalematrixbase$ is an estimate of $\lgtv$.
 
@@
 These extra eigenvalues control the rate at which probability decays as the matrix moves away from the subspace defined by $\lgtv$.
-\meta{Add something about how to set these.}
-
-Finally, the prior is completed by specifying a distribution over the possible values of the rank $\den(\rk)$.
+A simple choice is to set $\priorextratpval=\priortypval\idmat$ where $\priortypval$ is a prior estimate of a ``typical'' eigenvalue.
+Alternatively, $\priortypval$ could be made dependent on $\tvval$; for example it could be set to the average (arithmetic or geometric) of the current eigenvalues.
+In practice, uninformative priors will often be used.
+If $\priorcolumnvariance\to\infty$, then the effect of $\priortypval$ becomes negligible.
 
@@
 %
-Notice that this is no longer a singular Inverse Wishart distribution, since $\tvfull$ is $\rk\times\rk$.
+Notice that this is Inverse Wishart distribution is no longer singular, since $\tvfull$ is $\rk\times\rk$.
 
@@
 \begin{IEEEeqnarray}{rCl}
- \lgtm\mcnew &\sim& \matrixnormaldist{\lgtm\mcold}{\sigma_F^2\idmat}{\idmat}     ,
+ \lgtm\mcnew &\sim& \matrixnormaldist{\lgtm\mcold}{\sigma_F^2\idmat}{\idmat} \label{eq:transition_matrix_proposal}     ,
 \end{IEEEeqnarray}
@@
 
-\subsection{Reversible Jump for Rank Learning}
+\section{Reversible Jump for Rank Learning}
 
@@
 Thus, we will need to use reversible jump MCMC for this task \cite{Green1995,Green2009}.
+
+First we introduce a prior distribution for the rank of $\lgtv$,
+%
+\begin{IEEEeqnarray}{rCl}
+ \den(\rk) &=& \sum_{i=1}^{\lsd} \priorrankweight{i} \diracden{\rk}{i}     ,
+\end{IEEEeqnarray}
+%
+where $\diracden{j}{i}$ is a probability point mass at $j=i$, and $\sum_i \priorrankweight{i} = 1$. The target distribution is now,
+%
+\begin{IEEEeqnarray}{rCl}
+ \postden(\lgtm, \lgtv, \rk) &\propto& \den(\ob{1:\timax}|\lgtm, \lgtv) \den(\lgtm, \lgtv|\rk) \den(\rk)    .
+\end{IEEEeqnarray}
 
@@
 \begin{IEEEeqnarray}{rCl}
- C &=& \frac{\pi^{\half(\lsd-\rk)}}{\gammafun\left(\frac{\lsd-\rk}{2}\right)}
+ C &=& \frac{\pi^{\half(\lsd-\rk)}}{\gammafun\left(\frac{\lsd-\rk}{2}\right)}     .
 \end{IEEEeqnarray}
+%
+In order to generate such a vector, first draw a $(\lsd-\rk)$-dimensional sample from $\normaldist{\zmat}{\idmat}$, normalise it, and then multiply by $\tvvecorth$, which is an arbitrary matrix of orthonormal columns such that $\tvvec\tr\tvvecorth=\zmat$.
 
@@
 \begin{IEEEeqnarray}{rCl}
- \beta &=& \frac{ \postden(\lgtm, \lgtv\mcnew) \jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) }{ \postden(\lgtm,\lgtv) \ppslden{\eval{}}(\eval{\rk+1}|\lgtv) \ppslden{\evec{}}(\evec{\rk+1}|\lgtv) }     ,
+ \beta &=& \frac{ \postden(\lgtm, \lgtv\mcnew, \rk+1) \jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) }{ \postden(\lgtm,\lgtv,\rk) \ppslden{\eval{}}(\eval{\rk+1}|\lgtv) \ppslden{\evec{}}(\evec{\rk+1}|\lgtv) } \times \frac{P(\rk+1\to\rk)}{P(\rk\to\rk+1)}     ,
 \end{IEEEeqnarray}
 %
-where $\jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew)$ is the Jacobian of the transformation.
+where $\jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew)$ is the Jacobian of the transformation and $P(\rk+1\to\rk)$ is the probability of performing a move which changes the rank from $\rk$ to $\rk+1$.
 
@@
 
-\meta{I don't have $\rk$ in the distributions here. Sort that out.}
-
 \section{Simulations}
 
+In the previous sections, we have assumed that $\lgom$ and $\lgov$ are fixed and known a prior. In practice, $\lgov$ will often be unknown. In both the following examples we use models which have,
+\begin{IEEEeqnarray}{rCl}
+ \lgov &=& \xi_{\ob{}} \idmat     ,
+\end{IEEEeqnarray}
+%
+in which $\xi_{\ob{}}$ is unknown.
+This additional parameter may be learned in combination with the transition model by the MCMC sampler.
+This entails changing the target distribution to $\postden(\lgtm, \lgtv, \xi_{\ob{}}, \ls{1:\timax})$.
+An additional Gibbs move is introduced to draw from $\postden(\xi_{\ob{}} | \lgtm, \lgtv, \ls{1:\timax})$.
+This may be done exactly if we use a conjugate inverse gamma prior.
+
+Both simulations use the random walk Metropolis-Hastings kernels to alter $\lgtm$ and $\lgtv$, using the Cayley transformation method to propose changes to $\lgtv$.
+Each kernel has a single scalar algorithm parameter which determines the size of the proposed moves, $\sigma_Q^2$ in \eqref{eq:skewsymmetric_proposal} and $\sigma_F^2$ in \eqref{eq:transition_matrix_proposal}.
+In order to accelerate mixing, these are adjusted adaptively based on the acceptance rate using the method introduced by \cite{Roberts2009}.
+
+When learning a degenerate transition model, the Metropolis-Hastings moves are less efficient at exploring the parameter space than the Gibbs moves.
+In order to reduce the length of the required burn in, we initialise the sampler by running a small number of iterations ($100$ in both simulations) with a full rank model.
+This puts the parameters in broadly the right region of the parameter space before the rank is allowed to reduce and the less efficient Metropolis-Hastings steps are required.
+
 \subsection{A Toy Model}
-
-\subsection{Motion Capture Interpolation}
-
-%%% REWRITING %%%
-
-\section{Old Material}
-
-The only tuning required is to select the Metropolis-Hastings proposal distributions for $\mcmckern{\lgtp,1}$ and $\mcmckern{\lgtm,\ls{}}$.
-In both cases these can be reduced to a single scale parameter; $\Qpadding$ in \eqref{eq:padded_transition_precision} for $\mcmckern{\lgtm,\ls{}}$ and $\sigma_s$ in \eqref{eq:skewsymmetric_proposal} for $\mcmckern{\lgtp,1}$.
-An adaptive MCMC scheme, such as those discussed in \cite{Roberts2009}, can be used to modify these values as the chain runs for maximum efficiency.
-
-\section{Applications} \label{sec:applications}
-
-\subsection{Example} \label{sec:example_basic}
 
@@
  \lgtm & = & \begin{bmatrix}
-              0.9 & 0.7 & 0.7  & 0    \\
-              0   & 0.9 & -0.5 & 0.1  \\
+              0.95 & 0.8 & 0.8  & 0    \\
+              0   & 0.95 & -0.5 & 0.1  \\
               0   & 0   & 1.6  & -0.8 \\
@@
 \begin{IEEEeqnarray}{rCl}
- \lgtv & = & \begin{bmatrix}
-              1.0  & 0.5  & 0.5  & 0.5 \\
-              0.5  & 1.0  & 0.5  & 0.5 \\ 
-              0.5  & 0.5  & 1.0  & 0.5 \\
-              0.5  & 0.5  & 0.5  & 1.0
-             \end{bmatrix} \nonumber      .
+ \lgtv & = & \begin{bmatrix} \half & \frac{1}{\sqrt{2}} \\ \half & -\frac{1}{\sqrt{2}} \\ \half & 0 \\ \half &  0 \end{bmatrix} \begin{bmatrix} \frac{3}{2} & 0 \\ 0 & \frac{1}{2} \end{bmatrix} \begin{bmatrix} \half & \half & \half & \half \\ \frac{1}{\sqrt{2}} & -\frac{1}{\sqrt{2}} & 0 & 0 \end{bmatrix} \nonumber      ,
 \end{IEEEeqnarray}
 %
-For the observation model, $\lgom = \idmat$ and $\lgov = \idmat{}$.
+which has a rank of $2$. For the observation model, $\lgom = \idmat$ and $\lgov = 0.1 \idmat{}$.
 
-A data set with $\timax=100$ time instants simulated from the model is shown in figure~\ref{fig:data_fullrank}.
-Running the Gibbs sampler results in rapid convergence.
-The resulting posterior histograms are shown in figure~\ref{fig:basic_FQ_histograms}, using $5000$ samples following a burn in of $1000$ iterations.
+A data set with $\timax=200$ time instants simulated from the model is shown in figure~\ref{fig:toy-data}.
+
+MCMC was used to learn models with both full rank and singular covariance matrices, using $10000$ iterations, of which $5000$ are discarded as burn in.
+
+The hyperparameters were set to generally uninformative values. For the full rank model,
+%
+\begin{IEEEeqnarray}{rCl}
+ \priordof &=& \lsd - 1 \\
+ \priorscalematrix &=& \priordof \idmat \\
+ \priormeanmatrix &=& \zmat \\
+ \priorcolumnvariance &=& 100 \idmat      .
+\end{IEEEeqnarray}
+%
+Similarly, for the degenerate model,
+%
+\begin{IEEEeqnarray}{rCl}
+ \priorscalematrixbase &=& \idmat \\
+ \priormeanmatrix &=& \zmat \\
+ \priorcolumnvariance &=& 100 \idmat \\
+ \priortypval &=& 1     .
+\end{IEEEeqnarray}
+
+The simulation results demonstrate several important features of the degenerate model and the associated MCMC algorithm.
+\begin{itemize}
+ \item The sampler identifies the true rank of the covariance matrix within a few tens of iterations, and remains at this value thereafter.
+ \item Trace plots suggest that the sampler converges within a few thousand iterations. See figure~\ref{fig:toy-R-trace}.
+ \item Autocorrelation plots show that the sampler mixes more slowly than that learning a full rank covariance model. See figure~\ref{fig:toy-R-acf}.
+ \item The parameter posterior histograms appear consistent with the true values. See figure~\ref{fig:toy-FQ-hist}.
+ \item The state trajectory is estimated more accurately with a degenerate covariance, with an RMSE of $5.51$, compared to $6.01$ for the full rank model.
+\end{itemize}
 
@@
  \centering
-% \includegraphics[width=0.9\columnwidth]{../code/data_fullrank.pdf}
- \caption{An example simulated data set. States (solid line) and observations (dots) of the four state components.}
- \label{fig:data_fullrank}
+ \includegraphics[width=0.9\columnwidth]{figures/toy-state.pdf}
+ \caption{An example data set simulated from the toy model, showing evolution of the four state components over time.}
+ \label{fig:toy-data}
 \end{figure}
@@
  \centering
-% \subfloat[]{\includegraphics[width=0.7\columnwidth]{../code/basic_F.pdf} \label{subf:basic_FQ_histograms:F}} \\
-% \subfloat[]{\includegraphics[width=0.7\columnwidth]{../code/basic_Q.pdf} \label{subf:basic_FQ_histograms:Q}}
- \caption{MCMC Posterior histograms for \protect\subref{subf:basic_FQ_histograms:F} $\lgtm$ and \protect\subref{subf:basic_FQ_histograms:Q} $\lgtv$.
-Basic linear transition model.
-Vertical lines indicate true values}
- \label{fig:basic_FQ_histograms}
-\end{figure}
-
-\subsection{Example}
-
-The algorithm was tested using the transition matrix specified in section~\ref{sec:example_basic}, but with the transition covariance changed to,
-%
-\begin{IEEEeqnarray}{rCl}
- \lgdm\lgdm\tr & = & \begin{bmatrix}
-              0.26 &  0.33 &  0.26 & -0.42 \\
-              0.33 &  1.14 &  0.46 & -0.85 \\ 
-              0.26 &  0.46 &  0.29 & -0.48 \\
-             -0.42 & -0.85 & -0.48 &  0.81
-             \end{bmatrix} \nonumber      ,
-\end{IEEEeqnarray}
-%
-(showing $2$ decimal places) which has a rank of $2$.
-
-The MCMC algorithm was run using an adaptation scheme from \cite{Roberts2009}, adjusting the two proposal parameters based on the acceptance rate of the associated Metropolis-Hastings moves.
-Posterior histograms for $\lgtm$, $\lgtv$, and for the rank of $\lgtv$, are shown in figure~\ref{fig:degenerate_FQ_histograms} and~\ref{fig:degenerate_rank_histograms} for one MCMC run.
-
-\begin{figure}
- \centering
- \subfloat[]{\includegraphics[width=0.7\columnwidth]{../code/degenerate_F_with_rank.pdf} \label{subf:degenerate_FQ_histograms:F}} \\
- \subfloat[]{\includegraphics[width=0.7\columnwidth]{../code/degenerate_Q_with_rank.pdf} \label{subf:degenerate_FQ_histograms:Q}}
- \caption{MCMC Posterior histograms for \protect\subref{subf:degenerate_FQ_histograms:F} $\lgtm$ and \protect\subref{subf:degenerate_FQ_histograms:Q} $\lgtv$.
-Degenerate transition model.
-Unknown rank.}
- \label{fig:degenerate_FQ_histograms}
+ \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-basic-R-trace.pdf} \label{subf:toy-R-trace:basic}} \\
+ \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-R-trace.pdf} \label{subf:toy-R-trace:degenerate}}
+ \caption{MCMC trace plots for $\xi_{\ob{}}$ for the full rank transition model \protect\subref{subf:toy-R-trace:basic} and the degenerate transition model \protect\subref{subf:toy-R-trace:degenerate}.}
+ \label{fig:toy-R-trace}
 \end{figure}
@@
  \centering
- \includegraphics[width=0.7\columnwidth]{../code/degenerate_rank_with_rank.pdf} 
- \caption{MCMC Posterior histogram for $\rk$. Degenerate transition model.}
- \label{fig:degenerate_rank_histograms}
+ \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-basic-R-acf.pdf} \label{subf:toy-R-acf:basic}} \\
+ \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-R-acf.pdf} \label{subf:toy-R-acf:degenerate}}
+ \caption{MCMC autocorrelation plots for $\xi_{\ob{}}$ for the full rank transition model \protect\subref{subf:toy-R-acf:basic} and the degenerate transition model \protect\subref{subf:toy-R-acf:degenerate}.}
+ \label{fig:toy-R-acf}
 \end{figure}
 
-\subsection{Rigid-Body Modelling with Motion Capture} \label{sec:mocap}
+\begin{figure}
+ \centering
+ \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-F-hist.pdf} \label{subf:toy-FQ-hist:F}} \\
+ \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-Q-hist.pdf} \label{subf:toy-FQ-hist:Q}}
+ \caption{MCMC Posterior histograms for \protect\subref{subf:toy-FQ-hist:F} $\lgtm$ and \protect\subref{subf:toy-FQ-hist:Q} $\lgtv$, from the degenerate linear transition model.
+Vertical lines indicate true values}
+ \label{fig:toy-FQ-hist}
+\end{figure}
 
-In this section, the MCMC algorithm for degenerate transition models is applied to a simplified motion capture task.
+\subsection{Motion Capture Interpolation}
+
+In a more realistic example, the MCMC algorithms are applied to the problem of marker interpolation in a motion capture application.
 The data used is from \cite{Aristidou2013}.
@@
 Since these are all attached to the same rigid body, we expect there to be substantial correlation in their motion.
-The linear model used is slightly modified from that introduced before.
-The latent state $\ls{\ti}$ consists of two parts,
-%
-\begin{IEEEeqnarray}{rCl}
- \ls{\ti} & = & \begin{bmatrix}
-                 r_{\ti} \\ v_{\ti}
-                \end{bmatrix} \nonumber 
-\end{IEEEeqnarray}
-%
-where the vector $r_{\ti}$ denotes the positions of the markers and $v_{\ti}$ their corresponding velocities. A constant velocity model \cite{Li2003} is used with the following transition equation,
-%
-\begin{IEEEeqnarray}{rCl}
- \begin{bmatrix} r_{\ti} \\ v_{\ti} \end{bmatrix} & = & \begin{bmatrix} \idmat & \idmat \\ \zmat & \idmat \end{bmatrix} \begin{bmatrix} r_{\ti-1} \\ v_{\ti-1} \end{bmatrix} + \begin{bmatrix} \zmat \\ \lgdm \end{bmatrix} \tn{\ti} \nonumber \\
- \tn{\ti} & \sim & \normalden{\tn{\ti}}{0}{\idmat}      .
-\end{IEEEeqnarray}
-%
-Our objective is to learn $\lgdm$.
-Defining $\lgtp$ as before, the MCMC kernels derived in section~\ref{sec:degenerate_transition_models} can all be applied.
-The measured marker positions are modelled as independent noisy observations of $r_t$, i.e.
-with,
-%
-\begin{IEEEeqnarray}{rCl}
- \lgov & = & \frac{1}{\xi} \idmat \nonumber     .
-\end{IEEEeqnarray}
-%
-It is straightforward to Gibbs sample $\xi$; the posterior conditional is a Gamma distribution.
+Sometimes markers are occluded meaning that position measurements are missing.
+If a system model for the motion of the markers can be learned, then the missing marker locations can be inferred by sampling or Kalman smoothing.
 
-The original data is down-sampled to $5$Hz, and consists of $250$ time instants with occasional measurements missing due to occlusion. The full data set is shown in figure~\ref{fig:mocap_data}.
+The original data is down-sampled to $5$Hz, and consists of $250$ time instants with occasional measurements missing due to occlusion.
+Observations from one marker at a time are discarded in $4$ sections each of $20$ time instants (i.e.
+$4$s each).
+The remaining observations are used to learn various system models with MCMC, and performance is assessed by comparing the root-mean-square error (RMSE) of the inferred positions in the missing sections.
+The data set is shown in figure~\ref{fig:mocap_data}.
 
@@
  \centering
-% \includegraphics[width=0.9\columnwidth]{../code/mocap_data.pdf}
+ \includegraphics[width=0.9\columnwidth]{figures/mocap-data.pdf}
  \caption{Motion capture data. Showing the three position coordinates over time for each of the four head markers.}
@@
 
-In order to test the learning algorithm, we discard the observations from one marker at a time in $4$ sections each of $20$ time instants (i.e.
-$4$s each).
-The remaining observations are used to learn the system parameters with MCMC.
-Kalman smoothing with the resulting model is then used to estimate the marker positions in the sections of missing data, and a root-mean-square-error may be calculated by comparing the estimated marker positions to the true values.
+We assume that the latent state consists of $3$-dimensional position and velocity for each marker in Cartesian coordinates,
+%
+\begin{IEEEeqnarray}{rCl}
+ \ls{\ti} & = & \begin{bmatrix}
+                 \mathbf{r}_{\ti} \\ \mathbf{v}_{\ti}
+                \end{bmatrix} \nonumber     .
+\end{IEEEeqnarray}
+%
+The observation model is defined by the following matrices,
+%
+\begin{IEEEeqnarray}{rCl}
+ \lgom &=& \begin{bmatrix} \idmat & \zmat \end{bmatrix} \\
+ \lgov &=& \xi_{\ob{}} \idmat     .
+\end{IEEEeqnarray}
 
-The MCMC learning is run with $1000$ iterations, of which $500$ are discarded as burn-in.
-The sampler rapidly settles on a rank of $6$ and remains there for the rest of the chain.
-This accords with the fact that specifying the position of a solid body requires $6$ degrees of freedom.
-Figure~\ref{fig:mocap_Qhist} shows posterior histograms for a $6\times6$ block of the full covariance matrix.
+Three transition models are trained using MCMC:
+%
+\begin{itemize}
+ \item A near constant velocity motion (NCVM) model \cite{Li2003} in which the motion of each marker is assumed independent from the rest.
+ \item A standard linear model, with full rank $\lgtv$.
+ \item A degenerate linear model, with $\lgtv$ singular and of unknown rank.
+\end{itemize}
+%
+The near constant velocity model has the following system matrices,
+%
+\begin{IEEEeqnarray}{rCl}
+ \lgtm &=& \begin{bmatrix}
+            \idmat & \idmat \\ \zmat & \idmat
+           \end{bmatrix} \\
+ \lgtv &=& \xi_{\ls{}} \begin{bmatrix}
+            \frac{1}{3}\idmat & \half\idmat \\ \half\idmat & \idmat
+           \end{bmatrix}     .
+\end{IEEEeqnarray}
+%
+$\xi_{\ls{}}$ is an unknown which may be learned within the MCMC scheme using Gibbs moves.
 
-\begin{figure}
+The hyperparameters are set to the same uninformative values as for the toy model, with the following exception,
+%
+\begin{IEEEeqnarray}{rCl}
+ \priorscalematrix &=& 0.001 \priordof \idmat \\
+ \priorscalematrixbase &=& 0.001 \idmat \\
+ \priortypval &=& 0.01     .
+\end{IEEEeqnarray}
+
+Each MCMC algorithm is run for $20000$ iterations, of which the first $10000$ are discarded as burn-in.
+The sampler for the degenerate model displays very similar behaviour as with the toy model.
+In the first few hundred iterations the parameters alter rapidly, and the covariance rank is reduced to a value of $9$, where it remains thereafter.
+The Metropolis-Hastings moves continue to alter the system matrices until the chain appears to converge after about $5000$ iterations.
+
+For each model, an estimate of the position of the missing markers is made by taking the mean of the sampled state trajectories from the Markov chain.
+The resulting RMSE is then calculated.
+For comparison, a missing value singular value decomposition (MSVD) algorithm is also tested \cite{Srebro2003,Liu2006,Li2009}, in which we use components sufficient to capture $95\%$ of the energy in each iteration.
+The RMSE results are shown in table~\ref{tab:mocap-rmse}.
+
+\begin{table}
  \centering
-% \includegraphics[width=0.9\columnwidth]{../mocap/mocap_Qhist.pdf}
- \caption{Motion capture MCMC model learning. Posterior histograms for the elements of $\lgtv$ corresponding to the first two markers.}
- \label{fig:mocap_Qhist}
-\end{figure}
-
-Two other baseline methods are used for comparison: a Kalman smoother with $\lgdm=\idmat$, which corresponds to independent interpolation of each missing variable; and the missing value singular value decomposition algorithm \cite{Srebro2003,Liu2006,Li2009}, in which we use components sufficient to capture $95\%$ of the energy in each iteration.
-Results are shown in table~\ref{tab:mocap_rmse}.
-Figure~\ref{fig:missing_marker} shows one of the missing data sections and the estimates generated by the various methods.
-
-\begin{table}[t]
-\centering
-\begin{tabular}{l|c}
- Algorithm                              & RMSE \\
+ \begin{tabular}{l|l}
+  Model          & RMSE \\
   \hline
- Independent interpolation              & $0.505$ \\
- MSVD                                   & $0.101$ \\
- MCMC model mearning + Kalman Smoother  & $0.019$
+  NCVM           & 4.692 \\
+  Basic          & 0.154 \\
+  Degenerate     & 0.086 \\
+  MSVD           & 0.926
  \end{tabular}
-\caption{RMSE results for missing marker estimation.}
-\label{tab:mocap_rmse}
+ \caption{Root-mean-square error (RMSE) for estimated missing marker positions, in metres.}
+ \label{tab:mocap-rmse}
 \end{table}
 
-\begin{figure}
- \centering
-% \includegraphics[width=0.9\columnwidth]{../code/missing_marker.pdf}
- \caption{First position coordinate of marker $1$, with data missing from time instant $90$ to $110$, showing ground truth (solid), Kalman smoother estimate with the learned model (mean as dashed, $\pm$2 standard deviations as dotted), independent interpolation (dash-dotted, upper), and MSVD (dash-dotted, lower).}
- \label{fig:missing_marker}
-\end{figure}
-
-\section{Conclusions}
+\section{Discussion and Conclusions}
 
@@
 In this paper, we have introduced a number of new MCMC kernels for the linear system toolkit.
-New parameterisations have enabled us to formulate algorithms for learning transition models when it is known a priori that the model is (i) sparse or (ii) degenerate.
-The new algorithms are efficient and require minimal tuning, and have been verified with simulations.
+
+Although the Markov chain takes substantially longer to converge with the degenerate covariance, it should be noted that the estimated model is better than that obtained in the full rank case after just a few hundred iterations, in the sense that the RMSE for the inferred missing positions is lower.
+Furthermore, using the degenerate model, we obtain an explicit estimate of the covariance rank.
+
+\meta{ 
+\begin{itemize}
+ \item Slow but better model.
+ \item Generalises the basic case.
+ \item Hyperparameters
+ \item Future work: better MH moves.
+\end{itemize}
+}
 
@@
 
+%%% REWRITING %%%
+
+\section{Old Material}
+
+\section{Conclusions}
+
+ New parameterisations have enabled us to formulate algorithms for learning transition models when it is known a priori that the model is (i) sparse or (ii) degenerate.
+The new algorithms are efficient and require minimal tuning, and have been verified with simulations.
+
 \appendices
@@
 \section{A~Matrix~Factorisation for Positive~Semidefinite~Matrices} \label{app:givens-factorisation}
+
+\meta{Re-do this!}
 
@@
 \bibliographystyle{IEEEtran}
-\bibliography{/home/pete/Dropbox/PhD/bibliographies/Cleanbib}
-% \bibliography{/users/pete/Dropbox/PhD/Cleanbib}
-% \bibliography{/users/pete/Dropbox/PhD/OTbib}
-% \bibliography{/home/pete/Dropbox/PhD/OTbib}
+% \bibliography{/home/pete/Dropbox/PhD/bibliographies/Cleanbib}
+\bibliography{/home/pete/Dropbox/PhD/bibliographies/OTbib}
 \end{document}

### 2015-09-21 analyse_mocap.py

@@ -60,9 +60,11 @@ def mocap_rmse(truth, original, estimate):
     err = truth-estimate
     return la.norm(err[np.isnan(original)])
 
+figure_path = './notes/figures/'
+
 # Load the test data
 data_path = './mocap-data/'
-test_path = './results/'#'./results/N20000/'#'./results/N2000/'#
+test_path = './results/mocap-final/'
 markers_truth = np.genfromtxt(data_path+'downsampled_head_markers_truth.csv', delimiter=',')
 test_data_file = 'mocap-test-data.p'
 fh = open(test_path+test_data_file, 'rb')
@@ -72,7 +74,6 @@ markers = pickle.load(fh)
 naive_learner = load_learner(test_path+'mocap-mcmc-naive.p')
 basic_learner = load_learner(test_path+'mocap-mcmc-basic.p')
 degenerate_learner = load_learner(test_path+'mocap-mcmc-degenerate.p')
-#degenerate_learner = load_learner('longresults/intermediate-results-60000.p')
 
 # Get state estimates from each algorithm
 num_burn = 10000
@@ -98,6 +99,14 @@ print("NCVM           | {}".format(naive_rmse))
 print("Basic          | {}".format(basic_rmse))
 print("Degenerate     | {}".format(degenerate_rmse))
 print("MSVD           | {}".format(msvd_rmse))
+
+# Draw it
+fig, axs = plt.subplots(nrows=3,ncols=1)
+color_list = 'brgc'
+for mm in range(4):
+    for dd in range(3):
+        axs[dd].plot(markers[:,3*mm+dd], color=color_list[mm])
+fig.savefig(figure_path+'mocap-data.pdf', bbox_inches='tight')
 
 # Trace plots
 naive_learner.plot_chain_trace('R', dims=([0],[0]))

### 2015-09-21 learners_mcmc.py

@@ -259,6 +259,8 @@ class BaseMCMCLearner():
                 axs[idx].plot([0,len(samples)-1],[trueValue[coords[idx]]]*2,
                                                              'r', linewidth=2)
 
+        return fig, axs
+
 
     def plot_chain_histogram(self, paramName, numBurnIn=0, dims=None,
                                                trueModel=None, derived=False):
@@ -310,6 +312,8 @@ class BaseMCMCLearner():
                                                                   linewidth=2)
                 axs[idx].set_ylim(ylims)
 
+        return fig, axs
+
 
     def plot_chain_acf(self, paramName, numBurnIn=0, dims=None, nlags=30,
                                                                derived=False):
@@ -352,6 +356,8 @@ class BaseMCMCLearner():
             axs[idx].plot(acf, 'k')
             axs[idx].plot([0,nlags], [0,0], 'k:')
             axs[idx].set_xlim([0,nlags])
+
+        return fig, axs
 
 
 class MCMCLearnerObservationDiagonalCovarianceWithIGPrior():

### 2015-09-21 run_basic_learning.py

@@ -61,7 +61,7 @@ num_burn = int(num_iter/5)
 
 for ii in range(num_iter):
     print("Running iteration {} of {}.".format(ii+1,num_iter))
-    
+
     learner.sample_transition()
     learner.sample_observation_diagonal_covariance()
     learner.sample_state_trajectory()

### 2015-09-21 run_mocap_mcmc.py

@@ -38,7 +38,7 @@ np.random.seed(0)
 data_path = './mocap-data/'
 test_data_file = './results/mocap-test-data.p'
 
-model_type = 'degenerate'#'degenerate_augmented'#'degenerate'#'naive'#'basic'#
+model_type = 'naive'#'degenerate_augmented'#'degenerate'#'naive'#'basic'#
 num_iter = 20000
 
 num_burn = 10000#int(num_iter-10000)
@@ -86,7 +86,7 @@ Zmat = np.zeros((d,d))
 
 est_params['F'] = np.vstack((np.hstack((Imat,Imat)), np.hstack((Zmat,Imat))))
 
-est_params['Q'] = 0.001*np.vstack((np.hstack((Imat,Imat)), np.hstack((Imat,Imat))))
+est_params['Q'] = 0.001*np.vstack((np.hstack((Imat/3.0,Imat/2.0)), np.hstack((Imat/2.0,Imat/1.0))))
 est_params['val'], est_params['vec'] = la.eigh(est_params['Q'])
 est_params['rank'] = np.array([ds])
 #val,vec = la.eigh(est_params['Q'])
@@ -109,7 +109,7 @@ hyperparams['rPsi0'] = 0.001*np.identity(ds)
 hyperparams['Psi0'] = ds*hyperparams['rPsi0']
 hyperparams['M0'] = np.zeros((ds,ds))
 hyperparams['V0'] = 1E2*np.identity(ds)
-hyperparams['alpha'] = 0.001
+hyperparams['alpha'] = 0.01
 hyperparams['a0'] = 1
 hyperparams['b0'] = 0.001
 

### 2015-09-21 run_toy.py

@@ -1,0 +1,150 @@
+from copy import deepcopy
+import pickle
+
+import numpy as np
+from matplotlib import pyplot as plt
+
+from kalman import GaussianDensity
+from linear_models import BasicLinearModel, DegenerateLinearModel
+from learners_mcmc import (
+    BaseMCMCLearner,
+    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+    MCMCLearnerTransitionBasicModelWithMNIWPrior,
+    MCMCLearnerTransitionDegenerateModelWithMNIWPrior)
+
+# Create a learner class for this run
+class MCMCLearnerB(BaseMCMCLearner,
+                    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+                    MCMCLearnerTransitionBasicModelWithMNIWPrior):
+    pass
+
+class MCMCLearnerD(BaseMCMCLearner,
+                    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+                    MCMCLearnerTransitionDegenerateModelWithMNIWPrior):
+    pass
+
+
+plt.close('all')
+np.random.seed(0)
+
+test_data_file = 'toy-test-data.p'
+model_type = 'degenerate'
+K = 200
+num_warm = 100
+num_iter = 10000
+num_burn = int(num_iter/2)
+
+
+ds = 4
+do = 4
+params = dict()
+params['F'] = np.array([[0.95,0.8,0.8,0.0],[0,0.95,-0.5,0.1],[0,0,1.6,-0.8],[0.0,0.0,1.0,0.0]])
+params['rank'] = np.array([2])
+params['val'] = np.array([1.5,0.5])
+params['vec'] = np.array([[0.5,0.5,0.5,0.5],[1.0/np.sqrt(2),-1.0/np.sqrt(2),0.0,0.0]]).T
+params['Q'] = np.dot(params['vec'], np.dot(np.diag(params['val']), params['vec'].T))
+params['H'] = np.identity(do)
+params['R'] = 0.1*np.identity(do)
+prior = GaussianDensity(np.zeros(ds), 10*np.identity(ds))
+
+model = DegenerateLinearModel(ds, do, prior, params)
+state, observ = model.simulate_data(K)
+
+# Draw it
+fig, axs = plt.subplots(nrows=ds,ncols=1)
+for dd in range(ds):
+    axs[dd].plot(observ[:,dd])
+
+# Save the data
[…]

### 2015-09-22 notes/paper-degenerate-linear-systems.tex
Completed paper rewrite first draft.

@@
 
-The Jacobian of the eigendecomposition for a positive semidefinite matrix is \cite{Diaz-Garcia2006},
+The Jacobian of the eigendecomposition for a positive semi-definite matrix is \cite{Diaz-Garcia2006},
 %
@@
 
-\subsection{Motion Capture Interpolation}
+\subsection{Motion Capture Interpolation} \label{sec:mocap}
 
@@
 Markov chain Monte Carlo algorithms are an effective method for Bayesian learning of linear state space systems.
-In this paper, we have introduced a number of new MCMC kernels for the linear system toolkit.
+In this paper, we have introduced new MCMC kernels for learning linear state space models which may have a singular transition covariance matrix of unknown rank.
+Such degenerate models may provide a better mathematical description of some systems, when the number of independent components in the state transitions is lower than the number of state dimensions.
+When a degenerate model is appropriate, the state estimates it produces can be more accurate than those obtained from an equivalent full rank model, as we have shown in simulations.
+Furthermore, with the rank of the transition covariance matrix included in the model, the sampler provides an explicit estimate of the posterior distribution over this parameter.
 
-Although the Markov chain takes substantially longer to converge with the degenerate covariance, it should be noted that the estimated model is better than that obtained in the full rank case after just a few hundred iterations, in the sense that the RMSE for the inferred missing positions is lower.
-Furthermore, using the degenerate model, we obtain an explicit estimate of the covariance rank.
+The price for using a degenerate transition model is an increased computational cost of the MCMC simulation.
+The computational complexity is the same as the full rank algorithm, dominated by the Kalman filtering and backward sampling operations which are $\bigo{\lsd^3 \timax}$.
+However, the need for additional Metropolis-Hastings steps increases the cost per iteration.
+More significantly, the Markov chain mixes more slowly and thus requires a longer chain to be simulated to achieve the same effective sample size.
+An important avenue for future research will be designing more sophisticated Metropolis-Hastings moves to explore the parameters space more efficiently than the simple random walks used here.
 
-\meta{ 
-\begin{itemize}
- \item Slow but better model.
- \item Generalises the basic case.
- \item Hyperparameters
- \item Future work: better MH moves.
-\end{itemize}
-}
+The standard linear Gaussian state space models, with $\rank(\lgtv)=\lsd$, discussed in section~\ref{sec:linear_gaussian_models} are a special case of the more general class used thereafter, for which $\rank(\lgtv)$ was treated as unknown.
+Correspondingly, the MCMC algorithm for the general case reduces to the standard Gibbs sampler of section~\ref{sec:linear_gaussian_models} if we re-impose the constraint that $\rank(\lgtv)=\lsd$.
+In this case $\tvfull=\lgtv$, $\tvrot=\idmat$, the ``constrained'' Gibbs moves are no longer constrained, and the Metropolis-Hastings moves become redundant.
 
@@
 
-%%% REWRITING %%%
-
-\section{Old Material}
-
-\section{Conclusions}
-
- New parameterisations have enabled us to formulate algorithms for learning transition models when it is known a priori that the model is (i) sparse or (ii) degenerate.
-The new algorithms are efficient and require minimal tuning, and have been verified with simulations.
-
 \appendices
 
-\section{A~Matrix~Factorisation for Positive~Semidefinite~Matrices} \label{app:givens-factorisation}
+\section{A~Matrix~Factorisation for Positive~Semi-Definite~Matrices} \label{app:givens-factorisation}
 
-\meta{Re-do this!}
-
-For a $\lsd\times\lsd$, rank-$\rk$ positive semi-definite precision matrix $\lgtp$, we write the full eigendecomposition as,
-%
-\begin{IEEEeqnarray}{rCl}
- \lgtp & = & \tnmofull \begin{bmatrix}
-                        \tnmd & \zmat \\
-                        \zmat & \zmat
-                       \end{bmatrix} \tnmofull\tr \nonumber      .
-\end{IEEEeqnarray}
-%
-where $\tnmofull$ is a square orthogonal matrix and $\tnmd$ is a diagonal matrix of $\rk$ eigenvalues.% This is not a unique factorisation, since $\lsd-\rk$ of the columns of $\tnmofull$ do not contribute to $\lgtp$ at all.
-
-Next we consider factorising $\tnmofull$ using Givens rotations. A Givens rotation matrix has the following structure,
+A Givens rotation matrix has the following structure,
 %
@@
 where $\givrot{} \in [-\pi/2,\pi/2]$ is a rotation in the plane of the $i$ and $j$ coordinate directions.
-Any orthogonal matrix may be written as a product of $\half d(d-1)$ such rotations in the following way \cite{Anderson1987},
+Any orthogonal $d\times d$ matrix may be written as a product of $\half d(d-1)$ such rotations in the following way \cite{Anderson1987},
 %
 \begin{IEEEeqnarray}{rCl}
-\tnmosign &\times& \left[ \givmat{1}{2}{\givrot{1,2}}\right] \times \dotsm \nonumber\\
+\tvvec = \tvsign &\times& \left[ \givmat{1}{2}{\givrot{1,2}}\right] \times \dotsm \nonumber\\
 &\times& \left[\givmat{1}{d-1}{\givrot{1,d-1}} \dotsm \givmat{d-2}{d-1}{\givrot{d-2,d-1}} \right] \nonumber\\
@@
 %
-where $\tnmosign$ is a diagonal matrix in which the diagonal elements are $\pm1$.
-This factorisation may be derived (and also calculated) by iteratively multiplying the original matrix by a Givens rotation matrix such that one element is set to $0$.
-The matrix remaining once all the elements below the diagonal have been eliminated is $\tnmosign$, and the factorisation above results from a straightforward rearrangement.
+where $\tvsign$ is a diagonal matrix in which the diagonal elements are $\pm1$.
+This factorisation may be derived (and also calculated) by iteratively left multiplying the original matrix by a Givens rotation matrix such that one element is set to $0$.
+The matrix remaining once all the elements below the diagonal have been eliminated is $\tvsign$, and the factorisation above results from a straightforward rearrangement.
 See \cite{Anderson1987} for details.
+This factorisation is not unique.
+Any order of Givens matrices may be used provided each operation does not interfere with the elements zeroed by the preceding operations.
+Furthermore, we can also zero an element of the matrix by right multiplication by a Givens matrix, so the rotations may be split into two blocks, one each side of the sign matrix.
 
-The order of the rotation matrices in the Givens factorisation is not unique.
-By successively eliminating elements using both left and right multiplication with Givens matrices, it is possible in the same way to show that there is a unique factorisation,
+Now consider instead a $d \times r$ matrix of orthogonal columns.
+This can be factorised using the same mechanism.
+First right multiply by Givens matrices to zero the elements on the first $r$ rows below the main diagonal.
+Then left multiply by Givens matrices to zero all the elements on rows $r+1$ to $d$.
+The resulting factorisation is,
 %
+
 \begin{IEEEeqnarray}{rCl}
- \tnmofull & = & \tnmocross \tnmosign \tnmorow \tnmonull      ,
+ \tvvec & = & \tvcross \tvsign \tvrow      ,
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- \tnmonull & = & \left[\givmat{r+1}{r+2}{\givrot{r+1,r+2}}\right] \times \dotsm \nonumber \\
- & & \qquad\qquad \times \left[\givmat{r+1}{d}{\givrot{r+1,d}} \dots \givmat{d-1}{d}{\givrot{d-1,d}}\right] \nonumber \\
- \tnmorow  & = & \left[\givmat{1}{2}{\givrot{1,2}}\right] \times \dotsm \nonumber \\
- & & \qquad\qquad \times \left[\givmat{1}{r}{\givrot{1,r}} \dots \givmat{r-1}{r}{\givrot{r-1,r}}\right] \nonumber \\
- \tnmocross & = & \left[\givmat{1}{r+1}{\givrot{1,r+1}}\dots\givmat{1}{d}{\givrot{1,d}}\right] \times \dotsm \nonumber \\
- & & \qquad\qquad \times \left[\givmat{r}{r+1}{\givrot{r,r+1}} \dots \givmat{r}{d}{\givrot{r,d}}\right] \nonumber      ,
+ \tvrow  & = & \left[\givmat{1}{2}{\givrot{1,2}}\right] \times \dotsm \nonumber \\
+ & & \qquad\qquad \times \left[\givmat{1}{r}{\givrot{1,r}} \dots \givmat{r-1}{r}{\givrot{r-1,r}}\right] \\
+ \tvcross & = & \left[\givmat{1}{r+1}{\givrot{1,r+1}}\dots\givmat{1}{d}{\givrot{1,d}}\right] \times \dotsm \nonumber \\
+ & & \qquad\qquad \times \left[\givmat{r}{r+1}{\givrot{r,r+1}} \dots \givmat{r}{d}{\givrot{r,d}}\right]      ,
 \end{IEEEeqnarray}
 %
-and $\tnmosign$ is as before. The terms $\tnmonull$ and $\tnmorow$ have the following particular form,
+and $\tvsign$ is as before, with the following structure,
 %
 \begin{IEEEeqnarray}{rCl}
- \tnmonull & = & \begin{bmatrix}
-                  \idmat[r\times r] & \zmat[r\times(d-r)] \\
-                  \zmat[(d-r)\times r] & \tnmonullred
-                 \end{bmatrix} \nonumber \\
- \tnmorow & = & \begin{bmatrix}
-                  \tnmorowred & \zmat[r\times(d-r)] \\
-                  \zmat[(d-r)\times r] & \idmat[(d-r)\times(d-r)]
+ \tvsign & = & \begin{bmatrix}
+                  \tvsigndiag \\
+                  \zmat[(d-r)\times r]
                  \end{bmatrix} \nonumber      .
 \end{IEEEeqnarray}
-%
-Substituting into the eigendecomposition,
+
+Now consider a rank-$r$ positive semi-definite matrix $\psd$ for which the non-singular part of the eigendecomposition is,
 %
 \begin{IEEEeqnarray}{rCl}
- \lgtp & = & \tnmofull \begin{bmatrix}
-                        \tnmd & \zmat \\
-                        \zmat & \zmat
-                       \end{bmatrix} \tnmofull\tr \nonumber \\
- & = & \tnmocross \tnmosign \tnmorow \tnmonull \begin{bmatrix}
-                        \tnmd & \zmat \\
-                        \zmat & \zmat
-                       \end{bmatrix} \tnmonull\tr \tnmorow\tr \tnmosign\tr \tnmocross\tr \nonumber \\
- & = & \tnmocross \begin{bmatrix}
-                        \tnmosignred\tnmorowred\tnmd\tnmorowred\tr\tnmosignred\tr & \zmat \\
-                        \zmat & \zmat
-                       \end{bmatrix} \tnmocross\tr \nonumber      ,
+ \psd &=& \tvvec \tvval \tvvec\tr     . 
 \end{IEEEeqnarray}
 %
-where $\tnmosignred$ is the top-left $r\times r$ block of $\tnmosign$.
-Comparing with \eqref{eq:standard_givens}, $\tnmosignred\tnmorowred$ can represent any $r\times r$ orthogonal matrix and hence $\tnmosignred\tnmorowred\tnmd\tnmorowred\tr\tnmosignred\tr$ is a unique positive definite matrix, which we write $\tnmf$.
-Finally, define $\tnmocrossred$ as the first $r$ columns of $\tnmocross$, and we reach,
+Factorising the matrix of eigenvectors using the Givens method, we find,
+
+\begin{IEEEeqnarray}{rCl}
+ \psd & = & \tvcross \tvsign \tvrow \tvval \tvrow\tr \tvsign\tr \tvcross\tr  \nonumber \\
+ & = & \tvcross \begin{bmatrix}
+                  \tvsigndiag \tvrow \tvval \tvrow\tr \tvsigndiag\tr & \zmat \\
+                  \zmat & \zmat
+                \end{bmatrix} \tvcross\tr \nonumber \\
+ & = & \tvrot \tvfull \tvrot\tr \nonumber      ,
+\end{IEEEeqnarray}
+%
+in which $\tvrot$ consists of the first $r$ columns of $\tvcross$, and,
 %
 \begin{IEEEeqnarray}{rCl}
- \lgtp & = & \tnmocrossred \tnmf \tnmocrossred\tr      .
+ \tvfull &=& \tvsigndiag \tvrow \tvval \tvrow\tr \tvsigndiag\tr    .
 \end{IEEEeqnarray}
 %
-Note that the space of possible values for $\tnmocrossred$ is restricted by the Givens factorisation.
-In particular, note that it is defined by $r(d-r)$ independent parameters.
-However, since we keep $\tnmocrossred$ fixed when using this factorisation, this restriction does not matter.
+Comparing with \eqref{eq:standard_givens}, we see that $\tvsigndiag \tvrow$ is an unconstrained $r\times r$ orthogonal matrix and hence $\tvfull$ is a unique positive definite matrix.
 

### 2015-09-22 analyse_toy.py

@@ -1,0 +1,118 @@
+import pickle
+import numpy as np
+import scipy.linalg as la
+from matplotlib import pyplot as plt
+
+from kalman import GaussianDensity
+from linear_models import BasicLinearModel, DegenerateLinearModel
+from learners_mcmc import (
+    load_learner,
+    BaseMCMCLearner,
+    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+    MCMCLearnerTransitionBasicModelWithMNIWPrior,
+    MCMCLearnerTransitionDegenerateModelWithMNIWPrior)
+
+# Create learner classes
+class MCMCLearnerB(BaseMCMCLearner,
+                    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+                    MCMCLearnerTransitionBasicModelWithMNIWPrior):
+    pass
+
+class MCMCLearnerD(BaseMCMCLearner,
+                    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+                    MCMCLearnerTransitionDegenerateModelWithMNIWPrior):
+    pass
+
+def rmse(state, estimate):
+    return la.norm(state-estimate)
+
+figure_path = './notes/figures/'
+
+# Load the test data
+test_path = './results/toy-final/'
+test_data_file = 'toy-test-data.p'
+fh = open(test_path+test_data_file, 'rb')
+model,state,observ = pickle.load(fh)
+
+# Load the MCMC output
+basic_learner = load_learner(test_path+'toy-mcmc-basic.p')
+degenerate_learner = load_learner(test_path+'toy-mcmc-degenerate.p')
+
+# Get state estimates from each algorithm
+num_burn = 5000
+basic_mn, basic_sd = basic_learner.estimate_state_trajectory(
+                                                           numBurnIn=num_burn)
+degenerate_mn, degenerate_sd = degenerate_learner.estimate_state_trajectory(
+                                                           numBurnIn=num_burn)
+
+# Assess RMSE
+basic_rmse = rmse(state, basic_mn)
+degenerate_rmse = rmse(state, degenerate_mn)
+
+# Display results
+print("Model          | RMSE")
[…]

### 2015-09-22 sandbox.py

@@ -2,6 +2,99 @@ import numpy as np
 import scipy.linalg as la
 from scipy import stats
 
+# Givens matrix operations
+
+import numpy as np
+import scipy.linalg as la
+
+def givens_matrix(d, r, c, g):
+    assert(r!=c)
+    if r > c:
+        tmp = c
+        c = r
+        r = tmp
+    U = np.identity(d)
+    U[r,r] = np.cos(g)
+    U[c,c] = np.cos(g)
+    U[r,c] = -np.sin(g)
+    U[c,r] = np.sin(g)
+    return U
+
+def plane_rotation(x):
+    """Find the plane rotation and Givens angle to zero a component"""
+    x = x.copy()
+    r = la.norm(x)
+    x /= r
+    if x[0] < 0:
+        x *= -1
+    if r > 0:
+        G = np.hstack((x[:,np.newaxis],x[::-1,np.newaxis]))
+        G[1,0] *= -1
+    else:
+        G = np.identity(2)
+
+    return G
+
+
+def givensise(U):
+    """
+    Factorise a matrix of orthonormal columns into row space and cross
+    rotations and a sign matrix, such that U = Ur x E x Uc
+    """
+
+    # Copy it so we don't break it
+    E = U.copy()
+
+    # Get the shape
+    d,r = E.shape
+
+    # Create arrays for the two rotation matrices
+    Ur = np.identity(r)
+    Uc = np.identity(d)
+
+    # Row space loop
+    for rr in reversed(range(r)):
+        for cc in range(rr):
+
+            # Make a slice index for this row and column
+            slce = slice(cc,rr+1,rr-cc)
+
+            # Get the next pair of elements to be rotated
+            v = E[rr,[rr,cc]]
+
+            # Find the plane rotation to zero the first element
+            G = plane_rotation(v)
+
+            # Zero the element
+            E[:,slce] = np.dot(E[:,slce],G)
+
+            # Corresponding change in Ur
+            Ur[slce,:] = np.dot(G.T,Ur[slce,:])
+
+    # Cross loop
+    for cc in range(r):
+        for rr in range(r,d):
+
+            # Make a slice index for this row and column
+            slce = slice(cc,rr+1,rr-cc)
+
[…]

### 2015-09-22 notes/paper-degenerate-linear-systems.tex
Edited.

@@
 \usepackage{hyperref}
-\usepackage{subfig}
+\usepackage[caption=false]{subfig}
 \usepackage{cite}
@@
 In section~\ref{sec:linear_gaussian_models} we review the Gibbs sampler for basic linear Gaussian state space systems, focusing on the transition model.
-In section~\ref{sec:degenerate_transition_models} we describe modifications for degenerate transitions.
-Simulations on a toy model, and a motion capture application are presented in section~\ref{sec:applications}.
+In section~\ref{sec:degenerate_transition_models} we describe modifications for degenerate transitions, and in section~\ref{sec:rank_learning} we show how the rank of the transition covariance may be learned.
+Simulations on a toy model, and a motion capture application are presented in section~\ref{sec:simulations}.
 
@@
 %
-\begin{align}
-\den(\lgtv) &= \frac{ \determ{\priorscalematrix}^{\frac{\priordof}{2}} }{ 2^{\frac{\priordof}{2}} \gammafun\left(\frac{\priordof}{2}\right) } \determ{\lgtv}^{-\frac{\priordof+\lsd+1}{2}} \exp\left( -\half \trace\left[\lgtv\inv\priorscalematrix\right] \right) \\
-\den(\lgtm | \lgtv) &= \determ{2 \pi \lgtv}^{-\half} \determ{2 \pi \priorcolumnvariance}^{-\half} \exp\left(-\half \trace\left[ \lgtv\inv (\lgtm-\priormeanmatrix) \priorcolumnvariance\inv (\lgtm-\priormeanmatrix)\tr \right] \right)
-\end{align}
+\begin{IEEEeqnarray}{rCl}
+\den(\lgtv) &=& \frac{ \determ{\priorscalematrix}^{\frac{\priordof}{2}} }{ 2^{\frac{\priordof}{2}} \gammafun\left(\frac{\priordof}{2}\right) } \determ{\lgtv}^{-\frac{\priordof+\lsd+1}{2}} \exp\left( -\half \trace\left[\lgtv\inv\priorscalematrix\right] \right) \nonumber \\
+\\
+\den(\lgtm | \lgtv) &=& \determ{2 \pi \lgtv}^{-\half} \determ{2 \pi \priorcolumnvariance}^{-\half} \nonumber \\
+& & \times \exp\left(-\half \trace\left[ \lgtv\inv (\lgtm-\priormeanmatrix) \priorcolumnvariance\inv (\lgtm-\priormeanmatrix)\tr \right] \right) \nonumber \\
+\end{IEEEeqnarray}
 %
@@
  \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm,\lgtv) \propto \prod_{\ti=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) } \\
- \quad\quad & \propto & \determ{\lgtv}^{-\frac{\timax-1}{2}} \exp\left( -\half \sum_{\ti=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \lgtv\inv (\ls{\ti}-\lgtm\ls{\ti-1}) \right)      .
+ & \propto & \determ{\lgtv}^{-\frac{\timax-1}{2}} \exp\left( -\half \sum_{\ti=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \lgtv\inv (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber      .
 \end{IEEEeqnarray}
@@
 \begin{align}
- \postcolumnvariance\inv                 &= \priorcolumnvariance\inv + \suffstats{1} \\
- \postmeanmatrix \postcolumnvariance\inv &= \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2}\\
- \postdof                                &= \priordof + \suffstats{0} \\
+ \postcolumnvariance\inv                 &= \priorcolumnvariance\inv + \suffstats{1} \nonumber \\
+ \postmeanmatrix \postcolumnvariance\inv &= \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \nonumber \\
+ \postdof                                &= \priordof + \suffstats{0} \nonumber \\
  \postscalematrix                        &= \priorscalematrix + \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr - \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr    ,
@@
 %
-\begin{align}
- \suffstats{0} &= \timax - 1 \\
- \suffstats{1} &= \sum_{t=2}^{\timax} \ls{\ti-1}\ls{\ti-1}\tr \\
- \suffstats{2} &= \sum_{t=2}^{\timax} \ls{\ti}\ls{\ti-1}\tr \\
- \suffstats{3} &= \sum_{t=2}^{\timax} \ls{\ti}\ls{\ti}\tr      .
-\end{align}
+\begin{IEEEeqnarray}{rClCrCl}
+ \suffstats{0} &=& \timax - 1 & \qquad & \suffstats{1} &=& \sum_{t=2}^{\timax} \ls{\ti-1}\ls{\ti-1}\tr \nonumber \\
+ \suffstats{2} &=& \sum_{t=2}^{\timax} \ls{\ti}\ls{\ti-1}\tr & \qquad & \suffstats{3} &=& \sum_{t=2}^{\timax} \ls{\ti}\ls{\ti}\tr      .
+\end{IEEEeqnarray}
 
@@
 %
-\begin{IEEEeqnarray}{rClCl}
- \lgtv &=& \tvrot \tvfull \tvrot\tr &=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix} \begin{bmatrix}\tvfull & \zmat \\ \zmat & \zmat\end{bmatrix} \begin{bmatrix}\tvrot\tr \\ \tvrotorth\tr\end{bmatrix} \label{eq:givens_decomposition}
+\begin{IEEEeqnarray}{rCl}
+ \lgtv &=& \tvrot \tvfull \tvrot\tr %&=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix} \begin{bmatrix}\tvfull & \zmat \\ \zmat & \zmat\end{bmatrix} \begin{bmatrix}\tvrot\tr \\ \tvrotorth\tr\end{bmatrix} \label{eq:givens_decomposition}
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrixrank{\rk}}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrixrank{\rk} \right)    ,
+ \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrixrank{\rk}}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \nonumber \\
+ & & \qquad \times \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrixrank{\rk} \right)    ,
 \end{IEEEeqnarray}
@@
 \end{IEEEeqnarray}
-\begin{IEEEeqnarray}{rCl}
- \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) &\propto& \determ{\tvfull}^{-\half} \exp\left(-\half (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right)     .
+\begin{IEEEeqnarray}{l}
+ \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) \propto \determ{\tvfull}^{-\half}  \\
+ \times \exp\left(-\half (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber     .
 \end{IEEEeqnarray}
@@
 %
-\begin{align}
- \den(\ls{1:\timax}|\lgtm, \lgtv) &= \den(\ls{0}) \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) \nonumber \\
- &\propto \determ{\tvfull}^{-\half(\timax-1)} \exp\left( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber \\
- &= \determ{\tvfull}^{-\half(\timax-1)} \exp\left(-\half\trace\left[ \tvfull\inv \tvrot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \right]\right) \nonumber \\
- &= \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \tvrot\tr \lgtm \suffstats{1} \lgtm\tr \tvrot - \tvrot\tr \lgtm \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtm\tr \tvrot + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right) \\
- &= \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtmrot\tr + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right)
-\end{align}
+\begin{IEEEeqnarray}{rCl}
+ \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm, \lgtv) = \den(\ls{0}) \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) } \nonumber \\
+ &\propto& \determ{\tvfull}^{-\half(\timax-1)} \exp\Bigg( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \nonumber \\
+ && \qquad\qquad\qquad\qquad\qquad\qquad\qquad \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \Bigg) \nonumber \\
+ %&=& \determ{\tvfull}^{-\half(\timax-1)} \exp\left(-\half\trace\left[ \tvfull\inv \tvrot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \right]\right) \nonumber \\
+ %&=& \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \tvrot\tr \lgtm \suffstats{1} \lgtm\tr \tvrot - \tvrot\tr \lgtm \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtm\tr \tvrot + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right) \\
+ &=& \determ{\tvfull}^{-\half\suffstats{0}} \exp\bigg(-\half\trace \big[ \tvfull\inv ( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tvrot \nonumber \\
+ & & \qquad\qquad\qquad\qquad\qquad -\: \tvrot\tr \suffstats{2} \lgtmrot\tr + \tvrot\tr \suffstats{3} \tvrot ) \big] \bigg)
+\end{IEEEeqnarray}
 %
@@
 %
-\begin{align}
- \postcolumnvariance\inv                 &= \priorcolumnvariance\inv + \suffstats{1} \\
- \postmeanmatrix \postcolumnvariance\inv &= \tvrot\tr\left( \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \right) \\
- \postdof                                &= \rk + \suffstats{0} \\
- \postscalematrix                        &= (\tvrot\tr\priorscalematrix\inv\tvrot)\inv + \tvrot\tr\left( \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr \right)\tvrot - \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr
-\end{align}
+\begin{IEEEeqnarray}{rCl}
+ \postcolumnvariance\inv                 &=& \priorcolumnvariance\inv + \suffstats{1} \\
+ \postmeanmatrix \postcolumnvariance\inv &=& \tvrot\tr\left( \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \right) \\
+ \postdof                                &=& \rk + \suffstats{0} \\
+ \postscalematrix                        &=& (\tvrot\tr\priorscalematrix\inv\tvrot)\inv + \tvrot\tr\left( \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr \right)\tvrot \nonumber \\
+ & & \qquad\qquad\qquad\qquad\qquad -\: \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr
+\end{IEEEeqnarray}
 
@@
 
-\section{Reversible Jump for Rank Learning}
+\section{Reversible Jump for Rank Learning} \label{sec:rank_learning}
 
@@
 \begin{IEEEeqnarray}{rCl}
- \beta &=& \frac{ \postden(\lgtm, \lgtv\mcnew, \rk+1) \jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) }{ \postden(\lgtm,\lgtv,\rk) \ppslden{\eval{}}(\eval{\rk+1}|\lgtv) \ppslden{\evec{}}(\evec{\rk+1}|\lgtv) } \times \frac{P(\rk+1\to\rk)}{P(\rk\to\rk+1)}     ,
+ \beta &=& \frac{ \postden(\lgtm, \lgtv\mcnew, \rk+1) \jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) }{ \postden(\lgtm,\lgtv,\rk) \ppslden{\eval{}}(\eval{\rk+1}|\lgtv) \ppslden{\evec{}}(\evec{\rk+1}|\lgtv) } \nonumber \\
+ & & \qquad\qquad\qquad\qquad\qquad \times \frac{P(\rk+1\to\rk)}{P(\rk\to\rk+1)}     ,
 \end{IEEEeqnarray}
@@
 %
-\begin{IEEEeqnarray}{rCl}
- \jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) &=& 2^{-1} \eval{\rk+1}^{\lsd-\rk-1} \prod_{i=1}^{\rk}\eval{i}^{-1} \prod_{i=1}^{\rk}(\eval{i}-\eval{\rk+1})     .
+\begin{IEEEeqnarray}{l}
+ \jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) \nonumber \\
+ \qquad\qquad = 2^{-1} \eval{\rk+1}^{\lsd-\rk-1} \prod_{i=1}^{\rk}\eval{i}^{-1} \prod_{i=1}^{\rk}(\eval{i}-\eval{\rk+1})     .
 \end{IEEEeqnarray}
 
-\section{Simulations}
+\section{Simulations} \label{sec:simulations}
 
@@
 %
-\begin{IEEEeqnarray}{rCl}
- \priordof &=& \lsd - 1 \\
- \priorscalematrix &=& \priordof \idmat \\
- \priormeanmatrix &=& \zmat \\
- \priorcolumnvariance &=& 100 \idmat      .
+\begin{IEEEeqnarray}{rClCrCl}
+ \priordof &=& \lsd - 1 & \qquad & \priorscalematrix &=& \priordof \idmat \nonumber \\
+ \priormeanmatrix &=& \zmat & \qquad & \priorcolumnvariance &=& 100 \idmat      .
 \end{IEEEeqnarray}
@@
 %
-\begin{IEEEeqnarray}{rCl}
- \priorscalematrixbase &=& \idmat \\
- \priormeanmatrix &=& \zmat \\
- \priorcolumnvariance &=& 100 \idmat \\
- \priortypval &=& 1     .
+\begin{IEEEeqnarray}{rClCrCl}
+ \priorscalematrixbase &=& \idmat & \qquad & \priormeanmatrix &=& \zmat \nonumber \\
+ \priorcolumnvariance &=& 100 \idmat & \qquad & \priortypval &=& 1 \nonumber \\
+ \priorrankweight{i} &=& \frac{1}{\lsd}     .
 \end{IEEEeqnarray}
@@
  \centering
- \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-basic-R-acf.pdf} \label{subf:toy-R-acf:basic}} \\
- \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-R-acf.pdf} \label{subf:toy-R-acf:degenerate}}
+ \subfloat[]{\includegraphics[width=0.45\columnwidth]{figures/toy-basic-R-acf.pdf} \label{subf:toy-R-acf:basic}}
+ \subfloat[]{\includegraphics[width=0.45\columnwidth]{figures/toy-degenerate-R-acf.pdf} \label{subf:toy-R-acf:degenerate}}
  \caption{MCMC autocorrelation plots for $\xi_{\ob{}}$ for the full rank transition model \protect\subref{subf:toy-R-acf:basic} and the degenerate transition model \protect\subref{subf:toy-R-acf:degenerate}.}
@@
 \begin{IEEEeqnarray}{rCl}
- \priorscalematrix &=& 0.001 \priordof \idmat \\
- \priorscalematrixbase &=& 0.001 \idmat \\
+ \priorscalematrix &=& 0.001 \priordof \idmat \nonumber \\
+ \priorscalematrixbase &=& 0.001 \idmat \nonumber \\
  \priortypval &=& 0.01     .
@@
 \bibliographystyle{IEEEtran}
-% \bibliography{/home/pete/Dropbox/PhD/bibliographies/Cleanbib}
-\bibliography{/home/pete/Dropbox/PhD/bibliographies/OTbib}
+\bibliography{/home/pete/Dropbox/PhD/bibliographies/Cleanbib}
+% \bibliography{/home/pete/Dropbox/PhD/bibliographies/OTbib}
 \end{document}

### 2015-09-22 analyse_toy.py

@@ -71,8 +71,10 @@ degenerate_learner.plot_chain_trace('R', dims=([0],[0]), trueModel=model)
 plt.gcf().axes[0].set_ylim([0.05,0.15])
 plt.gcf().savefig(figure_path+'toy-degenerate-R-trace.pdf', bbox_inches='tight')
 basic_learner.plot_chain_acf('R', dims=([0],[0]), nlags=8)
+plt.gcf().set_size_inches(4,3, forward=True)
 plt.gcf().savefig(figure_path+'toy-basic-R-acf.pdf', bbox_inches='tight')
 degenerate_learner.plot_chain_acf('R', dims=([0],[0]), nlags=8)
+plt.gcf().set_size_inches(4,3, forward=True)
 plt.gcf().savefig(figure_path+'toy-degenerate-R-acf.pdf', bbox_inches='tight')
 
 # Histograms

### 2015-09-22 notes/paper-degenerate-linear-systems.tex
Additional minor edits.

@@
 \begin{IEEEeqnarray}{rCl}
- \lgtv &=& \tvrot \tvfull \tvrot\tr %&=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix} \begin{bmatrix}\tvfull & \zmat \\ \zmat & \zmat\end{bmatrix} \begin{bmatrix}\tvrot\tr \\ \tvrotorth\tr\end{bmatrix} \label{eq:givens_decomposition}
+ \lgtv &=& \tvrot \tvfull \tvrot\tr \label{eq:givens_decomposition} %&=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix} \begin{bmatrix}\tvfull & \zmat \\ \zmat & \zmat\end{bmatrix} \begin{bmatrix}\tvrot\tr \\ \tvrotorth\tr\end{bmatrix}
 \end{IEEEeqnarray}

### 2015-09-22 analyse_toy.py

@@ -66,8 +66,10 @@ fig.savefig(figure_path+'toy-state.pdf', bbox_inches='tight')
 # Trace and acf plots
 basic_learner.plot_chain_trace('R', dims=([0],[0]), trueModel=model)
 plt.gcf().axes[0].set_ylim([0.05,0.15])
+plt.gcf().set_size_inches(8,4, forward=True)
 plt.gcf().savefig(figure_path+'toy-basic-R-trace.pdf', bbox_inches='tight')
 degenerate_learner.plot_chain_trace('R', dims=([0],[0]), trueModel=model)
+plt.gcf().set_size_inches(8,4, forward=True)
 plt.gcf().axes[0].set_ylim([0.05,0.15])
 plt.gcf().savefig(figure_path+'toy-degenerate-R-trace.pdf', bbox_inches='tight')
 basic_learner.plot_chain_acf('R', dims=([0],[0]), nlags=8)

### 2015-09-23 notes/paper-degenerate-linear-systems.tex
Proofread and final edit.

@@
 An efficient Gibbs sampling algorithm exists for the basic formulation of linear model.
-We extend this to the more challenging situation where the transition model is degenerate.
+We extend this to the more challenging situation where the transition model is possibly degenerate.
 Appropriate Markov kernels are devised and demonstrated with simulations.
@@
 An important consideration is how to learn these fixed matrices from observed data.
-The parametric approach to this problem is to deduce the form of the system matrices in terms of a small number of scalar parameters by considering the physical mechanisms underlying the system, and then attempt to learn these.
+Sometimes it is possible to deduce the form of the system matrices in terms of a small number of scalar parameters by considering the physical mechanisms underlying the system, and then attempt to learn these.
 (See e.g.
-\cite{Kantas2009,Andrieu2010}.) However, in many systems there may not be an obvious parametric form, and it may be better to treat each system matrix in its entirety as an unknown variable to be learned.
-This type of non-parametric learning for linear Gaussian state space models is the focus of this paper.
+\cite{Kantas2009,Andrieu2010}.) However, in many systems there may not be an obvious parametric form, and it may be better to treat each system matrix in its entirety as an unknown variable to be learned; this is the case studied in this paper.
 There is a large body of research on this topic, which has generally concentrated on point estimate methods.
@@
 This is achieved by appropriate factorisations of the covariance matrix which allow us still to exploit the fast-mixing Gibbs kernel.
-Furthermore, reversible jump construction \cite{Green1995,Green2009} is introduced which enables the sampler to learn the rank of the transition covariance matrix.
+In addition, reversible jump moves \cite{Green1995,Green2009} are introduced which enable the sampler to learn explicitly the rank of the transition covariance matrix.
 
@@
 In section~\ref{sec:degenerate_transition_models} we describe modifications for degenerate transitions, and in section~\ref{sec:rank_learning} we show how the rank of the transition covariance may be learned.
-Simulations on a toy model, and a motion capture application are presented in section~\ref{sec:simulations}.
+Simulations on a toy model, and a motion capture application are presented in section~\ref{sec:simulations} with discussion and conclusions in section~\ref{sec:conclusions}.
 
@@
 In this paper we focus on learning the transition model, i.e.
-$\lgtm$ and $\lgtv$, and treat $\lgom$ and $\lgov$ as fixed and known throughout.
+$\lgtm$ and $\lgtv$, and for the most part treat $\lgom$ and $\lgov$ as fixed and known.
 Conditional on these matrices, state inference may be carried out analytically.
@@
 Furthermore, the Kalman filter may also be used to evaluate the marginal likelihood $\den(\ob{1:\timax}|\lgtm, \lgtv)$, and it is possible to draw posterior samples from the state posterior $\den(\ls{1:\timax}|\ob{1:\timax}, \lgtm, \lgtv)$ using the forward-filtering-backward-sampling method \cite{Chib1996}.
-The purpose of a learning algorithm is estimate unknown values of $\left\{\lgtm, \lgtv\right\}$ from a sequence of observations.
+The purpose of a learning algorithm is to estimate unknown values of $\left\{\lgtm, \lgtv\right\}$ from a sequence of observations.
 Within the Bayesian framework this means calculating or approximating the posterior distribution $\den(\lgtm, \lgtv|\ob{1:\timax})$, which we can achieve using MCMC.
 
-In order to conduct Bayesian learning, we first need to define prior distributions for the unknown system matrices $\lgtm$ and $\lgtv$.
+For Bayesian learning, we need to define prior distributions for the unknown system matrices $\lgtm$ and $\lgtv$.
 Although these should be selected to reflect prior belief about the system in question, there is substantial benefit in using conjugate priors, as these enable us to use Gibbs sampling moves.
@@
 in which $\tn{\ti}$ as dimension $\rk$, and $\lgdm$ is $\lsd\times\rk$.
-This is almost equivalent to the previous, with $\lgtv=\lgdm\lgdm\tr$.
+This is almost equivalent to the previous case, with $\lgtv=\lgdm\lgdm\tr$.
 However, whereas before $\lgtv$ was required to be positive definite, this is no longer the case.
@@
 Degenerate models such as these arise in tracking applications (see \cite{Maskell2004} for a discussion of this phenomenon); when non-Markovian models, such as autoregressive process, are converted to state space form; and when it is natural to parameterise a system with more latent states than there are degrees of freedom (see section~\ref{sec:mocap}).
-In this section, we introduce a parameterisation for such degenerate transition models which can uniquely represent any such system, and show how MCMC moves may be implemented for learning.
+In this section, we introduce suitable parameterisations for such degenerate transition models, and show how MCMC moves may be implemented for learning.
 Here we treat $\rk$ as fixed and known; in the following section mechanisms will be introduced to learn this too.
@@
 The matrix $\lgdm$ is not uniquely identifiable from the state sequence, because the distribution of $\lgdm\tn{\ti}$ is identical to that of $\lgdm \Xi \tn{\ti}$ where $\Xi$ is an arbitrary orthogonal matrix.
-To avoid this ambiguity, it is better to focus on learning $\lgtv$ as before, and to define $\lgdm$ as some matrix square root of $\lgtv$ (e.g.
+This ambiguity does not affect $\lgtv$, since,
+%
+\begin{IEEEeqnarray}{rClCl}
+ (\lgdm \Xi)(\lgdm \Xi)\tr &=& \lgdm\lgdm\tr &=& \lgtv     .
+\end{IEEEeqnarray}
+%
+For this reason, we treat $\lgtv$ as fundamental, as before.
+$\lgdm$ may be defined to be any matrix square root of $\lgtv$ (e.g.
 the Cholesky factor or the principal matrix square root).
@@
 We would prefer to attain the automated, fast-mixing properties displayed by the Gibbs sampler on the full rank model.
-In this section we show how a Gibbs kernel may still be applied to sample a component of the parameters, with the remainder handled using Metropolis-Hastings.
+We will show how a Gibbs kernel may still be applied to sample a component of the parameters, with the remainder handled using Metropolis-Hastings.
 
@@
 
-We can also write the ``full'' factorisation by finding a set of $\lsd-\rk$ additional orthonormal vectors and appending them to $\tvvec$, along with matching zeros to $\tvval$, as follows,
+Note that we can find a set of $\lsd-\rk$ additional vectors orthonormal to $\tvvec$ using the Gram-Schmidt procedure.
+We denote the matrix of these vectors $\tvvecorth$.
+This matrix is not unique in general.
+
+A further factorisation will also be useful, which we will refer to as the Givens decomposition (on account of its method of computation),
 %
 \begin{IEEEeqnarray}{rCl}
- \lgtv &=& \begin{bmatrix}\tvvec & \tvvecorth \end{bmatrix} \begin{bmatrix}\tvval & \zmat \\ \zmat & \zmat \end{bmatrix} \begin{bmatrix}\tvvec\tr \\ \tvvecorth\tr \end{bmatrix}
+ \lgtv &=& \tvrot \tvfull \tvrot\tr     , \label{eq:givens_decomposition} %&=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix} \begin{bmatrix}\tvfull & \zmat \\ \zmat & \zmat\end{bmatrix} \begin{bmatrix}\tvrot\tr \\ \tvrotorth\tr\end{bmatrix}
 \end{IEEEeqnarray}
 %
-Note that $\tvvecorth$ is not unique.
-
-A further factorisation will also be useful, which we will refer to as the Givens decomposition (on account of its method of computation).
-For the rank-$\rk$ positive semi-definite covariance matrix $\lgtp$,
-%
-\begin{IEEEeqnarray}{rCl}
- \lgtv &=& \tvrot \tvfull \tvrot\tr \label{eq:givens_decomposition} %&=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix} \begin{bmatrix}\tvfull & \zmat \\ \zmat & \zmat\end{bmatrix} \begin{bmatrix}\tvrot\tr \\ \tvrotorth\tr\end{bmatrix}
-\end{IEEEeqnarray}
-%
-where,
-%
-$\tvfull$ is a $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors subject to certain constraints.
+where $\tvfull$ is an $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors subject to certain constraints.
 The transformation from $\lgtv$ to $\{\tvfull, \tvrot\}$ is also bijective.
@@
 
-An appropriate prior distribution for the degenerate transition covariance matrix may be defined by setting the degrees of freedom of an inverse Wishart distribution to the desired rank, $\priordof=\rk$.
+An appropriate prior distribution for the singular transition covariance matrix may be defined by setting the degrees of freedom of an inverse Wishart distribution equal to the desired rank, $\priordof=\rk$.
 This results in a singular inverse Wishart distribution over the rank-$\rk$ positive semi-definite matrices \cite{Diaz-Garcia2006},
@@
 \begin{IEEEeqnarray}{rCl}
- \lgtv|\rk &\sim& \iwishartdist{\rk}{\priorscalematrixrank{\rk}}     .
+ \lgtv|\rk &\sim& \iwishartdist{\rk}{\priorscalematrix}     .
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrixrank{\rk}}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \nonumber \\
- & & \qquad \times \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrixrank{\rk} \right)    ,
+ \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrix}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \nonumber \\
+ & & \quad \times \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrix \right)    ,
 \end{IEEEeqnarray}
@@
 in which $\lgtv\pinv$ denotes the Moore-Penrose pseudoinverse of $\lgtv$.
-A prior estimate of $\lgtv$ is $\frac{1}{\rk}\priorscalematrixrank{\rk}$.
-Thus, in order to maintain consistency between the different ranks, we should set $\priorscalematrixrank{\rk} = \rk \priorscalematrixbase$, where $\priorscalematrixbase$ is an estimate of $\lgtv$.
 
 The matrix normal component of the prior is more problematic.
-Although the distribution it is still well-defined, the fact that $\lgtv$ is singular means that the space of possible values is restricted to a subspace of $\reals^{\lsd\times\lsd}$, placing an unwanted constraint on our model.
+Although the distribution is still well-defined, the fact that $\lgtv$ is singular means that the space of possible values is restricted to a subspace of $\reals^{\lsd\times\lsd}$, placing an unwanted constraint on our model.
 (See \cite{Diaz-Garcia2006} for a discussion of the singular matrix normal distribution.)
@@
 These extra eigenvalues control the rate at which probability decays as the matrix moves away from the subspace defined by $\lgtv$.
-A simple choice is to set $\priorextratpval=\priortypval\idmat$ where $\priortypval$ is a prior estimate of a ``typical'' eigenvalue.
+Since $\tvvecorth$ can be arbitrarily rotated, it will generally be most suitable to use $\priorextratpval=\priortypval\idmat$ where $\priortypval$ is a prior estimate of a ``typical'' eigenvalue.
 Alternatively, $\priortypval$ could be made dependent on $\tvval$; for example it could be set to the average (arithmetic or geometric) of the current eigenvalues.
@@
 As discussed, standard Gibbs sampling using $\postden(\lgtm, \lgtv|\ls{1:\timax})$ does not allow the Markov chain to fully explore the parameter space because of the constraint that each transition $(\ls{\ti}-\lgtm\ls{\ti-1})$ must lie in the $\rk$-dimensional column space of $\lgdm$.
-Nevertheless, we can sample from these distributions exactly, and doing so allows the chain to efficiently explore values of the transition matrix and transition covariance within the permitted regions.
+Nevertheless, we can sample from this distribution exactly, and doing so allows the chain to efficiently explore values of the transition matrix and transition covariance within the permitted regions.
 These constrained Gibbs sampling steps rely on the Givens decomposition \eqref{eq:givens_decomposition}.
@@
 \begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm, \lgtv) = \den(\ls{0}) \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) } \nonumber \\
+ \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm, \lgtv) \propto \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) } \nonumber \\
  &\propto& \determ{\tvfull}^{-\half(\timax-1)} \exp\Bigg( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \nonumber \\
@@
 %
-where $Z \in \reals^{\rk\times\lsd}$ and 
+where $Z \in \reals^{\rk\times(\timax-1)}$ and 
 %
@@
 
-\section{Reversible Jump for Rank Learning} \label{sec:rank_learning}
+\section{Unknown Covariance Matrix Rank} \label{sec:rank_learning}
 
@@
 
-First we introduce a prior distribution for the rank of $\lgtv$,
+\subsection{Priors with Unknown Rank}
+
+We must take care in setting $\priorscalematrix$ in $\den(\lgtv)$.
+In general, the scale of an inverse Wishart random variable depends on both the scale matrix and the degrees of freedom.
+If the same $\priorscalematrix$ were used for all $\rk$ then the prior would favour different scales with different possible values of the rank.
+In order to make the prior scale invariant across $\rk$, we replace $\priorscalematrix$ with $\priorscalematrixrank{\rk} = \rk \priorscalematrixbase$.
+$\priorscalematrixbase$ sets the scale of $\lgtv$ globally.
+
+We also introduce a prior distribution for the rank of $\lgtv$,
 %
@@
 \end{IEEEeqnarray}
+
+\subsection{Reversible Jump for Rank Learning}
 
@@
  \ppslden{\evec{}}(\evec{\rk+1}|\lgtv) = \begin{cases}
-                                          \frac{1}{C} & \evec{\rk+1} \in \evecset{\rk+1} \\
+                                          \frac{1}{C_{\rk+1}} & \evec{\rk+1} \in \evecset{\rk+1} \\
                                           0 & \text{ otherwise}    .
@@
 %
-$C$ is the volume of the manifold, which is half of a $(d-r-1)$-sphere (half because of the requirement that the sign of the first element be positive) \cite{Muirhead1982}. Hence,
+$C_{\rk+1}$ is the volume of $\evecset{\rk+1}$, which is half of a $(d-r-1)$-sphere (half because of the requirement that the sign of the first element be positive) \cite{Muirhead1982}. Hence,
 %
 \begin{IEEEeqnarray}{rCl}
- C &=& \frac{\pi^{\half(\lsd-\rk)}}{\gammafun\left(\frac{\lsd-\rk}{2}\right)}     .
+ C_{\rk+1} &=& \frac{\pi^{\half(\lsd-\rk)}}{\gammafun\left(\frac{\lsd-\rk}{2}\right)}     .
 \end{IEEEeqnarray}
 %
-In order to generate such a vector, first draw a $(\lsd-\rk)$-dimensional sample from $\normaldist{\zmat}{\idmat}$, normalise it, and then multiply by $\tvvecorth$, which is an arbitrary matrix of orthonormal columns such that $\tvvec\tr\tvvecorth=\zmat$.
+In order to generate such a vector, first draw a $(\lsd-\rk)$-dimensional sample from $\normaldist{\zmat}{\idmat}$, normalise it, and then multiply by $\tvvecorth$.
 
@@
 %
-where $\jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew)$ is the Jacobian of the transformation and $P(\rk+1\to\rk)$ is the probability of performing a move which changes the rank from $\rk$ to $\rk+1$.
+where $\jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew)$ is the Jacobian of the transformation and $P(\rk\to\rk+1)$ is the probability of performing a move which changes the rank from $\rk$ to $\rk+1$.
 
@@
 Each kernel has a single scalar algorithm parameter which determines the size of the proposed moves, $\sigma_Q^2$ in \eqref{eq:skewsymmetric_proposal} and $\sigma_F^2$ in \eqref{eq:transition_matrix_proposal}.
-In order to accelerate mixing, these are adjusted adaptively based on the acceptance rate using the method introduced by \cite{Roberts2009}.
+In order to accelerate mixing, and reduce the need for algorithm tuning, these are adjusted adaptively based on the acceptance rate using the method introduced by \cite{Roberts2009}.
 
@@
  \item Autocorrelation plots show that the sampler mixes more slowly than that learning a full rank covariance model. See figure~\ref{fig:toy-R-acf}.
- \item The parameter posterior histograms appear consistent with the true values. See figure~\ref{fig:toy-FQ-hist}.
+ \item The parameter posterior histograms are consistent with the true values. See figure~\ref{fig:toy-FQ-hist}.
  \item The state trajectory is estimated more accurately with a degenerate covariance, with an RMSE of $5.51$, compared to $6.01$ for the full rank model.
@@
  \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-R-trace.pdf} \label{subf:toy-R-trace:degenerate}}
- \caption{MCMC trace plots for $\xi_{\ob{}}$ for the full rank transition model \protect\subref{subf:toy-R-trace:basic} and the degenerate transition model \protect\subref{subf:toy-R-trace:degenerate}.}
+ \caption{MCMC trace plots for $\xi_{\ob{}}$ for \protect\subref{subf:toy-R-trace:basic} the full rank transition model and \protect\subref{subf:toy-R-trace:degenerate} the degenerate transition model.}
  \label{fig:toy-R-trace}
@@
  \subfloat[]{\includegraphics[width=0.45\columnwidth]{figures/toy-degenerate-R-acf.pdf} \label{subf:toy-R-acf:degenerate}}
- \caption{MCMC autocorrelation plots for $\xi_{\ob{}}$ for the full rank transition model \protect\subref{subf:toy-R-acf:basic} and the degenerate transition model \protect\subref{subf:toy-R-acf:degenerate}.}
+ \caption{MCMC autocorrelation plots for $\xi_{\ob{}}$ for \protect\subref{subf:toy-R-acf:basic} the full rank transition model and \protect\subref{subf:toy-R-acf:degenerate} the degenerate transition model.}
  \label{fig:toy-R-acf}
@@
 
-The hyperparameters are set to the same uninformative values as for the toy model, with the following exception,
+The hyperparameters are set to the same uninformative values as for the toy model, with the following exceptions,
 %
@@
 
-\section{Discussion and Conclusions}
+\section{Discussion and Conclusions} \label{sec:conclusions}
 
@@
 
-The standard linear Gaussian state space models, with $\rank(\lgtv)=\lsd$, discussed in section~\ref{sec:linear_gaussian_models} are a special case of the more general class used thereafter, for which $\rank(\lgtv)$ was treated as unknown.
+The standard linear Gaussian state space model, with $\rank(\lgtv)=\lsd$, discussed in section~\ref{sec:linear_gaussian_models} is a special case of the more general class with $\rank(\lgtv)$ unknown.
 Correspondingly, the MCMC algorithm for the general case reduces to the standard Gibbs sampler of section~\ref{sec:linear_gaussian_models} if we re-impose the constraint that $\rank(\lgtv)=\lsd$.
@@
 where $\tvsign$ is a diagonal matrix in which the diagonal elements are $\pm1$.
-This factorisation may be derived (and also calculated) by iteratively left multiplying the original matrix by a Givens rotation matrix such that one element is set to $0$.
+This factorisation may be derived (and also calculated) by iteratively left multiplying the original matrix by a (transposed) Givens rotation matrix such that one element is set to $0$.
 The matrix remaining once all the elements below the diagonal have been eliminated is $\tvsign$, and the factorisation above results from a straightforward rearrangement.
@@
 Then left multiply by Givens matrices to zero all the elements on rows $r+1$ to $d$.
-The resulting factorisation is,
+After rearranging, the resulting factorisation is,
 %
@@
 Factorising the matrix of eigenvectors using the Givens method, we find,
-
+%
 \begin{IEEEeqnarray}{rCl}

### 2015-09-23 run_mocap_mcmc.py

@@ -38,7 +38,7 @@ np.random.seed(0)
 data_path = './mocap-data/'
 test_data_file = './results/mocap-test-data.p'
 
-model_type = 'naive'#'degenerate_augmented'#'degenerate'#'naive'#'basic'#
+model_type = 'degenerate'#'degenerate_augmented'#'degenerate'#'naive'#'basic'#
 num_iter = 20000
 
 num_burn = 10000#int(num_iter-10000)

### 2015-09-23 notes/paper-degenerate-linear-systems.tex
Added impact statement

@@
 An efficient Gibbs sampling algorithm exists for the basic formulation of linear model.
-We extend this to the more challenging situation where the transition model is possibly degenerate.
+We extend this to the more challenging situation where the transition model is possibly degenerate, i.e.
+the transition covariance matrix is singular.
 Appropriate Markov kernels are devised and demonstrated with simulations.
@@
 
-The Gibbs sampler approach developed in \cite{Wills2012} is effective on the basic flavour of linear Gaussian state space models.
-However, it is not able to handle the case in which the transition model is degenerate, i.e.
-when the transition covariance matrix is not full rank.
+In some systems, the components of the latent state are not free to vary independently of each other.
+Such systems are best modelled by a degenerate transition model, in which the transition covariance matrix is singular.
+The Gibbs sampler of \cite{Wills2012} is not able to handle degenerate models.
 In this paper we extend the MCMC learning framework to encompass this scenario.
@@
 In addition, reversible jump moves \cite{Green1995,Green2009} are introduced which enable the sampler to learn explicitly the rank of the transition covariance matrix.
+The new algorithm is important because it allows us to perform Bayesian learning with a broader class of linear state space models, leading to better descriptions of real systems and improved estimation performance.
 

### 2016-01-08 notes/reviews/response_to_reviews.tex
Added reviews and template response

\documentclass{article}

% Formatting
\usepackage{geometry}\setlength\textwidth{6in}\setlength\textheight{8.5in}
\usepackage[usenames,dvipsnames]{color}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bbm}

\usepackage[scaled]{helvet}
\renewcommand*\familydefault{\sfdefault} %% Only if the base font of the document is to be sans serif
\usepackage[T1]{fontenc}

% Environments
\usepackage{IEEEtrantools}
\usepackage{algorithm}
\usepackage{algorithmic}

% References
\usepackage{natbib}%\def\bibfont{\footnotesize}

\newenvironment{review}[0]{\begin{itshape}\color{Gray}\noindent}{\end{itshape}\vspace{0.4cm}}
\newenvironment{response}[0]{\noindent}{\vspace{0.4cm}}
\newcommand{\meta}[1]{{\color{red}\em #1}}

%opening
\title{Bayesian Learning of Degenerate Linear~Gaussian~State~Space~Models using~Markov~Chain~Monte~Carlo \\ Response to Reviews}
\date{January 2016}

\begin{document}

\maketitle

Thank you for taking the time to review our paper. Here is our responses to the points raised.

\meta{
\begin{itemize}
 \item Reviewer 1
 \item Reviewer 2
 \item Reviewer 3
 \item Proof read
\end{itemize}

}

\section*{Response to Reviewer 1}

\begin{review}
The paper presents a Bayesian learning method based on Markov Chain Monte Carlo for linear Gaussian state space models. The model may have singular transition covariance matrix. The objective is to infer the unknown states and learn the model parameters given the sequence of observations. Gibbs sampling as well as Metropolis-Hastings methods are utilized.

The reviewer has the following comments:

A main concern is that the paper is not easy to follow. Many statements are mentioned without sufficient or clear justifications.
\end{review}

\begin{response}
 \meta{Contradicts other reivewer.}
\end{response}

\begin{review}
Such as: in Section II.B: ``The appropriate posterior distribution is not amenable to efficient MCMC sampling,...''.
\end{review}

\begin{response}
 \meta{Fix this}
\end{response}

\begin{review}
Can the model be extended to the case where the model parameters $H$ and $R$ in (6) are not fixed? Please add discussions.
\end{review}

\begin{response}
 \meta{Add discussion of identifiability.}
\end{response}

\begin{review}
It will be helpful to provide some relevant applications in Section I, explaining why a degenerate model will be more appropriate.
\end{review}

\begin{response}
 \meta{Add this discussion and examples}
\end{response}

\begin{review}
There is no complexity analysis.
\end{review}

\begin{response}
 \meta{Add this}
\end{response}

\begin{review}
The advantages over existing methods, such as those methods that use product of marginal distributions or other more recent approaches, have not been adequately demonstrated.
\end{review}

\begin{response}
 \meta{Is there such a thing? Easily conceived. Too much work to derive and implement.}
\end{response}

\begin{review}
The presentation needs to be improved:
(i) This includes correcting typos, such as: in abstract ``it is possible deduce''
\end{review}

\begin{response}
 \meta{fix this}
\end{response}

\begin{review}
(ii) Defining notations before/when using them, such as F and H in equations (1) and (2)
\end{review}

\begin{response}
 \meta{and this}
\end{response}

\begin{review}
Overall, the above concerns need to be addressed before a decision can be made.
\end{review}

\section*{Response to Reviewer 2}

\begin{review}
This manuscript studies the problem of Gibbs sampling in the estimation of degenerate linear Gaussian state space model, where the covariance matrix Q of noise in state transition function is singular. This manuscript proposes to decompose matrix Q and convert the problem to a general case conditioned on the known rank $r$ of $Q$. In order to estimate rank $r$, reversible jump MCMC is employed. There are several concerns as follows:

1.  In the manuscript, the authors assume a linear model in which the observation is only determined by the current state. However, in practice, sophisticated models where the observation is also related to the previous states and inputs are more commonly used. So this may limit the applicability of the proposed method.
\end{review}

\begin{response}
 \meta{Bullshit}
\end{response}

\begin{review}
2.  It is suggested to introduce some other related works on degenerate linear Gaussian state space model if any.
\end{review}

\begin{response}
 \meta{There are none.}
\end{response}

\begin{review}
3.  The $H$ and $R$ matrices are assumed to be known in the manuscript, which may not be the case in practice. It is suggested to consider the estimation of $H$ and $R$ in the degenerate model.
\end{review}

\begin{response}
 \meta{identifiability problems.}
\end{response}

\begin{review}
4.  This manuscript uses Givens decomposition to factorize matrix $Q$, and in section III B, it states that matrix $U$ in Givens decomposition is uniquely determined. It should be explained that how to determine matrix $U$ and the necessity of Givens decomposition.
\end{review}

\begin{response}
 \meta{Add more discussion of the Givens decomposition and add algorithm for generating it.}
\end{response}

\begin{review}
5.   The first step of Gibbs sampling in linear state space model is the estimation of states $x$ from observation $y$. It should be clarified that whether the singularity of matrix $Q$ will have any effect on this estimation. 
\end{review}

\begin{response}
 \meta{Detail Kalman smoothing with degenerate $Q$.}
\end{response}

\begin{review}
6.  It should be explained that how the probabilities of $P(r \to r+1)$ and $P(r+1 \to r)$ in Eq.(59) are determined in simulations.
\end{review}

\begin{response}
 \meta{Trivial}
\end{response}

\begin{review}
7.  In simulations, the Metropolis-Hastings step should also be applied to the full rank linear model if not. 
\end{review}

\begin{response}
 \meta{What? Could try the unknown rank algorithm on a full rank case to show it works.}
\end{response}

\begin{review}
8.  In eq. (24), function $exp()$ should be replaced by $exp(tr())$. In line 38 page 10, the first 'is' should be deleted.
\end{review}

\begin{response}
 \meta{Fix these.}
\end{response}

\begin{review}
I suggest the authors to address these problems and to submit it to IEEE Signal Processing Letters or do a resubmission.
\end{review}

\begin{response}
 \meta{Not appropriate for letters.}
\end{response}

\section*{Response to Reviewer 3}

\begin{review}
In this paper, the authors address the Bayesian learning of degenerate linear and Gaussian state space models. In this specific context, classical Gibbs sampler can not be used and therefore the authors propose an interesting and complete framework based on Markov Chain Monte Carlo (MCMC) to deal with this challenging problem. The paper is well written and organized. I have the following remarks:

Even if I found the paper well organized, I think that the authors should summarize the complete framework (RJ-MCMC) in an algorithm - in my opinion it will help the reader to understand it by summarizing all the steps of the proposed scheme in one place.
\end{review}

\begin{response}
 \meta{Do this. At least, list the steps of each kernel.}
\end{response}

\begin{review}
In Section III-C, the authors propose to use a collapsed Gibbs move to more efficiently exploit the target on the space of $F$ and $Q$, alternatively. I think the authors should discuss in more details why such collapsed Gibbs is important in the text. Moreover, the importance of such steps could be easily illustrated in the results section so the reader can appreciate the benefit of using such a strategy.
\end{review}

\begin{response}
 \meta{Good idea. Do this. Or perhaps just reference the paper more.}
\end{response}

\begin{review}
My main concern is regarding the simulation section.

Firstly, the comments are just a brief summary of results from the figures without any comments or discussions on why such results are obtained.
\end{review}

\begin{response}
 \meta{Add more discussion and analysis}
\end{response}

\begin{review}
There are no figures that show the evolution of the Markov chain OR the posterior approximation of the unknown rank $r$
The authors wrote that ``the sampler identifies the true rank.... and remains at this value thereafter'' but this can be worrying - this statement could also mean that the sampler is not mixing properly for this variable $r$. The author should be more precised by saying ``Trace plots suggest that the sampler converges within a few thousand iterations. See figure 2'' - The convergence showed in this figure is only related to the variable $\xi_y$.
\end{review}

\begin{response}
 \meta{Explain this and improve in the paper. Add simple example of changing rank to show that it is not stuck.}
\end{response}

\begin{review}
The authors could illustrate in the first toy example the benefit of the collapsed Gibbs
moves by showing the performance of the sampler with vs without this strategy.
\end{review}

\begin{response}
 \meta{Maybe...}
\end{response}

\begin{review}
I personally think that the proposed algorithm should be illustrated with at least a different set of parameters for the toy example (e.g. different values for $\xi_y$ thus allowing to have a curve of RMSE vs $\xi_y$ )
\end{review}

\begin{response}
 \meta{Ok... if it will make you happy. Obviously it will work for some parameters and not others.}
\end{response}

\begin{review}
List of minor remarks/typos:
\begin{itemize}
 \item Just after Eq. (16), the authors should be more precise on the following sentence to avoid any confusion: ``However, whereas before $Q$ was required to be positive definite, this is no longer the case'' (I would add) for unknown $G$.
 \item Please correct in Section V first sentence ... known a prior. by ... known a priori.
\end{itemize}
\end{review}

\begin{response}
 \meta{Fix all these.}
\end{response}

\end{document}

### 2016-01-08 notes/reviews/response_to_reviews.tex
Started changes and response to reviewers.

@@
 % Formatting
+\usepackage[parfill]{parskip}
 \usepackage{geometry}\setlength\textwidth{6in}\setlength\textheight{8.5in}
@@
  \item Reviewer 3
- \item Proof read
+ \item Proof read, aiming to (i) clarify (2) correct typos (c) define everything when first used
 \end{itemize}
@@
 
-A main concern is that the paper is not easy to follow. Many statements are mentioned without sufficient or clear justifications.
-\end{review}
-
-\begin{response}
- \meta{Contradicts other reivewer.}
-\end{response}
-
-\begin{review}
+A main concern is that the paper is not easy to follow.
+Many statements are mentioned without sufficient or clear justifications.
 Such as: in Section II.B: ``The appropriate posterior distribution is not amenable to efficient MCMC sampling,...''.
@@
 \begin{response}
- \meta{Fix this}
+ We apologise for this. We have looked through and tried to clarify some of what might be the more abstruse portions. In particular, we have re-written that sentence.
 \end{response}
@@
 \begin{response}
- \meta{Add this discussion and examples}
+We have modified the introduction to mention the example of tracking multiple points on a rigid body.
 \end{response}
@@
 \begin{response}
- \meta{Add this}
+We have added a brief discussion of the complexity (which is the same as standard Gibbs sampler for the full rank model).
 \end{response}
@@
 \begin{response}
- \meta{fix this}
+ We have corrected that typo and a few others which we found.
 \end{response}
@@
 \begin{review}
-(ii) Defining notations before/when using them, such as F and H in equations (1) and (2)
+(ii) Defining notations before/when using them, such as $F$ and $H$ in equations (1) and (2)
 \end{review}
@@
 \begin{response}
- \meta{and this}
+ We have added or moved a number of definitions, including those for $F$ and $H$.
 \end{response}
-
-\begin{review}
-Overall, the above concerns need to be addressed before a decision can be made.
-\end{review}
 

### 2016-01-08 notes/paper-degenerate-linear-systems.tex
Fixed typo in matrix normal distribution
Started changes and response to reviewers.

@@
 This can be achieved using Markov chain Monte Carlo.
-On some occasions it is possible deduce the form of the unknown system matrices in terms of a small number of scalar parameters, by considering the underlying physical processes involved.
+On some occasions it is possible to deduce the form of the unknown system matrices in terms of a small number of scalar parameters, by considering the underlying physical processes involved.
 Here we study the case where this is not possible, and the entire matrices must be treated as unknowns.
@@
 (See e.g.
-\cite{Kantas2009,Andrieu2010}.) However, in many systems there may not be an obvious parametric form, and it may be better to treat each system matrix in its entirety as an unknown variable to be learned; this is the case studied in this paper.
+\cite{Kantas2009,Andrieu2010}.) However, in many systems there may not be such an obvious parameterisation, and it may be better to treat each system matrix in its entirety as an unknown variable to be learned; this is the case studied in this paper.
 There is a large body of research on this topic, which has generally concentrated on point estimate methods.
@@
 
-In some systems, the components of the latent state are not free to vary independently of each other.
+In some systems, the components of the latent state may not be free to vary independently of each other.
 Such systems are best modelled by a degenerate transition model, in which the transition covariance matrix is singular.
+This is the case, for example, when the latent state comprises the positions and velocities of multiple points on the same rigid body, meaning that their motion is deterministically coupled.
 The Gibbs sampler of \cite{Wills2012} is not able to handle degenerate models.
@@
 \section{Basic Linear Gaussian Models} \label{sec:linear_gaussian_models}
+
+\subsection{Definition}
 State space models are used to represent time-varying systems, and consist of a latent state process $\{\ls{\ti}\}_{\ti=1:\timax}$ and a related observation process $\{\ob{\ti}\}_{\ti=1:\timax}$.
@@
 %
-where $\ls{\ti} \in \reals^{\lsd}$ and $\ob{\ti} \in \reals^{\obd}$. In addition, we assume the state disturbance and observation noise variables have a Gaussian distribution,
+where $\ls{\ti} \in \reals^{\lsd}$, $\ob{\ti} \in \reals^{\obd}$.
+$\lgtm \in \reals^{\lsd\times\lsd}$,$\lgom \in \reals^{\obd\times\lsd}$ are the transition and observation matrices.
+$\tn{\ti} \in \reals^{\lsd}$ and $\on{\ti} \in \reals^{\obd}$ are the state disturbance and observation noise variables and we additionally assume that these are drawn from a Gaussian distribution,
 %
@@
 
+\subsection{Identifiability}
 The system is parameterised by the four matrices $\left\{\lgtm, \lgom, \lgtv, \lgov\right\}$.
@@
 Conditional on these matrices, state inference may be carried out analytically.
+
+\subsection{Inference}
 Posterior filtering and smoothing densities may be calculated using the Kalman filter \cite{Kalman1960} and Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965}.
@@
 \\
-\den(\lgtm | \lgtv) &=& \determ{2 \pi \lgtv}^{-\half} \determ{2 \pi \priorcolumnvariance}^{-\half} \nonumber \\
+\den(\lgtm | \lgtv) &=& (2\pi)^{-\frac{\lsd^2}{2}} \determ{\lgtv}^{-\frac{\lsd}{2}} \determ{\priorcolumnvariance}^{-\frac{\lsd}{2}} \nonumber \\
 & & \times \exp\left(-\half \trace\left[ \lgtv\inv (\lgtm-\priormeanmatrix) \priorcolumnvariance\inv (\lgtm-\priormeanmatrix)\tr \right] \right) \nonumber \\
@@
 Our principal interest is to learn the system matrices $\lgtm$ and $\lgtv$ from a sequence of observations $\ob{1:\timax}$.
-The appropriate posterior distribution is not amenable to efficient MCMC sampling, so instead we introduce the latent state sequence $\ls{1:\timax}$ as a nuisance variable and target the joint posterior,
+The posterior distribution over these variables $\den(\lgtm, \lgtv | \ob{1:\timax})$ is not amenable to MCMC sampling because there is no analytic expression for the density.
+Instead we introduce the latent state sequence $\ls{1:\timax}$ as a nuisance variable and target the joint posterior,
 %
@@
 %
-To sample from $\mhrotppsl$, we draw $\half\lsd(\lsd-1)$ independent scalar random variables $\{s_{i,j}\}_{0<i<j<\lsd}$ from any zero-mean distribution; a nice choice is,
+To sample from $\mhrotppsl$, we draw $\half\lsd(\lsd-1)$ independent scalar random variables $\{s_{i,j}\}_{0<i<j<\lsd}$ from any zero-mean distribution; a suitable choice is,
 %
@@
 \end{IEEEeqnarray}
+
+\section{The Complete Algorithm}
+
+\meta{Pseudo-code here}
+
+The computational complexity of the algorithm is $\bigo{\lsd^3 \timax}$, the same as the standard Gibbs sampler for the full rank model.
+For moderate to large $\timax$, the limiting factor is the forward-filtering-backward-sampling steps used to draw from $\postden(\ls{1:\timax}|\lgtm, \lgtv)$.
 

### 2016-01-12 notes/reviews/response_to_reviews.tex
Done corrections for reviewer 1

@@
 \begin{itemize}
- \item Reviewer 1
  \item Reviewer 2
@@
 \begin{response}
- \meta{Add discussion of identifiability.}
+ It is straightforward to learn $R$ within the MCMC scheme, and indeed the existing simulations already do precisely this.
+We have left it out of the algorithm sections to avoid cluttering the presentation.
+If $H$ were also unknown then there would be problems with identifiability.
+Most learning algorithms tend to struggle in this case.
+Fixing $H$ is a design choice, with the appealing property that the components of the latent state can have a meaningful physical interpretation.
+We have added a discussion of these issues in section II.
 \end{response}
@@
 \begin{response}
- \meta{Is there such a thing? Easily conceived. Too much work to derive and implement.}
+ As far as we are aware, there are not yet any published algorithms for learning degenerate linear state space models.
+It would certainly be an interesting project to extend the variational approaches (such as that of Barber \& Chiappa, 2007) to the degenerate case.
+However, the modifications would be more than trivial, and we feel this would be better suited to some later publication.
 \end{response}
@@
 \begin{response}
- \meta{There are none.}
+ We have added a few references in the introduction to papers which use degenerate linear Gaussian transition models.
+As far as we are aware, there are not yet any published algorithms for learning degenerate linear state space models.
 \end{response}
@@
 \begin{response}
- \meta{identifiability problems.}
+ See response to the similar comment by reviewer 1.
 \end{response}
@@
 \begin{response}
- \meta{Detail Kalman smoothing with degenerate $Q$.}
+ \meta{No effect provided $P_1$ and $F$ not singular. Add this to paper.}
 \end{response}
@@
 \begin{response}
- \meta{Trivial}
+ These two probabilities are design parameters. We have clarified this and added their values to the simulations section.
 \end{response}
@@
 \begin{response}
- \meta{What? Could try the unknown rank algorithm on a full rank case to show it works.}
+ \meta{What? Does this mean include the random walk steps? Is that useful?}
 \end{response}
@@
 \begin{response}
- \meta{Fix these.}
+ We have corrected these two typos.
 \end{response}

### 2016-01-12 notes/paper-degenerate-linear-systems.tex
Done corrections for reviewer 1

@@
 This is the case, for example, when the latent state comprises the positions and velocities of multiple points on the same rigid body, meaning that their motion is deterministically coupled.
+\cite{Maskell2004,Gustafsson2002} consider the use of degenerate transition models for various tracking applications.
 The Gibbs sampler of \cite{Wills2012} is not able to handle degenerate models.
@@
 \subsection{Identifiability}
-The system is parameterised by the four matrices $\left\{\lgtm, \lgom, \lgtv, \lgov\right\}$.
-In this paper we focus on learning the transition model, i.e.
-$\lgtm$ and $\lgtv$, and for the most part treat $\lgom$ and $\lgov$ as fixed and known.
-Conditional on these matrices, state inference may be carried out analytically.
+The system is parameterised by four matrices: $\lgtm, \lgom, \lgtv, \lgov$.
+In some applications, it may be appropriate to treat all of these as completely unknown.
+However, this scenario can be troublesome for learning algorithms, because the system is not identifiable -- we obtain the same likelihood if the latent state is multiplied by an arbitrary invertible matrix, along with corresponding multiplications of $\lgtm$, $\lgtv$ and $\lgom$.
+(See e.g.
+\cite{Roweis1999}.) To resolve the parameter ambiguity, it is necessary to constrain the values of one or more of the system matrices.
+One option (that taken in \cite{Roweis1999}) is to set $\lgtv$ to the identity matrix.
+This entails that the nature of the latent states and their relationship to the observations are entirely unknown a priori.
+In this paper we consider a contrasting approach, where we resolve the ambiguity by specifying the nature of the latent states and setting $\lgom$ to a pre-determined value.
+This approach has the advantage that the resulting model is more easily interpreted, since we can choose the latent states to be physically meaningful quantities.
+For example, in the motion capture application studied in section~\ref{sec:simulations}, we choose the latent states to be the positions and velocities of the markers.
+
+For the rest of the paper we will focus on learning the matrices of the transition model, i.e.
+$\lgtm$ and $\lgtv$.
+It is straightforward include learning of the observation covariance matrix $\lgov$ by adding sampling steps to the MCMC scheme which draw from the appropriate conditional distribution, and this is demonstrated in the simulations in section~\ref{sec:simulations}.
 
 \subsection{Inference}
+Conditional on the system matrices, state inference may be carried out analytically.
 Posterior filtering and smoothing densities may be calculated using the Kalman filter \cite{Kalman1960} and Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965}.
@@
  \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrix}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \nonumber \\
- & & \quad \times \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrix \right)    ,
+ & & \times \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \trace\left[ \lgtv\pinv \priorscalematrix \right] \right)    ,
 \end{IEEEeqnarray}
@@
 %
-The subspace over which this is defined is,
+This is defined over the subspace,
 %
@@
 %
-Notice that this is Inverse Wishart distribution is no longer singular, since $\tvfull$ is $\rk\times\rk$.
+Notice that this Inverse Wishart distribution is no longer singular, since $\tvfull$ is $\rk\times\rk$.
 
@@
 %
-where $\jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew)$ is the Jacobian of the transformation and $P(\rk\to\rk+1)$ is the probability of performing a move which changes the rank from $\rk$ to $\rk+1$.
+where $\jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew)$ is the Jacobian of the transformation and $P(\rk\to\rk+1)$ is the chosen probability at each step of performing a move which changes the rank from $\rk$ to $\rk+1$.
 
@@
 \end{IEEEeqnarray}
+%
+The move probabilities were set to,
+%
+\begin{IEEEeqnarray}{rCl}
+ P(\rk \to\rk+1) = P(\rk\to\rk+1) = \half     .
+\end{IEEEeqnarray}
+\meta{CHECK THIS!}
 

### 2016-01-13 notes/reviews/response_to_reviews.tex
More corrections: ffbs, and other small bits

@@
 \begin{itemize}
+ \item different parameter settings for toy model
+ \item change the rank after burn in to show not stuck
+ \item plots
+ \item collapsed Gibbs
+ \item pseudo-code algorithm
+ \item Givens
  \item Reviewer 2
@@
 \begin{response}
-We have added a brief discussion of the complexity (which is the same as standard Gibbs sampler for the full rank model).
+There is a brief discussion of the computational complexity in the final section of the paper.
+There is not much to say -- it is of the same order for the full rank and degenerate models.
+We do not see much value in including a detailed operation count, since the dominating factor will be the difference in autocorrelation and the resulting difference in the necessary length of the chain, which will vary by model and data set.
 \end{response}
@@
 \begin{response}
- \meta{Bullshit}
+ There is a very large body of literature focussed on models where the observation depends only on the current latent state, including most of the papers we have cited.
+It is usually possible to convert such a non-Markovian model into a Markovian one by augmenting the latent state with additional components.
+We disagree that this should be considered a limitation of the proposed method.
 \end{response}
@@
 \begin{response}
- \meta{No effect provided $P_1$ and $F$ not singular. Add this to paper.}
+ The procedure for sampling the state trajectories is unchanged by the singularity of $Q$.
+We have noted this in the text and also added a short appendix discussing this point and providing the equations for the forward-filtering-backward-sampling algorithm.
 \end{response}
@@
 \begin{response}
- \meta{What? Does this mean include the random walk steps? Is that useful?}
+This comment is not entirely clear to us.
+We could indeed apply the random walk MH steps to the full rank model, but what would be the point?
+These steps are essential only in the degenerate case to ensure the reducibility of the Markov chain.
+The mixing due to the MH steps is much slower than that due to the Gibbs steps, so including them in the full rank algorithm has a negligible effect on the autocorrelation.
 \end{response}
@@
 \begin{response}
- \meta{Not appropriate for letters.}
+ We considered submitting this as a letters paper but we do not believe it meets, or could be made to meet, the requirements for either length or content.
 \end{response}

### 2016-01-13 notes/paper-degenerate-linear-systems.tex
More corrections: ffbs, and other small bits

@@
 The state conditional $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ can be sampled using the forward-filtering-backward-sampling algorithm \cite{Chib1996,Wills2012}, which works by first running a Kalman filter forwards through the data, then running backwards in time sampling each state conditional on those in the future.
+See appendix~\ref{app:state-sampling}.
 The parameter conditional may also be sampled directly.
@@
 
-As discussed, standard Gibbs sampling using $\postden(\lgtm, \lgtv|\ls{1:\timax})$ does not allow the Markov chain to fully explore the parameter space because of the constraint that each transition $(\ls{\ti}-\lgtm\ls{\ti-1})$ must lie in the $\rk$-dimensional column space of $\lgdm$.
-Nevertheless, we can sample from this distribution exactly, and doing so allows the chain to efficiently explore values of the transition matrix and transition covariance within the permitted regions.
+As discussed, standard Gibbs sampling where we alternately sample from the state and parameter conditionals does not allow the Markov chain to fully explore the parameter space because of the constraint that each transition $(\ls{\ti}-\lgtm\ls{\ti-1})$ must lie in the $\rk$-dimensional column space of $\lgdm$.
+Nevertheless, we can sample from these distributions exactly, and doing so allows the chain to efficiently explore values of the transition matrix and transition covariance within the permitted regions.
 These constrained Gibbs sampling steps rely on the Givens decomposition \eqref{eq:givens_decomposition}.
+
+Note that the forward-filtering-backward-sampling procedure can be used as before to draw from the state conditional.
+See appendix~\ref{app:state-sampling}.
+The parameter conditional, on the other hand, will require some modifications.
 
@@
 Once new values of $\lgtmrot$ and $\tvfull$ have been sampled, the corresponding values of $\lgtm$ and $\lgtv$ may be calculated using the existing $\lgtmorth$ and $\tvrot$.
+
+%TODO: move this
+\subsubsection{Forward Filtering Backward Sampling}
+
+The forward-filtering-backward-sampling procedure for drawing from $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ can be applied to a degenerate model without modification. The Kalman filter stage is
 
@@
 \meta{Pseudo-code here}
-
-The computational complexity of the algorithm is $\bigo{\lsd^3 \timax}$, the same as the standard Gibbs sampler for the full rank model.
-For moderate to large $\timax$, the limiting factor is the forward-filtering-backward-sampling steps used to draw from $\postden(\ls{1:\timax}|\lgtm, \lgtv)$.
 
@@
 
+\section{Sampling the State Trajectory} \label{app:state-sampling}
+
+The Kalman filter recursively calculates the state predictive and filtering distributions,
+%
+\begin{IEEEeqnarray}{rCl}
+ \den(\ls{\ti}|\ob{1:\ti-1}, \lgtm, \lgtv) &=& \normalden{\ls{\ti}}{\lspmn{\ti}}{\lspvr{\ti}} \\
+ \den(\ls{\ti}|\ob{1:\ti}, \lgtm, \lgtv) &=& \normalden{\ls{\ti}}{\lsfmn{\ti}}{\lsfvr{\ti} \vphantom{\lspvr{\ti}} }     .
+\end{IEEEeqnarray}
+%
+The recursion equations for the means and variances are well known. See e.g. \cite{Sarkka2013}.
+
+The forward-filtering-backward-sampling procedure draws a sample from $\den(\ls{1:\timax}|\ob{1:\timax}, \lgtm, \lgtv)$.
+This is achieved by first sampling from the final filtering distribution $\den(\ls{\timax}|\ob{1:\timax})$ and then for $\ti=\timax-1 , \dots , 1$ sampling from $\den(\ls{\ti}|\ls{\ti+1}, \ob{1:\ti}, \lgtm, \lgtv)$.
+Each of these factors is also a Gaussian,
+%
+\begin{IEEEeqnarray}{l}
+ \den(\ls{\ti}|\ls{\ti+1}, \ob{1:\ti}, \lgtm, \lgtv) \nonumber \\
+ \quad = \normalden{\ls{\ti}}{ \lsfmn{\ti} + K_t (\ls{\ti+1}-\lgtm\lsfmn{\ti}) }{ \lspvr{\ti} - K_t \lgtm \lsfvr{\ti} }      ,
+\end{IEEEeqnarray}
+%
+where $K_t = \lsfvr{\ti} \lgtm\tr (\lgtm \lsfvr{\ti} \lgtm\tr + \lgtv)\inv$. The proof is straightforward using standard identities for the Gaussian distribution \cite{Sarkka2013}.
+
+This procedure is still valid when $\lgtv$ is singular, provided the matrix $(\lgtm \lsfvr{\ti} \lgtm\tr + \lgtv)$ is invertible.
+A sufficient condition for this is that both the initial covariance $\lsfvr{1}$ and the transition matrix $\lgtm$ are both non-singular.
+With a matrix normal prior on $\lgtm$, the probability of it being singular is almost surely $0$.
+
+To derive the expression for $\den(\ls{\ti}|\ls{\ti+1}, \ob{1:\ti}, \lgtm, \lgtv)$ in the case where the transition covariance is singular, replace $\lgtv$ with $\lgtv + h I$ where $I$ is the identity matrix and $h$ a non-negative real scalar such that a transition density can be written down.
+Then the same algebraic steps can be followed as in the non-singular case.
+At the end, set $h=0$ and obtain the same formula.
+
 \section{A~Matrix~Factorisation for Positive~Semi-Definite~Matrices} \label{app:givens-factorisation}

### 2016-01-14 notes/reviews/response_to_reviews.tex
More corrections: Add pseudo-code. Re-arranged degenerate model section.

@@
  \item change the rank after burn in to show not stuck
+ \item leave out parameter conditional moves
  \item plots
- \item collapsed Gibbs
- \item pseudo-code algorithm
  \item Givens
@@
 There is a brief discussion of the computational complexity in the final section of the paper.
-There is not much to say -- it is of the same order for the full rank and degenerate models.
-We do not see much value in including a detailed operation count, since the dominating factor will be the difference in autocorrelation and the resulting difference in the necessary length of the chain, which will vary by model and data set.
+There is not much to say -- it is of the same order for the full rank and degenerate models but with a higher cost per iteration.
+We do not see much value in including a detailed operation count, since the dominating factor when considering processing time will be the difference in autocorrelation and the resulting difference in the necessary length of the chain, which will vary by model and data set.
 \end{response}
@@
 \begin{response}
- \meta{Do this. At least, list the steps of each kernel.}
+ A good point. We have added algorithm pseudo-code describing the steps of the complete MCMC kernel. It was an oversight to leave this out.
 \end{response}
@@
 \begin{response}
- \meta{Good idea. Do this. Or perhaps just reference the paper more.}
+ There is a misunderstanding here, probably due to the confusing terminology.
+The ``collapsed Gibbs move'' simply means the composition of first targeting $\pi(F, Q)$ with Metropolis-Hastings followed by a draw from $\pi(x_{1:T}|F, Q)$.
+These moves are essential for the sampler to be irreducible, and they are woefully inefficient.
+To speed things up, we also include exact draws from the the parameter conditional $\pi(F, Q|x_{1:T})$.
+These are inspired by the Gibbs sampler for the full rank algorithm, but cannot fully explore the parameter space, so we refer to them as constrained Gibbs moves.
+
+ It does not make sense to run the sampler without the Metropolis-Hastings/constrained-Gibbs moves.
+It will be stuck in whatever subspace we happen to initialise it in.
+We {\em could} run it without the draws from the parameter conditional, but our python implementation would take literally years to converge.
+
+ We have substantially re-arranged section III and added further explanation of these concepts. We hope it is clearer now.
 \end{response}
@@
 \begin{response}
- \meta{Maybe...}
+ \meta{Try running without parameter conditional draws on toy model. Very slow, right?!}
 \end{response}
@@
 List of minor remarks/typos:
-\begin{itemize}
- \item Just after Eq.
+
+Just after Eq.
 (16), the authors should be more precise on the following sentence to avoid any confusion: ``However, whereas before $Q$ was required to be positive definite, this is no longer the case'' (I would add) for unknown $G$.
- \item Please correct in Section V first sentence ... known a prior. by ... known a priori.
-\end{itemize}
 \end{review}
@@
 \begin{response}
- \meta{Fix all these.}
+ We have re-written this sentence. 
+\end{response}
+
+\begin{review}
+Please correct in Section V first sentence ... known a prior. by ... known a priori.
+\end{review}
+
+\begin{response}
+This has been corrected.
 \end{response}

### 2016-01-14 notes/paper-degenerate-linear-systems.tex
More corrections: Add pseudo-code. Re-arranged degenerate model section.

@@
 \usepackage{cite}
+\usepackage{setspace}
 
@@
  \tikzexternalize[prefix=tikz/]
+
+\renewcommand{\algorithmicrequire}{\textbf{Input:}}
+\renewcommand{\algorithmicensure}{\textbf{Output:}}
 
@@
 Our principal interest is to learn the system matrices $\lgtm$ and $\lgtv$ from a sequence of observations $\ob{1:\timax}$.
-The posterior distribution over these variables $\den(\lgtm, \lgtv | \ob{1:\timax})$ is not amenable to MCMC sampling because there is no analytic expression for the density.
+The posterior distribution over these variables $\postden(\lgtm, \lgtv) = \den(\lgtm, \lgtv | \ob{1:\timax})$ is not amenable to MCMC sampling because there is no analytic expression for the density.
 Instead we introduce the latent state sequence $\ls{1:\timax}$ as a nuisance variable and target the joint posterior,
@@
 in which $\tn{\ti}$ as dimension $\rk$, and $\lgdm$ is $\lsd\times\rk$.
-This is almost equivalent to the previous case, with $\lgtv=\lgdm\lgdm\tr$.
-However, whereas before $\lgtv$ was required to be positive definite, this is no longer the case.
+This is almost equivalent to the model in the previous section, with $\lgtv=\lgdm\lgdm\tr$.
+However, in the previous model $\lgtv$ was positive definite by definition, whereas here it may not be full rank.
 When this occurs, the conventional transition density $\den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgdm)$ is not defined, and the sampling operations underlying our Gibbs sampler are no longer valid.
@@
 
-There is an obvious way to avoid being constrained to a particular subspace by the sampled state trajectory; simply return to targeting the marginal posterior $\postden(\lgtm, \lgtv|\ob{1:\timax})$ and fall back to using a generic Metropolis-Hastings approach.
+There is an obvious way to avoid being constrained to a particular subspace by the sampled state trajectory; simply return to targeting the marginal posterior $\postden(\lgtm, \lgtv)$ and fall back to using any generic Metropolis-Hastings approach.
 However, with a complex and high dimensional parameter space, designing efficient proposal distributions for Metropolis-Hastings is difficult.
 We would prefer to attain the automated, fast-mixing properties displayed by the Gibbs sampler on the full rank model.
-We will show how a Gibbs kernel may still be applied to sample a component of the parameters, with the remainder handled using Metropolis-Hastings.
+
+The solution we propose here is to use Metropolis-Hastings steps targeting the marginal to ensure irreducibility of the sampler, combined with constrained Gibbs sampling steps, which can only modify a component of the parameters, to accelerate the mixing of the chain.
+As in the full rank case, the target distribution will be the posterior over parameters and the state trajectory $\postden(\lgtm, \lgtv, \ls{1:\timax})$.
+This means that each of the Metropolis-Hastings steps targeting the marginal posterior $\postden(\lgtm, \lgtv)$ must be followed by a draw from the state conditional $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ in order to complete the iteration.
+The two two steps together comprise a collapsed Gibbs move \cite{Dyk2008}.
+Note that the forward-filtering-backward-sampling procedure can be used without modification to draw from the state conditional.
+See appendix~\ref{app:state-sampling}.
+
+\subsection{Covariance Decompositions}
 
@@
 
+\subsection{Metropolis-Hastings for the Posterior Marginal}
+
+In this section we introduce two Metropolis-Hastings kernels which target the marginal posterior $\postden(\lgtm,\lgtv)$.
+Sampling from each of these should be followed by a draw of the state trajectory from $\postden(\ls{1:\timax}|\lgtm,\lgtv)$.
+Together, this constitutes a collapsed Gibbs move for the target joint posterior distribution \cite{Dyk2008}.
+
+These moves are needed in order to ensure that the sampler can attain any possible values in the parameter space.
+However, the resulting Markov chain are liable to mix very poorly, since the proposals are dumb random walks.
+In the following section, we will describe how to draw from the parameter conditional distribution exactly, which greatly accelerates the mixing of the chain.
+
+\subsubsection{Covariance Random Walk} \label{sec:mh-rw-transition-covariance}
+
+The covariance matrix is specified by its eigenvalues and eigenvectors.
+The sampler is already able to explore the possible eigenvalues using the constrained Gibbs moves, but the eigenvector matrix is partially fixed during this process.
+We can allow the sampler to reach any possible eigenvector matrix using a rotational random walk.
+Suppose the current value of the transition covariance is $\lgtv\mcold$, a new value is proposed using the following procedure:
+%
+\begin{itemize}
+ \item Sample an orthogonal matrix $\mhrot$ from some distribution $\mhrotppsl$.
+ \item Set $\lgtv\mcnew = \mhrot \lgtv\mcold \mhrot\tr$.
+\end{itemize}
+%
+The transformation is invertible and has a Jacobian of $1$.
+If in addition we require that $\mhrotppsl(\mhrot) = \mhrotppsl(\mhrot\tr)$, then the proposal distribution is symmetric and the acceptance probability is simply,
+%
+\begin{IEEEeqnarray}{rCl}
+ \mhap(\lgtv\mcold\to\lgtv\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtm, \lgtv\mcnew) }{ \postden(\lgtm,\lgtv\mcold) } \right\} \\
+  & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtv\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtv\mcold)} \times \frac{\den(\lgtm,\lgtv\mcnew)}{\den(\lgtm,\lgtv\mcold)} \right\} \nonumber     .
+\end{IEEEeqnarray}
+
+There are numerous ways to sample the rotation matrix $\mhrot$ from a suitable proposal distribution $\mhrotppsl$.
+For example, we could use the Cayley transform \cite{Leon2006}, a bijective mapping from the skew-symmetric matrices to the rotation matrices, defined by,
+%
+\begin{IEEEeqnarray}{rCl}
+ P(S) & = & (\idmat - S)\inv(I+S)     .
+\end{IEEEeqnarray}
+%
+To sample from $\mhrotppsl$, we draw $\half\lsd(\lsd-1)$ independent scalar random variables $\{s_{i,j}\}_{0<i<j<\lsd}$ from any zero-mean distribution; a suitable choice is,
+%
+\begin{IEEEeqnarray}{rCl}
+ s_{k,l} & \sim & \normaldist{0}{\sigma_Q^2} \label{eq:skewsymmetric_proposal}     .
+\end{IEEEeqnarray}
+%
+Use these to construct a skew-symmetric matrix $S$,
+%
+\begin{IEEEeqnarray}{rCl}
+ S_{k,l} & = & \begin{cases}
+                s_{k,l}  & k<l \\
+                -s_{l,k} & k>l \\
+                0        & k=l     ,
+               \end{cases}
+\end{IEEEeqnarray}
+%
+and then set $\mhrot=P(S)$. The Cayley transform has the property that $P(-S)=P(S)\inv=P(S)\tr$, which implies that $\mhrotppsl(\mhrot)=\mhrotppsl(\mhrot\tr)$ as required.
+
+An alternative is to use Givens rotations.
+First sample $i \in [1,\lsd]$, $j \in [1,\lsd]\setminus i$, and $\givrot{} \in [-\pi/2,\pi/2]$ from some zero-mean distribution.
+Form the Givens matrix $\givmat{i}{j}{\givrot{}}$ such that,
+%
+\begin{IEEEeqnarray}{rCl}
+ \left[\givmat{i}{j}{\givrot{}} - \idmat\right]_{k,l} & = & \begin{cases}
+                                                    \cos(\givrot{})-1 & k=l=i \text{ or } k=l=j \\
+                                                    \sin(\givrot{}) & k=i,l=j \\
+                                                    -\sin(\givrot{}) & k=j,l=i \\
+                                                    0 & \text{ otherwise,}
+                                                 \end{cases}
+\end{IEEEeqnarray}
+%
+and use $\mhrot=\givmat{i}{j}{\givrot{}}$. This also has the property that $\givmat{i}{j}{-\givrot{}} = \givmat{i}{j}{\givrot{}}\tr$, implying a symmetric proposal.
+
+\subsubsection{Transition Matrix Random Walk} \label{sec:mh-rw-transition-matrix}
+
+A more conventional Metropolis-Hastings kernel can be used for the transition matrix; for example, using a simple Gaussian random walk as the proposal distribution.
+If the current value is $\lgtm\mcold$, a new matrix is sampled from a matrix normal distribution,
+%
+\begin{IEEEeqnarray}{rCl}
+ \lgtm\mcnew &\sim& \matrixnormaldist{\lgtm\mcold}{\sigma_F^2\idmat}{\idmat} \label{eq:transition_matrix_proposal}     ,
+\end{IEEEeqnarray}
+%
+and accepted with probability,
+%
+\begin{IEEEeqnarray}{rCl}
+ \mhap(\lgtm\mcold\to\lgtm\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtm\mcnew, \lgtv) }{ \postden(\lgtm\mcold,\lgtv) } \right\} \\
+  & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm\mcnew,\lgtv)}{\den(\ob{1:\timax}|\lgtm\mcold,\lgtv)} \times \frac{\den(\lgtm\mcnew,\lgtv)}{\den(\lgtm\mcold,\lgtv)} \right\} \nonumber     .
+\end{IEEEeqnarray}
+
 \subsection{Constrained Gibbs Sampling for Degenerate Transitions}
@@
 These constrained Gibbs sampling steps rely on the Givens decomposition \eqref{eq:givens_decomposition}.
-
-Note that the forward-filtering-backward-sampling procedure can be used as before to draw from the state conditional.
-See appendix~\ref{app:state-sampling}.
-The parameter conditional, on the other hand, will require some modifications.
 
@@
 \begin{equation}
- \lgtmrot = \tvrot\tr \lgtm      .
+ \lgtmrot = \tvrot\tr \lgtm \label{eq:transition_matrix_free}      .
 \end{equation}
@@
 \begin{IEEEeqnarray}{rCl}
- \lgtmorth &=& (\idmat-\tvrot\tvrot\tr) \lgtm     ,
+ \lgtmorth &=& (\idmat-\tvrot\tvrot\tr) \lgtm \label{eq:transition_matrix_fixed}    ,
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- \lgtm &=& \tvrot \lgtmrot + \lgtmorth    .
+ \lgtm &=& \tvrot \lgtmrot + \lgtmorth \label{eq:transition_matrix_decomposition}    .
 \end{IEEEeqnarray}
@@
 \begin{align}
- \tvfull | \tvrot, \ls{1:\timax} &\sim \iwishartdist{\postdof}{\postscalematrix} \\
- \lgtmrot | \tvfull, \tvrot, \ls{1:\timax} &\sim \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}     ,
+ \tvfull | \tvrot, \ls{1:\timax} &\sim \iwishartdist{\postdof}{\postscalematrix} \nonumber \\
+ \lgtmrot | \tvfull, \tvrot, \ls{1:\timax} &\sim \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance} \label{eq:constrained_parameter_conditional}      ,
 \end{align}
@@
 
-%TODO: move this
-\subsubsection{Forward Filtering Backward Sampling}
-
-The forward-filtering-backward-sampling procedure for drawing from $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ can be applied to a degenerate model without modification. The Kalman filter stage is
-
-\subsection{Metropolis-Hastings on the Posterior Marginal}
-
-Gibbs sampling is very efficient and requires no algorithm tuning but it only allows the Markov chain to explore a subset of the parameter space.
-In this section we introduce two Metropolis-Hastings kernels which target the marginal posterior $\postden(\lgtm,\lgtv)$ and allow the sampler to fully explore the target distribution.
-These sampling operations are each followed by a draw of the state trajectory from $\postden(\ls{1:\timax}|\lgtm,\lgtv)$, thus constituting a collapsed Gibbs move \cite{Dyk2008}.
-
-\subsubsection{Covariance Random Walk}
-
-The covariance matrix is specified by its eigenvalues and eigenvectors.
-The sampler is already able to explore the possible eigenvalues using the constrained Gibbs moves, but the eigenvector matrix is partially fixed during this process.
-We can allow the sampler to reach any possible eigenvector matrix using a rotational random walk.
-Suppose the current value of the transition covariance is $\lgtv\mcold$, a new value is proposed using the following procedure:
-%
-\begin{itemize}
- \item Sample an orthogonal matrix $\mhrot$ from some distribution $\mhrotppsl$.
- \item Set $\lgtv\mcnew = \mhrot \lgtv\mcold \mhrot\tr$.
-\end{itemize}
-%
-The transformation is invertible and has a Jacobian of $1$.
-If in addition we require that $\mhrotppsl(\mhrot) = \mhrotppsl(\mhrot\tr)$, then the proposal distribution is symmetric and the acceptance probability is simply,
-%
-\begin{IEEEeqnarray}{rCl}
- \mhap(\lgtv\mcold\to\lgtv\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtm, \lgtv\mcnew) }{ \postden(\lgtm,\lgtv\mcold) } \right\} \\
-  & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtv\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtv\mcold)} \times \frac{\den(\lgtm,\lgtv\mcnew)}{\den(\lgtm,\lgtv\mcold)} \right\} \nonumber     .
-\end{IEEEeqnarray}
-
-There are numerous ways to sample the rotation matrix $\mhrot$ from a suitable proposal distribution $\mhrotppsl$.
-For example, we could use the Cayley transform \cite{Leon2006}, a bijective mapping from the skew-symmetric matrices to the rotation matrices, defined by,
-%
-\begin{IEEEeqnarray}{rCl}
- P(S) & = & (\idmat - S)\inv(I+S)     .
-\end{IEEEeqnarray}
-%
-To sample from $\mhrotppsl$, we draw $\half\lsd(\lsd-1)$ independent scalar random variables $\{s_{i,j}\}_{0<i<j<\lsd}$ from any zero-mean distribution; a suitable choice is,
-%
-\begin{IEEEeqnarray}{rCl}
- s_{k,l} & \sim & \normaldist{0}{\sigma_Q^2} \label{eq:skewsymmetric_proposal}     .
-\end{IEEEeqnarray}
-%
-Use these to construct a skew-symmetric matrix $S$,
-%
-\begin{IEEEeqnarray}{rCl}
- S_{k,l} & = & \begin{cases}
-                s_{k,l}  & k<l \\
-                -s_{l,k} & k>l \\
-                0        & k=l     ,
-               \end{cases}
-\end{IEEEeqnarray}
-%
-and then set $\mhrot=P(S)$. The Cayley transform has the property that $P(-S)=P(S)\inv=P(S)\tr$, which implies that $\mhrotppsl(\mhrot)=\mhrotppsl(\mhrot\tr)$ as required.
-
-An alternative is to use Givens rotations.
-First sample $i \in [1,\lsd]$, $j \in [1,\lsd]\setminus i$, and $\givrot{} \in [-\pi/2,\pi/2]$ from some zero-mean distribution.
-Form the Givens matrix $\givmat{i}{j}{\givrot{}}$ such that,
-%
-\begin{IEEEeqnarray}{rCl}
- \left[\givmat{i}{j}{\givrot{}} - \idmat\right]_{k,l} & = & \begin{cases}
-                                                    \cos(\givrot{})-1 & k=l=i \text{ or } k=l=j \\
-                                                    \sin(\givrot{}) & k=i,l=j \\
-                                                    -\sin(\givrot{}) & k=j,l=i \\
-                                                    0 & \text{ otherwise,}
-                                                 \end{cases}
-\end{IEEEeqnarray}
-%
-and use $\mhrot=\givmat{i}{j}{\givrot{}}$. This also has the property that $\givmat{i}{j}{-\givrot{}} = \givmat{i}{j}{\givrot{}}\tr$, implying a symmetric proposal.
-
-\subsubsection{Transition Matrix Random Walk}
-
-A more conventional Metropolis-Hastings kernel can be used for the transition matrix; for example, using a simple Gaussian random walk as the proposal distribution.
-If the current value is $\lgtm\mcold$, a new matrix is sampled from a matrix normal distribution,
-%
-\begin{IEEEeqnarray}{rCl}
- \lgtm\mcnew &\sim& \matrixnormaldist{\lgtm\mcold}{\sigma_F^2\idmat}{\idmat} \label{eq:transition_matrix_proposal}     ,
-\end{IEEEeqnarray}
-%
-and accepted with probability,
-%
-\begin{IEEEeqnarray}{rCl}
- \mhap(\lgtm\mcold\to\lgtm\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtm\mcnew, \lgtv) }{ \postden(\lgtm\mcold,\lgtv) } \right\} \\
-  & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm\mcnew,\lgtv)}{\den(\ob{1:\timax}|\lgtm\mcold,\lgtv)} \times \frac{\den(\lgtm\mcnew,\lgtv)}{\den(\lgtm\mcold,\lgtv)} \right\} \nonumber     .
-\end{IEEEeqnarray}
-
 \section{Unknown Covariance Matrix Rank} \label{sec:rank_learning}
@@
 
-\subsection{Reversible Jump for Rank Learning}
+\subsection{Reversible Jump for Rank Learning} \label{sec:rjmcmc-covariance-rank}
 
@@
 \begin{IEEEeqnarray}{rCl}
- \evecset{\rk+1} &=& \{\evec{} : |\evec{}|=1,\:\tvvec\tr\evec{}=0,\:\sign(\evec{\rk+1,1}=+1 \}    .
+ \evecset{\rk+1} &=& \{\evec{} : |\evec{}|=1,\:\tvvec\tr\evec{}=0,\:\sign(\evec{\rk+1,1})=+1 \}    .
 \end{IEEEeqnarray}
@@
 
-\meta{Pseudo-code here}
+There are many ways in which the various Markov kernels could be combined to produce a complete MCMC sampler.
+Our suggested scheme is shown in algorithm~\ref{alg:complete-degenerate-model-algorithm}.
+A Markov chain is generated by repeatedly sampling according to this procedure and storing the values returned each iteration.
+Steps~\ref{step:mh-start} to~\ref{step:mh-end} constitute the collapsed Gibbs move, and steps~\ref{step:gibbs-start} to~\ref{step:gibbs-end} comprise a draw from the constrained parameter conditional.
+Each iteration requires a pass of the Kalman filter to calculate the marginal likelihood $\den(\ob{1:\timax}|\lgtm\mcint, \lgtv\mcint)$, plus an additional pass for each Metropolis-Hastings step conducted.
+
+\begin{algorithm}
+\small
+\onehalfspacing
+\begin{algorithmic}[1]
+\REQUIRE $\lgtm\mcold, \lgtv\mcold, \ls{1:\timax}\mcold$.
+\FOR{each of the Metropolis-Hastings kernels targeting $\postden(\lgtm, \lgtv)$, detailed in sections~\ref{sec:mh-rw-transition-matrix},~\ref{sec:mh-rw-transition-covariance},~and~\ref{sec:rjmcmc-covariance-rank}}
+%
+\STATE \label{step:mh-start} Propose new parameter values $\lgtm\mcnew, \lgtv\mcnew$.
+\STATE With the appropriate acceptance probability, set $\lgtm\mcold \leftarrow \lgtm\mcnew$, $\lgtv\mcold \leftarrow \lgtv\mcnew$.
+\STATE \label{step:mh-end} Sample $\ls{1:\timax}\mcold \sim \postden(\ls{1:\timax}|\lgtm\mcold, \lgtv\mcold)$.
+%
+\STATE \label{step:gibbs-start} Calculate $\tvfull\mcold, \tvrot\mcold$, the Givens decomposition of $\lgtv\mcold$.
+\STATE Calculate $\lgtmrot\mcold, \lgtmorth\mcold$, the orthogonal components of $\lgtm\mcold$, using \eqref{eq:transition_matrix_free} \& \eqref{eq:transition_matrix_fixed}.
+\STATE Sample $\lgtmrot\mcint, \tvfull\mcint \sim \postden(\lgtmrot,\tvfull|\lgtmorth\mcold,\tvrot\mcold,\ls{1:\timax}\mcold)$, the constrained parameter conditional given in \eqref{eq:constrained_parameter_conditional}.
+\STATE \label{step:gibbs-end} Calculate $\lgtm\mcint$ and $\lgtv\mcint$ from $\lgtmrot\mcint, \lgtmorth\mcold, \tvfull\mcint, \tvrot\mcold$ using \eqref{eq:givens_decomposition} \& \eqref{eq:transition_matrix_decomposition}.
+%
+\ENDFOR
+%
+\ENSURE $\lgtm\mcint, \lgtv\mcint, \ls{1:\timax}\mcint$.
+\end{algorithmic}
+\caption{\small MCMC kernel for degenerate linear state space models}
+\label{alg:complete-degenerate-model-algorithm}
+\end{algorithm}
 
@@
 
-In the previous sections, we have assumed that $\lgom$ and $\lgov$ are fixed and known a prior. In practice, $\lgov$ will often be unknown. In both the following examples we use models which have,
+In the previous sections, we have assumed that $\lgom$ and $\lgov$ are fixed and known a priori. In practice, $\lgov$ will often be unknown. In both the following examples we use models which have,
 \begin{IEEEeqnarray}{rCl}
@@
  \den(\ls{\ti}|\ls{\ti+1}, \ob{1:\ti}, \lgtm, \lgtv) \nonumber \\
+ \quad \propto \den(\ls{\ti+1}|\ls{\ti}, \lgtm, \lgtv) \den(\ls{\ti}| \ob{1:\ti}, \lgtm, \lgtv) \nonumber \\
  \quad = \normalden{\ls{\ti}}{ \lsfmn{\ti} + K_t (\ls{\ti+1}-\lgtm\lsfmn{\ti}) }{ \lspvr{\ti} - K_t \lgtm \lsfvr{\ti} }      ,
@@
 where $\tvsign$ is a diagonal matrix in which the diagonal elements are $\pm1$.
-This factorisation may be derived (and also calculated) by iteratively left multiplying the original matrix by a (transposed) Givens rotation matrix such that one element is set to $0$.
+This factorisation may be derived (and also calculated) by repeatedly left-multiplying the original matrix by a (transposed) Givens rotation matrix such that one element is set to $0$.
 The matrix remaining once all the elements below the diagonal have been eliminated is $\tvsign$, and the factorisation above results from a straightforward rearrangement.

### 2016-01-15 notes/reviews/response_to_reviews.tex
Added Givens decomposition pseudo-code and additional explanation

@@
 \begin{itemize}
- \item different parameter settings for toy model
- \item change the rank after burn in to show not stuck
- \item leave out parameter conditional moves
+ \item Add trace plots for the transition matrix.
+ \item change the observation noise so the rank is not so certain.
+ \item Add rank posterior histograms.
+ \item leave out parameter conditional moves?
  \item plots
- \item Givens
- \item Reviewer 2
  \item Reviewer 3
@@
 \begin{response}
- \meta{Add more discussion of the Givens decomposition and add algorithm for generating it.}
+ We have added more explanation of the Givens decomposition, and also algorithm pseudo-code for its calculation.
 \end{response}

### 2016-01-15 notes/paper-degenerate-linear-systems.tex
Added Givens decomposition pseudo-code and additional explanation

@@
 
-A further factorisation will also be useful, which we will refer to as the Givens decomposition (on account of its method of computation),
-%
-\begin{IEEEeqnarray}{rCl}
- \lgtv &=& \tvrot \tvfull \tvrot\tr     , \label{eq:givens_decomposition} %&=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix} \begin{bmatrix}\tvfull & \zmat \\ \zmat & \zmat\end{bmatrix} \begin{bmatrix}\tvrot\tr \\ \tvrotorth\tr\end{bmatrix}
-\end{IEEEeqnarray}
-%
-where $\tvfull$ is an $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors subject to certain constraints.
+A second decomposition will also be useful. Any positive semi-definite can be factorised in the following way,
+%
+\begin{IEEEeqnarray}{rCl}
+ \lgtv &=& \tvrot \tvfull \tvrot\tr     , \label{eq:givens_decomposition}
+\end{IEEEeqnarray}
+%
+where $\tvfull$ is an $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors with a particular structure.
 The transformation from $\lgtv$ to $\{\tvfull, \tvrot\}$ is also bijective.
-See appendix~\ref{app:givens-factorisation} for details.
+This factorisation is constructed using Givens rotations and thus we refer to it as the Givens decomposition.
+A Givens rotation matrix has the following structure,
+%
+\begin{IEEEeqnarray}{rCl}
+ \left[\givmat{i}{j}{\givrot{}} - \idmat\right]_{k,l} & = & \begin{cases}
+                                                    \cos(\givrot{})-1 & k=l=i \text{ or } k=l=j \\
+                                                    \sin(\givrot{}) & k=i,l=j \\
+                                                    -\sin(\givrot{}) & k=j,l=i \\
+                                                    0 & \text{ otherwise}     ,
+                                                 \end{cases}
+\end{IEEEeqnarray}
+%
+where $\givrot{} \in [-\pi/2,\pi/2]$ is a rotation in the plane of the $i$ and $j$ coordinate directions.
+
+The Givens decomposition is important because $\tvrot$ is uniquely determined by the state trajectory, whereas $\tvfull$ is free to vary.
+We provide here the algorithm for conducting the Givens decomposition.
+The computational complexity is $\bigo{\rk\times(\lsd-\rk)}$.
+(Note that multiplication by a Givens matrix is $\bigo{1}$ due to sparsity.) Details and further discussion is included in appendix~\ref{app:givens-factorisation}.
+
+\begin{algorithm}
+\onehalfspacing
+\small
+\begin{algorithmic}[1]
+ \REQUIRE Eigvenvalues and eigenvectors $\tvval,\tvvec$ of positive semi-definite matrix $\lgtv$.
+ \STATE Set $\tvcross \leftarrow \idmat[\lsd\times\lsd]$, $\tvrow \leftarrow \idmat[\rk\times\rk]$, $\tvsign \leftarrow \tvvec$.
+ \FOR{$i = \rk,\dots,1$}
+   \FOR{$j = 1,\dots,\rk-1$}
+      \STATE Calculate the rotation $\givrot{i,j} = \tan^{-1}(\tvsign_{i,j} / \tvsign_{i,i})$.
+      \STATE Construct the Givens matrix $\givmat{i}{j}{\givrot{i,j}}$.
+      \STATE Set $\tvsign \leftarrow \tvsign \: \givmat{i}{j}{\givrot{i,j}}$, $\tvrow \leftarrow \givmat{i}{j}{\givrot{i,j}}\tr \tvrow$.
+   \ENDFOR
+ \ENDFOR
+  \FOR{$j = 1,\dots,\rk$}
+   \FOR{$i = \lsd,\dots,\rk+1$}
+      \STATE Calculate the rotation $\givrot{i,j} = \tan^{-1}(\tvsign_{j,j} / \tvsign_{i,j})$.
+      \STATE Construct the Givens matrix $\givmat{i}{j}{\givrot{i,j}}$.
+      \STATE Set $\tvsign \leftarrow \givmat{i}{j}{\givrot{i,j}} \tvsign$, $\tvcross \leftarrow \tvcross \givmat{i}{j}{\givrot{i,j}}\tr$.
+   \ENDFOR
+ \ENDFOR
+ \STATE Extract $\tvsigndiag$, the first $\rk$ rows of $\tvsign$, and $\tvrot$, the first $\rk$ columns of $\tvcross$.
+ \STATE Calculate $\tvfull = \tvsigndiag \tvrow \tvval \tvrow\tr \tvsigndiag\tr$.
+ \ENSURE $\tvrot, \tvfull$.
+\end{algorithmic}
+\caption{Givens decomposition}
+\label{alg:givens-decomposition}
+\end{algorithm}
 
@@
 \end{IEEEeqnarray}
-\meta{CHECK THIS!}
 
@@
 \begin{IEEEeqnarray}{rCl}
-\tvvec = \tvsign &\times& \left[ \givmat{1}{2}{\givrot{1,2}}\right] \times \dotsm \nonumber\\
+\tvvec = \tvsigndiag &\times& \left[ \givmat{1}{2}{\givrot{1,2}}\right] \times \dotsm \nonumber\\
 &\times& \left[\givmat{1}{d-1}{\givrot{1,d-1}} \dotsm \givmat{d-2}{d-1}{\givrot{d-2,d-1}} \right] \nonumber\\
@@
 %
-where $\tvsign$ is a diagonal matrix in which the diagonal elements are $\pm1$.
+where $\tvsigndiag$ is a diagonal matrix in which the diagonal elements are $\pm1$.
 This factorisation may be derived (and also calculated) by repeatedly left-multiplying the original matrix by a (transposed) Givens rotation matrix such that one element is set to $0$.
-The matrix remaining once all the elements below the diagonal have been eliminated is $\tvsign$, and the factorisation above results from a straightforward rearrangement.
+The matrix remaining once all the elements below the diagonal have been eliminated is $\tvsigndiag$, and the factorisation above results from a straightforward rearrangement.
 See \cite{Anderson1987} for details.
@@
 %
-and $\tvsign$ is as before, with the following structure,
+and $\tvsign$ has the following structure,
 %
@@
 \end{IEEEeqnarray}
-
-Now consider a rank-$r$ positive semi-definite matrix $\psd$ for which the non-singular part of the eigendecomposition is,
+%
+where $\tvsigndiag$, as before, is diagonal with elements $\pm1$.
+
+Finally consider a rank-$r$ positive semi-definite matrix $\psd$ for which the non-singular part of the eigendecomposition is,
 %
@@
 %
-Comparing with \eqref{eq:standard_givens}, we see that $\tvsigndiag \tvrow$ is an unconstrained $r\times r$ orthogonal matrix and hence $\tvfull$ is a unique positive definite matrix.
+Comparing with \eqref{eq:standard_givens}, we see that $\tvsigndiag \tvrow$ can represent any $r\times r$ orthogonal matrix and hence $\tvfull$ is a unique positive definite matrix.
 

### 2016-01-18 notes/reviews/response_to_reviews.tex
Improved simulations. Added second toy model.

@@
 
-Thank you for taking the time to review our paper. Here is our responses to the points raised.
-
-\meta{
-\begin{itemize}
- \item Add trace plots for the transition matrix.
- \item change the observation noise so the rank is not so certain.
- \item Add rank posterior histograms.
- \item leave out parameter conditional moves?
- \item plots
- \item Reviewer 3
- \item Proof read, aiming to (i) clarify (2) correct typos (c) define everything when first used
-\end{itemize}
-
-}
+Thank you for taking the time to review our paper. We are sure it is a better paper as a result of your scrutiny and the consequent changes. Here is our responses to the points raised.
 
@@
 
- It does not make sense to run the sampler without the Metropolis-Hastings/constrained-Gibbs moves.
+ It does not make sense to run the sampler without the Metropolis-Hastings moves.
 It will be stuck in whatever subspace we happen to initialise it in.
-We {\em could} run it without the draws from the parameter conditional, but our python implementation would take literally years to converge.
+Similarly, the constrained parameter conditional draws are necessary for sampling the covariance eigenvalues.
+In short, there is no step we could leave out without invalidating the sampler.
 
@@
 \begin{response}
- \meta{Add more discussion and analysis}
+ We have made efforts to improve the simulations section. Our aims here are:
+ \begin{itemize}
+  \item To show that the sampler converges and mixes satisfactorily.
+  \item To show that the resulting posterior approximations are consistent with the true values.
+  \item To compare using the degenerate model with the full rank model (less error but slower mixing).
+  \item To demonstrate the utility of the algorithm on a practical application.
+ \end{itemize}
+We have added a paragraph at the start of the section explaining these aims and added some additional comments.
+We deliberately reserve the more general discussion of the algorithm for the final section, where we think it better belongs.
 \end{response}
@@
 \begin{response}
- \meta{Explain this and improve in the paper. Add simple example of changing rank to show that it is not stuck.}
+We a bit zealous with trying to save space in the first draft by omitting certain figures.
+We have added trace plots and posterior histograms for $F$ and $Q$.
+We believe that the reason the sampler sits at one value of the rank throughout is simply that the posterior histogram is tightly concentrated on this value.
+However, as you rightly point out, this could be a sign that it is mixing poorly.
+We have added a second toy model in which the rank is less clearly determined by the observations, in order to demonstrate that the sampler is able to change values after convergence.
 \end{response}
@@
 \begin{review}
-The authors could illustrate in the first toy example the benefit of the collapsed Gibbs
-moves by showing the performance of the sampler with vs without this strategy.
+The authors could illustrate in the first toy example the benefit of the collapsed Gibbs moves by showing the performance of the sampler with vs without this strategy.
 \end{review}
@@
 \begin{response}
- \meta{Try running without parameter conditional draws on toy model. Very slow, right?!}
+ See earlier comments. It is not possible to omit any of the moves without the sampler becoming reducible, so this would not be a useful demonstration.
 \end{response}
@@
 \begin{response}
- \meta{Ok... if it will make you happy. Obviously it will work for some parameters and not others.}
+We do not feel that such a plot would be particularly informative, since this model is artificial and somewhat arbitrary.
+Inevitably, higher noise leads to larger errors...
+However we agree with the idea that the algorithm should be tested with some different parameter settings, which we have achieved by adding a second toy model (which also illustrates learning a posterior distribution for the rank with greater uncertainty).
 \end{response}

### 2016-01-18 notes/paper-degenerate-linear-systems.tex
Improved simulations. Added second toy model.

@@
 
-More recently, Bayesian approaches have been developed for non-parametric learning of linear Gaussian state space models, in which full posterior distributions for the system parameters are calculated.
+More recently, Bayesian approaches have been developed for learning linear Gaussian state space models, in which full posterior distributions for the system parameters are calculated.
 Since this is not possible analytically (see discussion in \cite{Beal2003}), approximations must be employed.
@@
 This is the case, for example, when the latent state comprises the positions and velocities of multiple points on the same rigid body, meaning that their motion is deterministically coupled.
-\cite{Maskell2004,Gustafsson2002} consider the use of degenerate transition models for various tracking applications.
+In \cite{Maskell2004,Gustafsson2002}, the authors consider the use of degenerate transition models for various tracking applications.
 The Gibbs sampler of \cite{Wills2012} is not able to handle degenerate models.
@@
 where $\ls{\ti} \in \reals^{\lsd}$, $\ob{\ti} \in \reals^{\obd}$.
-$\lgtm \in \reals^{\lsd\times\lsd}$,$\lgom \in \reals^{\obd\times\lsd}$ are the transition and observation matrices.
+$\lgtm \in \reals^{\lsd\times\lsd}$ and $\lgom \in \reals^{\obd\times\lsd}$ are the transition and observation matrices.
 $\tn{\ti} \in \reals^{\lsd}$ and $\on{\ti} \in \reals^{\obd}$ are the state disturbance and observation noise variables and we additionally assume that these are drawn from a Gaussian distribution,
@@
 \begin{IEEEeqnarray}{rCl}
- \tn{\ti} & \sim & \normalden{\tn{\ti}}{0}{\lgtv} \\
- \on{\ti} & \sim & \normalden{\on{\ti}}{0}{\lgov}     ,
+ \tn{\ti} & \sim & \normaldist{0}{\lgtv} \\
+ \on{\ti} & \sim & \normaldist{0}{\lgov}     ,
 \end{IEEEeqnarray}
@@
 \cite{Roweis1999}.) To resolve the parameter ambiguity, it is necessary to constrain the values of one or more of the system matrices.
-One option (that taken in \cite{Roweis1999}) is to set $\lgtv$ to the identity matrix.
+One option (that taken in \cite{Roweis1999}, for example) is to set $\lgtv$ to the identity matrix.
 This entails that the nature of the latent states and their relationship to the observations are entirely unknown a priori.
@@
 This means that each of the Metropolis-Hastings steps targeting the marginal posterior $\postden(\lgtm, \lgtv)$ must be followed by a draw from the state conditional $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ in order to complete the iteration.
-The two two steps together comprise a collapsed Gibbs move \cite{Dyk2008}.
+The two steps together comprise a collapsed Gibbs move \cite{Dyk2008}.
 Note that the forward-filtering-backward-sampling procedure can be used without modification to draw from the state conditional.
@@
 %
-where $\tvfull$ is an $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors with a particular structure.
-The transformation from $\lgtv$ to $\{\tvfull, \tvrot\}$ is also bijective.
+where $\tvfull$ is an $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors.
+In general, this factorisation is not unique.
+However, here we employ a particular procedure for generating $\tvfull$ and $\tvrot$ which does yield a unique result, ensuring that the transformation from $\lgtv$ to $\{\tvfull, \tvrot\}$ is also bijective.
 This factorisation is constructed using Givens rotations and thus we refer to it as the Givens decomposition.
@@
                                                     0 & \text{ otherwise}     ,
-                                                 \end{cases}
+                                                 \end{cases} \label{eq:givens_matrix}
 \end{IEEEeqnarray}
@@
 where $\givrot{} \in [-\pi/2,\pi/2]$ is a rotation in the plane of the $i$ and $j$ coordinate directions.
+The procedure for the Givens decomposition is shown in algorithm~\ref{alg:givens-decomposition}
 
 The Givens decomposition is important because $\tvrot$ is uniquely determined by the state trajectory, whereas $\tvfull$ is free to vary.
-We provide here the algorithm for conducting the Givens decomposition.
 The computational complexity is $\bigo{\rk\times(\lsd-\rk)}$.
@@
  \FOR{$i = \rk,\dots,1$}
-   \FOR{$j = 1,\dots,\rk-1$}
+   \FOR{$j = 1,\dots,i-1$}
       \STATE Calculate the rotation $\givrot{i,j} = \tan^{-1}(\tvsign_{i,j} / \tvsign_{i,i})$.
@@
    \FOR{$i = \lsd,\dots,\rk+1$}
-      \STATE Calculate the rotation $\givrot{i,j} = \tan^{-1}(\tvsign_{j,j} / \tvsign_{i,j})$.
+      \STATE Calculate the rotation $\givrot{i,j} = \tan^{-1}(\tvsign_{i,j} / \tvsign_{j,j})$.
       \STATE Construct the Givens matrix $\givmat{i}{j}{\givrot{i,j}}$.
@@
 
-\subsection{Metropolis-Hastings for the Posterior Marginal}
-
-In this section we introduce two Metropolis-Hastings kernels which target the marginal posterior $\postden(\lgtm,\lgtv)$.
-Sampling from each of these should be followed by a draw of the state trajectory from $\postden(\ls{1:\timax}|\lgtm,\lgtv)$.
-Together, this constitutes a collapsed Gibbs move for the target joint posterior distribution \cite{Dyk2008}.
-
-These moves are needed in order to ensure that the sampler can attain any possible values in the parameter space.
-However, the resulting Markov chain are liable to mix very poorly, since the proposals are dumb random walks.
-In the following section, we will describe how to draw from the parameter conditional distribution exactly, which greatly accelerates the mixing of the chain.
-
-\subsubsection{Covariance Random Walk} \label{sec:mh-rw-transition-covariance}
-
-The covariance matrix is specified by its eigenvalues and eigenvectors.
-The sampler is already able to explore the possible eigenvalues using the constrained Gibbs moves, but the eigenvector matrix is partially fixed during this process.
-We can allow the sampler to reach any possible eigenvector matrix using a rotational random walk.
-Suppose the current value of the transition covariance is $\lgtv\mcold$, a new value is proposed using the following procedure:
-%
-\begin{itemize}
- \item Sample an orthogonal matrix $\mhrot$ from some distribution $\mhrotppsl$.
- \item Set $\lgtv\mcnew = \mhrot \lgtv\mcold \mhrot\tr$.
-\end{itemize}
-%
-The transformation is invertible and has a Jacobian of $1$.
-If in addition we require that $\mhrotppsl(\mhrot) = \mhrotppsl(\mhrot\tr)$, then the proposal distribution is symmetric and the acceptance probability is simply,
-%
-\begin{IEEEeqnarray}{rCl}
- \mhap(\lgtv\mcold\to\lgtv\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtm, \lgtv\mcnew) }{ \postden(\lgtm,\lgtv\mcold) } \right\} \\
-  & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtv\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtv\mcold)} \times \frac{\den(\lgtm,\lgtv\mcnew)}{\den(\lgtm,\lgtv\mcold)} \right\} \nonumber     .
-\end{IEEEeqnarray}
-
-There are numerous ways to sample the rotation matrix $\mhrot$ from a suitable proposal distribution $\mhrotppsl$.
-For example, we could use the Cayley transform \cite{Leon2006}, a bijective mapping from the skew-symmetric matrices to the rotation matrices, defined by,
-%
-\begin{IEEEeqnarray}{rCl}
- P(S) & = & (\idmat - S)\inv(I+S)     .
-\end{IEEEeqnarray}
-%
-To sample from $\mhrotppsl$, we draw $\half\lsd(\lsd-1)$ independent scalar random variables $\{s_{i,j}\}_{0<i<j<\lsd}$ from any zero-mean distribution; a suitable choice is,
-%
-\begin{IEEEeqnarray}{rCl}
- s_{k,l} & \sim & \normaldist{0}{\sigma_Q^2} \label{eq:skewsymmetric_proposal}     .
-\end{IEEEeqnarray}
-%
-Use these to construct a skew-symmetric matrix $S$,
-%
-\begin{IEEEeqnarray}{rCl}
- S_{k,l} & = & \begin{cases}
-                s_{k,l}  & k<l \\
-                -s_{l,k} & k>l \\
-                0        & k=l     ,
-               \end{cases}
-\end{IEEEeqnarray}
-%
-and then set $\mhrot=P(S)$. The Cayley transform has the property that $P(-S)=P(S)\inv=P(S)\tr$, which implies that $\mhrotppsl(\mhrot)=\mhrotppsl(\mhrot\tr)$ as required.
-
-An alternative is to use Givens rotations.
-First sample $i \in [1,\lsd]$, $j \in [1,\lsd]\setminus i$, and $\givrot{} \in [-\pi/2,\pi/2]$ from some zero-mean distribution.
-Form the Givens matrix $\givmat{i}{j}{\givrot{}}$ such that,
-%
-\begin{IEEEeqnarray}{rCl}
- \left[\givmat{i}{j}{\givrot{}} - \idmat\right]_{k,l} & = & \begin{cases}
-                                                    \cos(\givrot{})-1 & k=l=i \text{ or } k=l=j \\
-                                                    \sin(\givrot{}) & k=i,l=j \\
-                                                    -\sin(\givrot{}) & k=j,l=i \\
-                                                    0 & \text{ otherwise,}
-                                                 \end{cases}
-\end{IEEEeqnarray}
-%
-and use $\mhrot=\givmat{i}{j}{\givrot{}}$. This also has the property that $\givmat{i}{j}{-\givrot{}} = \givmat{i}{j}{\givrot{}}\tr$, implying a symmetric proposal.
-
-\subsubsection{Transition Matrix Random Walk} \label{sec:mh-rw-transition-matrix}
-
-A more conventional Metropolis-Hastings kernel can be used for the transition matrix; for example, using a simple Gaussian random walk as the proposal distribution.
-If the current value is $\lgtm\mcold$, a new matrix is sampled from a matrix normal distribution,
-%
-\begin{IEEEeqnarray}{rCl}
- \lgtm\mcnew &\sim& \matrixnormaldist{\lgtm\mcold}{\sigma_F^2\idmat}{\idmat} \label{eq:transition_matrix_proposal}     ,
-\end{IEEEeqnarray}
-%
-and accepted with probability,
-%
-\begin{IEEEeqnarray}{rCl}
- \mhap(\lgtm\mcold\to\lgtm\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtm\mcnew, \lgtv) }{ \postden(\lgtm\mcold,\lgtv) } \right\} \\
-  & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm\mcnew,\lgtv)}{\den(\ob{1:\timax}|\lgtm\mcold,\lgtv)} \times \frac{\den(\lgtm\mcnew,\lgtv)}{\den(\lgtm\mcold,\lgtv)} \right\} \nonumber     .
-\end{IEEEeqnarray}
-
 \subsection{Constrained Gibbs Sampling for Degenerate Transitions}
@@
  \LS{2:\timax} &=& (\tvrot \lgtmrot + \lgtmorth) \LS{1:\timax-1} + \tvrot Z \nonumber \\
- \Rightarrow \LS{2:\timax} &=& \lgtmorth \LS{1:\timax-1} + \tvrot Z' \nonumber      .
-\end{IEEEeqnarray}
-%
+ \Rightarrow \LS{2:\timax} &=& \lgtmorth \LS{1:\timax-1} + \tvrot Z' \nonumber      ,
+\end{IEEEeqnarray}
+%
+where $Z'$ is a second arbitrary matrix in the same space as $Z$.
 Notice that $\lgtmrot$ and $\tvfull$ do not appear in this equation, and thus may be freely altered by the sampler.
@@
 We will need the distribution on $\{\lgtmrot, \tvfull\}$ implied by the prior we chose for $\{\lgtm,\lgtv\}$.
-These follow from simple properties of the wishart and matrix normal distributions (see \cite{Muirhead1982}).
+These follow from simple properties of the Wishart and matrix normal distributions (see \cite{Muirhead1982}).
 For the matrix normal part,
@@
 
+\subsection{Metropolis-Hastings for the Posterior Marginal}
+
+In this section we introduce two Metropolis-Hastings kernels which target the marginal posterior $\postden(\lgtm,\lgtv)$.
+Sampling from each of these should be followed by a draw of the state trajectory from $\postden(\ls{1:\timax}|\lgtm,\lgtv)$.
+Together, this constitutes a collapsed Gibbs move for the target joint posterior distribution \cite{Dyk2008}.
+These moves are needed in order to ensure that the sampler can attain any possible values in the parameter space.
+
+\subsubsection{Covariance Random Walk} \label{sec:mh-rw-transition-covariance}
+
+The covariance matrix is specified by its eigenvalues and eigenvectors.
+The sampler is already able to explore the possible eigenvalues using the constrained Gibbs moves, but the eigenvector matrix is partially fixed during this process.
+We can allow the sampler to reach any possible eigenvector matrix using a rotational random walk.
+Suppose the current value of the transition covariance is $\lgtv\mcold$, a new value is proposed using the following procedure:
+%
+\begin{itemize}
+ \item Sample an orthogonal matrix $\mhrot$ from some distribution $\mhrotppsl$.
+ \item Set $\lgtv\mcnew = \mhrot \lgtv\mcold \mhrot\tr$.
+\end{itemize}
+%
+The transformation is invertible and has a Jacobian of $1$.
+If in addition we require that $\mhrotppsl(\mhrot) = \mhrotppsl(\mhrot\tr)$, then the proposal distribution is symmetric and the acceptance probability is simply,
+%
+\begin{IEEEeqnarray}{rCl}
+ \mhap(\lgtv\mcold\to\lgtv\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtm, \lgtv\mcnew) }{ \postden(\lgtm,\lgtv\mcold) } \right\} \\
+  & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtv\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtv\mcold)} \times \frac{\den(\lgtm,\lgtv\mcnew)}{\den(\lgtm,\lgtv\mcold)} \right\} \nonumber     .
+\end{IEEEeqnarray}
+
+There are numerous ways to sample the rotation matrix $\mhrot$ from a suitable proposal distribution $\mhrotppsl$.
+For example, we could use the Cayley transform \cite{Leon2006}, a bijective mapping from the skew-symmetric matrices to the rotation matrices, defined by,
+%
+\begin{IEEEeqnarray}{rCl}
+ P(S) & = & (\idmat - S)\inv(I+S)     .
+\end{IEEEeqnarray}
+%
+To sample from $\mhrotppsl$, we draw $\half\lsd(\lsd-1)$ independent scalar random variables $\{s_{i,j}\}_{0<i<j<\lsd}$ from any zero-mean distribution; a suitable choice is,
+%
+\begin{IEEEeqnarray}{rCl}
+ s_{k,l} & \sim & \normaldist{0}{\sigma_Q^2} \label{eq:skewsymmetric_proposal}     .
+\end{IEEEeqnarray}
+%
+Use these to construct a skew-symmetric matrix $S$,
+%
+\begin{IEEEeqnarray}{rCl}
+ S_{k,l} & = & \begin{cases}
+                s_{k,l}  & k<l \\
+                -s_{l,k} & k>l \\
+                0        & k=l     ,
+               \end{cases}
+\end{IEEEeqnarray}
+%
+and then set $\mhrot=P(S)$. The Cayley transform has the property that $P(-S)=P(S)\inv=P(S)\tr$, which implies that $\mhrotppsl(\mhrot)=\mhrotppsl(\mhrot\tr)$ as required.
+An alternative is to use Givens rotations.
+First sample $i \in [1,\lsd]$, $j \in [1,\lsd]\setminus i$, and $\givrot{} \in [-\pi/2,\pi/2]$ from some zero-mean distribution.
+Form the Givens matrix $\givmat{i}{j}{\givrot{}}$ according to \eqref{eq:givens_matrix} and use $\mhrot=\givmat{i}{j}{\givrot{}}$.
+This also has the property that $\givmat{i}{j}{-\givrot{}} = \givmat{i}{j}{\givrot{}}\tr$, implying a symmetric proposal.
+
+\subsubsection{Transition Matrix Random Walk} \label{sec:mh-rw-transition-matrix}
+
+A more conventional Metropolis-Hastings kernel can be used for the transition matrix; for example, using a simple Gaussian random walk as the proposal distribution.
+If the current value is $\lgtm\mcold$, a new matrix is sampled from a matrix normal distribution,
+%
+\begin{IEEEeqnarray}{rCl}
+ \lgtm\mcnew &\sim& \matrixnormaldist{\lgtm\mcold}{\sigma_F^2\idmat}{\idmat} \label{eq:transition_matrix_proposal}     ,
+\end{IEEEeqnarray}
+%
+and accepted with probability,
+%
+\begin{IEEEeqnarray}{rCl}
+ \mhap(\lgtm\mcold\to\lgtm\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtm\mcnew, \lgtv) }{ \postden(\lgtm\mcold,\lgtv) } \right\} \\
+  & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm\mcnew,\lgtv)}{\den(\ob{1:\timax}|\lgtm\mcold,\lgtv)} \times \frac{\den(\lgtm\mcnew,\lgtv)}{\den(\lgtm\mcold,\lgtv)} \right\} \nonumber     .
+\end{IEEEeqnarray}
+
 \section{Unknown Covariance Matrix Rank} \label{sec:rank_learning}
@@
 Steps~\ref{step:mh-start} to~\ref{step:mh-end} constitute the collapsed Gibbs move, and steps~\ref{step:gibbs-start} to~\ref{step:gibbs-end} comprise a draw from the constrained parameter conditional.
-Each iteration requires a pass of the Kalman filter to calculate the marginal likelihood $\den(\ob{1:\timax}|\lgtm\mcint, \lgtv\mcint)$, plus an additional pass for each Metropolis-Hastings step conducted.
+Each iteration requires two passes of the Kalman filter, in order to calculate the marginal likelihoods $\den(\ob{1:\timax}|\lgtm\mcint, \lgtv\mcint)$ for the Metropolis-Hastings acceptance probability.
 
@@
 \REQUIRE $\lgtm\mcold, \lgtv\mcold, \ls{1:\timax}\mcold$.
-\FOR{each of the Metropolis-Hastings kernels targeting $\postden(\lgtm, \lgtv)$, detailed in sections~\ref{sec:mh-rw-transition-matrix},~\ref{sec:mh-rw-transition-covariance},~and~\ref{sec:rjmcmc-covariance-rank}}
+\STATE Randomly select one of the Metropolis-Hastings kernels targeting $\postden(\lgtm, \lgtv)$, detailed in sections~\ref{sec:mh-rw-transition-matrix},~\ref{sec:mh-rw-transition-covariance},~and~\ref{sec:rjmcmc-covariance-rank}.
 %
 \STATE \label{step:mh-start} Propose new parameter values $\lgtm\mcnew, \lgtv\mcnew$.
-\STATE With the appropriate acceptance probability, set $\lgtm\mcold \leftarrow \lgtm\mcnew$, $\lgtv\mcold \leftarrow \lgtv\mcnew$.
+\STATE With the appropriate acceptance probability, set $\lgtm\mcold \leftarrow \lgtm\mcnew$ and $\lgtv\mcold \leftarrow \lgtv\mcnew$.
 \STATE \label{step:mh-end} Sample $\ls{1:\timax}\mcold \sim \postden(\ls{1:\timax}|\lgtm\mcold, \lgtv\mcold)$.
 %
-\STATE \label{step:gibbs-start} Calculate $\tvfull\mcold, \tvrot\mcold$, the Givens decomposition of $\lgtv\mcold$.
+\STATE \label{step:gibbs-start} Calculate $\tvfull\mcold, \tvrot\mcold$, the Givens decomposition of $\lgtv\mcold$, using algorithm~\ref{alg:givens-decomposition}.
 \STATE Calculate $\lgtmrot\mcold, \lgtmorth\mcold$, the orthogonal components of $\lgtm\mcold$, using \eqref{eq:transition_matrix_free} \& \eqref{eq:transition_matrix_fixed}.
@@
 \STATE \label{step:gibbs-end} Calculate $\lgtm\mcint$ and $\lgtv\mcint$ from $\lgtmrot\mcint, \lgtmorth\mcold, \tvfull\mcint, \tvrot\mcold$ using \eqref{eq:givens_decomposition} \& \eqref{eq:transition_matrix_decomposition}.
-%
-\ENDFOR
 %
@@
 
+We have so far assumed that $\lgov$ is fixed and known a priori.
+In practice, $\lgov$ will often be unknown.
+We can learn it within the MCMC scheme by targeting the posterior distribution $\postden(\lgtm, \lgtv, \lgov, \ls{1:\timax})$.
+An additional Gibbs move is introduced to draw from $\postden(\lgov | \lgtm, \lgtv, \ls{1:\timax})$, which can be sampled exactly if we use a conjugate inverse Wishart prior.
+
+An alternative, which we employ in the following simulations, is to assume the following structure,
+%
+\begin{IEEEeqnarray}{rCl}
+ \lgov &=& \xi_{\ob{}} \idmat     ,
+\end{IEEEeqnarray}
+%
+in which $\xi_{\ob{}} \in \reals^+$ is an unknown scalar variance.
+Sampling from the appropriate conditional distribution $\postden(\xi_{\ob{}} | \lgtm, \lgtv, \ls{1:\timax})$ can also be achieved exactly, this time using an inverse gamma prior.
+
 \section{Simulations} \label{sec:simulations}
 
-In the previous sections, we have assumed that $\lgom$ and $\lgov$ are fixed and known a priori. In practice, $\lgov$ will often be unknown. In both the following examples we use models which have,
-\begin{IEEEeqnarray}{rCl}
- \lgov &=& \xi_{\ob{}} \idmat     ,
-\end{IEEEeqnarray}
-%
-in which $\xi_{\ob{}}$ is unknown.
-This additional parameter may be learned in combination with the transition model by the MCMC sampler.
-This entails changing the target distribution to $\postden(\lgtm, \lgtv, \xi_{\ob{}}, \ls{1:\timax})$.
-An additional Gibbs move is introduced to draw from $\postden(\xi_{\ob{}} | \lgtm, \lgtv, \ls{1:\timax})$.
-This may be done exactly if we use a conjugate inverse gamma prior.
-
-Both simulations use the random walk Metropolis-Hastings kernels to alter $\lgtm$ and $\lgtv$, using the Cayley transformation method to propose changes to $\lgtv$.
-Each kernel has a single scalar algorithm parameter which determines the size of the proposed moves, $\sigma_Q^2$ in \eqref{eq:skewsymmetric_proposal} and $\sigma_F^2$ in \eqref{eq:transition_matrix_proposal}.
+In this section we seek to demonstrate the efficacy of MCMC for learning degenerate linear state space models.
+The algorithm is applied to two toy models with simulated data in order to demonstrate that the sampler converges and generates a plausible posterior distribution, and also that the results are more accurate than those achieved by simply approximating the covariance matrix as full rank and using a standard Gibbs sampler.
+Then we use the new algorithm to learn a degenerate model for interpolation in a motion capture application, and demonstrate improved accuracy over competing methods.
+
+All the simulations use the random walk Metropolis-Hastings kernels to alter $\lgtm$ and $\lgtv$, using the Cayley transformation method to propose changes to $\lgtv$.
+During each iteration, one of the three kernels is chosen, each with probability $\frac{1}{3}$.
+For the reversible jump moves, the probability of either adding or removing an eigenvalue is $\half$.
+Each of the random walk kernels has a single scalar algorithm parameter which determines the size of the proposed moves, $\sigma_Q^2$ in \eqref{eq:skewsymmetric_proposal} and $\sigma_F^2$ in \eqref{eq:transition_matrix_proposal}.
 In order to accelerate mixing, and reduce the need for algorithm tuning, these are adjusted adaptively based on the acceptance rate using the method introduced by \cite{Roberts2009}.
@@
 
-\subsection{A Toy Model}
-
-We demonstrate the efficacy of the Gibbs sampler with a simple simulation example on a 4-dimensional model. The following transition matrix is used,
+\subsection{Toy Model 1}
+
+First we consider a 4-dimensional model, with transition matrix,
 %
@@
 %
-which has a rank of $2$. For the observation model, $\lgom = \idmat$ and $\lgov = 0.1 \idmat{}$.
+which has a rank of $2$. For the observation model, $\lgom = \idmat$ and $\lgov = 0.03 \idmat{}$.
 
@@
 \end{IEEEeqnarray}
-%
-The move probabilities were set to,
-%
-\begin{IEEEeqnarray}{rCl}
- P(\rk \to\rk+1) = P(\rk\to\rk+1) = \half     .
-\end{IEEEeqnarray}
-
-The simulation results demonstrate several important features of the degenerate model and the associated MCMC algorithm.
-\begin{itemize}
- \item The sampler identifies the true rank of the covariance matrix within a few tens of iterations, and remains at this value thereafter.
- \item Trace plots suggest that the sampler converges within a few thousand iterations. See figure~\ref{fig:toy-R-trace}.
- \item Autocorrelation plots show that the sampler mixes more slowly than that learning a full rank covariance model. See figure~\ref{fig:toy-R-acf}.
- \item The parameter posterior histograms are consistent with the true values. See figure~\ref{fig:toy-FQ-hist}.
- \item The state trajectory is estimated more accurately with a degenerate covariance, with an RMSE of $5.51$, compared to $6.01$ for the full rank model.
-\end{itemize}
+
+Trace plots and posterior histograms for $\lgtm$ and $\lgtv$ learned using the degenerate model are shown in figures~\ref{fig:toy-F-trace-hist} and~\ref{fig:toy-Q-trace-hist}.
+These seem to demonstrate that the Markov chain mixes reasonably well, taking a few thousand iterations to converge.
+() Furthermore, the resulting approximations of the posterior distribution are consistent with the true values.
+
+Next we compare the results from full rank and degenerate models.
+The state trajectory is estimated more accurately using the degenerate model, with an RMSE of $3.15$, compared to $3.81$ for the full rank model.
+However, the Markov chain for the degenerate model also mixes a little more slowly, as illustrated by the autocorrelation plots in figure~\ref{fig:toy-R-acf}.
+
+The algorithm correctly identifies the rank of the covariance matrix.
+However, its behaviour may be cause for concern.
+The sampler settles on the true value within a few tens of iterations and then never changes.
+With a discrete variable such as rank this may simply indicate that the posterior distribution highly concentrated on this value.
+However, it may also be a symptom of poor mixing for this variable.
 
@@
  \includegraphics[width=0.9\columnwidth]{figures/toy-state.pdf}
- \caption{An example data set simulated from the toy model, showing evolution of the four state components over time.}
+ \caption{An example data set simulated from toy model 1, showing evolution of the four state components over time.}
  \label{fig:toy-data}
@@
  \centering
- \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-basic-R-trace.pdf} \label{subf:toy-R-trace:basic}} \\
- \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-R-trace.pdf} \label{subf:toy-R-trace:degenerate}}
- \caption{MCMC trace plots for $\xi_{\ob{}}$ for \protect\subref{subf:toy-R-trace:basic} the full rank transition model and \protect\subref{subf:toy-R-trace:degenerate} the degenerate transition model.}
- \label{fig:toy-R-trace}
+ \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-F-trace.pdf} \label{subf:toy-F-trace:degenerate}} \\
+ \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-F-hist.pdf} \label{subf:toy-F-hist:degenerate}}
+ \caption{Toy Model 1: MCMC \protect\subref{subf:toy-F-trace:degenerate} trace plot and \protect\subref{subf:toy-F-hist:degenerate} histogram for the components of $\lgtm$, for the degenerate transition model.}
+ \label{fig:toy-F-trace-hist}
+\end{figure}
+
+\begin{figure}
+ \centering
+ \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-Q-trace.pdf} \label{subf:toy-Q-trace:degenerate}} \\
+ \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-Q-hist.pdf} \label{subf:toy-Q-hist:degenerate}}
+ \caption{Toy Model 1: MCMC \protect\subref{subf:toy-Q-trace:degenerate} trace plot and \protect\subref{subf:toy-Q-hist:degenerate} histogram for the components of $\lgtv$, for the degenerate transition model.}
+ \label{fig:toy-Q-trace-hist}
 \end{figure}
@@
  \subfloat[]{\includegraphics[width=0.45\columnwidth]{figures/toy-degenerate-R-acf.pdf} \label{subf:toy-R-acf:degenerate}}
- \caption{MCMC autocorrelation plots for $\xi_{\ob{}}$ for \protect\subref{subf:toy-R-acf:basic} the full rank transition model and \protect\subref{subf:toy-R-acf:degenerate} the degenerate transition model.}
+ \caption{Toy Model 1: MCMC autocorrelation plots for $\xi_{\ob{}}$ for \protect\subref{subf:toy-R-acf:basic} the full rank transition model and \protect\subref{subf:toy-R-acf:degenerate} the degenerate transition model.}
  \label{fig:toy-R-acf}
@@
 
+\subsection{Toy Model 2}
+
+We consider a second toy example in which the covariance matrix rank is less precisely determined by the observations in order to investigate the ability of the sampler to explore the posterior distribution of this variable.
+The latent state is 6-dimensional with $\lgtm=0.99 I$, and a covariance matrix of rank $3$ sampled from the prior,
+%
+\begin{IEEEeqnarray}{rCl}
+ \lgtv &=& A A\tr \nonumber 
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
+ A &=& \begin{bmatrix}
+  1.76 & 0.40 & 0.98 \\
+  2.24 & 1.87 &-0.98 \\
+  0.95 &-0.15 &-0.10 \\
+  0.41 & 0.14 & 1.45 \\
+  0.76 & 0.12 & 0.44 \\
+  0.33 & 1.49 &-0.21
+ \end{bmatrix}     .
+\end{IEEEeqnarray}
+%
+For the observation model, $\lgom = \idmat$ and $\lgov = \xi_{\ob{}} \idmat$, with $\xi_{\ob{}}=1$.
+
+A data set with $\timax=100$ time instants simulated from the model is shown in figure~\ref{fig:toy2-data}.
+
+The MCMC algorithm was used to learn a degenerate model.
+All algorithm parameters were the same as the previous example.
+The trace plot for the covariance matrix rank and the posterior histogram are shown in figure~\ref{fig:toy2-rank-trace-hist}.
+The posterior approximation places significant weight on the true value.
+Reducing the value of $\xi_{\ob{}}$ leads to an increased posterior weight on the correct value of the rank.
+As noise variance is increased, so does the uncertainty in the rank, because the noise masks the effect of the transition covariance corresponding to the smaller eigenvalue components.
+
 \begin{figure}
  \centering
- \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-F-hist.pdf} \label{subf:toy-FQ-hist:F}} \\
- \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-Q-hist.pdf} \label{subf:toy-FQ-hist:Q}}
- \caption{MCMC Posterior histograms for \protect\subref{subf:toy-FQ-hist:F} $\lgtm$ and \protect\subref{subf:toy-FQ-hist:Q} $\lgtv$, from the degenerate linear transition model.
+ \includegraphics[width=0.9\columnwidth]{figures/toy2-state.pdf}
+ \caption{An example data set simulated from toy model 2, showing evolution of the six state components over time.}
+ \label{fig:toy2-data}
+\end{figure}
+
+\begin{figure}
+ \centering
+ \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy2-degenerate-rank-trace.pdf} \label{subf:toy2-rank-trace:degenerate}} \\
+ \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy2-degenerate-rank-hist.pdf} \label{subf:toy2-rank-hist:degenerate}}
+ \caption{Toy Model 2: MCMC \protect\subref{subf:toy2-rank-trace:degenerate} trace plot and \protect\subref{subf:toy2-rank-hist:degenerate} histogram for $\rank(\lgtv)$ with the degenerate transition model.}
+ \label{fig:toy2-rank-trace-hist}
+\end{figure}
+
+% \begin{figure}
+%  \centering
+%  \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-basic-R-trace.pdf} \label{subf:toy-R-trace:basic}} \\
+%  \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-R-trace.pdf} \label{subf:toy-R-trace:degenerate}}
+%  \caption{MCMC trace plots for $\xi_{\ob{}}$ for \protect\subref{subf:toy-R-trace:basic} the full rank transition model and \protect\subref{subf:toy-R-trace:degenerate} the degenerate transition model.}
+%  \label{fig:toy-R-trace}
+% \end{figure}
+
+% \begin{figure}
+%  \centering
+%  \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-F-hist.pdf} \label{subf:toy-FQ-hist:F}} \\
+%  \subfloat[]{\includegraphics[width=0.9\columnwidth]{figures/toy-degenerate-Q-hist.pdf} \label{subf:toy-FQ-hist:Q}}
+%  \caption{MCMC Posterior histograms for \protect\subref{subf:toy-FQ-hist:F} $\lgtm$ and \protect\subref{subf:toy-FQ-hist:Q} $\lgtv$, from the degenerate linear transition model.
 Vertical lines indicate true values}
- \label{fig:toy-FQ-hist}
-\end{figure}
+%  \label{fig:toy-FQ-hist}
+% \end{figure}
 
@@
 
-In a more realistic example, the MCMC algorithms are applied to the problem of marker interpolation in a motion capture application.
+In a practical example, the MCMC algorithms are applied to the problem of marker interpolation in a motion capture application.
 The data used is from \cite{Aristidou2013}.
@@
 More significantly, the Markov chain mixes more slowly and thus requires a longer chain to be simulated to achieve the same effective sample size.
-An important avenue for future research will be designing more sophisticated Metropolis-Hastings moves to explore the parameters space more efficiently than the simple random walks used here.
+An important avenue for future research will be designing more sophisticated Metropolis-Hastings moves to explore the parameter space more efficiently than the simple random walks used here.
 

### 2016-01-18 analyse_toy.py

@@ -120,3 +120,37 @@ for ii in range(ds):
         ax.set_yticklabels([])
         ax.set_xticks([Qmin[ii,jj],Qmax[ii,jj]])
 plt.gcf().savefig(figure_path+'toy-degenerate-Q-hist.pdf', bbox_inches='tight')
+
+
+
+degen_F_samples = np.array([pp['F'] for pp in degenerate_learner.chain_model])
+
+spread = [0.025,0.25,0.7,0.7]
+
+fig, axs, coords =  degenerate_learner._create_2d_plot_axes(model.parameters['F'])
+for idx in np.ndindex(coords.shape):
+    samples = [pp[coords[idx]] for pp in degen_F_samples]
+    axs[idx].plot(samples, color='k')
+    xlims = axs[idx].get_xlim()
+    axs[idx].plot(xlims, [model.parameters['F'][coords[idx]]]*2, 'r',
+                                                          linewidth=2)
+    axs[idx].set_xlim(xlims)
+    ylims = axs[idx].get_ylim()
+    axs[idx].locator_params(nbins=4)
+    axs[idx].set_ylim(model.parameters['F'][coords[idx]]+spread[idx[1]]*np.array([-1,+1]))
+plt.tight_layout()
+plt.gcf().savefig(figure_path+'toy-degenerate-F-trace.pdf', bbox_inches='tight')
+
+fig, axs, coords =  degenerate_learner._create_2d_plot_axes(model.parameters['Q'])
+for idx in np.ndindex(coords.shape):
+    samples = [pp[coords[idx]] for pp in degen_Q_samples]
+    axs[idx].plot(samples, color='k')
+    xlims = axs[idx].get_xlim()
+    axs[idx].plot(xlims, [model.parameters['Q'][coords[idx]]]*2, 'r',
+                                                          linewidth=2)
+    axs[idx].set_xlim(xlims)
+    ylims = axs[idx].get_ylim()
+    axs[idx].locator_params(nbins=2)
+    axs[idx].set_ylim([-0.2,1.5])
+plt.tight_layout()
+plt.gcf().savefig(figure_path+'toy-degenerate-Q-trace.pdf', bbox_inches='tight')

### 2016-01-18 analyse_toy2.py

@@ -1,0 +1,79 @@
+import pickle
+import numpy as np
+import scipy.linalg as la
+from matplotlib import pyplot as plt
+
+from kalman import GaussianDensity
+from linear_models import BasicLinearModel, DegenerateLinearModel
+from learners_mcmc import (
+    load_learner,
+    BaseMCMCLearner,
+    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+    MCMCLearnerTransitionBasicModelWithMNIWPrior,
+    MCMCLearnerTransitionDegenerateModelWithMNIWPrior)
+
+# Create learner classes
+class MCMCLearnerB(BaseMCMCLearner,
+                    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+                    MCMCLearnerTransitionBasicModelWithMNIWPrior):
+    pass
+
+class MCMCLearnerD(BaseMCMCLearner,
+                    MCMCLearnerObservationDiagonalCovarianceWithIGPrior,
+                    MCMCLearnerTransitionDegenerateModelWithMNIWPrior):
+    pass
+
+def rmse(state, estimate):
+    return la.norm(state-estimate)
+
+figure_path = './notes/figures/'
+
+# Load the test data
+test_path = './results/toy2-final/'
+test_data_file = 'toy2-test-data.p'
+fh = open(test_path+test_data_file, 'rb')
+model,state,observ = pickle.load(fh)
+
+# Load the MCMC output
+degenerate_learner = load_learner(test_path+'toy2-mcmc-degenerate.p')
+
+# Get state estimates from each algorithm
+num_burn = 5000
+degenerate_mn, degenerate_sd = degenerate_learner.estimate_state_trajectory(
+                                                           numBurnIn=num_burn)
+
+# Draw state
+ds = degenerate_learner.model.ds
+fig, axs = plt.subplots(nrows=ds,ncols=1,sharex=True)
+for dd in range(ds):
+    axs[dd].plot(state[:,dd], 'k')
+    axs[dd].locator_params(axis='y',nbins=2)
+fig.savefig(figure_path+'toy2-state.pdf', bbox_inches='tight')
+
+# Rank plots
+degenerate_learner.plot_chain_trace('rank', numBurnIn=num_burn, trueModel=model)
[…]

### 2016-01-18 givens_test.py

@@ -1,0 +1,59 @@
+import numpy as np
+import scipy.linalg as la
+
+d = 4
+r = 2
+
+A = np.random.randn(d,d)
+V = la.qr(A)[0][:,:r]
+
+L = np.diag([2,1])
+Q = np.dot(V, np.dot(L, V.T))
+
+Uc = np.identity(d)
+Ur = np.identity(r)
+E = V.copy()
+
+def givmat(d,i,j,g):
+    G = np.identity(d)
+    G[i,i] = np.cos(g)
+    G[j,j] = np.cos(g)
+    G[i,j] = -np.sin(g)
+    G[j,i] = np.sin(g)
+    return G
+
+for i in reversed(range(r)):
+    for j in range(i):
+
+        angle = np.arctan(E[i,j]/E[i,i])
+        G = givmat(r,i,j,angle)
+        E = np.dot(E, G)
+        Ur = np.dot(G.T, Ur)
+
+        print(E)
+        print('')
+
+for j in range(r):
+    for i in reversed(range(r,d)):
+
+        angle = np.arctan(E[i,j]/E[j,j])
+        G = givmat(d,i,j,angle)
+        E = np.dot(G, E)
+        Uc = np.dot(Uc, G.T)
+
+        print(E)
+        print('')
+
+print(V)
+print(np.dot(Uc, np.dot(E, Ur)))
+print('')
+
+U = Uc[:,:r]
+Ed = E[:r,:r]
+EUr = np.dot(Ed,Ur)
+D = np.dot(EUr, np.dot(L, EUr.T))
+
+Qr = np.dot(U, np.dot(D, U.T))
+
+print(Q)
+print(Qr)

### 2016-01-18 run_toy2.py

@@ -2,6 +2,7 @@ from copy import deepcopy
 import pickle
 
 import numpy as np
+import scipy.linalg as la
 from matplotlib import pyplot as plt
 
 from kalman import GaussianDensity
@@ -27,24 +28,28 @@ class MCMCLearnerD(BaseMCMCLearner,
 plt.close('all')
 np.random.seed(0)
 
-test_data_file = 'toy-test-data.p'
-model_type = 'basic'
-K = 200
+test_data_file = 'toy2-test-data.p'
+model_type = 'degenerate'
+K = 100
 num_warm = 100
 num_iter = 10000
 num_burn = int(num_iter/2)
 
 
-ds = 4
-do = 4
+ds = 6
+do = 6
 params = dict()
-params['F'] = np.array([[0.95,0.8,0.8,0.0],[0,0.95,-0.5,0.1],[0,0,1.6,-0.8],[0.0,0.0,1.0,0.0]])
-params['rank'] = np.array([2])
-params['val'] = np.array([1.5,0.5])
-params['vec'] = np.array([[0.5,0.5,0.5,0.5],[1.0/np.sqrt(2),-1.0/np.sqrt(2),0.0,0.0]]).T
-params['Q'] = np.dot(params['vec'], np.dot(np.diag(params['val']), params['vec'].T))
+params['F'] = 0.99*np.identity(ds)
+params['rank'] = np.array([3])
+
+A = np.round(np.random.randn(ds,params['rank']),2)
+print(A)
+Q = np.dot(A,A.T)
+params['val'], params['vec'] = la.eigh(Q)
+params['Q'] = Q
+
 params['H'] = np.identity(do)
-params['R'] = 0.03*np.identity(do)
+params['R'] = 10*np.identity(do)
 prior = GaussianDensity(np.zeros(ds), 10*np.identity(ds))
 
 model = DegenerateLinearModel(ds, do, prior, params)
@@ -62,8 +67,8 @@ fh.close()
 
 est_params = deepcopy(params)
 est_params['F'] = 0.5*np.identity(ds)
-est_params['rank'] = np.array([4])
-est_params['val'] = np.array([4.0,3.0,2.0,1.0])
+est_params['rank'] = np.array([ds])
+est_params['val'] = 1 + 0.2*np.random.randn(ds)
 est_params['vec'] = np.identity(ds)
 est_params['Q'] = np.dot(est_params['vec'], np.dot(np.diag(est_params['val']), est_params['vec'].T))
[…]
